---
title: "Investigating strain, domain and developmental stage effects"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
      embed-resources: true
      standalone: true
code-fold: true
toc: true
toc-depth: 5
toc-expand: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| message: false
#| warning: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(enrichR)
library(DT)
library(here)

set.seed(123)
select <- dplyr::select 
rename <- dplyr::rename 
filter <- dplyr::filter 
summarize <- dplyr::summarize 
count <- dplyr::count

domain_colors <- c(PM = rgb(252,206,93, maxColorValue = 255), 
                   MAX = rgb(163,82,157, maxColorValue = 255), 
                   POST = rgb(158,156,201, maxColorValue = 255))

strain_colors <- c(B6 = "#555555", CAST = "#009E73")
```

```{r}
#| label: helper-functions
#| message: false
#| warning: false

create_dt <- function(x){
  format_num <- function(val) {
    case_when(
      is.na(val) ~ "NA",
      abs(val) < 0.01 ~ formatC(val, digits = 2),
      TRUE ~ formatC(round(val, 2))
    )
  }
  x |> 
    mutate(across(where(is.numeric), format_num)) |>
    DT::datatable(
      extensions = 'Buttons',
      rownames = FALSE, 
      filter = "top",
      options = list(dom = 'Blfrtip',
                     buttons = c('copy', 'csv', 'excel'),
                     pageLength = 5, 
                     scrollX = TRUE))
}

# Helper function for consistent filtering of DESeq2 results
filter_deseq_results <- function(res, 
                                  min_basemean = NULL,
                                  min_basemean_quantile = 0.1,
                                  max_lfcse = 3, 
                                  max_abs_lfc = 20, 
                                  min_abs_lfc = 0.5, 
                                  max_padj = 0.05) {
  
  # Convert to data frame
  res_df <- as.data.frame(res) |>
    rownames_to_column("gene_id")
  
  # Set baseMean threshold - use quantile if not specified
 if(is.null(min_basemean)) {
    min_basemean <- quantile(res_df$baseMean, min_basemean_quantile, na.rm = TRUE)
  }
  
  # Apply filters
  res_filtered <- res_df |>
    filter(
      baseMean > min_basemean,
      lfcSE < max_lfcse,
      abs(log2FoldChange) < max_abs_lfc,
      abs(log2FoldChange) > min_abs_lfc,
      padj < max_padj
    )
  
  return(res_filtered)
}

# Helper to get passing gene IDs from a filtered result
get_passing_genes <- function(res, ...) {
  filter_deseq_results(res, ...) |> pull(gene_id)
}
get_normalized_counts <- function(dds, genes, vsd = NULL) {
  if (is.null(vsd)) {
    # Fallback: compute if not pre-computed (slower)
    vsd <- vst(dds, blind = FALSE)
  }
  assay(vsd)[genes, , drop = FALSE] |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    pivot_longer(-gene_id, names_to = "sample", values_to = "count") |>
    left_join(colData(dds) |> as.data.frame() |> rownames_to_column("sample"), by = "sample") |>
    left_join(all_genes_palate |> select(gene_id, symbol), by = "gene_id")
}

run_enrichment_hybrid_v2 <- function(gene_ids, 
                                 gene_reference = all_genes_palate,
                                 description = "Gene Set",
                                 max_term_size = 1000,
                                 min_term_size = 4,
                                 retries = 3,
                                 retry_wait = 5) {
  
  databases <- c(
     "GO_Biological_Process_2025",
      "GO_Cellular_Component_2025",
      "GO_Molecular_Function_2025",
      "GWAS_Catalog_2025", 
      "Human_Phenotype_Ontology",
      "JASPAR_PWM_Mouse_2025",
      "MGI_Mammalian_Phenotype_Level_4_2021",
      "CORUM",
      "KEGG_2019_Mouse",
      "Tabula_Muris",
      # "GTEx_Tissues_V8_2023", # results too broad, not useful
      # "PerturbAtlas_MouseGenePerturbationSigs", # too big, parser can't handle it!
      "Mouse_Gene_Atlas",
      # "DSigDB",
      "Jensen_DISEASES_Curated_2025",
      "KOMP2_Mouse_Phenotypes_2022",
      "WikiPathways_2024_Mouse"
  )
  
  # Convert gene IDs to symbols
  query_symbols <- gene_reference |>
    filter(gene_id %in% unique(gene_ids)) |>
    pull(symbol) |>
    unique()
  
  # Get background symbols
  background_symbols <- gene_reference$symbol
  
  # Skip if too few genes
  if(length(query_symbols) < 5) {
    message(paste0("Skipping ", description, ": fewer than 5 genes"))
    return(NULL)
  }
  
  # Query each database individually with tryCatch + retry
  result_enrichr <- list()
  
  for (db in databases) {
    res <- NULL
    for (attempt in seq_len(retries)) {
      res <- tryCatch(
        enrichr(
          genes      = query_symbols,
          databases  = db,
          background = background_symbols,
          include_overlap = TRUE
        ),
        error = function(e) {
          message(sprintf("  [%s] Attempt %d failed: %s", db, attempt, e$message))
          NULL
        }
      )
      if (!is.null(res)) break
      if (attempt < retries) {
        message(sprintf("  [%s] Retrying in %ds...", db, retry_wait))
        Sys.sleep(retry_wait)
      }
    }
    
    if (is.null(res)) {
      message(sprintf("  [%s] Skipped after %d failed attempts.", db, retries))
    } else {
      result_enrichr <- c(result_enrichr, res)
    }
  }
  
  # Format Enrichr results
  enrichr_df <- NULL
  if (length(result_enrichr) > 0) {
    enrichr_dfs <- lapply(names(result_enrichr), function(db_name) {
      db_result <- result_enrichr[[db_name]]
      
      if (is.null(db_result) || nrow(db_result) == 0 || all(is.na(db_result$Term))) {
        return(NULL)
      }
      
      db_result |>
        filter(Adjusted.P.value < 0.05) |>
        mutate(
          gene_set         = description,
          source           = db_name,
          term_name        = Term,
          p_value          = Adjusted.P.value,
          query_size       = length(query_symbols),
          intersection_size = as.numeric(gsub("/.*", "", Overlap)),
          term_size        = as.numeric(gsub(".*/", "", Overlap)),
          intersection     = Genes
        ) |>
        filter(term_size < max_term_size, term_size > min_term_size) |>
        select(gene_set, source, term_name, p_value,
               query_size, term_size, intersection_size, intersection)
    })
    
    enrichr_dfs <- enrichr_dfs[!sapply(enrichr_dfs, is.null)]
    if (length(enrichr_dfs) > 0) {
      enrichr_df <- bind_rows(enrichr_dfs)
    }
  }
  
  combined_result <- enrichr_df
  
  if (!is.null(combined_result) && nrow(combined_result) > 0) {
    return(list(result = combined_result))
  } else {
    return(NULL)
  }
}

# plotting-functions
plot_main_strain_effect <- function(gene_ids, direction, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + 
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = direction, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = strain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) + geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) + 
      scale_fill_manual(values = strain_colors) +
      labs(x = "", y = "") + theme_pubclean(base_size = 16) + theme(legend.position = "none")
    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

# ── Helper: build strain effect bar data ──────────────────────────
get_strain_diffs <- function(eff_data) {
  tibble(
    AP_domain = factor(c("PM", "MAX", "POST"), levels = c("PM", "MAX", "POST")),
    effects   = c(eff_data$strain_effect_PM,
                  eff_data$strain_effect_MAX,
                  eff_data$strain_effect_POST)
  )
}

# ── FUNCTION 1: Base plot (boxplot + effect bars) ──────────────────
# Used by: PM/MAX/POST strongest, static, opposite directions
plot_domain_strain_base <- function(gene_ids, pattern_desc, count_data,
                                    n = 5, eff_data_full = domain_strain_patterns) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    eff_data    <- eff_data_full |> filter(gene_id == gid)
    if (nrow(gene_data) == 0 || nrow(eff_data) == 0) return(NULL)

    gene_name   <- unique(gene_data$symbol)[1]
    strain_diffs <- get_strain_diffs(eff_data)

    p_box <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain),
                 position = position_jitterdodge(jitter.width = 0.15),
                 alpha = 0.4, size = 1.5) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values  = strain_colors) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc,
           x = "Domain", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom",
            plot.title = element_text(face = "bold"))

    p_bar <- ggplot(strain_diffs, aes(x = AP_domain, y = effects, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      scale_fill_manual(values = domain_colors) +
      labs(y = "Strain Effect\n(log2FC CAST vs B6)", x = "") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    p_box + p_bar + plot_layout(widths = c(2.5, 1))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}

# ── FUNCTION 2: Profile plot (mean±SE lines) + effect bars ────────
# Used by: high domain specificity, opposite directions
plot_domain_strain_profile <- function(gene_ids, pattern_desc, count_data,
                                       n = 5, eff_data_full = domain_strain_patterns) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    eff_data     <- eff_data_full |> filter(gene_id == gid)
    if (nrow(gene_data) == 0 || nrow(eff_data) == 0) return(NULL)

    gene_name    <- unique(gene_data$symbol)[1]
    strain_diffs <- get_strain_diffs(eff_data)

    summ <- gene_data |>
      group_by(AP_domain, strain) |>
      summarize(mn = mean(count, na.rm = TRUE),
                se = sd(count, na.rm = TRUE) / sqrt(n()),
                .groups = "drop")

    p_profile <- ggplot(summ, aes(x = AP_domain, y = mn,
                                   color = strain, group = strain)) +
      geom_line(linewidth = 1.2) +
      geom_pointrange(aes(ymin = mn - se, ymax = mn + se), size = 0.8) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc,
           x = "Domain", y = "Mean Normalized Counts (±SE)") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom",
            plot.title = element_text(face = "bold"))

    p_bar <- ggplot(strain_diffs, aes(x = AP_domain, y = effects, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      scale_fill_manual(values = domain_colors) +
      labs(y = "Strain Effect\n(log2FC CAST vs B6)", x = "") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    p_profile + p_bar + plot_layout(widths = c(2.5, 1))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}


# ── FUNCTION 3: Boxplot + strain-faceted temporal plot (single row) 
# Used by: time-affected pattern
# Left  : boxplot per domain colored by strain
# Right : loess over time, faceted by strain, colored by domain
plot_domain_strain_temporal <- function(gene_ids, pattern_desc, count_data,
                                        n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) return(NULL)

    gene_name <- unique(gene_data$symbol)[1]

    p_box <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain),
                 position = position_jitterdodge(jitter.width = 0.15),
                 alpha = 0.4, size = 1.5) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values  = strain_colors) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc,
           x = "Domain", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom",
            plot.title = element_text(face = "bold"))

    p_loess <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count,
                                      color = AP_domain)) +
      geom_point(alpha = 0.5, size = 1.5) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 1.2) +
      facet_wrap(~ strain, nrow = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts", color = "Domain") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom",
            strip.text = element_text(face = "bold"))

    p_box + p_loess + plot_layout(widths = c(1, 2.5))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}



# ── FUNCTION 4: Overview lm + strain-faceted domain-colored loess ─
# Used by: uniform, diverging CAST/B6 faster time patterns
# Left : all data together, color = strain, shape = domain, lm smooth
#         — shows the overall temporal divergence between strains
# Right: faceted by strain, color = domain, loess smooth
#         — shows whether all domains move in parallel within each strain
plot_time_strain_overview <- function(gene_ids, pattern_desc, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) return(NULL)

    gene_name <- unique(gene_data$symbol)[1]

    # Left: all domains together, strain as color, domain as shape
    p_overview <- ggplot(gene_data,
                         aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(aes(shape = AP_domain), alpha = 0.5, size = 2.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1.5) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      scale_shape_manual(values = c(PM = 16, MAX = 17, POST = 15)) +
      labs(title = gene_name, subtitle = pattern_desc,
           x = "LB stage (centered)", y = "Normalized Counts",
           shape = "Domain") +
      theme_pubclean(base_size = 16) +
      theme(legend.position   = "bottom",
            plot.title        = element_text(face = "bold"))

    # Right: faceted by strain, domain as color, loess
    p_facet <- ggplot(gene_data,
                      aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 1.5) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 1.2) +
      facet_wrap(~ strain, nrow = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts", color = "Domain") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom",
            strip.text      = element_text(face = "bold"))

    p_overview + p_facet + plot_layout(widths = c(1, 2.5))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}


plot_time_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data$symbol)[1]
    
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(aes(shape = AP_domain), alpha = 0.5, size = 2.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1.5) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "right", plot.title = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) + geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "none", strip.text = element_text(face = "bold"))
    
    p1 + p2 + plot_layout(nrow = 1, widths = c(1, 2.5))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect")
}


plot_domain_specificity <- function(gene_ids, domains_affected, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain), position = position_jitterdodge(jitter.width = 0.1), alpha = 0.4, size = 1.5) +
      scale_y_log10(labels = scales::comma) + scale_fill_manual(values = strain_colors) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = paste("Strain effect in", domains_affected), x = "Domain", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) + geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "none", strip.text = element_text(face = "bold"))
    p1 / p2 + plot_layout(heights = c(1, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1,  guides = "collect") & theme( legend.position = "bottom")
}
```

```{r}
#| label: load-data
#| message: false
#| warning: false

sample_details <- read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 
load("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/palate_gbrs_emase_counts_v4.RData")

contam_samples <- c("B6_12.5_PM_1","B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat_nondo[, !colnames(expr_mat_nondo) %in% contam_samples]
rm(expr_mat_do, expr_mat_nondo)

all_genes_palate <- read_tsv(here("_data","ENSMUSGid_to_symbol_v105.txt"),
                              col_names = c("gene_id","symbol","gene_biotype","chromosome","start","end"), 
                              skip = 1) |>  
  mutate(midpoint = (start + end) / 2) |>
  filter(gene_id %in% rownames(raw_expr_mat))

sample_annot <- sample_details |>
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |>
  separate(Strain_stock, into = c("strain","tmp"), remove = FALSE, sep = "_") |>
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, 
         palate_closure, L_R_palate) |>
  mutate(strain = if_else(strain == "B6J", "B6", strain)) |>
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |>
  select(-temp) |>
  mutate(Palate_ori = if_else(Palate_ori == "L", "Left", "Right"))
```

# Differential gene expression analysis overview

```{r}
#| label: prep-run-deseq2
#| message: false
#| warning: false

# ── Prep data for DESeq2 ──────────────────────────────────────────────────────
b6_cast_samples <- sample_annot |>
  filter(strain %in% c("B6", "CAST")) |>
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = if_else(is.na(rep), "1", rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

coldata <- b6_cast_samples |>
  select(sample, unique_sampleid, sex, ind, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(coldata), drop = FALSE])

strain_dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~1)
strain_dds_coll <- collapseReplicates(strain_dds, strain_dds$unique_sampleid)

min_LB_stage <- min(colData(strain_dds_coll)$LB_stage)
colData(strain_dds_coll)$LB_stage_from_start <- colData(strain_dds_coll)$LB_stage - min_LB_stage
colData(strain_dds_coll)$sex <- factor(colData(strain_dds_coll)$sex)
colData(strain_dds_coll)$strain <- factor(colData(strain_dds_coll)$strain, levels = c("B6", "CAST"))
colData(strain_dds_coll)$ind <- factor(colData(strain_dds_coll)$ind)


# ── Fit all the models  ─────────────────────────────────────────────
# 1. Additive model -- no interactions
strain_dds_additive <- strain_dds_coll
design(strain_dds_additive) <- ~ sex + AP_domain + LB_stage_from_start  + strain
strain_dds_additive <- DESeq(strain_dds_additive, parallel = TRUE)

# 2. Two-way model -- no threeway interactions
strain_dds_twoway <- strain_dds_coll
design(strain_dds_twoway) <- ~ sex + AP_domain + LB_stage_from_start  +
                                   strain + strain:AP_domain + strain:LB_stage_from_start
strain_dds_twoway <- DESeq(strain_dds_twoway, parallel = TRUE)

# 3. Full model
design(strain_dds_coll) <- ~ sex + AP_domain + LB_stage_from_start  + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start
strain_dds_full <- DESeq(strain_dds_coll, parallel = TRUE)
coef_names <- resultsNames(strain_dds_full)

# ── LRT tests  ────────────────────────────────────────────
# 1. Any strain effect at all
strain_dds_lrt <- DESeq(strain_dds_additive, test = "LRT", 
                        reduced = ~ sex + AP_domain + LB_stage_from_start  ,
                        parallel = TRUE)
res_strain_lrt <- results(strain_dds_lrt)
genes_w_strain_eff <- res_strain_lrt |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05 )


# 2. Domain × Strain interaction
dds_domain_lrt <- DESeq(strain_dds_twoway, test = "LRT",
                        reduced = ~ sex + AP_domain + LB_stage_from_start  + 
                                   strain + strain:LB_stage_from_start,
                        parallel = TRUE)
res_domain_lrt <- results(dds_domain_lrt)
genes_w_strain_by_domain_eff <- res_domain_lrt |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05)

# 3. Time × Strain interaction
dds_time_lrt <- DESeq(strain_dds_twoway, test = "LRT",
                      reduced = ~ sex + AP_domain + LB_stage_from_start  +
                                 strain + strain:AP_domain,
                      parallel = TRUE)
res_time_lrt <- results(dds_time_lrt)
genes_w_strain_by_time_eff <- res_time_lrt |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05)


# 4. Three-way interaction (domain × time × strain)
dds_3way_lrt <- DESeq(strain_dds_full, test = "LRT",
                      reduced = ~ sex + AP_domain + LB_stage_from_start  + 
                                 strain + strain:AP_domain + strain:LB_stage_from_start,
                      parallel = TRUE)
res_3way_lrt <- results(dds_3way_lrt)
genes_w_3way_eff <- res_3way_lrt |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05)

# ── Effects  ────────────────────────────────────────────
# 1. Extract main strain effect (additive model) 
res_strain_main <- results(strain_dds_additive, name = "strain_CAST_vs_B6") 

# 2. Extract domain specific strain effects (two way model)
res_strain_PM  <- results(strain_dds_twoway, name = "strain_CAST_vs_B6")
res_strain_MAX <- results(strain_dds_twoway, 
                          contrast = list(c("strain_CAST_vs_B6", "AP_domainMAX.strainCAST")))
res_strain_POST <- results(strain_dds_twoway, 
                           contrast = list(c("strain_CAST_vs_B6", "AP_domainPOST.strainCAST")))

# 3a. Extract overall time slope
res_time_int_twoway <- results(strain_dds_twoway, name = "LB_stage_from_start.strainCAST")
# 3b. Extract the time slope for PM domain at time = 0
res_time_int_full <- results(strain_dds_full, name = "LB_stage_from_start.strainCAST")

# 4. Domain × strain interaction coefficients (how CAST effect in MAX/POST differs from PM)
res_domain_int_MAX <- results(strain_dds_twoway, name = "AP_domainMAX.strainCAST") 
res_domain_int_POST <- results(strain_dds_twoway, name = "AP_domainPOST.strainCAST")

# 5. Three-way interaction coefficients (how temporal slope differs by domain)
res_3way_MAX <- results(strain_dds_full, name = "AP_domainMAX.LB_stage_from_start.strainCAST") 
res_3way_POST <- results(strain_dds_full, name = "AP_domainPOST.LB_stage_from_start.strainCAST") 

# ── Apply quality & LRT filters ──────────────────────────────────────────────────
# Main strain effect: from additive model, including genes with interactions
res_strain_main_filtered <- filter_deseq_results(res_strain_main) |> 
  filter(gene_id %in% genes_w_strain_eff$gene_id)
genes_main_strain_pass <- res_strain_main_filtered$gene_id

# Strain effect in each domain (from full model)
res_strain_PM_filtered <- filter_deseq_results(res_strain_PM) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_strain_MAX_filtered <- filter_deseq_results(res_strain_MAX) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_strain_POST_filtered <- filter_deseq_results(res_strain_POST) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)

genes_strain_PM_pass <- res_strain_PM_filtered$gene_id
genes_strain_MAX_pass <- res_strain_MAX_filtered$gene_id
genes_strain_POST_pass <- res_strain_POST_filtered$gene_id

# Domain × strain interaction coefficients (differences from PM)
res_domain_int_MAX_filtered <- filter_deseq_results(res_domain_int_MAX, min_abs_lfc = 0.3) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_domain_int_POST_filtered <- filter_deseq_results(res_domain_int_POST, min_abs_lfc = 0.3) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)

genes_domain_int_MAX_pass <- res_domain_int_MAX_filtered$gene_id
genes_domain_int_POST_pass <- res_domain_int_POST_filtered$gene_id

# Time × strain interaction (slope — different scale)
res_time_int_filtered <- filter_deseq_results(res_time_int_twoway, min_abs_lfc = 0.1, max_lfcse = 0.5) |> 
  filter( gene_id %in% genes_w_strain_by_time_eff$gene_id)
genes_time_int_pass <- res_time_int_filtered$gene_id

# Three-way coefficients
res_3way_MAX_filtered <- filter_deseq_results(res_3way_MAX, min_abs_lfc = 0.1, max_lfcse = 0.5) |> 
  filter( gene_id %in% genes_w_3way_eff$gene_id)
res_3way_POST_filtered <- filter_deseq_results(res_3way_POST, min_abs_lfc = 0.1, max_lfcse = 0.5) |> 
  filter( gene_id %in% genes_w_3way_eff$gene_id)

genes_3way_MAX_pass <- res_3way_MAX_filtered$gene_id
genes_3way_POST_pass <- res_3way_POST_filtered$gene_id


genes_with_any_interaction <- unique( c(genes_w_strain_by_time_eff$gene_id,
                                     genes_w_strain_by_domain_eff$gene_id,
                                     genes_w_3way_eff$gene_id)
                                     )
# ── Filter summary ─────────────────────────────────────────────────────────
filter_summary <- tibble(
  Result = c("Main strain effect (additive model)",
             "Strain in PM", "Strain in MAX", "Strain in POST",
             "Domain int (MAX)", "Domain int (POST)", "Time int", 
             "3-way (MAX)", "3-way (POST)"),
  Total_tested = c(nrow(res_strain_main), nrow(res_strain_PM), nrow(res_strain_MAX),
                   nrow(res_strain_POST), nrow(res_domain_int_MAX), nrow(res_domain_int_POST),
                   nrow(res_time_int_twoway), nrow(res_3way_MAX), nrow(res_3way_POST)),
  Passing_filter = c(length(genes_main_strain_pass), length(genes_strain_PM_pass),
                     length(genes_strain_MAX_pass), length(genes_strain_POST_pass),
                     length(genes_domain_int_MAX_pass), length(genes_domain_int_POST_pass),
                     length(genes_time_int_pass), length(genes_3way_MAX_pass),
                     length(genes_3way_POST_pass)),
 Notes = c(paste0("Includes ", length(genes_with_any_interaction), " genes with domain, time, or three-way interactions"),
            rep("From two-way model", 3),
            rep("Interaction coefficient", 2),
            "Slope coefficient", rep("Three-way coefficient", 2))
)

# ── Combine results into classified table ──────────────────────────────────
strain_effects_classified <- tibble::tibble(
  gene_id = rownames(res_strain_lrt),
  baseMean = res_strain_lrt$baseMean,
  
  # LRT significance T/F
  any_strain_sig = res_strain_lrt$padj < 0.05, 
  domain_int_sig = res_domain_lrt$padj < 0.05,
  time_int_sig = res_time_lrt$padj < 0.05,
  threeway_int_sig = res_3way_lrt$padj < 0.05,
  
  # LRT p-values
  any_strain_pval = res_strain_lrt$padj,
  domain_int_pval = res_domain_lrt$padj,
  time_int_pval = res_time_lrt$padj,
  threeway_int_pval = res_3way_lrt$padj,
  
  # Main strain effect (additive model — marginal, averaged over domain and time)
  main_strain_lfc = res_strain_main$log2FoldChange,
  main_strain_lfcSE = res_strain_main$lfcSE,
  main_strain_pval = res_strain_main$padj,
  main_strain_pass = gene_id %in% genes_main_strain_pass,
  has_any_interaction = gene_id %in% genes_with_any_interaction,
  
  # Strain effect within each domain (two-way model)
  strain_effect_PM = res_strain_PM$log2FoldChange,
  strain_effect_PM_lfcSE = res_strain_PM$lfcSE,
  strain_effect_PM_pval = res_strain_PM$padj,
  strain_effect_PM_pass = gene_id %in% genes_strain_PM_pass,
  
  strain_effect_MAX = res_strain_MAX$log2FoldChange,
  strain_effect_MAX_lfcSE = res_strain_MAX$lfcSE,
  strain_effect_MAX_pval = res_strain_MAX$padj,
  strain_effect_MAX_pass = gene_id %in% genes_strain_MAX_pass,
  
  strain_effect_POST = res_strain_POST$log2FoldChange,
  strain_effect_POST_lfcSE = res_strain_POST$lfcSE,
  strain_effect_POST_pval = res_strain_POST$padj,
  strain_effect_POST_pass = gene_id %in% genes_strain_POST_pass,
  
  # Domain × strain interaction coefficients (two-way model)
  domain_int_MAX_lfc = res_domain_int_MAX$log2FoldChange,
  domain_int_MAX_lfcSE = res_domain_int_MAX$lfcSE,
  domain_int_MAX_pval = res_domain_int_MAX$padj,
  domain_int_MAX_pass = gene_id %in% genes_domain_int_MAX_pass,
  
  domain_int_POST_lfc = res_domain_int_POST$log2FoldChange,
  domain_int_POST_lfcSE = res_domain_int_POST$lfcSE,
  domain_int_POST_pval = res_domain_int_POST$padj,
  domain_int_POST_pass = gene_id %in% genes_domain_int_POST_pass,
  
  # Time × strain interaction (two-way model)
  # time_int_slope = res_time_int$log2FoldChange,
  time_int_slope_avg = res_time_int_twoway$log2FoldChange,    # from two-way model 
  time_int_slope_lfcSE = res_time_int_twoway$lfcSE,
  time_int_slope_pval = res_time_int_twoway$padj,
  time_int_pass = gene_id %in% genes_time_int_pass,
  time_int_slope_PM  = res_time_int_full$log2FoldChange,      # from full model 
  
  # Three-way interaction coefficients (full model)
  threeway_MAX_slope = res_3way_MAX$log2FoldChange,
  threeway_MAX_slope_lfcSE = res_3way_MAX$lfcSE,
  threeway_MAX_slope_pval = res_3way_MAX$padj,
  threeway_MAX_pass = gene_id %in% genes_3way_MAX_pass,
  
  threeway_POST_slope = res_3way_POST$log2FoldChange,
  threeway_POST_slope_lfcSE = res_3way_POST$lfcSE,
  threeway_POST_slope_pval = res_3way_POST$padj,
  threeway_POST_pass = gene_id %in% genes_3way_POST_pass
) |>
  tidyr::replace_na(list(
    any_strain_sig = FALSE, domain_int_sig = FALSE,
    time_int_sig = FALSE, threeway_int_sig = FALSE,
    main_strain_pass = FALSE, has_any_interaction = FALSE,
    strain_effect_PM_pass = FALSE, strain_effect_MAX_pass = FALSE,
    strain_effect_POST_pass = FALSE,
    domain_int_MAX_pass = FALSE, domain_int_POST_pass = FALSE,
    time_int_pass = FALSE, threeway_MAX_pass = FALSE, threeway_POST_pass = FALSE
  )) |>
  dplyr::left_join(all_genes_palate |> dplyr::select(gene_id, symbol, gene_biotype)) |>
  dplyr::mutate(
    # Overall filter: significant strain LRT, passes QC
    passes_strain_filter = any_strain_sig & main_strain_pass, #& !has_any_interaction,
    
    # Main strain pattern
    # Note: min_abs_lfc = 0.5 filter ensures no genes fall below 0.5
    main_strain_pattern = dplyr::case_when(
      !passes_strain_filter ~ "No effect / filtered",
      main_strain_lfc > 1 ~ "CAST >> B6 (strong)",
      main_strain_lfc > 0.5 ~ "CAST > B6 (moderate)",
      main_strain_lfc < -1 ~ "B6 >> CAST (strong)",
      main_strain_lfc < -0.5 ~ "B6 > CAST (moderate)"
    ),
    
    # Domain with largest strain effect (only for genes WITH domain interaction)
    max_domain_effect = dplyr::case_when(
      !domain_int_sig ~ NA_character_,
      is.na(strain_effect_PM) | is.na(strain_effect_MAX) | is.na(strain_effect_POST) ~ NA_character_,
      abs(strain_effect_PM) >= abs(strain_effect_MAX) & 
        abs(strain_effect_PM) >= abs(strain_effect_POST) ~ "PM",
      abs(strain_effect_MAX) >= abs(strain_effect_POST) & 
        abs(strain_effect_MAX) >= abs(strain_effect_PM) ~ "MAX",
      abs(strain_effect_POST) >= abs(strain_effect_MAX) & 
        abs(strain_effect_POST) >= abs(strain_effect_PM) ~ "POST"
    ),
    
    max_domain_direction = dplyr::case_when(
      is.na(max_domain_effect) ~ NA_character_,
      max_domain_effect == "PM" & strain_effect_PM > 0 ~ "PM: CAST > B6",
      max_domain_effect == "PM" & strain_effect_PM < 0 ~ "PM: B6 > CAST",
      max_domain_effect == "MAX" & strain_effect_MAX > 0 ~ "MAX: CAST > B6",
      max_domain_effect == "MAX" & strain_effect_MAX < 0 ~ "MAX: B6 > CAST",
      max_domain_effect == "POST" & strain_effect_POST > 0 ~ "POST: CAST > B6",
      max_domain_effect == "POST" & strain_effect_POST < 0 ~ "POST: B6 > CAST"
    ),
    
    domain_consistency = dplyr::case_when(
      is.na(strain_effect_PM) | is.na(strain_effect_MAX) | is.na(strain_effect_POST) ~ NA_character_,
      sign(strain_effect_PM) == sign(strain_effect_MAX) & 
        sign(strain_effect_MAX) == sign(strain_effect_POST) ~ "Consistent",
      TRUE ~ "Variable"
    ),
    
    time_pattern = dplyr::case_when(
      !time_int_sig ~ "No time interaction",
      !time_int_pass ~ "Filtered",
      is.na(time_int_slope_avg) ~ NA_character_,
      time_int_slope_avg > 0.1 ~ "Diverging (CAST changing faster than B6)",
      time_int_slope_avg < -0.1 ~ "Diverging (B6 changing faster than CAST)",
      TRUE ~ "Parallel - Both changing similarly over time"
    ),
    
    # Overall pattern type — now mutually exclusive by construction
    pattern_type = dplyr::case_when(
      !any_strain_sig ~ "No strain effect",
      main_strain_pass ~ "Main strain effect",
      threeway_int_sig ~ "Three-way interaction",
      domain_int_sig & time_int_sig ~ "Domain + Time interactions",
      domain_int_sig ~ "Domain × Strain interaction",
      time_int_sig ~ "Time × Strain interaction",
      TRUE ~ "Filtered (QC)"
    )
  )

# Filtered dataset: genes with consistent strain effect (no interactions)
strain_effects_filtered <- strain_effects_classified |>
  filter(passes_strain_filter)


# # source the pattern identification script
source("DEgene_pattern_identification_CAST_B6.R")
# ── Load pattern classification results ────────────────────────────────────
pattern_summary <- read_csv(here("strain_pattern_summary.csv"))
main_patterns <- read_csv(here("main_strain_effect_patterns.csv"))
domain_strain_patterns <- read_csv(here("domain_strain_patterns.csv"))
time_patterns <- read_csv(here("time_strain_patterns.csv"))
threeway_patterns <- read_csv(here("threeway_patterns.csv"))
gene_assignments <- read_csv(here("strain_gene_pattern_assignments.csv"))

# ── Extract gene lists ─────────────────────────────────────────────────────

# Main strain effects (from additive model - average across contexts)
genes_CAST_high_strong <- main_patterns |>
  filter(main_strain_pattern == "CAST >> B6 (strong)") |> pull(gene_id)
genes_CAST_high_mod <- main_patterns |>
  filter(main_strain_pattern == "CAST > B6 (moderate)") |> pull(gene_id)
genes_B6_high_strong <- main_patterns |>
  filter(main_strain_pattern == "B6 >> CAST (strong)") |> pull(gene_id)
genes_B6_high_mod <- main_patterns |>
  filter(main_strain_pattern == "B6 > CAST (moderate)") |> pull(gene_id)
genes_main_only <- main_patterns |>
  filter(!has_any_interaction) |> pull(gene_id)

# Domain × strain patterns (which domain shows strongest strain response)
genes_PM_strongest_strain <- domain_strain_patterns |>
  filter(strongest_domain == "PM") |> pull(gene_id)
genes_MAX_strongest_strain <- domain_strain_patterns |>
  filter(strongest_domain == "MAX") |> pull(gene_id)
genes_POST_strongest_strain <- domain_strain_patterns |>
  filter(strongest_domain == "POST") |> pull(gene_id)
genes_opposite_directions <- domain_strain_patterns |>
  filter(direction_pattern == "Opposite directions") |> pull(gene_id)
genes_high_domain_specificity <- domain_strain_patterns |>
  filter(specificity_strength == "High") |> pull(gene_id)

# Domain × strain: temporal stability
genes_domain_strain_static <- domain_strain_patterns |>
  filter(time_status == "Static") |> pull(gene_id)
genes_domain_strain_time_affected <- domain_strain_patterns |>
  filter(time_status == "Main time effect present") |> pull(gene_id)

# Time × strain patterns (temporal dynamics)
genes_diverging_CAST_faster <- time_patterns |>
  filter(base_slope > 0.1) |> pull(gene_id)
genes_diverging_B6_faster <- time_patterns |>
  filter(base_slope < -0.1) |> pull(gene_id)

# Time × strain: spatial uniformity
genes_time_uniform        <- time_patterns |> filter(!domain_int_sig, !threeway_int_sig) |> pull(gene_id)
genes_time_domain_affected <- time_patterns |> filter(domain_int_sig, !threeway_int_sig)  |> pull(gene_id)
# genes_time_domain_specific removed — handled in threeway section

# Three-way patterns (domain-specific temporal dynamics)
genes_threeway_complex <- threeway_patterns |>
  filter(str_detect(pattern_type, "Complex")) |> pull(gene_id)
genes_threeway_initial <- threeway_patterns |>
  filter(str_detect(pattern_type, "Primarily initial")) |> pull(gene_id)
genes_threeway_dynamic <- threeway_patterns |>
  filter(str_detect(pattern_type, "Primarily dynamic")) |> pull(gene_id)

# ── Pre-compute VST for plotting ──────────────────────────────────────────
vsd_global <- vst(strain_dds_full, blind = FALSE)
```


```{r}
#| label: test-ind-cont
#| eval: false

library(variancePartition)

# 1. Transform counts properly
vsd <- vst(strain_dds_coll, blind = FALSE)

# 2. Run variance partition
form <- ~ (1|ind) + (1|AP_domain) + (1|strain) + LB_stage_from_start

# Use all genes (or sample if too slow)
varPart <- fitExtractVarPartModel(
  exprObj = assay(vsd),
  formula = form,
  data = as.data.frame(colData(strain_dds_coll))
)

# 3. Visualize
plotVarPart(varPart)

# 4. Get specific numbers
vp_summary <- as.data.frame(varPart) %>%
  summarize(
    ind_median = median(ind),
    ind_mean = mean(ind),
        strain_median = median(strain),
    strain_mean = mean(strain),
    domain_median = median(AP_domain),
    domain_mean = mean(AP_domain),
    stage_median = median(LB_stage_from_start),
    stage_mean = mean(LB_stage_from_start)
  )

print(vp_summary)

# 5. Look at genes where individual matters most
high_ind_variance <- as.data.frame(varPart) %>%
  rownames_to_column("gene_id") %>%
  arrange(desc(ind)) %>%
  head(20) %>%
  left_join(all_genes_palate)

cat("\nGenes with highest individual variance:\n")
print(high_ind_variance %>% select(symbol, ind, AP_domain, LB_stage_from_start))

saveRDS(varPart, here("_data/B6_CAST_variancePartition.Rds"))
```



This analysis investigates how gene expression differs between B6 and CAST mouse strains across three palate domains (PM, MAX, POST) and developmental time. I use a statistical model that decomposes strain effects into:

- **Main strain effect**: Average CAST vs B6 difference
- **Domain × Strain interaction**: How strain effect differs between domains
- **Time × Strain interaction**: How strain effect changes per unit developmental time
- **Three-way interaction**: Domain-specific changes in temporal strain dynamics

**Sample preparation and model fitting**

I merge technical replicates arising from re-sequencing of several samples using the `collapseReplicates()` function from DESeq2. Then calculate the relative or centered limb bud (LB) stage (`LB_stage_from_start`) by subtracting the minimum observed stage so that 0 represents the earliest developmental timepoint in the dataset. The domains are collected from the same individual resulting in a repeated measure setup however variance partition analysis showed that individual-level variation accounted for a negligible proportion of transcriptional variance (median < 0.01%, mean 1.5%), therefore I omittede the nesting as an effect in the DeSeq2 model (`ind:AP_domain`). 

I use three statistical models to extract strain and interaction effects. The **additive model** (`~ sex + AP_domain + LB_stage_from_start + strain`) contains no strain interaction terms and is used to estimate the marginal strain effect — the average CAST vs B6 log2 fold change across all domains and timepoints. The **two-way interaction model** (`~ sex + AP_domain + LB_stage_from_start + strain + strain:AP_domain + strain:LB_stage_from_start`) includes domain and time interactions with strain but omits the three-way term; it is used to extract domain-specific strain effects averaged over developmental time and the overall temporal slope averaged across domains. The **full interaction model** (`~ sex + AP_domain + LB_stage_from_start + strain + strain:AP_domain + strain:LB_stage_from_start + strain:AP_domain:LB_stage_from_start`) additionally includes the three-way interaction term and is used to extract three-way interaction coefficients.

**Identifying significant effects with LRT**

I use likelihood ratio tests (LRT) to identify genes with significant strain-related effects. The **overall strain LRT** compares the additive model to a reduced model lacking the strain term (`~ sex + AP_domain + LB_stage_from_start `), identifying any genes with strain-related variation. The **domain × strain LRT** uses the two-way model as the full model and removes only `strain:AP_domain`, testing for static domain-specific strain differences independently of temporal dynamics. The **time × strain LRT** similarly uses the two-way model and removes only `strain:LB_stage_from_start`, testing for temporally diverging strain effects independently of domain structure. The **three-way LRT** compares the full model to the two-way model as the reduced model, testing whether temporal strain dynamics differ across domains. Using the two-way model (rather than the full model) as the basis for the domain and time LRTs ensures that these tests are cleanly separated — a gene with only a three-way interaction will not appear in the domain or time gene lists.

**Extracting effect sizes**

The LRT tests identify **which** genes have significant effects but do not provide effect sizes. To get the magnitude and direction of effects, I extract coefficients using Wald tests:

- **Main strain effect**: From the additive model — represents the average CAST vs B6 log2 fold change across all domains and timepoints.
- **Domain-specific strain effects**: From the two-way model using contrasts. The strain effect in PM is the `strain_CAST_vs_B6` coefficient (PM is the reference level). The strain effect in MAX combines the main effect with the `strainCAST:AP_domainMAX` interaction term, and similarly for POST. Using the two-way model rather than the full model ensures these estimates represent the strain effect in each domain averaged over developmental time, rather than conditional on time = 0.
- **Domain interaction coefficients**: From the two-way model — `strainCAST:AP_domainMAX` and `strainCAST:AP_domainPOST` indicate how much the strain effect in MAX and POST differs from PM, averaged over time.
- **Overall time interaction slope**: From the two-way model — `strainCAST:LB_stage_from_start` represents the average change in log2FC(CAST/B6) per unit of developmental time, averaged across all domains.
- **PM-specific time slope**: From the full model — `strainCAST:LB_stage_from_start` in the full model represents the temporal slope conditional on the PM reference domain, used together with three-way coefficients to reconstruct per-domain slopes.
- **Three-way coefficients**: From the full model — `strainCAST:AP_domainMAX:LB_stage_from_start` and `strainCAST:AP_domainPOST:LB_stage_from_start` indicate how the temporal slope in MAX and POST deviates from the PM slope.

**Quality filtering**

I apply quality filters to each result object individually, removing genes with low expression (baseMean below the 10th percentile), high uncertainty (lfcSE ≥ 1.5 for effect estimates, ≥ 0.5 for slope coefficients), unrealistically large effects (|log2FC| ≥ 10), or very small effects (|log2FC| < 0.5 for main effects, < 0.3 for interaction coefficients, < 0.1 for slope coefficients). This per-result filtering ensures that each effect estimate meets quality standards appropriate for its scale.

**Results summary**

From this analysis, I have identified **`r sum(strain_effects_classified$any_strain_sig, na.rm = TRUE)`** genes with significant strain effects (padj < 0.05). Of these, **`r nrow(strain_effects_filtered)`** have a main strain effect — they pass quality filters on the additive model's strain coefficient and the strain effect LRT. These significant genes are further classified into interaction categories based on their LRT results.

**Pattern classification**

I use the `DEgene_pattern_identification_CAST_B6.R` script to classify differentially expressed genes into expression pattern categories. The classification is hierarchical: genes are first separated into those with consistent effects versus those with interactions, and interaction genes are further sub-classified by the nature of their interaction.

**Domain × Strain interaction patterns** are assigned to genes with a significant domain × strain LRT (padj < 0.05) and at least one domain interaction coefficient that passes quality filters. These genes are classified by which domain shows the largest strain effect magnitude (PM, MAX, or POST) and whether the strain effect direction is consistent or variable across domains. A subset of particular interest are genes with opposite directions — where the strain effect reverses sign between domains. Additionally, these patterns are characterized by their **temporal stability**: (1) Static patterns show stable domain differences throughout development (no time or three-way interactions); (2) Time-affected patterns may shift as both strains change over time (time interaction present, but no three-way interaction); (3) Time-varying patterns actively change their spatial structure during development (three-way interaction significant).

**Time × Strain interaction patterns** are assigned to genes with a significant time × strain LRT and an overall time interaction slope (from the two-way model) that passes quality filters. These are classified as diverging with CAST increasing faster (positive slope > 0.1), diverging with B6 increasing faster (negative slope < -0.1), or parallel/weak. Strong divergence (|slope| > 0.2) is distinguished from moderate divergence. Additionally, these patterns are characterized by their **spatial uniformity**: (1) Uniform patterns show the same temporal dynamics across all domains (no domain or three-way interactions); (2) Domain-affected patterns have domains that differ at t=0 but show parallel temporal trajectories (domain interaction present, no three-way interaction); (3) Domain-specific patterns show different temporal dynamics in different domains (three-way interaction present). For genes with domain-specific dynamics, per-domain slopes are reconstructed by combining the PM-specific slope from the full model with the three-way coefficients.

**Three-way interaction patterns** are assigned to genes with a significant three-way LRT and at least one three-way coefficient passing quality filters. Per-domain temporal slopes are reconstructed by combining the PM-specific slope from the full model with the three-way adjustment coefficients for MAX and POST. These genes are classified based on **whether domains differ primarily in their initial state (t=0), their temporal dynamics (slopes), or both**: (1) Complex patterns show substantial differences across both initial effects and temporal trajectories; (2) Primarily initial patterns have domains that differ at t=0 but follow similar temporal trajectories; (3) Primarily dynamic patterns start similar but diverge over time with domain-specific slopes. Within each pattern type, genes are further characterized by which domain is most distinct when considering both initial state and temporal dynamics combined.

```{r}
#| label: pattern-summary-table
#| message: false
#| warning: false

pattern_summary |> 
  create_dt()
```

# Main Strain Effects

These genes show expression differences between CAST and B6. The log2FC values represent the CAST vs B6 effect estimated from the model. 

:::: {#main-strain-tabs .panel-tabset .nav-pills}

## CAST >> B6 (Strong)

Genes with strong upregulation in CAST (log2FC > 1). Here are some examples:

```{r}
#| label: main-CAST-strong-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6


count_data <- get_normalized_counts(strain_dds_full, genes_CAST_high_strong, vsd = vsd_global)
plot_main_strain_effect(genes_CAST_high_strong, "CAST >> B6 (strong)", count_data, n = 2)


```

Here is the full list of genes:
```{r}
#| label: main-CAST-strong-table
#| message: false
#| warning: false

main_patterns |> 
  filter(gene_id %in% genes_CAST_high_strong) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```

Here is the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-CAST-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_strong_ora <- run_enrichment_hybrid_v2(genes_CAST_high_strong, 
                                                    description = "CAST >> B6 (strong)")

```

```{r}
#| label: main-CAST-strong-ora-table
#| message: false
#| warning: false

if (!is.null(genes_CAST_high_strong_ora) && !is.null(genes_CAST_high_strong_ora$result)) {
  genes_CAST_high_strong_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}

```


## CAST > B6 (Moderate)

Genes with moderate upregulation in CAST (0.5 < log2FC ≤ 1).

```{r}
#| label: main-CAST-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_normalized_counts(strain_dds_full, 
                                    genes_CAST_high_mod, vsd = vsd_global)

plot_main_strain_effect(genes_CAST_high_mod, "CAST > B6 (moderate)", 
                        count_data, n = 2)

```


```{r}
#| label: main-CAST-mod-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_CAST_high_mod) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-CAST-mod-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_mod_ora <- run_enrichment_hybrid_v2(genes_CAST_high_mod,
                                                 description = "CAST > B6 (moderate)")

```

```{r}
#| label: main-CAST-mod-ora-table
#| message: false
#| warning: false

if (!is.null(genes_CAST_high_mod_ora) && !is.null(genes_CAST_high_mod_ora$result)) {
  genes_CAST_high_mod_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}

```

## B6 >> CAST (Strong)

Genes with strong upregulation in B6 (log2FC < -1).

```{r}
#| label: main-B6-strong-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_normalized_counts(strain_dds_full, 
                                    genes_B6_high_strong, 
                                    vsd = vsd_global)
plot_main_strain_effect(genes_B6_high_strong, "B6 >> CAST (strong)", count_data, n = 2)

```

```{r}
#| label: main-B6-strong-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_B6_high_strong) |>
  select(gene_id,symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-B6-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true


genes_B6_high_strong_ora <- run_enrichment_hybrid_v2(genes_B6_high_strong,
                                                  description = "B6 >> CAST (strong)")

```

```{r}
#| label: main-B6-strong-ora-table
#| message: false
#| warning: false


if (!is.null(genes_B6_high_strong_ora) && !is.null(genes_B6_high_strong_ora$result)) {
  genes_B6_high_strong_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}

```


## B6 > CAST (Moderate)

Genes with moderate upregulation in B6 (-1 ≤ log2FC < -0.5).

```{r}
#| label: main-B6-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_B6_high_mod, vsd = vsd_global)
plot_main_strain_effect(genes_B6_high_mod, "B6 > CAST (moderate)", count_data, n = 2)

```

```{r}
#| label: main-B6-mod-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_B6_high_mod) |>
  select(gene_id,symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-B6-mod-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_B6_high_mod_ora <- run_enrichment_hybrid_v2(genes_B6_high_mod, 
                                               description = "B6 > CAST (moderate)")
  
```

```{r}
#| label: main-B6-mod-ora-table
#| message: false
#| warning: false

if (!is.null(genes_B6_high_mod_ora) && !is.null(genes_B6_high_mod_ora$result)) {
  genes_B6_high_mod_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
  
```

::::

# Domain × Strain Interactions

These genes show strain differences that vary across palate domains. The analysis focuses on which domain shows the strongest strain response and whether strain effects are consistent in direction across domains.

:::: {#domain-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-strain-overview
#| message: false
#| warning: false

domain_strain_patterns |> 
  count(domain_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = domain_pattern, `N Genes` = n) |> 
  create_dt()

```

## PM: Strongest Effect

Genes where the strain effect is strongest in the PM domain. 

```{r}
#| label: domain-PM-strongest-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_PM_strongest_strain, vsd = vsd_global)
plot_domain_strain_base(genes_PM_strongest_strain, "Strongest strain effect in PM", count_data, n = 2)

```

```{r}
#| label: domain-PM-strongest-table
#| message: false
#| warning: false

domain_strain_patterns |> 
  filter(gene_id %in% genes_PM_strongest_strain) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_strain_range, specificity_strength, direction_pattern, time_status) |> 
  create_dt()

```

```{r}
#| label: domain-PM-strongest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in PM
genes_PM_strongest_up <- domain_strain_patterns |> 
  filter(gene_id %in% genes_PM_strongest_strain, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_PM_strongest_down <- domain_strain_patterns |> 
  filter(gene_id %in% genes_PM_strongest_strain, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_PM_strongest_effect_ora_up <- run_enrichment_hybrid_v2(
  genes_PM_strongest_up, 
  description = "PM strongest effect - Up in CAST"
)

genes_PM_strongest_effect_ora_down <- run_enrichment_hybrid_v2(
  genes_PM_strongest_down, 
  description = "PM strongest effect - Down in CAST"
)
  
```

```{r}
#| label: domain-PM-strongest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_PM_strongest_effect_ora_up$result)) genes_PM_strongest_effect_ora_up$result,
  if (!is.null(genes_PM_strongest_effect_ora_down$result)) genes_PM_strongest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}
  
```

## MAX: Strongest Effect

Genes where the strain effect is strongest in the MAX domain. 

```{r}
#| label: domain-MAX-strongest-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_MAX_strongest_strain, vsd = vsd_global)
plot_domain_strain_base(genes_MAX_strongest_strain, "Strongest strain effect in MAX", count_data, n = 2)

```

```{r}
#| label: domain-MAX-strongest-table
#| message: false
#| warning: false

domain_strain_patterns |> 
  filter(gene_id %in% genes_MAX_strongest_strain) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_strain_range, specificity_strength, direction_pattern, time_status) |> 
  create_dt()

```

```{r}
#| label: domain-MAX-strongest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in MAX
genes_MAX_strongest_up <- domain_strain_patterns |> 
  filter(gene_id %in% genes_MAX_strongest_strain, strain_effect_MAX > 0) |> 
  pull(gene_id)

genes_MAX_strongest_down <- domain_strain_patterns |> 
  filter(gene_id %in% genes_MAX_strongest_strain, strain_effect_MAX < 0) |> 
  pull(gene_id)

genes_MAX_strongest_effect_ora_up <- run_enrichment_hybrid_v2(
  genes_MAX_strongest_up, 
  description = "MAX strongest effect - Up in CAST"
)

genes_MAX_strongest_effect_ora_down <- run_enrichment_hybrid_v2(
  genes_MAX_strongest_down, 
  description = "MAX strongest effect - Down in CAST"
)
  
```

```{r}
#| label: domain-MAX-strongest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_MAX_strongest_effect_ora_up$result)) genes_MAX_strongest_effect_ora_up$result,
  if (!is.null(genes_MAX_strongest_effect_ora_down$result)) genes_MAX_strongest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}
  
```

## POST: Strongest Effect

Genes where the strain effect is strongest in the POST domain. 

```{r}
#| label: domain-POST-strongest-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_POST_strongest_strain, vsd = vsd_global)
plot_domain_strain_base(genes_POST_strongest_strain, "Strongest strain effect in POST", count_data, n = 2)

```

```{r}
#| label: domain-POST-strongest-table
#| message: false
#| warning: false

domain_strain_patterns |> 
  filter(gene_id %in% genes_POST_strongest_strain) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_strain_range, specificity_strength, direction_pattern, time_status) |> 
  create_dt()

```

```{r}
#| label: domain-POST-strongest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in POST
genes_POST_strongest_up <- domain_strain_patterns |> 
  filter(gene_id %in% genes_POST_strongest_strain, strain_effect_POST > 0) |> 
  pull(gene_id)

genes_POST_strongest_down <- domain_strain_patterns |> 
  filter(gene_id %in% genes_POST_strongest_strain, strain_effect_POST < 0) |> 
  pull(gene_id)

genes_POST_strongest_effect_ora_up <- run_enrichment_hybrid_v2(
  genes_POST_strongest_up, 
  description = "POST strongest effect - Up in CAST"
)

genes_POST_strongest_effect_ora_down <- run_enrichment_hybrid_v2(
  genes_POST_strongest_down, 
  description = "POST strongest effect - Down in CAST"
)

```

```{r}
#| label: domain-POST-strongest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_POST_strongest_effect_ora_up$result)) genes_POST_strongest_effect_ora_up$result,
  if (!is.null(genes_POST_strongest_effect_ora_down$result)) genes_POST_strongest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## High Domain Specificity

Genes with particularly strong domain-specific strain effects (range > 1.0 log2FC). These show dramatic differences in how strongly strain effects manifest across domains.

```{r}
#| label: domain-high-specificity-plot
#| message: false
#| warning: false
#| fig-height: 15
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_high_domain_specificity, vsd = vsd_global)
plot_domain_specificity(genes_high_domain_specificity, "High domain specificity", count_data, n = 2)

```

```{r}
#| label: domain-high-specificity-table
#| message: false
#| warning: false

domain_strain_patterns |> 
  filter(gene_id %in% genes_high_domain_specificity) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_strain_range, strongest_domain, weakest_domain, direction_pattern, time_status) |> 
  arrange(desc(domain_strain_range)) |>
  create_dt()

```

```{r}
#| label: domain-high-specificity-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_high_domain_specificity_ora <- run_enrichment_hybrid_v2(
  genes_high_domain_specificity, 
  description = "High domain specificity"
)

```

```{r}
#| label: domain-high-specificity-ora-table
#| message: false
#| warning: false

if (!is.null(genes_high_domain_specificity_ora$result)) {
  genes_high_domain_specificity_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Opposite Directions

Genes where the strain effect reverses direction across domains — for example, higher in CAST in one domain but higher in B6 in another. These are particularly interesting as they suggest domain-specific regulatory divergence between strains.

```{r}
#| label: domain-opposite-plot
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 10

count_data <- get_normalized_counts(strain_dds_full, genes_opposite_directions, vsd = vsd_global)
plot_domain_strain_profile(genes_opposite_directions, "Opposite directions across domains", count_data, n = 2)

```

```{r}
#| label: domain-opposite-table
#| message: false
#| warning: false

domain_strain_patterns |> 
  filter(gene_id %in% genes_opposite_directions) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_strain_range, strongest_domain, direction_pattern, time_status) |> 
  create_dt()

```

```{r}
#| label: domain-opposite-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction using PM as reference
genes_opposite_up_in_PM <- domain_strain_patterns |> 
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_opposite_down_in_PM <- domain_strain_patterns |> 
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_opposite_directions_ora_up <- run_enrichment_hybrid_v2(
  genes_opposite_up_in_PM, 
  description = "Opposite directions - Up in CAST in PM"
)

genes_opposite_directions_ora_down <- run_enrichment_hybrid_v2(
  genes_opposite_down_in_PM, 
  description = "Opposite directions - Down in CAST in PM"
)
 
```

```{r}
#| label: domain-opposite-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_opposite_directions_ora_up$result)) genes_opposite_directions_ora_up$result,
  if (!is.null(genes_opposite_directions_ora_down$result)) genes_opposite_directions_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Static Pattern

Domain differences that remain stable over developmental time (no time or three-way interaction). These represent fixed spatial differences in strain effects.

```{r}
#| label: domain-static-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_domain_strain_static, vsd = vsd_global)
plot_domain_strain_temporal(genes_domain_strain_static, "Static domain pattern", count_data, n = 2)
```

```{r}
#| label: domain-static-table
#| message: false
#| warning: false

domain_strain_patterns |> 
  filter(gene_id %in% genes_domain_strain_static) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         strongest_domain, direction_pattern, specificity_strength, time_status) |> 
  create_dt()

```

## Time-Affected Pattern

Domain differences where main time effects are present, so the spatial pattern may shift over development, but without domain-specific temporal dynamics (no three-way interaction).

```{r}
#| label: domain-time-affected-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_domain_strain_time_affected, vsd = vsd_global)
plot_domain_strain_temporal(genes_domain_strain_time_affected, "Time-affected domain pattern", count_data, n = 2)

```

```{r}
#| label: domain-time-affected-table
#| message: false
#| warning: false

domain_strain_patterns |> 
  filter(gene_id %in% genes_domain_strain_time_affected) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         strongest_domain, direction_pattern, specificity_strength, time_status) |> 
  create_dt()

```

::::

# Time × Strain Interactions

These genes show strain differences that change over developmental time. The temporal slope coefficient represents the change in log2FC(CAST/B6) per unit of developmental time.

:::: {#time-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: time-strain-overview
#| message: false
#| warning: false

time_patterns |> 
  count(detailed_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |> 
  create_dt()

```

## Diverging: CAST Faster

Genes where strains diverge over developmental time, with CAST expression increasing (or decreasing less) relative to B6 as development progresses. The positive time interaction slope indicates that the log2FC(CAST/B6) becomes more positive over time.

```{r}
#| label: time-CAST-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_diverging_CAST_faster, vsd = vsd_global)
plot_time_strain_interaction(genes_diverging_CAST_faster, "Diverging: CAST changing faster", count_data, n = 2)

```

```{r}
#| label: time-CAST-faster-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster) |>
  select(gene_id, symbol, base_slope, time_int_pval, main_strain_lfc, 
         trajectory_pattern, spatial_pattern, threeway_int_sig) |>
  create_dt()

```

```{r}
#| label: time-CAST-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_CAST_faster_up <- time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc > 0) |> 
  pull(gene_id)

genes_CAST_faster_down <- time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc < 0) |> 
  pull(gene_id)

genes_diverging_CAST_faster_ora_up <- run_enrichment_hybrid_v2(
  genes_CAST_faster_up, 
  description = "CAST faster - Up in CAST at baseline"
)

genes_diverging_CAST_faster_ora_down <- run_enrichment_hybrid_v2(
  genes_CAST_faster_down, 
  description = "CAST faster - Down in CAST at baseline"
)

```

```{r}
#| label: time-CAST-faster-ora-table
#| message: false
#| warning: false

combined_diverging_cast_ora_results <- bind_rows(
  genes_diverging_CAST_faster_ora_up$result,
  genes_diverging_CAST_faster_ora_down$result
) 

if (!is.null(combined_diverging_cast_ora_results) && nrow(combined_diverging_cast_ora_results) > 0) {
  combined_diverging_cast_ora_results |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Diverging: B6 Faster

Genes where strains diverge over developmental time, with B6 expression increasing (or decreasing less) relative to CAST as development progresses. The negative time interaction slope indicates that the log2FC(CAST/B6) becomes more negative over time.

```{r}
#| label: time-B6-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_diverging_B6_faster, vsd = vsd_global)
plot_time_strain_interaction(genes_diverging_B6_faster, "Diverging: B6 changing faster", count_data, n = 2)

```

```{r}
#| label: time-B6-faster-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster) |>
  select(gene_id, symbol, base_slope, time_int_pval, main_strain_lfc, 
         trajectory_pattern, spatial_pattern,threeway_int_sig) |> 
  create_dt()

```

```{r}
#| label: time-B6-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_B6_faster_up <- time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc > 0) |> 
  pull(gene_id)

genes_B6_faster_down <- time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc < 0) |> 
  pull(gene_id)

genes_diverging_B6_faster_ora_up <- run_enrichment_hybrid_v2(
  genes_B6_faster_up, 
  description = "B6 faster - Up in CAST at baseline"
)

genes_diverging_B6_faster_ora_down <- run_enrichment_hybrid_v2(
  genes_B6_faster_down, 
  description = "B6 faster - Down in CAST at baseline"
)

```

```{r}
#| label: time-B6-faster-ora-table
#| message: false
#| warning: false

combined_diverging_b6_ora_results <- bind_rows(
  genes_diverging_B6_faster_ora_up$result,
  genes_diverging_B6_faster_ora_down$result
) 

if (!is.null(combined_diverging_b6_ora_results) && nrow(combined_diverging_b6_ora_results) > 0) {
  combined_diverging_b6_ora_results |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Uniform Across Domains

Temporal strain dynamics that are consistent across all palate domains (no domain or three-way interaction). These represent spatially uniform temporal changes.

```{r}
#| label: time-uniform-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_time_uniform, vsd = vsd_global)
plot_time_strain_interaction(genes_time_uniform, "Uniform temporal dynamics", count_data, n = 2)

```

```{r}
#| label: time-uniform-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_time_uniform) |>
  select(gene_id, symbol, base_slope, main_strain_lfc, trajectory_pattern, spatial_pattern, threeway_int_sig) |> 
  create_dt()

```

## Domain-Affected

Temporal dynamics where domains differ at t=0 but show parallel trajectories over time (domain interaction present, but no three-way interaction).

```{r}
#| label: time-domain-affected-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_time_domain_affected, vsd = vsd_global)
plot_domain_strain_temporal(genes_time_domain_affected, "Domain-affected temporal dynamics", count_data, n = 2)

```

```{r}
#| label: time-domain-affected-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_time_domain_affected) |>
  select(gene_id, symbol, base_slope, main_strain_lfc, trajectory_pattern, spatial_pattern, threeway_int_sig) |> 
  create_dt()

```

::::

# Three-Way Interactions

These genes show domain-specific strain differences that also change over developmental time. We classify them based on whether domains primarily differ in their initial state (t=0), their temporal dynamics (slopes), or both.

:::: {#threeway-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: threeway-overview
#| message: false
#| warning: false

threeway_patterns |> 
  count(detailed_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |> 
  create_dt()

```

## Complex Pattern

Genes where domains differ in BOTH their initial state (t=0) AND their temporal dynamics. These show substantial differences across both dimensions.

```{r}
#| label: threeway-complex-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_threeway_complex, vsd = vsd_global)
plot_time_strain_overview(genes_threeway_complex, "Complex: Differ in initial state & dynamics", count_data, n = 2)

```

```{r}
#| label: threeway-complex-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex) |>
  select(gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST, 
         init_range, slope_range, most_distinct_domain) |> 
  create_dt()

```

```{r}
#| label: threeway-complex-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_complex_PM <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "PM") |> 
  pull(gene_id)

genes_complex_MAX <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "MAX") |> 
  pull(gene_id)

genes_complex_POST <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "POST") |> 
  pull(gene_id)

genes_threeway_complex_ora_PM <- run_enrichment_hybrid_v2(
  genes_complex_PM, 
  description = "Complex pattern - PM most distinct"
)

genes_threeway_complex_ora_MAX <- run_enrichment_hybrid_v2(
  genes_complex_MAX, 
  description = "Complex pattern - MAX most distinct"
)

genes_threeway_complex_ora_POST <- run_enrichment_hybrid_v2(
  genes_complex_POST, 
  description = "Complex pattern - POST most distinct"
)

```

```{r}
#| label: threeway-complex-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_complex_ora_PM$result)) {
    genes_threeway_complex_ora_PM$result |> 
      mutate(distinct_domain = paste0("PM (n=", length(genes_complex_PM), ")"))
  },
  if (!is.null(genes_threeway_complex_ora_MAX$result)) {
    genes_threeway_complex_ora_MAX$result |> 
      mutate(distinct_domain = paste0("MAX (n=", length(genes_complex_MAX), ")"))
  },
  if (!is.null(genes_threeway_complex_ora_POST$result)) {
    genes_threeway_complex_ora_POST$result |> 
      mutate(distinct_domain = paste0("POST (n=", length(genes_complex_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(distinct_domain, everything()) |> 
    create_dt()
}

```

## Primarily Initial Differences

Genes where domains differ mainly at t=0, but show similar temporal trajectories. The spatial pattern is established early and maintained over time.

```{r}
#| label: threeway-initial-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_threeway_initial, vsd = vsd_global)
plot_time_strain_overview(genes_threeway_initial, "Primarily initial differences", count_data, n = 2)

```

```{r}
#| label: threeway-initial-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial) |>
  select(gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST, 
         init_range, slope_range, most_distinct_domain) |> 
  create_dt()

```

```{r}
#| label: threeway-initial-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_initial_PM <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "PM") |> 
  pull(gene_id)

genes_initial_MAX <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "MAX") |> 
  pull(gene_id)

genes_initial_POST <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "POST") |> 
  pull(gene_id)

genes_threeway_initial_ora_PM <- run_enrichment_hybrid_v2(
  genes_initial_PM, 
  description = "Initial differences - PM most distinct"
)

genes_threeway_initial_ora_MAX <- run_enrichment_hybrid_v2(
  genes_initial_MAX, 
  description = "Initial differences - MAX most distinct"
)

genes_threeway_initial_ora_POST <- run_enrichment_hybrid_v2(
  genes_initial_POST, 
  description = "Initial differences - POST most distinct"
)

```

```{r}
#| label: threeway-initial-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_initial_ora_PM$result)) {
    genes_threeway_initial_ora_PM$result |> 
      mutate(distinct_domain = paste0("PM (n=", length(genes_initial_PM), ")"))
  },
  if (!is.null(genes_threeway_initial_ora_MAX$result)) {
    genes_threeway_initial_ora_MAX$result |> 
      mutate(distinct_domain = paste0("MAX (n=", length(genes_initial_MAX), ")"))
  },
  if (!is.null(genes_threeway_initial_ora_POST$result)) {
    genes_threeway_initial_ora_POST$result |> 
      mutate(distinct_domain = paste0("POST (n=", length(genes_initial_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(distinct_domain, everything()) |> 
    create_dt()
}

```

## Primarily Dynamic Differences

Genes where domains start similar at t=0 but diverge over time with different temporal trajectories. The spatial pattern emerges during development.

```{r}
#| label: threeway-dynamic-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_threeway_dynamic, vsd = vsd_global)
plot_time_strain_overview(genes_threeway_dynamic, "Primarily dynamic differences", count_data, n = 2)

```

```{r}
#| label: threeway-dynamic-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic) |>
  select(gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST, 
         init_range, slope_range, most_distinct_domain) |> 
  create_dt()

```

```{r}
#| label: threeway-dynamic-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_dynamic_PM <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "PM") |> 
  pull(gene_id)

genes_dynamic_MAX <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "MAX") |> 
  pull(gene_id)

genes_dynamic_POST <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "POST") |> 
  pull(gene_id)

genes_threeway_dynamic_ora_PM <- run_enrichment_hybrid_v2(
  genes_dynamic_PM, 
  description = "Dynamic differences - PM most distinct"
)

genes_threeway_dynamic_ora_MAX <- run_enrichment_hybrid_v2(
  genes_dynamic_MAX, 
  description = "Dynamic differences - MAX most distinct"
)

genes_threeway_dynamic_ora_POST <- run_enrichment_hybrid_v2(
  genes_dynamic_POST, 
  description = "Dynamic differences - POST most distinct"
)

```

```{r}
#| label: threeway-dynamic-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_dynamic_ora_PM$result)) {
    genes_threeway_dynamic_ora_PM$result |> 
      mutate(distinct_domain = paste0("PM (n=", length(genes_dynamic_PM), ")"))
  },
  if (!is.null(genes_threeway_dynamic_ora_MAX$result)) {
    genes_threeway_dynamic_ora_MAX$result |> 
      mutate(distinct_domain = paste0("MAX (n=", length(genes_dynamic_MAX), ")"))
  },
  if (!is.null(genes_threeway_dynamic_ora_POST$result)) {
    genes_threeway_dynamic_ora_POST$result |> 
      mutate(distinct_domain = paste0("POST (n=", length(genes_dynamic_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(distinct_domain, everything()) |> 
    create_dt()
}

```

::::

# Comprehensive Tables

This section provides downloadable tables summarizing all analysis results.

:::: {#comprehensive-tables-tabs .panel-tabset .nav-pills}

## Gene-Level Pattern Assignments

All genes with their assigned expression patterns across strain, domain, and time.

```{r}
#| label: comprehensive-gene-assignments
#| message: false
#| warning: false

gene_assignments |>
  left_join(all_genes_palate |> select(gene_id, symbol, gene_biotype, chromosome)) |>
  select(gene_id, symbol, gene_biotype, chromosome, everything()) |>
  create_dt()

```

## Full Strain Effects Classification

Complete results table with all coefficients and classifications.

```{r}
#| label: comprehensive-strain-effects
#| message: false
#| warning: false

strain_effects_classified |>
  filter(any_strain_sig) |>
  select(
    gene_id, symbol, gene_biotype, baseMean,
    # Pattern classifications
    pattern_type, main_strain_pattern, 
    # Main strain effect
    main_strain_lfc, main_strain_pval,
    # Domain-specific effects at t=0
    strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    strain_effect_PM_pass, strain_effect_MAX_pass, strain_effect_POST_pass,
    # Time effects
    time_int_slope_avg, time_pattern,
    # Significance flags
    domain_int_sig, time_int_sig, threeway_int_sig
  ) |>
  create_dt()
```

## Combined ORA Results

All over-representation analysis results combined into a single table.

```{r}
#| label: comprehensive-ora-combine
#| message: false
#| warning: false

# Combine all ORA results into one table
all_ora_results <- bind_rows(
  # Main strain effects
  if (!is.null(genes_CAST_high_strong_ora$result)) 
    genes_CAST_high_strong_ora$result |> 
      mutate(category = "Main Effect", pattern = "CAST >> B6 (strong)", direction = "Up in CAST"),
  if (!is.null(genes_CAST_high_mod_ora$result)) 
    genes_CAST_high_mod_ora$result |> 
      mutate(category = "Main Effect", pattern = "CAST > B6 (moderate)", direction = "Up in CAST"),
  if (!is.null(genes_B6_high_strong_ora$result)) 
    genes_B6_high_strong_ora$result |> 
      mutate(category = "Main Effect", pattern = "B6 >> CAST (strong)", direction = "Down in CAST"),
  if (!is.null(genes_B6_high_mod_ora$result)) 
    genes_B6_high_mod_ora$result |> 
      mutate(category = "Main Effect", pattern = "B6 > CAST (moderate)", direction = "Down in CAST"),
  
  # Domain effects - PM strongest
  if (!is.null(genes_PM_strongest_effect_ora_up$result)) 
    genes_PM_strongest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "PM strongest effect", direction = "Up in CAST"),
  if (!is.null(genes_PM_strongest_effect_ora_down$result)) 
    genes_PM_strongest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "PM strongest effect", direction = "Down in CAST"),
  
  # Domain effects - MAX strongest
  if (!is.null(genes_MAX_strongest_effect_ora_up$result)) 
    genes_MAX_strongest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "MAX strongest effect", direction = "Up in CAST"),
  if (!is.null(genes_MAX_strongest_effect_ora_down$result)) 
    genes_MAX_strongest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "MAX strongest effect", direction = "Down in CAST"),
  
  # Domain effects - POST strongest
  if (!is.null(genes_POST_strongest_effect_ora_up$result)) 
    genes_POST_strongest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "POST strongest effect", direction = "Up in CAST"),
  if (!is.null(genes_POST_strongest_effect_ora_down$result)) 
    genes_POST_strongest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "POST strongest effect", direction = "Down in CAST"),
  
  # Domain effects - High specificity
  if (!is.null(genes_high_domain_specificity_ora$result)) 
    genes_high_domain_specificity_ora$result |> 
      mutate(category = "Domain Effect", pattern = "High domain specificity", direction = "All"),
  
  # Domain effects - Opposite directions
  if (!is.null(genes_opposite_directions_ora_up$result)) 
    genes_opposite_directions_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "Opposite directions", direction = "Up in CAST (PM)"),
  if (!is.null(genes_opposite_directions_ora_down$result)) 
    genes_opposite_directions_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "Opposite directions", direction = "Down in CAST (PM)"),
  
  # Time effects - CAST faster
  if (!is.null(genes_diverging_CAST_faster_ora_up$result)) 
    genes_diverging_CAST_faster_ora_up$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: CAST faster", direction = "Up in CAST"),
  if (!is.null(genes_diverging_CAST_faster_ora_down$result)) 
    genes_diverging_CAST_faster_ora_down$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: CAST faster", direction = "Down in CAST"),
  
  # Time effects - B6 faster
  if (!is.null(genes_diverging_B6_faster_ora_up$result)) 
    genes_diverging_B6_faster_ora_up$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: B6 faster", direction = "Up in CAST"),
  if (!is.null(genes_diverging_B6_faster_ora_down$result)) 
    genes_diverging_B6_faster_ora_down$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: B6 faster", direction = "Down in CAST"),
  
  # Three-way effects - Complex pattern
  if (!is.null(genes_threeway_complex_ora_PM$result)) 
    genes_threeway_complex_ora_PM$result |> 
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "PM most distinct"),
  if (!is.null(genes_threeway_complex_ora_MAX$result)) 
    genes_threeway_complex_ora_MAX$result |> 
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "MAX most distinct"),
  if (!is.null(genes_threeway_complex_ora_POST$result)) 
    genes_threeway_complex_ora_POST$result |> 
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "POST most distinct"),
  
  # Three-way effects - Primarily initial
  if (!is.null(genes_threeway_initial_ora_PM$result)) 
    genes_threeway_initial_ora_PM$result |> 
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "PM most distinct"),
  if (!is.null(genes_threeway_initial_ora_MAX$result)) 
    genes_threeway_initial_ora_MAX$result |> 
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "MAX most distinct"),
  if (!is.null(genes_threeway_initial_ora_POST$result)) 
    genes_threeway_initial_ora_POST$result |> 
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "POST most distinct"),
  
  # Three-way effects - Primarily dynamic
  if (!is.null(genes_threeway_dynamic_ora_PM$result)) 
    genes_threeway_dynamic_ora_PM$result |> 
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "PM most distinct"),
  if (!is.null(genes_threeway_dynamic_ora_MAX$result)) 
    genes_threeway_dynamic_ora_MAX$result |> 
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "MAX most distinct"),
  if (!is.null(genes_threeway_dynamic_ora_POST$result)) 
    genes_threeway_dynamic_ora_POST$result |> 
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "POST most distinct")
)

```

```{r}
#| label: comprehensive-ora-table
#| message: false
#| warning: false

all_ora_results |>
  filter(p_value < 0.01) |>
  select(category, pattern, direction, source, term_name, p_value, 
         intersection_size, term_size, intersection) |>
  arrange(category, pattern, direction, p_value) |>
  create_dt()

```

::::

# Session Info

```{r}
#| label: session-info
#| message: false
#| warning: false

sessionInfo()
```