---
title: "Investigating strain, domain and developmental stage effects"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
      embed-resources: true
      standalone: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
code-fold: true
toc: true
toc-depth: 5
toc-expand: true
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r setup}
#| message: false
#| warning: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)
# load packages
suppressPackageStartupMessages(library(tidyverse))

library(pheatmap)
library(plotly)
library(GGally)
library(ggpubr)
library(ggplot2)
library(corrplot)
library(simplifyEnrichment)
library(patchwork)

library(DESeq2)
library(DEGreport)
library(gprofiler2)
set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e108_eg55_p17/")
library(BioNERO)
set.seed(123) # for reproducibility
# library(WGCNA)

library(DT)

library(here)

select <- dplyr::select # I am adding this explicitly
rename <- dplyr::rename # I am adding this explicitly
filter <- dplyr::filter # I am adding this explicitly
summarize <- dplyr::summarize # I am adding this explicitly

# functions
create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                rownames = FALSE, 
                filter="top",
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel'),
                               pageLength = 5, 
                               scrollX= TRUE
                               ))
  
}

# reading in the sample details
sample_details <- read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 

# domain colors
domain_colors <- c(PM=rgb(252,206,93, maxColorValue = 255), 
                   MAX=rgb(163,82,157, maxColorValue = 255), 
                   POST=rgb(158,156,201, maxColorValue = 255)
                   )

# strain colors
strain_colors <- c( B6 = "#555555", 
                    CAST = "#009E73",
                    DO = rgb(163,82,157, maxColorValue = 255))



```

```{r plotting-functions}
# ===== FUNCTION 1: COMPARE UP VS DOWN IN SAME DOMAIN =====
plot_domain_direction_comparison <- function(genes_up, genes_down, 
                                             domain_name, count_data,
                                             n_each = 2) {
  
  gene_col <- if("symbol" %in% names(count_data)) "symbol" else "symbol"
  
  # Get examples
  up_examples <- genes_up[1:min(n_each, length(genes_up))]
  down_examples <- genes_down[1:min(n_each, length(genes_down))]
  
  # Create plots for UP genes
  plots_up <- lapply(up_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("↑ in ", domain_name),
           x = "Development stage",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold", size = 11),
            plot.subtitle = element_text(color = "darkgreen"))
  })
  
  # Create plots for DOWN genes
  plots_down <- lapply(down_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("↓ in ", domain_name),
           x = "Development stage",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold", size = 11),
            plot.subtitle = element_text(color = "darkred"))
  })
  
  # Remove NULLs
  plots_up <- plots_up[!sapply(plots_up, is.null)]
  plots_down <- plots_down[!sapply(plots_down, is.null)]
  
  # Combine
  all_plots <- c(plots_up, plots_down)
  if(length(all_plots) == 0) return(NULL)
  
  if(length(all_plots) == 2) {
    combined <- all_plots[[1]] / all_plots[[2]]
  } else if(length(all_plots) == 4) {
    combined <- (all_plots[[1]] | all_plots[[2]]) / (all_plots[[3]] | all_plots[[4]])
  } else {
    combined <- wrap_plots(all_plots, ncol = 2)
  }
  
  combined <- combined +
    plot_annotation(
      title = paste0(domain_name, ": Higher (↑) vs Lower (↓)"),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 2: PLOT DOMAIN HIERARCHY =====
plot_domain_hierarchy <- function(genes_highest, domain_name, count_data, n = 3) {
  
  gene_col <- if("symbol" %in% names(count_data)) "symbol" else "symbol"
  
  example_genes <- genes_highest[1:min(n, length(genes_highest))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Create both trajectory and boxplot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("Highest in ", domain_name),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8),
            plot.title = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = AP_domain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values = domain_colors) +
      labs(x = "", y = "") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none")
    
    return(p1 + p2 + plot_layout(widths = c(2, 1)))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Genes Highest in ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 3: PLOT A-P GRADIENTS =====
plot_AP_gradient <- function(genes_decreasing, genes_increasing, count_data, n_each = 3) {
  
  gene_col <- if("symbol" %in% names(count_data)) "symbol" else "symbol"
  
  # Get examples
  dec_examples <- genes_decreasing[1:min(n_each, length(genes_decreasing))]
  inc_examples <- genes_increasing[1:min(n_each, length(genes_increasing))]
  
  # Function to create gradient plot
  make_gradient_plot <- function(gid, gradient_type) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Summary by domain
    domain_summary <- gene_data |> 
      group_by(AP_domain) |> 
      summarize(mean_count = mean(count, na.rm = TRUE), .groups = "drop")
    
    ggplot(domain_summary, aes(x = AP_domain, y = mean_count, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_point(data = gene_data, aes(y = count), alpha = 0.3, size = 1.5,
                 position = position_jitter(width = 0.1)) +
      scale_fill_manual(values = domain_colors) +
      scale_y_log10(labels = scales::comma) +
      labs(title = gene_name,
           subtitle = ifelse(gradient_type == "dec", "PM > MAX > POST", "POST > MAX > PM"),
           x = "",
           y = "Expression") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = ifelse(gradient_type == "dec", 
                                                        "darkred", "darkgreen")))
  }
  
  # Create plots
  plots_dec <- lapply(dec_examples, function(gid) make_gradient_plot(gid, "dec"))
  plots_inc <- lapply(inc_examples, function(gid) make_gradient_plot(gid, "inc"))
  
  # Remove NULLs
  plots_dec <- plots_dec[!sapply(plots_dec, is.null)]
  plots_inc <- plots_inc[!sapply(plots_inc, is.null)]
  
  # Combine
  all_plots <- c(plots_dec, plots_inc)
  if(length(all_plots) == 0) return(NULL)
  
  combined <- wrap_plots(all_plots, ncol = 3)
  combined <- combined +
    plot_annotation(
      title = "Anterior-Posterior Expression Gradients",
      subtitle = "Top: A→P decreasing | Bottom: A→P increasing",
      theme = theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12)
      )
    )
  
  return(combined)
}

# ===== FUNCTION 4: PLOT DOMAIN-SPECIFIC DYNAMICS =====
plot_domain_specific_dynamics <- function(gene_ids, domain_name, direction,
                                          count_data, slope_data, n = 3) {
  
  gene_col <- if("symbol" %in% names(count_data)) "symbol" else "symbol"
  
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    # Get count data
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Get slope data
    gene_slopes <- slope_data |> 
      filter(gene_id == gid) |> 
      mutate(Domain = factor(Domain, levels = c("PM", "MAX", "POST")))
    
    # Trajectory plot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(direction, " in ", domain_name, " only"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"))
    
    # Slope plot
    if(nrow(gene_slopes) > 0) {
      p2 <- ggplot(gene_slopes, aes(x = Domain, y = log2FoldChange, fill = Domain)) +
        geom_col(alpha = 0.8) +
        geom_errorbar(aes(ymin = log2FoldChange - lfcSE,
                          ymax = log2FoldChange + lfcSE),
                      width = 0.3) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_text(aes(label = ifelse(padj < 0.001, "***",
                                     ifelse(padj < 0.01, "**",
                                            ifelse(padj < 0.05, "*", "ns")))),
                  vjust = -0.5, size = 4) +
        scale_fill_manual(values = domain_colors) +
        labs(y = "Slope", x = "") +
        theme_pubclean(base_size = 10) +
        theme(legend.position = "none")
      
      return(p1 + p2 + plot_layout(widths = c(2, 1)))
    } else {
      return(p1)
    }
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Domain-Specific Dynamics: ", domain_name, " ", direction),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 5: PLOT DIVERGING PATTERNS =====
plot_diverging_detailed <- function(genes_faster, genes_opposite, 
                                    domain_name, count_data, n_each = 2) {
  
  gene_col <- if("symbol" %in% names(count_data)) "symbol" else "symbol"
  
  faster_examples <- genes_faster[1:min(n_each, length(genes_faster))]
  opposite_examples <- genes_opposite[1:min(n_each, length(genes_opposite))]
  
  # Plot "faster" pattern
  plots_faster <- lapply(faster_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(domain_name, " increases faster"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = "darkgreen", size = 9))
  })
  
  # Plot "opposite" pattern
  plots_opposite <- lapply(opposite_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(domain_name, " ↑, other ↓"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = "purple", size = 9))
  })
  
  # Remove NULLs
  plots_faster <- plots_faster[!sapply(plots_faster, is.null)]
  plots_opposite <- plots_opposite[!sapply(plots_opposite, is.null)]
  
  all_plots <- c(plots_faster, plots_opposite)
  if(length(all_plots) == 0) return(NULL)
  
  if(length(all_plots) <= 2) {
    combined <- wrap_plots(all_plots, ncol = 1)
  } else {
    combined <- wrap_plots(all_plots, ncol = 2)
  }
  
  combined <- combined +
    plot_annotation(
      title = paste0("Diverging Patterns: ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 6: PLOT SIMPLE TRAJECTORIES =====
plot_simple_trajectories <- function(gene_ids, count_data, title_text, n = 5) {
  
  gene_col <- if("symbol" %in% names(count_data)) "symbol" else "symbol"
  
  # Sample if needed
  if(length(gene_ids) > n) {
    set.seed(123)
    gene_ids <- sample(gene_ids, n)
  }
  
  plots <- lapply(gene_ids, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8),
            plot.title = element_text(face = "bold", size = 11))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 3)
  combined <- combined +
    plot_annotation(
      title = title_text,
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}
```

```{r load_data_emase}
#| message: false
#| warning: false

# loading the count data
load("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/palate_gbrs_emase_counts_v4.RData") # raw_expr_mat, expr_mat_do, expr_mat_nondo 
expr_mat <- expr_mat_nondo
contam_samples <- c("B6_12.5_PM_1","B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat[ ,!colnames(expr_mat) %in% contam_samples]
do_expr_mat <- expr_mat_do
remove_samples <- c("DO_13.5_MAX_6", "DO_13.5_MAX_14", "DO_13.5_MAX_14_2",
                    "DO_12.5_MAX_1" , "DO_12.5_MAX_2" ,"DO_13.5_MAX_97", "DO_13.5_MAX_96") # 96 & 97 are also 12.5 samples
do_expr_mat <- do_expr_mat[ ,!colnames(do_expr_mat) %in% remove_samples]

expr_mat_all <- cbind( expr_mat[intersect(rownames(expr_mat),rownames(do_expr_mat)),,drop=F],
                       do_expr_mat[intersect(rownames(expr_mat),rownames(do_expr_mat)),,drop=F])
rm(expr_mat_do, expr_mat_nondo)

# Get v105 annotations from biomart
all.genes_v105 <- read_tsv(here("_data","ENSMUSGid_to_symbol_v105.txt"),
                           col_names = c("gene_id","symbol","gene_biotype","chromosome","start","end"), skip =1) |>  
  mutate(midpoint = (start+end)/2)

all_genes_palate <- all.genes_v105 %>%
  filter(gene_id %in% rownames(raw_expr_mat))

all_genes_palate_filtered <- all.genes_v105 %>%
  filter(gene_id %in% rownames(expr_mat_all))

# sample annotations
sample_annot <- sample_details %>% 
  #as_tibble( rownames = "sample") |> 
  filter( !sample %in% contam_samples & sample %in% colnames(expr_mat)) %>% 
  separate(Strain_stock, into = c("strain","tmp"), remove = F, sep = "_") %>% 
  select(sample,sex, AP_domain,Gest_stage, strain,batch, Harvest_date,LB_stage, palate_closure, L_R_palate) %>% 
  mutate( strain = case_when(strain =="B6J"~"B6",
                             strain !="B6J"~strain)) |> 
  mutate( LB_stage_binned = case_when(
    ( LB_stage < 12.25 )~"exclude",
    (between(LB_stage, 13.25, 13.5))~"exclude",
    (between(LB_stage, 12.25,12.75))~"1",
    (between(LB_stage, 13.5, 14))~"2",
    (between(LB_stage, 14.25, 14.75))~"3",
    (between(LB_stage, 15.1, 15.6))~"4"
  )) |> 
  mutate( LB_stage_binned = as_factor(LB_stage_binned)) |> 
  separate( L_R_palate, into = c("temp","Palate_ori" ), sep = "_") |> 
  select(-temp) |> 
  mutate( Palate_ori = ifelse( Palate_ori == "L", "Left","Right")) 
  

sample_annot_all <- sample_details %>% 
  #as_tibble( rownames = "sample") |> 
  filter( !sample %in% contam_samples & sample %in% colnames(expr_mat_all)) %>% 
  separate(Strain_stock, into = c("strain","tmp"), remove = F, sep = "_") %>% 
  select(sample,sex, AP_domain,Gest_stage, strain,batch, Harvest_date,LB_stage, palate_closure,L_R_palate) %>% 
  mutate( strain = case_when(strain =="B6J"~"B6",
                             strain !="B6J"~strain)) |> 
  separate( L_R_palate, into = c("temp","Palate_ori" ), sep = "_") |> 
  select(-temp) |> 
  mutate( Palate_ori = ifelse( Palate_ori == "L", "Left","Right")) |> 
  mutate( LB_stage_binned = case_when(
    ( LB_stage < 12.25 )~"exclude",
    (between(LB_stage, 13.25, 13.5))~"exclude",
    (between(LB_stage, 12.25,12.75))~"1",
    (between(LB_stage, 13.5, 14))~"2",
    (between(LB_stage, 14.25, 14.75))~"3",
    (between(LB_stage, 15.1, 15.6))~"4"
  )) 

# Genes from Ian:
pheno_gene_list <- readxl::read_xlsx(here("_data","Supp6_BoneGeneList.xlsx")) |> 
  select( symbol = gene_name) |> 
  distinct() |> 
  mutate(type = "pheno", type_long = "bone phenotype gene list (Sabik et al 2020, Table S6)")

skeletal_gene_table <- readxl::read_xlsx(here("_data","Skeletal_stem_progenitor_markers.xlsx")) |> 
  dplyr::rename("symbol" = "gene") |> 
  mutate(type = "MSC", type_long = "MSC/skeletal progenitor markers")

figure_genes <- c(MSC =c("Axin2","Gli1","Meis2","Lepr","Prrx1"),
           skeletal = c("Sox9","Runx2","Sp7","Nfatc1","Bcl11b","Msx2","Msx1","Egr1","Egr2"),
           osteoclast = c("Spi1","Tnfsf11","Tnfrsf11","PParg","Fos","Junb","Csf1","Csf1r","Ostf1",
                          "Acp5","Mmp9","Ctsk","Timp1","Atp6v0d2","Adgre1","Sfrp4","Dcstamp","Ocstamp","Htra1","Tcirg1","Itgb3","Gnptab"),
           osteoblast = c("Alpl","Ibsp","Phospho1","Vdr","Sparc","Panx3","Enpp1","Bglap","Bglap2","Clecl11a","Sost","Dmp1","Col1a1","Col1a2","Bgn","Dcn","Thbs2")
           
           ) |> 
  as_tibble(rownames = "type" ) |> 
  rename( symbol = value  ) |> 
  mutate(type = gsub("[0-9]","",type)) |> 
  mutate( type_long = case_when(type =="MSC"~"MSC/skeletal progenitor markers",
                                type =="skeletal"~"Skeletal speciation markers",
                                type =="osteoclast"~"Osteoclast markers",
                                type =="osteoblast"~"osteoblast/osteocyte markers"
                                )) 

morpho_genes <- readxl::read_xlsx(path = here("_data","abnormal craniofacial morphology-MGIBatchReport.xlsx")) |> 
  select(symbol = Symbol, mgi_id = `MGI Gene/Marker ID`, gene_biotype = `Feature Type`) 

# add gene names to MGI ids for genes
# checked the overlap
# length(intersect(mgi_id_annots$symbol, all_genes_palate$symbol)) # 16958/17146
# most of the missing ones are RPXX-YY genes or Gms
mgi_id_annots <- read_delim( file = here("_data","MRK_List1.rpt"),
                             #MGI Marker Accession ID	Chromosome	cM Position	Genome Coordinate Start	Genome Coordinate End	Genome Strand	Marker Symbol	Status	Marker Name	Marker Type	Feature Types (|-delimted)	Marker Synonyms (|-delimited)
                             col_names = c("mgi_id",
                                           "chromosome",
                                           "pos_cM",
                                           "start",
                                           "end",
                                           "strand",
                                           "symbol",
                                           "status",
                                           "marker_name",
                                           "marker_type",
                                           "feature_types",
                                           "marker_synms"), skip = 1) 

# getting the gene list for mammalian phenotypes
mp_genes <- read_delim( file = here("_data","MGI_PhenoGenoMP.rpt"),
                        # Allelic Composition	Allele Symbol(s)	Genetic Background	Mammalian Phenotype ID	PubMed ID	MGI Marker Accession ID (comma-delimited)
                        col_names = c("allelic_comp",
                                      "allele_symbol",
                                      "genetic_background",
                                      "mp_id",
                                      "pubmed_id",
                                      "mgi_id")) |> 
  select(mp_id, mgi_id) |> 
  distinct() |> 
  separate_longer_delim( mgi_id, delim = "|") |> 
  left_join(mgi_id_annots |> 
              select(mgi_id, symbol)) 
# add mammalian phenotype definitions
mp_defs <- read_delim( file = here("_data","VOC_MammalianPhenotype.rpt"),
                       col_names = c("mp_id","term_name","term_details")) 



genes <- full_join( figure_genes, skeletal_gene_table ) |> 
  #full_join( pheno_gene_list) |> 
  left_join(all_genes_palate)
# 
# genes |> 
#   group_by(symbol) |> 
#   mutate(n = n()) |> 
#   filter(n > 1) |> 
#   arrange(symbol)

tfs <- read_tsv(file = here("_data/search_result_from_tfcheckpoint.txt")) |> 
  filter( !is.na(Gene_Name)) |> 
  select( symbol = Gene_symbol) |> 
  mutate( symbol = tools::toTitleCase(tolower(symbol)) ) |> 
  distinct() |> 
  left_join( all_genes_palate)

```


# Comparing B6 and CAST strains across AP domains and developmental time

::: {#B6_CAST_DESeq2_results .panel-tabset .nav-pills}

## Genes differentially expressed between B6 and CAST (no interaction with time or domain included)

```{r}
#| label: b6_cast_comparison
#| warning: false  
#| message: false

b6_cast_samples <- sample_annot %>%
  filter(strain %in% c("B6","CAST")) %>%
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = F) %>%
  select(-e,-d) %>%
  mutate( rep = ifelse( is.na(rep),1,rep)) %>%
  unite( "unique_sampleid" , c(s,ind,AP_domain,Gest_stage),sep ="_", remove = F)

coldata <- b6_cast_samples %>%
  select(sample,unique_sampleid, sex,ind, rep, AP_domain, Gest_stage, strain, LB_stage) %>%
  #filter( rep ==1) %>%
  filter(!is.na(sex) ) %>%
  mutate( AP_domain = factor(AP_domain, levels = c("PM","MAX","POST")) ) %>%
  column_to_rownames("sample")

cts <- round(raw_expr_mat[all_genes_palate$gene_id,rownames(coldata), drop = F])

# let's make the dds object then collapse the technical replicates
strain_dds <- DESeqDataSetFromMatrix( countData = cts,
                                      colData = coldata,
                                      design = ~1)

strain_dds_coll <- collapseReplicates(strain_dds, strain_dds$unique_sampleid)

# https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#levels-without-samples
# the domains are from the same animal! So I need to add that to the covariate matrix + try to add that info to deseq design. see:
# https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#group-specific-condition-effects-individuals-nested-within-groups
# I am not adding interactions for sex, mainly because we can't really parse them out and also I don't believe there to be significant interactions
strain_simple_model <- model.matrix(~ strain + AP_domain + LB_stage + AP_domain:ind,
                                  data = colData(strain_dds_coll))
# all.zero <- apply(strain_simple_model, 2, function(x) all(x==0))
# idx <- which(all.zero)
# strain_simple_model <- strain_simple_model[,-idx]
strain_dds_simple <- DESeq(strain_dds_coll, betaPrior = FALSE, full=strain_simple_model)


# too complex! I will do this within each domain instead.
# strain_comp_model <-  model.matrix(~ strain + AP_domain + LB_stage + AP_domain:ind +
#                                      strain:AP_domain + strain:LB_stage + LB_stage:AP_domain +
#                                      strain:AP_domain:LB_stage,
#                                   data = colData(strain_dds_coll))
# strain_dds_comp <- DESeq(strain_dds_coll, betaPrior = FALSE, full=strain_comp_model)
# 
# # Run LRT:
# strain_dds_comp_simple_lrt <- DESeq(strain_dds_coll, test="LRT",
#                                 full= strain_comp_model,
#                                 reduced = strain_simple_model)


strain_gene_counts <-  counts(strain_dds_simple, normalized=TRUE) %>%
  as_tibble( rownames = "gene_id") %>%
  pivot_longer(2:110, names_to = "sample", values_to = "count") %>%
  left_join( colData(strain_dds_simple) %>%
               as_tibble( rownames = "sample")
             )
```


```{r}
#| label: b6_cast_strain_eff_ora
#| cache: true
#| warning: false
#| message: false

strain_eff_genes <- results(strain_dds_simple, name = "strainCAST" ) %>%
  as_tibble( rownames = "gene_id") %>%
  filter( padj < 0.05, abs(log2FoldChange) > .5) %>%
  left_join(all_genes_palate) %>%
  mutate( Comparison = ifelse( `log2FoldChange` < 0,
          "Lower expression in CAST",
          "Higher expression in CAST")
  )

g.strain_eff_cast_up <- gost( query = unique( (strain_eff_genes %>%
                                                filter(log2FoldChange > 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_eff_cast_up$result <- g.strain_eff_cast_up$result %>% filter(term_size < 660)


g.strain_eff_cast_down <- gost( query = unique( (strain_eff_genes %>%
                                                filter(log2FoldChange < 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_eff_cast_down$result <- g.strain_eff_cast_down$result %>% filter(term_size < 660)

```

There are `r formatC(length(unique(strain_eff_genes$gene_id)), big.mark = ",")` genes that show a strain effect (i.e differentially expressed between CAST and B6). Note that we are not including any interaction terms and some of these genes are likely to show variability in their difference between the strains over time or across domains.

```{r}
#| label: b6_cast_strain_eff_genes_ex_plot
#| warning: false
#| message: false
#| fig-width: 20
#| fig-height: 4
strain_eff_genes  |> 
  inner_join(genes) |> 
  select(gene_id,symbol,Comparison,type_long) |> 
  distinct() |> 
  left_join(strain_gene_counts) |> 
  ggplot()+
  aes(x = strain, 
      y = count, 
      col = strain,
      group=interaction(gene_id,strain)
      )+
  #geom_boxplot( width = 0.2)+
  geom_point()+
  geom_boxplot(outlier.size = 0)+
  geom_jitter( aes(col = strain, shape =strain))+
  #geom_smooth(method = "lm")+
  scale_color_manual(values = strain_colors)+
  theme_pubclean()+
  theme(legend.position = "none")+
  scale_y_log10()+
  ylab("Normalized gene expression counts (au)")+
  #ggtitle("Domain effect (no interaction)")+
  facet_wrap(~symbol, scales = "free", nrow = 1)

```


```{r}
#| label: b6_cast_strain_eff_genes_table
#| warning: false
#| message: false

strain_eff_genes  %>%
  select(Comparison, gene_id, symbol, chromosome , padj, log2FoldChange) %>%
  mutate( padj = formatC(padj, digits = 2,format = "e"),
          log2FoldChange = formatC(log2FoldChange, digits = 2, format = "g")) %>%
  create_dt()

```


```{r}
#| label: b6_cast_strain_eff_genes_ora_table
#| warning: false
#| message: false

g.strain_eff_cast_up$result %>%
  mutate( Category = "Higher expression in CAST") %>%
  rbind(
    g.strain_eff_cast_down$result %>%
      mutate(Category = "Lower expression in CAST")
  )  %>%
  select( Category, term_name, source, p_value, term_size, intersection_size,intersection) %>%  # intersecion: adding the list of genes identified as overlapping with each category
  mutate_if( is.numeric, formatC, digits =2) %>%
  create_dt()

```


### Genes with different expression and trends over time in each domain between B6 and CAST

Since the full model is too complex leading to many pairwise comparisons, I instead decided to run the comparison between the simple model without interactions and the complex one with interactions for each AP domain. Mainly addressing two questions:

Which genes are differentially expressed between B6 and CAST in PM/MAX/POST domains that is consistent over developmental time (no interaction between strain and LB stage)?

Which genes are differentially expressed in PM/MAX/POST domains between B6 and CAST AND show different trends over developmental time (with interaction between strain and LB stage)?

```{r}
#| label: b6_cast_comparison_over_domains
#| message: false
#| warning: false

# pm domain
# subset samples to pm domain
b6_cast_samples_pm <- sample_annot %>%
  filter(strain %in% c("B6","CAST")) %>%
  filter( AP_domain == "PM") %>%
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = F) %>%
  select(-e,-d) %>%
  mutate( rep = ifelse( is.na(rep),1,rep)) %>%
  unite( "unique_sampleid" , c(s,ind,AP_domain,Gest_stage),sep ="_", remove = F)
coldata_pm <- b6_cast_samples_pm %>%
  select(sample,unique_sampleid, sex,ind, rep, AP_domain, Gest_stage, strain, LB_stage) %>%
  #filter( rep ==1) %>%
  filter(!is.na(sex) ) %>%
  #mutate( AP_domain = factor(AP_domain, levels = c("PM","MAX","POST")) ) %>%
  column_to_rownames("sample")
cts_pm <- round(raw_expr_mat[all_genes_palate$gene_id,rownames(coldata_pm), drop = F])
# let's make the dds object then collapse the technical replicates
strain_dds_pm <- DESeqDataSetFromMatrix( countData = cts_pm,
                                         colData = coldata_pm,
                                         design = ~1)
strain_dds_coll_pm <- collapseReplicates(strain_dds_pm, strain_dds_pm$unique_sampleid)

# prep model wo interaction
strain_simple_model_pm <- model.matrix(~ strain + LB_stage + strain:ind,
                                       data = colData(strain_dds_coll_pm))
all.zero <- apply(strain_simple_model_pm, 2, function(x) all(x==0))
idx <- which(all.zero)
strain_simple_model_pm <- strain_simple_model_pm[,-idx]
strain_dds_simple_pm <- DESeq(strain_dds_coll_pm, betaPrior = FALSE, full=strain_simple_model_pm)

# prep model with interaction
strain_int_model_pm <- model.matrix(~ strain + LB_stage + strain:ind + strain:LB_stage,
                                    data = colData(strain_dds_coll_pm))
all.zero <- apply(strain_int_model_pm, 2, function(x) all(x==0))
idx <- which(all.zero)
strain_int_model_pm <- strain_int_model_pm[,-idx]
strain_dds_int_pm <- DESeq(strain_dds_coll_pm, betaPrior = FALSE, full=strain_int_model_pm) # will get the logfold change + p values from this model

# run lrt
# comparing the full model with the interaction term between stage + domain to the reduced model without the interaction.
strain_dds_pm_lrt <- DESeq(strain_dds_coll_pm, test="LRT",
                                full= strain_int_model_pm,
                                reduced = strain_simple_model_pm)

# resultsNames(strain_dds_pm_lrt)
# Let's get the gene list that has a significant difference with/without the interaction term hence is likely to show an interaction effect!
strain_dds_pm_lrt_genes <- results( strain_dds_pm_lrt) %>%
  as_tibble( rownames = "gene_id") %>%
  filter( padj < 0.05) %>%
  left_join( all_genes_palate)

# resultsNames(strain_dds_simple_pm)
# get genes
# see this link here: https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#time-course-experiments
# from what I can decipher: 
# using strainCAST+strainCAST.LB_stage will give me genes where the difference between CAST and B6 are changing over time. 
# I am only interested in the interaction term answering the last question above! Note, that is the same term we are getting the results for from the LRT too! 
strain_by_time_pm_genes <- results(strain_dds_int_pm, contrast = list(c("strainCAST.LB_stage")) ) %>%
  as_tibble( rownames = "gene_id") %>%
  filter( gene_id %in% strain_dds_pm_lrt_genes$gene_id,
          padj < 0.05,
          abs(log2FoldChange) > 0.5
          ) |> 
  left_join(all_genes_palate)
# Genes with strain effects without an interaction with LB stage:
strain_pm_genes <- results( strain_dds_simple_pm, contrast = list(c("strainCAST"))) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(
    ! gene_id %in% strain_dds_pm_lrt_genes$gene_id,
    padj < 0.05,
    abs(log2FoldChange) > 0.5
  ) |> 
  left_join(all_genes_palate) |> 
  mutate( Comparison = case_when( log2FoldChange > 0 ~ "Higher expression in CAST",
                                  log2FoldChange < 0 ~ "Lower expression in CAST")
          )
# get gene counts
strain_dds_pm_gene_counts <- counts(strain_dds_int_pm, normalized=TRUE)[unique(strain_dds_pm_lrt_genes$gene_id),] %>%
  as_tibble( rownames = "gene_id") %>%
  pivot_longer(2:ncol(.), names_to = "sample", values_to = "count") %>%
  left_join(
    colData(strain_dds_int_pm) %>%
      as_tibble(rownames = "sample") %>%
      select(sample,strain, LB_stage, AP_domain)
  )

# max domain
# subset samples to max domain
b6_cast_samples_max <- sample_annot %>%
  filter(strain %in% c("B6","CAST")) %>%
  filter( AP_domain == "MAX") %>%
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = F) %>%
  select(-e,-d) %>%
  mutate( rep = ifelse( is.na(rep),1,rep)) %>%
  unite( "unique_sampleid" , c(s,ind,AP_domain,Gest_stage),sep ="_", remove = F)
coldata_max <- b6_cast_samples_max %>%
  select(sample,unique_sampleid, sex,ind, rep, AP_domain, Gest_stage, strain, LB_stage) %>%
  #filter( rep ==1) %>%
  filter(!is.na(sex) ) %>%
  #mutate( AP_domain = factor(AP_domain, levels = c("PM","MAX","POST")) ) %>%
  column_to_rownames("sample")
cts_max <- round(raw_expr_mat[all_genes_palate$gene_id,rownames(coldata_max), drop = F])
# let's make the dds object then collapse the technical replicates
strain_dds_max <- DESeqDataSetFromMatrix( countData = cts_max,
                                         colData = coldata_max,
                                         design = ~1)
strain_dds_coll_max <- collapseReplicates(strain_dds_max, strain_dds_max$unique_sampleid)

# prep model wo interaction
strain_simple_model_max <- model.matrix(~ strain + LB_stage + strain:ind,
                                       data = colData(strain_dds_coll_max))
all.zero <- apply(strain_simple_model_max, 2, function(x) all(x==0))
idx <- which(all.zero)
strain_simple_model_max <- strain_simple_model_max[,-idx]
strain_dds_simple_max <- DESeq(strain_dds_coll_max, betaPrior = FALSE, full=strain_simple_model_max)

# prep model with interaction
strain_int_model_max <- model.matrix(~ strain + LB_stage + strain:ind + strain:LB_stage,
                                    data = colData(strain_dds_coll_max))
all.zero <- apply(strain_int_model_max, 2, function(x) all(x==0))
idx <- which(all.zero)
strain_int_model_max <- strain_int_model_max[,-idx]
strain_dds_int_max <- DESeq(strain_dds_coll_max, betaPrior = FALSE, full=strain_int_model_max) # will get the logfold change + p values from this model

# run lrt
# comparing the full model with the interaction term between stage + domain to the reduced model without the interaction.
strain_dds_max_lrt <- DESeq(strain_dds_coll_max, test="LRT",
                           full= strain_int_model_max,
                           reduced = strain_simple_model_max)

# resultsNames(strain_dds_max_lrt)
# Let's get the gene list that has a significant difference with/without the interaction term hence is likely to show an interaction effect!
strain_dds_max_lrt_genes <- results( strain_dds_max_lrt) %>%
  as_tibble( rownames = "gene_id") %>%
  filter( padj < 0.05) %>%
  left_join( all_genes_palate)

# resultsNames(strain_dds_simple_max)
# get genes
strain_by_time_max_genes <- results(strain_dds_int_max, contrast = list(c("strainCAST.LB_stage")) ) %>%
  as_tibble( rownames = "gene_id") %>%
  filter( gene_id %in% strain_dds_max_lrt_genes$gene_id,
          padj < 0.05,
          abs(log2FoldChange) > 0.5
  ) %>%
  # mutate(
  #   Comparison = ifelse( log2FoldChange >0, "Higher expression in CAST", "Lower expression in CAST")
  # ) %>%
  left_join(all_genes_palate)
# Genes with strain effects without an interaction with LB stage:
strain_max_genes <- results( strain_dds_simple_max, contrast = list(c("strainCAST"))) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(
    ! gene_id %in% strain_dds_max_lrt_genes$gene_id,
    padj < 0.05,
    abs(log2FoldChange) > 0.5
  ) |> 
  left_join(all_genes_palate) |> 
  mutate( Comparison = case_when( log2FoldChange > 0 ~ "Higher expression in CAST",
                                  log2FoldChange < 0 ~ "Lower expression in CAST")
          )
# get gene counts
strain_dds_max_gene_counts <- counts(strain_dds_int_max, normalized=TRUE)[unique(strain_dds_max_lrt_genes$gene_id),] %>%
  as_tibble( rownames = "gene_id") %>%
  pivot_longer(2:ncol(.), names_to = "sample", values_to = "count") %>%
  left_join(
    colData(strain_dds_int_max) %>%
      as_tibble(rownames = "sample") %>%
      select(sample,strain, LB_stage, AP_domain)
  )

# post domain
# subset samples to post domain
b6_cast_samples_post <- sample_annot %>%
  filter(strain %in% c("B6","CAST")) %>%
  filter( AP_domain == "POST") %>%
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = F) %>%
  select(-e,-d) %>%
  mutate( rep = ifelse( is.na(rep),1,rep)) %>%
  unite( "unique_sampleid" , c(s,ind,AP_domain,Gest_stage),sep ="_", remove = F)
coldata_post <- b6_cast_samples_post %>%
  select(sample,unique_sampleid, sex,ind, rep, AP_domain, Gest_stage, strain, LB_stage) %>%
  #filter( rep ==1) %>%
  filter(!is.na(sex) ) %>%
  #mutate( AP_domain = factor(AP_domain, levels = c("PM","MAX","POST")) ) %>%
  column_to_rownames("sample")
cts_post <- round(raw_expr_mat[all_genes_palate$gene_id,rownames(coldata_post), drop = F])
# let's make the dds object then collapse the technical replicates
strain_dds_post <- DESeqDataSetFromMatrix( countData = cts_post,
                                         colData = coldata_post,
                                         design = ~1)
strain_dds_coll_post <- collapseReplicates(strain_dds_post, strain_dds_post$unique_sampleid)

# prep model wo interaction
strain_simple_model_post <- model.matrix(~ strain + LB_stage + strain:ind,
                                       data = colData(strain_dds_coll_post))
all.zero <- apply(strain_simple_model_post, 2, function(x) all(x==0))
idx <- which(all.zero)
strain_simple_model_post <- strain_simple_model_post[,-idx]
strain_dds_simple_post <- DESeq(strain_dds_coll_post, betaPrior = FALSE, full=strain_simple_model_post)

# prep model with interaction
strain_int_model_post <- model.matrix(~ strain + LB_stage + strain:ind + strain:LB_stage,
                                    data = colData(strain_dds_coll_post))
all.zero <- apply(strain_int_model_post, 2, function(x) all(x==0))
idx <- which(all.zero)
strain_int_model_post <- strain_int_model_post[,-idx]
strain_dds_int_post <- DESeq(strain_dds_coll_post, betaPrior = FALSE, full=strain_int_model_post) # will get the logfold change + p values from this model

# run lrt
# comparing the full model with the interaction term between stage + domain to the reduced model without the interaction.
strain_dds_post_lrt <- DESeq(strain_dds_coll_post, test="LRT",
                           full= strain_int_model_post,
                           reduced = strain_simple_model_post)

# resultsNames(strain_dds_post_lrt)
# Let's get the gene list that has a significant difference with/without the interaction term hence is likely to show an interaction effect!
strain_dds_post_lrt_genes <- results( strain_dds_post_lrt) %>%
  as_tibble( rownames = "gene_id") %>%
  filter( padj < 0.05) %>%
  left_join( all_genes_palate)

# resultsNames(strain_dds_simple_post)
# get genes
strain_by_time_post_genes <- results(strain_dds_int_post, contrast = list(c("strainCAST.LB_stage")) ) %>%
  as_tibble( rownames = "gene_id") %>%
  filter( gene_id %in% strain_dds_post_lrt_genes$gene_id,
          padj < 0.05,
          abs(log2FoldChange) > 0.5
  ) %>%
  left_join(all_genes_palate)
# Genes with strain effects without an interaction with LB stage:
strain_post_genes <- results( strain_dds_simple_post, contrast = list(c("strainCAST"))) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(
    ! gene_id %in% strain_dds_post_lrt_genes$gene_id,
    padj < 0.05,
    abs(log2FoldChange) > 0.5
  ) |> 
  left_join(all_genes_palate) |> 
  mutate( Comparison = case_when( log2FoldChange > 0 ~ "Higher expression in CAST",
                                  log2FoldChange < 0 ~ "Lower expression in CAST")
          )
# get gene counts
strain_dds_post_gene_counts <- counts(strain_dds_int_post, normalized=TRUE)[unique(strain_dds_post_lrt_genes$gene_id),] %>%
  as_tibble( rownames = "gene_id") %>%
  pivot_longer(2:ncol(.), names_to = "sample", values_to = "count") %>%
  left_join(
    colData(strain_dds_int_post) %>%
      as_tibble(rownames = "sample") %>%
      select(sample,strain, LB_stage, AP_domain)
  )

```

::: {#B6_CAST_DESeq2_results2 .panel-tabset .nav-pills}

#### PM domain

There are `r formatC(length(unique(strain_pm_genes$gene_id)), big.mark = ",")`genes that are differentially expressed between B6 and CAST in the PM domain that do not have an interaction with developmental time (i.e. the difference between the strains are consistent over time). On the other hand, there are `r length(unique(strain_by_time_pm_genes$gene_id))` genes differentially expressed in the PM domain between B6 and CAST that also show difference in the trends over developmental time.

##### Genes differentially expressed between B6 and CAST (no interaction with LB stage)

```{r}
#| warning: false
#| message: false
#| fig-width: 20
#| fig-height: 4

strain_pm_genes |> 
  inner_join(genes) |> 
  select(gene_id,symbol,Comparison,type_long) |> 
  distinct() |> 
  left_join(strain_gene_counts) |> 
  filter(AP_domain =="PM") |> 
  ggplot()+
  aes(x = LB_stage, 
      y = count, 
      col = strain,
      group=interaction(gene_id,strain)
      )+
  #geom_boxplot( width = 0.2)+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_color_manual(values = strain_colors)+
  theme_pubclean()+
  scale_y_log10()+
  ylab("Normalized gene expression counts (au)")+
  #ggtitle("Domain effect (no interaction)")+
  facet_wrap(~AP_domain+symbol, scales = "free", nrow = 1)
```

As you can see in the examples above, the trend over time for each gene is highly similar over developmental time.

```{r}
#| label: strain_pm_ora
#| warning: false
#| message: false
#| cache: true

# ora
g.strain_pm_up_genes <- gost( query = unique( (strain_pm_genes %>%
                                                filter(log2FoldChange > 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_pm_up_genes$result <- g.strain_pm_up_genes$result %>% filter(term_size < 660)


g.strain_pm_down_genes <- gost( query = unique( (strain_pm_genes %>%
                                                filter(log2FoldChange < 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_pm_down_genes$result <- g.strain_pm_down_genes$result %>% filter(term_size < 660)

```

```{r}
#| warning: false
#| message: false

strain_pm_genes  %>%
  left_join( all_genes_palate) |> 
  select(gene_id, symbol, chromosome , padj, log2FoldChange) %>%
  mutate( Comparison = ifelse( log2FoldChange >0, "Higher expression in CAST in PM","Lower expression in CAST in PM")) |> 
  mutate( padj = formatC(padj, digits = 2,format = "e"),
          log2FoldChange = formatC(log2FoldChange, digits = 2, format = "g")) %>%
  create_dt()

```

```{r}
#| warning: false
#| message: false

g.strain_pm_up_genes$result %>%
  mutate( Category = "Log2FoldChange >0") %>%
  rbind(
    g.strain_pm_down_genes$result %>%
      mutate(Category = "Log2FoldChange <0")
  )  %>%
  select( Category, term_name, source, p_value, term_size, intersection_size,intersection) %>%  # intersecion: adding the list of genes identified as overlapping with each category
  mutate_if( is.numeric, formatC, digits =2) %>%
  create_dt()

```


##### Genes differentially expressed with different trends over time (with strain by LB stage interaction)

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4

strain_by_time_pm_genes |> 
  inner_join(genes) |> 
  select(gene_id,symbol,type_long,log2FoldChange) |> 
  distinct() |> 
  left_join(strain_gene_counts) |> 
  filter(AP_domain =="PM") |> 
  mutate( log2foldchange = round(log2FoldChange,2)) |> 
  ggplot()+
  aes(x = LB_stage, 
      y = count, 
      col = strain,
      group=interaction(gene_id,strain)
      )+
  #geom_boxplot( width = 0.2)+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_color_manual(values = strain_colors)+
  theme_pubclean()+
  scale_y_log10()+
  ylab("Normalized gene expression counts (au)")+
  #ggtitle("Domain effect (no interaction)")+
  facet_wrap(~AP_domain+symbol+log2foldchange, scales = "free", nrow = 2, ncol = 3)

```

The example genes here show different trends over time, the slopes of the CAST line is higher than the B6 line on the left for *Itgb3* leading to a positive fold change whereas the opposite is observed for *Mmp9* leading to a negative fold change. Currently I am grouping genes based on the sign of the fold changes although I am not sure if that biologically makes sense. Let me know which genes you would like to investigate further and we can look at them closer instead.

```{r}
#| label: strain_by_time_pm_ora
#| warning: false
#| message: false
#| cache: true


# ora
g.strain_by_time_pm_up_genes <- gost( query = unique( (strain_by_time_pm_genes %>%
                                                filter(log2FoldChange > 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_by_time_pm_up_genes$result <- g.strain_by_time_pm_up_genes$result %>% filter(term_size < 660)


g.strain_by_time_pm_down_genes <- gost( query = unique( (strain_by_time_pm_genes %>%
                                                filter(log2FoldChange < 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_by_time_pm_down_genes$result <- g.strain_by_time_pm_down_genes$result %>% filter(term_size < 660)
```

```{r}

strain_by_time_pm_genes  %>%
  left_join( all_genes_palate) |> 
  select(gene_id, symbol, chromosome , padj, log2FoldChange) %>%
  mutate( padj = formatC(padj, digits = 2,format = "e"),
          log2FoldChange = formatC(log2FoldChange, digits = 2, format = "g")) %>%
  create_dt()

```

```{r}
g.strain_by_time_pm_up_genes$result %>%
  mutate( Category = "Log2FoldChange >0") %>%
  rbind(
    g.strain_by_time_pm_down_genes$result %>%
      mutate(Category = "Log2FoldChange <0")
  )  %>%
  select( Category, term_name, source, p_value, term_size, intersection_size,intersection) %>%  # intersecion: adding the list of genes identified as overlapping with each category
  mutate_if( is.numeric, formatC, digits =2) %>%
  create_dt()

```

#### MAX domain

There are `r formatC(length(unique(strain_max_genes$gene_id)),big.mark = ",")` genes that are differentially expressed between B6 and CAST in the MAX domain that do not have an interaction with developmental time (i.e. the difference between the strains are consistent over time). On the other hand, there are `r length(unique(strain_by_time_max_genes$gene_id))` genes differentially expressed in the MAX domain between B6 and CAST that also show difference in the trends over developmental time.

##### Genes differentially expressed between B6 and CAST (no interaction with LB stage)

```{r}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 4
strain_max_genes |> 
  inner_join(genes) |> 
  select(gene_id,symbol,Comparison,type_long) |> 
  distinct() |> 
  left_join(strain_gene_counts) |> 
  filter(AP_domain =="MAX") |> 
  ggplot()+
  aes(x = LB_stage, 
      y = count, 
      col = strain,
      group=interaction(gene_id,strain)
      )+
  #geom_boxplot( width = 0.2)+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_color_manual(values = strain_colors)+
  theme_pubclean()+
  scale_y_log10()+
  ylab("Normalized gene expression counts (au)")+
  #ggtitle("Domain effect (no interaction)")+
  facet_wrap(~AP_domain+symbol, scales = "free", nrow =1)
```

As you can see in the examples above, the trend over time for each gene is highly similar over developmental time.

```{r}
#| label: strain_max_ora
#| warning: false
#| message: false
#| cache: true

# ora
g.strain_max_up_genes <- gost( query = unique( (strain_max_genes %>%
                                                filter(log2FoldChange > 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_max_up_genes$result <- g.strain_max_up_genes$result %>% filter(term_size < 660)


g.strain_max_down_genes <- gost( query = unique( (strain_max_genes %>%
                                                filter(log2FoldChange < 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_max_down_genes$result <- g.strain_max_down_genes$result %>% filter(term_size < 660)

```

```{r}
#| warning: false
#| message: false

strain_max_genes  %>%
  left_join( all_genes_palate) |> 
  select(gene_id, symbol, chromosome , padj, log2FoldChange) %>%
  mutate( Comparison = ifelse( log2FoldChange >0, "Higher expression in CAST in MAX","Lower expression in CAST in MAX")) |> 
  mutate( padj = formatC(padj, digits = 2,format = "e"),
          log2FoldChange = formatC(log2FoldChange, digits = 2, format = "g")) %>%
  create_dt()

```

```{r}
#| warning: false
#| message: false

g.strain_max_up_genes$result %>%
  mutate( Category = "Log2FoldChange >0") %>%
  rbind(
    g.strain_max_down_genes$result %>%
      mutate(Category = "Log2FoldChange <0")
  )  %>%
  select( Category, term_name, source, p_value, term_size, intersection_size,intersection) %>%  # intersecion: adding the list of genes identified as overlapping with each category
  mutate_if( is.numeric, formatC, digits =2) %>%
  create_dt()

```

##### Genes differentially expressed with different trends over time (with strain by LB stage interaction)

```{r}
#| warning: false
#| message: false
#| fig-width: 16
#| fig-height: 4
strain_by_time_max_genes |> 
  inner_join(genes) |> 
  select(gene_id,symbol,type_long,log2FoldChange) |> 
  distinct() |> 
  left_join(strain_gene_counts) |> 
  filter(AP_domain =="MAX") |> 
  mutate( log2foldchange = round(log2FoldChange,2)) |> 
  ggplot()+
  aes(x = LB_stage, 
      y = count, 
      col = strain,
      group=interaction(gene_id,strain)
      )+
  #geom_boxplot( width = 0.2)+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_color_manual(values = strain_colors)+
  theme_pubclean()+
  scale_y_log10()+
  ylab("Normalized gene expression counts (au)")+
  #ggtitle("Domain effect (no interaction)")+
  facet_wrap(~AP_domain+symbol+log2foldchange, scales = "free", nrow = 1)
```

The example genes here show different trends over time, where the slopes of the CAST line is higher than the B6 line it leads to a positive fold change (ex: Itgb3) and where the opposite is observed leads to a negative fold change (ex: Acp5, Dcn, Mmp9). Currently I am grouping genes based on the sign of the fold changes although I am not sure if that biologically makes sense. Let me know which genes you would like to investigate further and we can look at them closer instead.

```{r}
#| label: strain_by_time_max_ora
#| warning: false
#| message: false
#| cache: true

# ora
g.strain_by_time_max_up_genes <- gost( query = unique( (strain_by_time_max_genes %>%
                                                         filter(log2FoldChange > 0))$symbol),
                                      organism = "mmusculus",
                                      domain_scope = "custom",
                                      custom_bg = all_genes_palate$symbol,
                                      evcodes = TRUE
)
if( length(g.strain_by_time_max_up_genes$result) != 0) g.strain_by_time_max_up_genes$result <- g.strain_by_time_max_up_genes$result %>% filter(term_size < 660)


g.strain_by_time_max_down_genes <- gost( query = unique( (strain_by_time_max_genes %>%
                                                           filter(log2FoldChange < 0))$symbol),
                                        organism = "mmusculus",
                                        domain_scope = "custom",
                                        custom_bg = all_genes_palate$symbol,
                                        evcodes = TRUE
)
g.strain_by_time_max_down_genes$result <- g.strain_by_time_max_down_genes$result %>% filter(term_size < 660)
```


```{r}
#| warning: false
#| message: false

strain_by_time_max_genes  %>%
  select(gene_id, symbol, chromosome , padj, log2FoldChange) %>%
  mutate( padj = formatC(padj, digits = 2,format = "e"),
          log2FoldChange = formatC(log2FoldChange, digits = 2, format = "g")) %>%
  create_dt()

```

```{r}
#| warning: false
#| message: false

g.strain_by_time_max_up_genes$result %>%
  mutate( Category = "Log2FoldChange >0") %>%
  rbind(
    g.strain_by_time_max_down_genes$result %>%
      mutate(Category = "Log2FoldChange <0")
  )  %>%
  select( Category, term_name, source, p_value, term_size, intersection_size,intersection) %>%  # intersecion: adding the list of genes identified as overlapping with each category
  mutate_if( is.numeric, formatC, digits =2) %>%
  create_dt()

```


#### POST domain

There are `r formatC(length(unique(strain_post_genes$gene_id)),big.mark = ",")` genes that are differentially expressed between B6 and CAST in the POST domain that do not have an interaction with developmental time (i.e. the difference between the strains are consistent over time). On the other hand, there are `r length(unique(strain_by_time_post_genes$gene_id))` genes differentially expressed in the POST domain between B6 and CAST that also show difference in the trends over developmental time.

##### Genes differentially expressed between B6 and CAST (no interaction with LB stage)

```{r}
#| warning: false
#| message: false
#| fig-width: 20
#| fig-height: 4
strain_post_genes |> 
  inner_join(genes) |> 
  select(gene_id,symbol,Comparison,type_long) |> 
  distinct() |> 
  left_join(strain_gene_counts) |> 
  filter(AP_domain =="POST") |> 
  ggplot()+
  aes(x = LB_stage, 
      y = count, 
      col = strain,
      group=interaction(gene_id,strain)
      )+
  #geom_boxplot( width = 0.2)+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_color_manual(values = strain_colors)+
  theme_pubclean()+
  scale_y_log10()+
  ylab("Normalized gene expression counts (au)")+
  #ggtitle("Domain effect (no interaction)")+
  facet_wrap(~AP_domain+symbol, scales = "free", nrow = 1)

```

As you can see in the examples above, the trend over time for each gene is highly similar over developmental time. What is going on with Bglap2? I think it is expressed at a very low level and it is being picked up here because CAST samples don’t really have much expression.

```{r}
#| label: strain_post_ora
#| warning: false
#| message: false
#| cache: true

# ora
g.strain_post_up_genes <- gost( query = unique( (strain_post_genes %>%
                                                filter(log2FoldChange > 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_post_up_genes$result <- g.strain_post_up_genes$result %>% filter(term_size < 660)


g.strain_post_down_genes <- gost( query = unique( (strain_post_genes %>%
                                                filter(log2FoldChange < 0))$symbol),
                           organism = "mmusculus",
                           domain_scope = "custom",
                           custom_bg = all_genes_palate$symbol,
                           evcodes = TRUE
                           )
g.strain_post_down_genes$result <- g.strain_post_down_genes$result %>% filter(term_size < 660)

```

```{r}
#| warning: false
#| message: false

strain_post_genes  %>%
  left_join( all_genes_palate) |> 
  select(gene_id, symbol, chromosome , padj, log2FoldChange) %>%
  mutate( Comparison = ifelse( log2FoldChange >0, "Higher expression in CAST in POST","Lower expression in CAST in POST")) |> 
  mutate( padj = formatC(padj, digits = 2,format = "e"),
          log2FoldChange = formatC(log2FoldChange, digits = 2, format = "g")) %>%
  create_dt()

```

```{r}
#| warning: false
#| message: false

g.strain_post_up_genes$result %>%
  mutate( Category = "Log2FoldChange >0") %>%
  rbind(
    g.strain_post_down_genes$result %>%
      mutate(Category = "Log2FoldChange <0")
  )  %>%
  select( Category, term_name, source, p_value, term_size, intersection_size,intersection) %>%  # intersecion: adding the list of genes identified as overlapping with each category
  mutate_if( is.numeric, formatC, digits =2) %>%
  create_dt()

```


##### Genes differentially expressed with different trends over time (with strain by LB stage interaction)

```{r}
#| warning: false
#| message: false
#| fig-width: 16
#| fig-height: 4
strain_by_time_post_genes |> 
  #inner_join(genes) |> 
  slice_sample(n = 4) |> 
  select(gene_id,symbol,log2FoldChange) |> 
  distinct() |> 
  left_join(strain_gene_counts) |> 
  filter(AP_domain =="POST") |> 
  mutate( log2foldchange = round(log2FoldChange,2)) |> 
  ggplot()+
  aes(x = LB_stage, 
      y = count, 
      col = strain,
      group=interaction(gene_id,strain)
      )+
  #geom_boxplot( width = 0.2)+
  geom_point()+
  geom_smooth(method = "lm")+
  scale_color_manual(values = strain_colors)+
  theme_pubclean()+
  scale_y_log10()+
  ylab("Normalized gene expression counts (au)")+
  #ggtitle("Domain effect (no interaction)")+
  facet_wrap(~AP_domain+symbol+log2foldchange, scales = "free", nrow = 1)

```


The example genes here show different trends over time, where the slopes of the CAST line is higher than the B6 line it leads to a positive fold change and where the opposite is observed leads to a negative fold change. Currently I am grouping genes based on the sign of the fold changes although I am not sure if that biologically makes sense. Let me know which genes you would like to investigate further and we can look at them closer instead.

```{r}
#| label: strain_by_time_post_ora
#| warning: false
#| message: false
#| cache: true


# ora
g.strain_by_time_post_up_genes <- gost( query = unique( (strain_by_time_post_genes %>%
                                                         filter(log2FoldChange > 0))$symbol),
                                      organism = "mmusculus",
                                      domain_scope = "custom",
                                      custom_bg = all_genes_palate$symbol,
                                      evcodes = TRUE
)
g.strain_by_time_post_up_genes$result <- g.strain_by_time_post_up_genes$result %>% filter(term_size < 660)


g.strain_by_time_post_down_genes <- gost( query = unique( (strain_by_time_post_genes %>%
                                                           filter(log2FoldChange < 0))$symbol),
                                        organism = "mmusculus",
                                        domain_scope = "custom",
                                        custom_bg = all_genes_palate$symbol,
                                        evcodes = TRUE
)
g.strain_by_time_post_down_genes$result <- g.strain_by_time_post_down_genes$result %>% filter(term_size < 660)
```

```{r}
#| warning: false
#| message: false

strain_by_time_post_genes  %>%
  select(gene_id, symbol, chromosome , padj, log2FoldChange) %>%
  mutate( padj = formatC(padj, digits = 2,format = "e"),
          log2FoldChange = formatC(log2FoldChange, digits = 2, format = "g")) %>%
  create_dt()

```


```{r}
#| warning: false
#| message: false

g.strain_by_time_post_up_genes$result %>%
  mutate( Category = "Log2FoldChange >0") %>%
  rbind(
    g.strain_by_time_post_down_genes$result %>%
      mutate(Category = "Log2FoldChange <0")
  )  %>%
  select( Category, term_name, source, p_value, term_size, intersection_size,intersection) %>%  # intersecion: adding the list of genes identified as overlapping with each category
  mutate_if( is.numeric, formatC, digits =2) %>%
  create_dt()

```

:::


:::




