---
title: "Investigating strain, domain and developmental stage effects"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
      embed-resources: true
      standalone: true
code-fold: true
toc: true
toc-depth: 5
toc-expand: true
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| message: false
#| warning: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(gprofiler2)
set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e108_eg55_p17/")
library(DT)
library(here)

set.seed(123)
select <- dplyr::select 
rename <- dplyr::rename 
filter <- dplyr::filter 
summarize <- dplyr::summarize 
count <- dplyr::count

domain_colors <- c(PM = rgb(252,206,93, maxColorValue = 255), 
                   MAX = rgb(163,82,157, maxColorValue = 255), 
                   POST = rgb(158,156,201, maxColorValue = 255))

strain_colors <- c(B6 = "#555555", CAST = "#009E73")


# helper-functions
create_dt <- function(x){
  format_num <- function(val) {
    case_when(
      is.na(val) ~ NA_real_,
      abs(val) < 0.01 ~ signif(val, 2),
      TRUE ~ round(val, 2)
    )
  }
  x |> 
    mutate(across(where(is.numeric), format_num)) |>
    DT::datatable(
      extensions = 'Buttons',
      rownames = FALSE, 
      filter = "top",
      options = list(dom = 'Blfrtip',
                     buttons = c('copy', 'csv', 'excel'),
                     pageLength = 10, 
                     scrollX = TRUE))
}

get_normalized_counts <- function(dds, genes) {
  vsd <- vst(dds, blind = FALSE)
  assay(vsd)[genes, , drop = FALSE] |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    pivot_longer(-gene_id, names_to = "sample", values_to = "count") |>
    left_join(colData(dds) |> as.data.frame() |> rownames_to_column("sample"), by = "sample") |>
    left_join(all_genes_palate |> select(gene_id, symbol), by = "gene_id")
}

run_enrichment_hybrid <- function(gene_ids, 
                                  sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC"),
                                  pval_threshold = 0.05) {
  gene_symbols <- all_genes_palate |>
    filter(gene_id %in% gene_ids) |>
    pull(symbol) |> unique() |> na.omit()
  
  if(length(gene_symbols) < 5) return(NULL)
  
  tryCatch({
    gost_result <- gost(query = gene_symbols, organism = "mmusculus",
                        sources = sources, significant = TRUE,
                        user_threshold = pval_threshold,
                        correction_method = "fdr", evcodes = TRUE)
    if(is.null(gost_result$result) || nrow(gost_result$result) == 0) return(NULL)
    gost_result
  }, error = function(e) NULL)
}

format_enrichment_result <- function(gost_result, max_terms = 50) {
  if(is.null(gost_result$result)) return(NULL)
  gost_result$result |>
    select(Source = source, Term_ID = term_id, Term_Name = term_name,
           P_Value = p_value, Term_Size = term_size, Query_Size = query_size,
           Intersection_Size = intersection_size, Genes = intersection) |>
    arrange(P_Value) |> head(max_terms) |>
    mutate(P_Value = signif(P_Value, 3),
           Gene_Ratio = paste0(Intersection_Size, "/", Query_Size))
}


# plotting-functions
plot_main_strain_effect <- function(gene_ids, direction, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = direction, x = "Development stage", y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
            strip.text = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = strain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values = strain_colors) +
      labs(x = "", y = "") +
      theme_pubclean(base_size = 10) + theme(legend.position = "none")
    
    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

plot_domain_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    
    strain_diffs <- gene_data |> 
      group_by(AP_domain, strain) |> 
      summarize(mean_count = mean(count, na.rm = TRUE), .groups = "drop") |> 
      pivot_wider(names_from = strain, values_from = mean_count) |> 
      mutate(pct_diff = (CAST - B6) / B6 * 100)
    
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "Development", y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
            strip.text = element_text(face = "bold"))
    
    p2 <- ggplot(strain_diffs, aes(x = AP_domain, y = pct_diff, fill = AP_domain)) +
      geom_col(alpha = 0.8) + geom_hline(yintercept = 0, linetype = "dashed") +
      scale_fill_manual(values = domain_colors) +
      labs(y = "% Diff\n(CAST-B6)", x = "") +
      theme_pubclean(base_size = 10) + theme(legend.position = "none")
    
    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

plot_time_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(aes(shape = AP_domain), alpha = 0.5, size = 2.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1.5) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "Development stage", y = "Counts") +
      theme_pubclean(base_size = 11) + theme(legend.position = "right", plot.title = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(x = "Development", y = "Counts") +
      theme_pubclean(base_size = 9) +
      theme(legend.position = "none", strip.text = element_text(face = "bold"))
    
    p1 / p2 + plot_layout(heights = c(2, 1))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

plot_threeway_interaction <- function(gene_ids, pattern_desc, count_data, n = 4) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "Development", y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
            strip.text = element_text(face = "bold"))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

plot_domain_specificity <- function(gene_ids, domains_affected, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    
    p1 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain), position = position_jitterdodge(jitter.width = 0.1),
                 alpha = 0.4, size = 1.5) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values = strain_colors) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = paste("Strain effect in", domains_affected),
           x = "Domain", y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(x = "Development", y = "Counts") +
      theme_pubclean(base_size = 9) +
      theme(legend.position = "none", strip.text = element_text(face = "bold"))
    
    p1 / p2 + plot_layout(heights = c(1, 1))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}
```

```{r}
#| label: load-data
#| message: false
#| warning: false

sample_details <- read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 

load("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/palate_gbrs_emase_counts_v4.RData")

contam_samples <- c("B6_12.5_PM_1","B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat_nondo[, !colnames(expr_mat_nondo) %in% contam_samples]
rm(expr_mat_do, expr_mat_nondo)

all_genes_palate <- read_tsv(here("_data","ENSMUSGid_to_symbol_v105.txt"),
                              col_names = c("gene_id","symbol","gene_biotype","chromosome","start","end"), 
                              skip = 1) |>  
  mutate(midpoint = (start + end) / 2) |>
  filter(gene_id %in% rownames(raw_expr_mat))

sample_annot <- sample_details |>
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |>
  separate(Strain_stock, into = c("strain","tmp"), remove = FALSE, sep = "_") |>
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, 
         palate_closure, L_R_palate) |>
  mutate(strain = if_else(strain == "B6J", "B6", strain)) |>
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |>
  select(-temp) |>
  mutate(Palate_ori = if_else(Palate_ori == "L", "Left", "Right"))
```

# Comparing B6 and CAST strains across AP domains and developmental time

This analysis investigates how gene expression differs between B6 and CAST mouse strains across three palate domains (PM, MAX, POST) and developmental time. We decompose strain effects into:

- **Main strain effect**: Genes consistently higher/lower in one strain across all conditions
- **Domain × Strain interaction**: Strain differences that vary by palate domain
- **Time × Strain interaction**: Strain differences that change over developmental time
- **Three-way interaction**: Domain-specific strain effects that also change over time

```{r}
#| label: prep-run-deseq2
#| message: false
#| warning: false

b6_cast_samples <- sample_annot |>
  filter(strain %in% c("B6", "CAST")) |>
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = if_else(is.na(rep), "1", rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

coldata <- b6_cast_samples |>
  select(sample, unique_sampleid, sex, ind, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(coldata), drop = FALSE])

strain_dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~1)
strain_dds_coll <- collapseReplicates(strain_dds, strain_dds$unique_sampleid)

min_LB_stage <- min(colData(strain_dds_coll)$LB_stage)
colData(strain_dds_coll)$LB_stage_from_start <- colData(strain_dds_coll)$LB_stage - min_LB_stage
colData(strain_dds_coll)$sex <- factor(colData(strain_dds_coll)$sex)
colData(strain_dds_coll)$strain <- factor(colData(strain_dds_coll)$strain, levels = c("B6", "CAST"))
colData(strain_dds_coll)$ind <- factor(colData(strain_dds_coll)$ind)


# run-deseq2
design(strain_dds_coll) <- ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start

strain_dds_lrt <- DESeq(strain_dds_coll, test = "LRT", 
                        reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind,
                        parallel = TRUE)
res_strain_all <- results(strain_dds_lrt)

res_strain_main <- results(strain_dds_lrt, contrast = c("strain", "CAST", "B6"), test = "Wald")

dds_strain_domain <- DESeq(strain_dds_coll, test = "LRT",
                           full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                  strain + strain:AP_domain + strain:LB_stage_from_start + 
                                  strain:AP_domain:LB_stage_from_start,
                           reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                      strain + strain:LB_stage_from_start,
                           parallel = TRUE)
res_strain_domain_int <- results(dds_strain_domain)

dds_strain_time <- DESeq(strain_dds_coll, test = "LRT",
                         full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                strain + strain:AP_domain + strain:LB_stage_from_start + 
                                strain:AP_domain:LB_stage_from_start,
                         reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                    strain + strain:AP_domain,
                         parallel = TRUE)
res_strain_time_int <- results(dds_strain_time)

dds_strain_3way <- DESeq(strain_dds_coll, test = "LRT",
                         full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                strain + strain:AP_domain + strain:LB_stage_from_start + 
                                strain:AP_domain:LB_stage_from_start,
                         reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                    strain + strain:AP_domain + strain:LB_stage_from_start,
                         parallel = TRUE)
res_strain_3way_int <- results(dds_strain_3way)

# categorize-genes
strain_effects <- tibble(
  gene_id = rownames(res_strain_all),
  any_strain = res_strain_all$padj < 0.05,
  main_effect = res_strain_main$padj < 0.05,
  domain_int = res_strain_domain_int$padj < 0.05,
  time_int = res_strain_time_int$padj < 0.05,
  three_way_int = res_strain_3way_int$padj < 0.05
) |>
  replace_na(list(any_strain = FALSE, main_effect = FALSE, 
                  domain_int = FALSE, time_int = FALSE, three_way_int = FALSE)) |>
  mutate(pattern = case_when(
    !any_strain ~ "No strain effect",
    three_way_int ~ "3-way interaction",
    domain_int & time_int ~ "Domain×strain + Time×strain",
    domain_int ~ "Domain×strain interaction",
    time_int ~ "Time×strain interaction",
    main_effect ~ "Main strain effect only",
    TRUE ~ "Other"
  ))

strain_effects_annotated <- strain_effects |>
  left_join(all_genes_palate |> select(gene_id, symbol, gene_biotype, chromosome), by = "gene_id") |>
  mutate(
    any_strain_pval = res_strain_all$padj[match(gene_id, rownames(res_strain_all))],
    main_effect_lfc = res_strain_main$log2FoldChange[match(gene_id, rownames(res_strain_main))],
    main_effect_pval = res_strain_main$padj[match(gene_id, rownames(res_strain_main))],
    domain_int_pval = res_strain_domain_int$padj[match(gene_id, rownames(res_strain_domain_int))],
    time_int_pval = res_strain_time_int$padj[match(gene_id, rownames(res_strain_time_int))],
    three_way_int_pval = res_strain_3way_int$padj[match(gene_id, rownames(res_strain_3way_int))]
  )

genes_main_only <- strain_effects_annotated |> filter(pattern == "Main strain effect only")
genes_domain_int <- strain_effects_annotated |> filter(domain_int & !time_int & !three_way_int)
genes_time_int <- strain_effects_annotated |> filter(time_int & !domain_int & !three_way_int)
genes_domain_time <- strain_effects_annotated |> filter(domain_int & time_int & !three_way_int)
genes_three_ways <- strain_effects_annotated |> filter(three_way_int)

# load-patterns
# source(here("DEgene_pattern_identification_CAST_B6.R"))
pattern_summary <- read_csv(here("strain_pattern_summary.csv"))
gene_assignments <- read_csv(here("strain_gene_pattern_assignments.csv"))

main_patterns <- if(file.exists(here("main_strain_effect_patterns.csv"))) 
  read_csv(here("main_strain_effect_patterns.csv")) else NULL
domain_patterns <- if(file.exists(here("domain_strain_patterns.csv"))) 
  read_csv(here("domain_strain_patterns.csv")) else NULL
time_patterns <- if(file.exists(here("time_strain_patterns.csv"))) 
  read_csv(here("time_strain_patterns.csv")) else NULL
threeway_patterns <- if(file.exists(here("threeway_patterns.csv"))) 
  read_csv(here("threeway_patterns.csv")) else NULL
domain_specificity <- if(file.exists(here("domain_specificity_patterns.csv"))) 
  read_csv(here("domain_specificity_patterns.csv")) else NULL

genes_CAST_high_strong <- gene_assignments |> filter(CAST_high_strong) |> pull(gene_id)
genes_CAST_high_mod <- gene_assignments |> filter(CAST_high_mod) |> pull(gene_id)
genes_B6_high_strong <- gene_assignments |> filter(B6_high_strong) |> pull(gene_id)
genes_B6_high_mod <- gene_assignments |> filter(B6_high_mod) |> pull(gene_id)

genes_PM_biggest_diff <- gene_assignments |> filter(PM_biggest_diff) |> pull(gene_id)
genes_MAX_biggest_diff <- gene_assignments |> filter(MAX_biggest_diff) |> pull(gene_id)
genes_POST_biggest_diff <- gene_assignments |> filter(POST_biggest_diff) |> pull(gene_id)
genes_PM_CAST_high <- gene_assignments |> filter(PM_CAST_high) |> pull(gene_id)
genes_MAX_CAST_high <- gene_assignments |> filter(MAX_CAST_high) |> pull(gene_id)
genes_POST_CAST_high <- gene_assignments |> filter(POST_CAST_high) |> pull(gene_id)
genes_opposite_directions <- gene_assignments |> filter(Opposite_directions) |> pull(gene_id)

genes_both_increase_CAST_faster <- gene_assignments |> filter(Both_inc_CAST_faster) |> pull(gene_id)
genes_both_decrease_CAST_faster <- gene_assignments |> filter(Both_dec_CAST_faster) |> pull(gene_id)
genes_both_increase_B6_faster <- gene_assignments |> filter(Both_inc_B6_faster) |> pull(gene_id)
genes_both_decrease_B6_faster <- gene_assignments |> filter(Both_dec_B6_faster) |> pull(gene_id)
genes_B6_up_CAST_down <- gene_assignments |> filter(B6_up_CAST_down) |> pull(gene_id)
genes_CAST_up_B6_down <- gene_assignments |> filter(CAST_up_B6_down) |> pull(gene_id)
genes_CAST_dynamic_B6_flat <- gene_assignments |> filter(CAST_dynamic_B6_flat) |> pull(gene_id)
genes_B6_dynamic_CAST_flat <- gene_assignments |> filter(B6_dynamic_CAST_flat) |> pull(gene_id)

genes_PM_biggest_slope_diff <- gene_assignments |> filter(ThreeWay_PM_biggest) |> pull(gene_id)
genes_MAX_biggest_slope_diff <- gene_assignments |> filter(ThreeWay_MAX_biggest) |> pull(gene_id)
genes_POST_biggest_slope_diff <- gene_assignments |> filter(ThreeWay_POST_biggest) |> pull(gene_id)
genes_consistent_slope_diffs <- gene_assignments |> filter(ThreeWay_consistent) |> pull(gene_id)

if(!is.null(domain_specificity)) {
  genes_strain_PM_only <- domain_specificity |> filter(specificity_pattern == "PM only") |> pull(gene_id)
  genes_strain_MAX_only <- domain_specificity |> filter(specificity_pattern == "MAX only") |> pull(gene_id)
  genes_strain_POST_only <- domain_specificity |> filter(specificity_pattern == "POST only") |> pull(gene_id)
  genes_strain_all_domains <- domain_specificity |> filter(specificity_pattern == "All domains") |> pull(gene_id)
}
```

## Overview


The analysis identified **`r sum(strain_effects_annotated$any_strain)`** genes with significant strain effects (padj < 0.05).


```{r}
#| label: overview-summary
#| message: false
#| warning: false

strain_effects_annotated |>
  filter(any_strain) |>
  count(pattern) |>
  arrange(desc(n)) |>
  rename(`Pattern Type` = pattern, `Number of Genes` = n) |>
  create_dt()
```


## Main Strain Effects

These genes show consistent expression differences between CAST and B6 across all domains and developmental stages.

::::::::::: {#main-strain-tabs .panel-tabset .nav-pills}

### CAST >> B6 (Strong)

Genes with strong upregulation in CAST (log2FC > 1).

:::::::: {#main-CAST-strong-subtabs .panel-tabset}

#### Plots

```{r}
#| label: main-CAST-strong-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_CAST_high_strong) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_CAST_high_strong)
  plot_main_strain_effect(genes_CAST_high_strong, "CAST >> B6 (strong)", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: main-CAST-strong-table
#| message: false
#| warning: false

if(!is.null(main_patterns) && length(genes_CAST_high_strong) > 0) {
  main_patterns |> filter(gene_id %in% genes_CAST_high_strong) |>
    select(symbol, main_effect_lfc, strain_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: main-CAST-strong-enrich
#| message: false
#| warning: false

if(length(genes_CAST_high_strong) > 0) {
  enrich <- run_enrichment_hybrid(genes_CAST_high_strong)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### CAST > B6 (Moderate)

Genes with moderate upregulation in CAST (0.5 < log2FC ≤ 1).

:::::::: {#main-CAST-mod-subtabs .panel-tabset}

#### Plots

```{r}
#| label: main-CAST-mod-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_CAST_high_mod) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_CAST_high_mod)
  plot_main_strain_effect(genes_CAST_high_mod, "CAST > B6 (moderate)", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: main-CAST-mod-table
#| message: false
#| warning: false

if(!is.null(main_patterns) && length(genes_CAST_high_mod) > 0) {
  main_patterns |> filter(gene_id %in% genes_CAST_high_mod) |>
    select(symbol, main_effect_lfc, strain_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: main-CAST-mod-enrich
#| message: false
#| warning: false

if(length(genes_CAST_high_mod) > 0) {
  enrich <- run_enrichment_hybrid(genes_CAST_high_mod)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### B6 >> CAST (Strong)

Genes with strong upregulation in B6 (log2FC < -1).

:::::::: {#main-B6-strong-subtabs .panel-tabset}

#### Plots

```{r}
#| label: main-B6-strong-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_B6_high_strong) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_B6_high_strong)
  plot_main_strain_effect(genes_B6_high_strong, "B6 >> CAST (strong)", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: main-B6-strong-table
#| message: false
#| warning: false

if(!is.null(main_patterns) && length(genes_B6_high_strong) > 0) {
  main_patterns |> filter(gene_id %in% genes_B6_high_strong) |>
    select(symbol, main_effect_lfc, strain_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: main-B6-strong-enrich
#| message: false
#| warning: false

if(length(genes_B6_high_strong) > 0) {
  enrich <- run_enrichment_hybrid(genes_B6_high_strong)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### B6 > CAST (Moderate)

Genes with moderate upregulation in B6 (-1 ≤ log2FC < -0.5).

:::::::: {#main-B6-mod-subtabs .panel-tabset}

#### Plots

```{r}
#| label: main-B6-mod-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_B6_high_mod) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_B6_high_mod)
  plot_main_strain_effect(genes_B6_high_mod, "B6 > CAST (moderate)", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: main-B6-mod-table
#| message: false
#| warning: false

if(!is.null(main_patterns) && length(genes_B6_high_mod) > 0) {
  main_patterns |> filter(gene_id %in% genes_B6_high_mod) |>
    select(symbol, main_effect_lfc, strain_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: main-B6-mod-enrich
#| message: false
#| warning: false

if(length(genes_B6_high_mod) > 0) {
  enrich <- run_enrichment_hybrid(genes_B6_high_mod)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

:::::::::::

## Domain × Strain Interactions

These genes show strain differences that vary across palate domains.

::::::::::: {#domain-strain-tabs .panel-tabset .nav-pills}

### Overview

```{r}
#| label: domain-strain-overview
#| message: false
#| warning: false

if(!is.null(domain_patterns)) {
  domain_patterns |> count(detailed_pattern) |> arrange(desc(n)) |>
    rename(Pattern = detailed_pattern, `N Genes` = n) |> create_dt()
}
```

### PM: Biggest Difference

Genes where the largest strain difference occurs in the premaxilla (PM) domain.

:::::::: {#domain-PM-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-PM-biggest-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_PM_biggest_diff) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_PM_biggest_diff)
  plot_domain_strain_interaction(genes_PM_biggest_diff, "Biggest strain diff in PM", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-PM-biggest-table
#| message: false
#| warning: false

if(!is.null(domain_patterns) && length(genes_PM_biggest_diff) > 0) {
  domain_patterns |> filter(gene_id %in% genes_PM_biggest_diff) |>
    select(symbol, diff_PM, diff_MAX, diff_POST, max_diff_direction) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-PM-biggest-enrich
#| message: false
#| warning: false

if(length(genes_PM_biggest_diff) > 0) {
  enrich <- run_enrichment_hybrid(genes_PM_biggest_diff)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### MAX: Biggest Difference

Genes where the largest strain difference occurs in the maxillary (MAX) domain.

:::::::: {#domain-MAX-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-MAX-biggest-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_MAX_biggest_diff) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_MAX_biggest_diff)
  plot_domain_strain_interaction(genes_MAX_biggest_diff, "Biggest strain diff in MAX", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-MAX-biggest-table
#| message: false
#| warning: false

if(!is.null(domain_patterns) && length(genes_MAX_biggest_diff) > 0) {
  domain_patterns |> filter(gene_id %in% genes_MAX_biggest_diff) |>
    select(symbol, diff_PM, diff_MAX, diff_POST, max_diff_direction) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-MAX-biggest-enrich
#| message: false
#| warning: false

if(length(genes_MAX_biggest_diff) > 0) {
  enrich <- run_enrichment_hybrid(genes_MAX_biggest_diff)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### POST: Biggest Difference

Genes where the largest strain difference occurs in the posterior (POST) domain.

:::::::: {#domain-POST-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-POST-biggest-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_POST_biggest_diff) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_POST_biggest_diff)
  plot_domain_strain_interaction(genes_POST_biggest_diff, "Biggest strain diff in POST", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-POST-biggest-table
#| message: false
#| warning: false

if(!is.null(domain_patterns) && length(genes_POST_biggest_diff) > 0) {
  domain_patterns |> filter(gene_id %in% genes_POST_biggest_diff) |>
    select(symbol, diff_PM, diff_MAX, diff_POST, max_diff_direction) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-POST-biggest-enrich
#| message: false
#| warning: false

if(length(genes_POST_biggest_diff) > 0) {
  enrich <- run_enrichment_hybrid(genes_POST_biggest_diff)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### Opposite Directions

Genes where strain effects go in opposite directions across domains.

:::::::: {#domain-opposite-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-opposite-plot
#| message: false
#| warning: false
#| fig.height: 16

if(length(genes_opposite_directions) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_opposite_directions)
  plot_domain_strain_interaction(genes_opposite_directions, "Opposite strain effects across domains", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-opposite-table
#| message: false
#| warning: false

if(!is.null(domain_patterns) && length(genes_opposite_directions) > 0) {
  domain_patterns |> filter(gene_id %in% genes_opposite_directions) |>
    select(symbol, diff_PM, diff_MAX, diff_POST, consistency) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-opposite-enrich
#| message: false
#| warning: false

if(length(genes_opposite_directions) > 0) {
  enrich <- run_enrichment_hybrid(genes_opposite_directions)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

:::::::::::

## Time × Strain Interactions

These genes show strain differences that change over developmental time.

::::::::::: {#time-strain-tabs .panel-tabset .nav-pills}

### Overview

```{r}
#| label: time-strain-overview
#| message: false
#| warning: false

if(!is.null(time_patterns)) {
  time_patterns |> count(trajectory_pattern) |> arrange(desc(n)) |>
    rename(Pattern = trajectory_pattern, `N Genes` = n) |> create_dt()
}
```

### Both Increase, CAST Faster

Both strains increase over time, but CAST increases faster.

:::::::: {#time-both-inc-CAST-subtabs .panel-tabset}

#### Plots

```{r}
#| label: time-both-inc-CAST-plot
#| message: false
#| warning: false
#| fig.height: 18

if(length(genes_both_increase_CAST_faster) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_both_increase_CAST_faster)
  plot_time_strain_interaction(genes_both_increase_CAST_faster, "Both ↑, CAST faster", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: time-both-inc-CAST-table
#| message: false
#| warning: false

if(!is.null(time_patterns) && length(genes_both_increase_CAST_faster) > 0) {
  time_patterns |> filter(gene_id %in% genes_both_increase_CAST_faster) |>
    select(symbol, slope_CAST, slope_B6, slope_diff, trajectory_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: time-both-inc-CAST-enrich
#| message: false
#| warning: false

if(length(genes_both_increase_CAST_faster) > 0) {
  enrich <- run_enrichment_hybrid(genes_both_increase_CAST_faster)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### Both Decrease, CAST Faster

Both strains decrease over time, but CAST decreases faster.

:::::::: {#time-both-dec-CAST-subtabs .panel-tabset}

#### Plots

```{r}
#| label: time-both-dec-CAST-plot
#| message: false
#| warning: false
#| fig.height: 18

if(length(genes_both_decrease_CAST_faster) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_both_decrease_CAST_faster)
  plot_time_strain_interaction(genes_both_decrease_CAST_faster, "Both ↓, CAST faster", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: time-both-dec-CAST-table
#| message: false
#| warning: false

if(!is.null(time_patterns) && length(genes_both_decrease_CAST_faster) > 0) {
  time_patterns |> filter(gene_id %in% genes_both_decrease_CAST_faster) |>
    select(symbol, slope_CAST, slope_B6, slope_diff, trajectory_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: time-both-dec-CAST-enrich
#| message: false
#| warning: false

if(length(genes_both_decrease_CAST_faster) > 0) {
  enrich <- run_enrichment_hybrid(genes_both_decrease_CAST_faster)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### CAST Increases, B6 Decreases

Strains diverge: CAST expression increases while B6 decreases.

:::::::: {#time-CAST-up-B6-down-subtabs .panel-tabset}

#### Plots

```{r}
#| label: time-CAST-up-B6-down-plot
#| message: false
#| warning: false
#| fig.height: 18

if(length(genes_CAST_up_B6_down) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_CAST_up_B6_down)
  plot_time_strain_interaction(genes_CAST_up_B6_down, "CAST ↑, B6 ↓", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: time-CAST-up-B6-down-table
#| message: false
#| warning: false

if(!is.null(time_patterns) && length(genes_CAST_up_B6_down) > 0) {
  time_patterns |> filter(gene_id %in% genes_CAST_up_B6_down) |>
    select(symbol, slope_CAST, slope_B6, slope_diff, trajectory_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: time-CAST-up-B6-down-enrich
#| message: false
#| warning: false

if(length(genes_CAST_up_B6_down) > 0) {
  enrich <- run_enrichment_hybrid(genes_CAST_up_B6_down)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### B6 Increases, CAST Decreases

Strains diverge: B6 expression increases while CAST decreases.

:::::::: {#time-B6-up-CAST-down-subtabs .panel-tabset}

#### Plots

```{r}
#| label: time-B6-up-CAST-down-plot
#| message: false
#| warning: false
#| fig.height: 18

if(length(genes_B6_up_CAST_down) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_B6_up_CAST_down)
  plot_time_strain_interaction(genes_B6_up_CAST_down, "B6 ↑, CAST ↓", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: time-B6-up-CAST-down-table
#| message: false
#| warning: false

if(!is.null(time_patterns) && length(genes_B6_up_CAST_down) > 0) {
  time_patterns |> filter(gene_id %in% genes_B6_up_CAST_down) |>
    select(symbol, slope_CAST, slope_B6, slope_diff, trajectory_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: time-B6-up-CAST-down-enrich
#| message: false
#| warning: false

if(length(genes_B6_up_CAST_down) > 0) {
  enrich <- run_enrichment_hybrid(genes_B6_up_CAST_down)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### CAST Dynamic, B6 Flat

CAST shows dynamic expression changes while B6 remains relatively flat.

:::::::: {#time-CAST-dynamic-subtabs .panel-tabset}

#### Plots

```{r}
#| label: time-CAST-dynamic-plot
#| message: false
#| warning: false
#| fig.height: 18

if(length(genes_CAST_dynamic_B6_flat) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_CAST_dynamic_B6_flat)
  plot_time_strain_interaction(genes_CAST_dynamic_B6_flat, "CAST dynamic, B6 flat", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: time-CAST-dynamic-table
#| message: false
#| warning: false

if(!is.null(time_patterns) && length(genes_CAST_dynamic_B6_flat) > 0) {
  time_patterns |> filter(gene_id %in% genes_CAST_dynamic_B6_flat) |>
    select(symbol, slope_CAST, slope_B6, trajectory_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: time-CAST-dynamic-enrich
#| message: false
#| warning: false

if(length(genes_CAST_dynamic_B6_flat) > 0) {
  enrich <- run_enrichment_hybrid(genes_CAST_dynamic_B6_flat)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### B6 Dynamic, CAST Flat

B6 shows dynamic expression changes while CAST remains relatively flat.

:::::::: {#time-B6-dynamic-subtabs .panel-tabset}

#### Plots

```{r}
#| label: time-B6-dynamic-plot
#| message: false
#| warning: false
#| fig.height: 18

if(length(genes_B6_dynamic_CAST_flat) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_B6_dynamic_CAST_flat)
  plot_time_strain_interaction(genes_B6_dynamic_CAST_flat, "B6 dynamic, CAST flat", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: time-B6-dynamic-table
#| message: false
#| warning: false

if(!is.null(time_patterns) && length(genes_B6_dynamic_CAST_flat) > 0) {
  time_patterns |> filter(gene_id %in% genes_B6_dynamic_CAST_flat) |>
    select(symbol, slope_CAST, slope_B6, trajectory_pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: time-B6-dynamic-enrich
#| message: false
#| warning: false

if(length(genes_B6_dynamic_CAST_flat) > 0) {
  enrich <- run_enrichment_hybrid(genes_B6_dynamic_CAST_flat)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

:::::::::::

## Three-Way Interactions

These genes show domain-specific strain differences that also change over developmental time.

::::::::::: {#threeway-tabs .panel-tabset .nav-pills}

### Overview

```{r}
#| label: threeway-overview
#| message: false
#| warning: false

if(!is.null(threeway_patterns)) {
  threeway_patterns |> count(pattern) |> arrange(desc(n)) |>
    rename(Pattern = pattern, `N Genes` = n) |> create_dt()
}
```

### PM: Biggest Slope Difference

Genes where the strain difference in developmental trajectory is largest in PM.

:::::::: {#threeway-PM-subtabs .panel-tabset}

#### Plots

```{r}
#| label: threeway-PM-plot
#| message: false
#| warning: false
#| fig.height: 14

if(length(genes_PM_biggest_slope_diff) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_PM_biggest_slope_diff)
  plot_threeway_interaction(genes_PM_biggest_slope_diff, "PM: largest slope diff", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: threeway-PM-table
#| message: false
#| warning: false

if(!is.null(threeway_patterns) && length(genes_PM_biggest_slope_diff) > 0) {
  threeway_patterns |> filter(gene_id %in% genes_PM_biggest_slope_diff) |>
    select(symbol, slope_diff_PM, slope_diff_MAX, slope_diff_POST, pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: threeway-PM-enrich
#| message: false
#| warning: false

if(length(genes_PM_biggest_slope_diff) > 0) {
  enrich <- run_enrichment_hybrid(genes_PM_biggest_slope_diff)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### MAX: Biggest Slope Difference

Genes where the strain difference in developmental trajectory is largest in MAX.

:::::::: {#threeway-MAX-subtabs .panel-tabset}

#### Plots

```{r}
#| label: threeway-MAX-plot
#| message: false
#| warning: false
#| fig.height: 14

if(length(genes_MAX_biggest_slope_diff) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_MAX_biggest_slope_diff)
  plot_threeway_interaction(genes_MAX_biggest_slope_diff, "MAX: largest slope diff", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: threeway-MAX-table
#| message: false
#| warning: false

if(!is.null(threeway_patterns) && length(genes_MAX_biggest_slope_diff) > 0) {
  threeway_patterns |> filter(gene_id %in% genes_MAX_biggest_slope_diff) |>
    select(symbol, slope_diff_PM, slope_diff_MAX, slope_diff_POST, pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: threeway-MAX-enrich
#| message: false
#| warning: false

if(length(genes_MAX_biggest_slope_diff) > 0) {
  enrich <- run_enrichment_hybrid(genes_MAX_biggest_slope_diff)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### POST: Biggest Slope Difference

Genes where the strain difference in developmental trajectory is largest in POST.

:::::::: {#threeway-POST-subtabs .panel-tabset}

#### Plots

```{r}
#| label: threeway-POST-plot
#| message: false
#| warning: false
#| fig.height: 14

if(length(genes_POST_biggest_slope_diff) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_POST_biggest_slope_diff)
  plot_threeway_interaction(genes_POST_biggest_slope_diff, "POST: largest slope diff", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: threeway-POST-table
#| message: false
#| warning: false

if(!is.null(threeway_patterns) && length(genes_POST_biggest_slope_diff) > 0) {
  threeway_patterns |> filter(gene_id %in% genes_POST_biggest_slope_diff) |>
    select(symbol, slope_diff_PM, slope_diff_MAX, slope_diff_POST, pattern) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: threeway-POST-enrich
#| message: false
#| warning: false

if(length(genes_POST_biggest_slope_diff) > 0) {
  enrich <- run_enrichment_hybrid(genes_POST_biggest_slope_diff)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### Consistent Across Domains

Genes where strain slope differences are consistent across all domains.

:::::::: {#threeway-consistent-subtabs .panel-tabset}

#### Plots

```{r}
#| label: threeway-consistent-plot
#| message: false
#| warning: false
#| fig.height: 14

if(length(genes_consistent_slope_diffs) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_consistent_slope_diffs)
  plot_threeway_interaction(genes_consistent_slope_diffs, "Consistent slope diffs", count_data, n = 4)
}
```

#### Gene List

```{r}
#| label: threeway-consistent-table
#| message: false
#| warning: false

if(!is.null(threeway_patterns) && length(genes_consistent_slope_diffs) > 0) {
  threeway_patterns |> filter(gene_id %in% genes_consistent_slope_diffs) |>
    select(symbol, slope_diff_PM, slope_diff_MAX, slope_diff_POST, consistency) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: threeway-consistent-enrich
#| message: false
#| warning: false

if(length(genes_consistent_slope_diffs) > 0) {
  enrich <- run_enrichment_hybrid(genes_consistent_slope_diffs)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

:::::::::::

## Domain-Specific Strain Effects

For genes with domain × strain interactions, we identify which specific domains show strain effects.

::::::::::: {#domain-specific-tabs .panel-tabset .nav-pills}

### Overview

```{r}
#| label: domain-specific-overview
#| message: false
#| warning: false

if(!is.null(domain_specificity)) {
  domain_specificity |> count(specificity_pattern) |> arrange(desc(n)) |>
    rename(`Specificity Pattern` = specificity_pattern, `Number of Genes` = n) |> create_dt()
}
```

### PM Only

Strain differences are present only in the premaxilla domain.

:::::::: {#domain-specific-PM-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-specific-PM-only-plot
#| message: false
#| warning: false
#| fig.height: 16

if(exists("genes_strain_PM_only") && length(genes_strain_PM_only) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_strain_PM_only)
  plot_domain_specificity(genes_strain_PM_only, "PM only", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-specific-PM-only-table
#| message: false
#| warning: false

if(!is.null(domain_specificity) && exists("genes_strain_PM_only") && length(genes_strain_PM_only) > 0) {
  domain_specificity |> filter(gene_id %in% genes_strain_PM_only) |>
    select(symbol, specificity_pattern, effect_PM, effect_MAX, effect_POST) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-specific-PM-only-enrich
#| message: false
#| warning: false

if(exists("genes_strain_PM_only") && length(genes_strain_PM_only) > 0) {
  enrich <- run_enrichment_hybrid(genes_strain_PM_only)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### MAX Only

Strain differences are present only in the maxillary domain.

:::::::: {#domain-specific-MAX-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-specific-MAX-only-plot
#| message: false
#| warning: false
#| fig.height: 16

if(exists("genes_strain_MAX_only") && length(genes_strain_MAX_only) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_strain_MAX_only)
  plot_domain_specificity(genes_strain_MAX_only, "MAX only", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-specific-MAX-only-table
#| message: false
#| warning: false

if(!is.null(domain_specificity) && exists("genes_strain_MAX_only") && length(genes_strain_MAX_only) > 0) {
  domain_specificity |> filter(gene_id %in% genes_strain_MAX_only) |>
    select(symbol, specificity_pattern, effect_PM, effect_MAX, effect_POST) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-specific-MAX-only-enrich
#| message: false
#| warning: false

if(exists("genes_strain_MAX_only") && length(genes_strain_MAX_only) > 0) {
  enrich <- run_enrichment_hybrid(genes_strain_MAX_only)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### POST Only

Strain differences are present only in the posterior palate domain.

:::::::: {#domain-specific-POST-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-specific-POST-only-plot
#| message: false
#| warning: false
#| fig.height: 16

if(exists("genes_strain_POST_only") && length(genes_strain_POST_only) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_strain_POST_only)
  plot_domain_specificity(genes_strain_POST_only, "POST only", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-specific-POST-only-table
#| message: false
#| warning: false

if(!is.null(domain_specificity) && exists("genes_strain_POST_only") && length(genes_strain_POST_only) > 0) {
  domain_specificity |> filter(gene_id %in% genes_strain_POST_only) |>
    select(symbol, specificity_pattern, effect_PM, effect_MAX, effect_POST) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-specific-POST-only-enrich
#| message: false
#| warning: false

if(exists("genes_strain_POST_only") && length(genes_strain_POST_only) > 0) {
  enrich <- run_enrichment_hybrid(genes_strain_POST_only)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

### All Domains

Strain differences are present in all three AP domains.

:::::::: {#domain-specific-all-subtabs .panel-tabset}

#### Plots

```{r}
#| label: domain-specific-all-plot
#| message: false
#| warning: false
#| fig.height: 16

if(exists("genes_strain_all_domains") && length(genes_strain_all_domains) > 0) {
  count_data <- get_normalized_counts(strain_dds_lrt, genes_strain_all_domains)
  plot_domain_specificity(genes_strain_all_domains, "All domains", count_data, n = 5)
}
```

#### Gene List

```{r}
#| label: domain-specific-all-table
#| message: false
#| warning: false

if(!is.null(domain_specificity) && exists("genes_strain_all_domains") && length(genes_strain_all_domains) > 0) {
  domain_specificity |> filter(gene_id %in% genes_strain_all_domains) |>
    select(symbol, specificity_pattern, effect_PM, effect_MAX, effect_POST) |> create_dt()
}
```

#### Enrichment

```{r}
#| label: domain-specific-all-enrich
#| message: false
#| warning: false

if(exists("genes_strain_all_domains") && length(genes_strain_all_domains) > 0) {
  enrich <- run_enrichment_hybrid(genes_strain_all_domains)
  if(!is.null(enrich)) format_enrichment_result(enrich) |> create_dt()
}
```

::::::::

:::::::::::

## Session Info

```{r}
#| label: session-info
#| message: false
#| warning: false

sessionInfo()
```