---
title: "Investigating strain, domain and developmental stage effects"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
      embed-resources: true
      standalone: true
code-fold: true
toc: true
toc-depth: 5
toc-expand: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| message: false
#| warning: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(gprofiler2)
set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e108_eg55_p17/")
library(enrichR)
library(DT)
library(here)

set.seed(123)
select <- dplyr::select 
rename <- dplyr::rename 
filter <- dplyr::filter 
summarize <- dplyr::summarize 
count <- dplyr::count

domain_colors <- c(PM = rgb(252,206,93, maxColorValue = 255), 
                   MAX = rgb(163,82,157, maxColorValue = 255), 
                   POST = rgb(158,156,201, maxColorValue = 255))

strain_colors <- c(B6 = "#555555", CAST = "#009E73")
```

```{r}
#| label: helper-functions
#| message: false
#| warning: false

create_dt <- function(x){
  format_num <- function(val) {
    case_when(
      is.na(val) ~ "NA",
      abs(val) < 0.01 ~ formatC(val, digits = 2),
      TRUE ~ formatC(round(val, 2))
    )
  }
  x |> 
    mutate(across(where(is.numeric), format_num)) |>
    DT::datatable(
      extensions = 'Buttons',
      rownames = FALSE, 
      filter = "top",
      options = list(dom = 'Blfrtip',
                     buttons = c('copy', 'csv', 'excel'),
                     pageLength = 5, 
                     scrollX = TRUE))
}

# Helper function for consistent filtering of DESeq2 results
filter_deseq_results <- function(res, 
                                  min_basemean = NULL,
                                  min_basemean_quantile = 0.1,
                                  max_lfcse = 3, 
                                  max_abs_lfc = 20, 
                                  min_abs_lfc = 0.5, 
                                  max_padj = 0.05) {
  
  # Convert to data frame
  res_df <- as.data.frame(res) |>
    rownames_to_column("gene_id")
  
  # Set baseMean threshold - use quantile if not specified
 if(is.null(min_basemean)) {
    min_basemean <- quantile(res_df$baseMean, min_basemean_quantile, na.rm = TRUE)
  }
  
  # Apply filters
  res_filtered <- res_df |>
    filter(
      baseMean > min_basemean,
      lfcSE < max_lfcse,
      abs(log2FoldChange) < max_abs_lfc,
      abs(log2FoldChange) > min_abs_lfc,
      padj < max_padj
    )
  
  return(res_filtered)
}

# Helper to get passing gene IDs from a filtered result
get_passing_genes <- function(res, ...) {
  filter_deseq_results(res, ...) |> pull(gene_id)
}
get_normalized_counts <- function(dds, genes, vsd = NULL) {
  if (is.null(vsd)) {
    # Fallback: compute if not pre-computed (slower)
    vsd <- vst(dds, blind = FALSE)
  }
  assay(vsd)[genes, , drop = FALSE] |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    pivot_longer(-gene_id, names_to = "sample", values_to = "count") |>
    left_join(colData(dds) |> as.data.frame() |> rownames_to_column("sample"), by = "sample") |>
    left_join(all_genes_palate |> select(gene_id, symbol), by = "gene_id")
}

# Function to run enrichment with both gprofiler2 AND Enrichr
run_enrichment_hybrid <- function(gene_ids, 
                                 gene_reference = all_genes_palate,
                                 description = "Gene Set",
                                 max_term_size = 660,
                                 min_term_size = 4) {
  
  # Convert gene IDs to symbols
  query_symbols <- gene_reference |>
    filter(gene_id %in% unique(gene_ids)) |>
    pull(symbol) |>
    unique()
  
  # Get background symbols
  background_symbols <- gene_reference$symbol
  
  # Skip if too few genes
  if(length(query_symbols) < 5) {
    message(paste0("Skipping ", description, ": fewer than 5 genes"))
    return(NULL)
  }
  
  # Query 1: gprofiler2 for GO, KEGG, Reactome, WikiPathways
  result_gprofiler <- gost(
    query = query_symbols,
    organism = "mmusculus",
    domain_scope = "custom",
    custom_bg = background_symbols,
    #sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP"),
    evcodes = TRUE
  )
  
  # Query 2: Enrichr for Mammalian Phenotype (full database)
  result_enrichr <- enrichr(
    genes = query_symbols,
    databases = c("MGI_Mammalian_Phenotype_Level_4_2021"),
                  #"MGI_Mammalian_Phenotype_Level_4_2019"),
    background =  background_symbols,
    include_overlap = TRUE
  )
  
  # Format gprofiler results
  gprofiler_df <- NULL
  if(!is.null(result_gprofiler$result)) {
    gprofiler_df <- result_gprofiler$result |>
      filter(term_size < max_term_size) |>
      filter(term_size > min_term_size) |>
      mutate(gene_set = description) |>
      select(gene_set, source, term_id, term_name, p_value, 
             query_size, term_size, intersection_size, intersection)
  }
  
  # Format Enrichr results to match gprofiler format
  enrichr_df <- NULL
  if(!is.null(result_enrichr) && length(result_enrichr) > 0) {
    # Process each database separately
    enrichr_dfs <- lapply(names(result_enrichr), function(db_name) {
      db_result <- result_enrichr[[db_name]]
      
      # Check if result has data (not empty or all NA)
      if(is.null(db_result) || nrow(db_result) == 0 || all(is.na(db_result$Term))) {
        return(NULL)
      }
      
      db_result |>
        filter(Adjusted.P.value < 0.05) |>
        mutate(
          gene_set = description,
          source = "Mammalian Phenotype",
          term_id = Term,
          term_name = Term,
          p_value = Adjusted.P.value,
          query_size = length(query_symbols),
          intersection_size = as.numeric(gsub("/.*", "", Overlap)),  # numerator (X from "X/Y")
          term_size = as.numeric(gsub(".*/", "", Overlap)),          # denominator (Y from "X/Y")
          intersection = Genes
        ) |>
        filter(term_size < max_term_size) |>
        filter(term_size > min_term_size) |>
        select(gene_set, source, term_id, term_name, p_value, 
               query_size, term_size, intersection_size, intersection)
    })
    
    # Remove NULL entries and combine
    enrichr_dfs <- enrichr_dfs[!sapply(enrichr_dfs, is.null)]
    if(length(enrichr_dfs) > 0) {
      enrichr_df <- bind_rows(enrichr_dfs)
    }
  }
  
  # Combine results
  combined_result <- bind_rows(gprofiler_df, enrichr_df)
  
  if(nrow(combined_result) > 0) {
    return(list(result = combined_result))
  } else {
    return(NULL)
  }
}

format_enrichment_result <- function(gost_result, max_terms = 50) {
  if(is.null(gost_result$result)) return(NULL)
  gost_result$result |>
    select(Source = source, Term_ID = term_id, Term_Name = term_name,
           P_Value = p_value, Term_Size = term_size, Query_Size = query_size,
           Intersection_Size = intersection_size, Genes = intersection) |>
    arrange(P_Value) |> head(max_terms) |>
    mutate(P_Value = signif(P_Value, 3),
           Gene_Ratio = paste0(Intersection_Size, "/", Query_Size))
}

# plotting-functions
plot_main_strain_effect <- function(gene_ids, direction, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + 
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = direction, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = strain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) + geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) + 
      scale_fill_manual(values = strain_colors) +
      labs(x = "", y = "") + theme_pubclean(base_size = 16) + theme(legend.position = "none")
    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

plot_domain_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    eff_data <- domain_patterns |> 
      filter( gene_id == gid)
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data$symbol)[1]
    
    # strain_diffs <- gene_data |> 
    #   group_by(AP_domain, strain) |> 
    #   summarize(mean_count = mean(count, na.rm = TRUE), .groups = "drop") |> 
    #   pivot_wider(names_from = strain, values_from = mean_count) |> 
    #   mutate(pct_diff = (CAST - B6) / B6 * 100)
    # plot the effects instead of the diff
    
    strain_diffs <- eff_data |> 
      select(gene_id, symbol,
                 strain_effect_PM, strain_effect_MAX, strain_effect_POST,
                 strain_effect_PM_pval, strain_effect_MAX_pval, strain_effect_POST_pval) |>
      pivot_longer(
            cols = -c(gene_id, symbol),
            names_to = c("AP_domain", "metric"),
            names_pattern = "strain_effect_(PM|MAX|POST)(_pval)?$",
            values_to = "value"
          ) |>
      mutate(metric = case_when(
            is.na(metric) | metric == "" ~ "effects",
            TRUE ~ "pvalues"
          )) |> 
      pivot_wider(
            names_from = metric,
            values_from = value
          ) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + 
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
    
    p2 <- ggplot(strain_diffs, aes(x = AP_domain, y = effects, fill = AP_domain)) +
      geom_col(alpha = 0.8) + 
      geom_hline(yintercept = 0, linetype = "dashed") +
      scale_fill_manual(values = domain_colors) + 
      labs(y = "Strain Effects (CAST vs B6)", x = "") +
      theme_pubclean(base_size = 16) + 
      theme(legend.position = "none")
    
    p1 + p2 + plot_layout(widths = c(2.5, 1))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  
  if(length(plots) == 0) return(NULL)
  
  wrap_plots(plots, ncol = 1)
}

plot_time_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data$symbol)[1]
    
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(aes(shape = AP_domain), alpha = 0.5, size = 2.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1.5) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "right", plot.title = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) + geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "none", strip.text = element_text(face = "bold"))
    
    p1 + p2 + plot_layout(nrow = 1, widths = c(1, 2.5))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect")
}

plot_threeway_interaction <- function(gene_ids, pattern_desc, count_data, n = 4) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.5, size = 2) + geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect") & theme( legend.position = "bottom")
}

plot_domain_specificity <- function(gene_ids, domains_affected, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain), position = position_jitterdodge(jitter.width = 0.1), alpha = 0.4, size = 1.5) +
      scale_y_log10(labels = scales::comma) + scale_fill_manual(values = strain_colors) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = paste("Strain effect in", domains_affected), x = "Domain", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) + geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "none", strip.text = element_text(face = "bold"))
    p1 / p2 + plot_layout(heights = c(1, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1,  guides = "collect") & theme( legend.position = "bottom")
}
```

```{r}
#| label: load-data
#| message: false
#| warning: false

sample_details <- read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 
load("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/palate_gbrs_emase_counts_v4.RData")

contam_samples <- c("B6_12.5_PM_1","B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat_nondo[, !colnames(expr_mat_nondo) %in% contam_samples]
rm(expr_mat_do, expr_mat_nondo)

all_genes_palate <- read_tsv(here("_data","ENSMUSGid_to_symbol_v105.txt"),
                              col_names = c("gene_id","symbol","gene_biotype","chromosome","start","end"), 
                              skip = 1) |>  
  mutate(midpoint = (start + end) / 2) |>
  filter(gene_id %in% rownames(raw_expr_mat))

sample_annot <- sample_details |>
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |>
  separate(Strain_stock, into = c("strain","tmp"), remove = FALSE, sep = "_") |>
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, 
         palate_closure, L_R_palate) |>
  mutate(strain = if_else(strain == "B6J", "B6", strain)) |>
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |>
  select(-temp) |>
  mutate(Palate_ori = if_else(Palate_ori == "L", "Left", "Right"))
```

# Differential gene expression analysis overview

```{r}
#| label: prep-run-deseq2
#| message: false
#| warning: false

# ── Prep data for DESeq2 ──────────────────────────────────────────────────────
b6_cast_samples <- sample_annot |>
  filter(strain %in% c("B6", "CAST")) |>
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = if_else(is.na(rep), "1", rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

coldata <- b6_cast_samples |>
  select(sample, unique_sampleid, sex, ind, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(coldata), drop = FALSE])

strain_dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~1)
strain_dds_coll <- collapseReplicates(strain_dds, strain_dds$unique_sampleid)

min_LB_stage <- min(colData(strain_dds_coll)$LB_stage)
colData(strain_dds_coll)$LB_stage_from_start <- colData(strain_dds_coll)$LB_stage - min_LB_stage
colData(strain_dds_coll)$sex <- factor(colData(strain_dds_coll)$sex)
colData(strain_dds_coll)$strain <- factor(colData(strain_dds_coll)$strain, levels = c("B6", "CAST"))
colData(strain_dds_coll)$ind <- factor(colData(strain_dds_coll)$ind)

# ── Fit full interaction model ──────────────────────────────────────────────
design(strain_dds_coll) <- ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start

strain_dds_full <- DESeq(strain_dds_coll, parallel = TRUE)
coef_names <- resultsNames(strain_dds_full)

# ── Fit additive model (no strain interactions) ────────────────────────────
# The strain coefficient here represents the marginal CAST vs B6 effect,
# averaged across all domains and developmental timepoints.
strain_dds_additive <- strain_dds_coll
design(strain_dds_additive) <- ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + strain
strain_dds_additive <- DESeq(strain_dds_additive, parallel = TRUE)

# ── LRT tests  ────────────────────────────────────────────
# These tests will get us the genes with strain and strain interaction effects.
# Any strain effect at all
strain_dds_lrt <- DESeq(strain_dds_additive, test = "LRT", 
                         full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                               strain,
                        reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind ,
                        parallel = TRUE)
res_any_strain <- results(strain_dds_lrt)
genes_w_strain_eff <- res_any_strain |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05 )

# Domain × Strain interaction
dds_domain_lrt <- DESeq(strain_dds_coll, test = "LRT",
                        full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                               strain + strain:AP_domain + strain:LB_stage_from_start + 
                               strain:AP_domain:LB_stage_from_start,
                        reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                   strain + strain:LB_stage_from_start,
                        parallel = TRUE)
res_domain_lrt <- results(dds_domain_lrt)
genes_w_strain_by_domain_eff <- res_domain_lrt |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05)
# Time × Strain interaction
dds_time_lrt <- DESeq(strain_dds_coll, test = "LRT",
                      full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start,
                      reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                 strain + strain:AP_domain,
                      parallel = TRUE)
res_time_lrt <- results(dds_time_lrt)
genes_w_strain_by_time_eff <- res_time_lrt |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05)
# Three-way interaction (domain × time × strain)
dds_3way_lrt <- DESeq(strain_dds_coll, test = "LRT",
                      full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start,
                      reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                 strain + strain:AP_domain + strain:LB_stage_from_start,
                      parallel = TRUE)
res_3way_lrt <- results(dds_3way_lrt)
genes_w_3way_eff <- res_3way_lrt |> 
  as_tibble( rownames = "gene_id") |> 
  filter( padj < 0.05)
# ── Extract main strain effect (additive model) ────────────────────────────
res_strain_main <- results(strain_dds_additive, name = "strain_CAST_vs_B6") 

# ── Extract interaction effects (full model) ───────────────────────────────
# Domain × strain interaction coefficients (how CAST effect in MAX/POST differs from PM)
res_domain_int_MAX <- results(strain_dds_full, name = "AP_domainMAX.strainCAST") 
res_domain_int_POST <- results(strain_dds_full, name = "AP_domainPOST.strainCAST")

# Time × strain interaction (change in log2FC per unit developmental time)
res_time_int <- results(strain_dds_full, name = "LB_stage_from_start.strainCAST") 

# Three-way interaction coefficients (how temporal slope differs by domain)
res_3way_MAX <- results(strain_dds_full, name = "AP_domainMAX.LB_stage_from_start.strainCAST") 
res_3way_POST <- results(strain_dds_full, name = "AP_domainPOST.LB_stage_from_start.strainCAST") 

# ── Extract domain-specific strain effects (full model contrasts) ──────────
# Strain effect in PM = strain coefficient from full model (PM is reference, time=0)
res_strain_PM <- results(strain_dds_full, name = "strain_CAST_vs_B6") 
# Strain effect in MAX = main + strainCAST:AP_domainMAX
res_strain_MAX <- results(strain_dds_full,
                          contrast = list(c("strain_CAST_vs_B6", "AP_domainMAX.strainCAST"))) 
# Strain effect in POST = main + strainCAST:AP_domainPOST  
res_strain_POST <- results(strain_dds_full,
                           contrast = list(c("strain_CAST_vs_B6", "AP_domainPOST.strainCAST"))) 


# ── Extract time-specific strain effects (full model contrasts) ──────────
# Note: Since time is a continuous variable I can't get an overall effect like I do with the domains. That is why I am only going to use the interaction term.


# ── Apply quality & LRT filters ──────────────────────────────────────────────────
# Main strain effect: from additive model, including genes with interactions
res_strain_main_filtered <- filter_deseq_results(res_strain_main) |> 
  filter(gene_id %in% genes_w_strain_eff$gene_id)
genes_main_strain_pass <- res_strain_main_filtered$gene_id

# Strain effect in each domain (from full model)
res_strain_PM_filtered <- filter_deseq_results(res_strain_PM) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_strain_MAX_filtered <- filter_deseq_results(res_strain_MAX) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_strain_POST_filtered <- filter_deseq_results(res_strain_POST) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)

genes_strain_PM_pass <- res_strain_PM_filtered$gene_id
genes_strain_MAX_pass <- res_strain_MAX_filtered$gene_id
genes_strain_POST_pass <- res_strain_POST_filtered$gene_id

# Domain × strain interaction coefficients (differences from PM)
res_domain_int_MAX_filtered <- filter_deseq_results(res_domain_int_MAX, min_abs_lfc = 0.3) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_domain_int_POST_filtered <- filter_deseq_results(res_domain_int_POST, min_abs_lfc = 0.3) |> 
  filter( gene_id %in% genes_w_strain_by_domain_eff$gene_id)

genes_domain_int_MAX_pass <- res_domain_int_MAX_filtered$gene_id
genes_domain_int_POST_pass <- res_domain_int_POST_filtered$gene_id

# Time × strain interaction (slope — different scale)
res_time_int_filtered <- filter_deseq_results(res_time_int, min_abs_lfc = 0.1, max_lfcse = 0.5) |> 
  filter( gene_id %in% genes_w_strain_by_time_eff$gene_id)
genes_time_int_pass <- res_time_int_filtered$gene_id

# Three-way coefficients
res_3way_MAX_filtered <- filter_deseq_results(res_3way_MAX, min_abs_lfc = 0.1, max_lfcse = 0.5) |> 
  filter( gene_id %in% genes_w_3way_eff$gene_id)
res_3way_POST_filtered <- filter_deseq_results(res_3way_POST, min_abs_lfc = 0.1, max_lfcse = 0.5) |> 
  filter( gene_id %in% genes_w_3way_eff$gene_id)

genes_3way_MAX_pass <- res_3way_MAX_filtered$gene_id
genes_3way_POST_pass <- res_3way_POST_filtered$gene_id


genes_with_any_interaction <- unique( c(genes_w_strain_by_time_eff$gene_id,
                                     genes_w_strain_by_domain_eff$gene_id,
                                     genes_w_3way_eff$gene_id)
                                     )
# ── Filter summary ─────────────────────────────────────────────────────────
filter_summary <- tibble(
  Result = c("Main strain effect (additive model)",
             "Strain in PM", "Strain in MAX", "Strain in POST",
             "Domain int (MAX)", "Domain int (POST)", "Time int", 
             "3-way (MAX)", "3-way (POST)"),
  Total_tested = c(nrow(res_strain_main), nrow(res_strain_PM), nrow(res_strain_MAX),
                   nrow(res_strain_POST), nrow(res_domain_int_MAX), nrow(res_domain_int_POST),
                   nrow(res_time_int), nrow(res_3way_MAX), nrow(res_3way_POST)),
  Passing_filter = c(length(genes_main_strain_pass), length(genes_strain_PM_pass),
                     length(genes_strain_MAX_pass), length(genes_strain_POST_pass),
                     length(genes_domain_int_MAX_pass), length(genes_domain_int_POST_pass),
                     length(genes_time_int_pass), length(genes_3way_MAX_pass),
                     length(genes_3way_POST_pass)),
 Notes = c(paste0("Includes ", length(genes_with_any_interaction), " genes with domain, time, or three-way interactions"),
            rep("From full model", 3),
            rep("Interaction coefficient", 2),
            "Slope coefficient", rep("Three-way coefficient", 2))
)

# ── Combine results into classified table ──────────────────────────────────
strain_effects_classified <- tibble::tibble(
  gene_id = rownames(res_any_strain),
  baseMean = res_any_strain$baseMean,
  
  # LRT significance T/F
  any_strain_sig = res_any_strain$padj < 0.05, 
  domain_int_sig = res_domain_lrt$padj < 0.05,
  time_int_sig = res_time_lrt$padj < 0.05,
  threeway_int_sig = res_3way_lrt$padj < 0.05,
  
  # LRT p-values
  any_strain_pval = res_any_strain$padj,
  domain_int_pval = res_domain_lrt$padj,
  time_int_pval = res_time_lrt$padj,
  threeway_int_pval = res_3way_lrt$padj,
  
  # Main strain effect (additive model — marginal, averaged over domain and time)
  main_strain_lfc = res_strain_main$log2FoldChange,
  main_strain_lfcSE = res_strain_main$lfcSE,
  main_strain_pval = res_strain_main$padj,
  main_strain_pass = gene_id %in% genes_main_strain_pass,
  has_any_interaction = gene_id %in% genes_with_any_interaction,
  
  # Strain effect within each domain (full model)
  strain_effect_PM = res_strain_PM$log2FoldChange,
  strain_effect_PM_lfcSE = res_strain_PM$lfcSE,
  strain_effect_PM_pval = res_strain_PM$padj,
  strain_effect_PM_pass = gene_id %in% genes_strain_PM_pass,
  
  strain_effect_MAX = res_strain_MAX$log2FoldChange,
  strain_effect_MAX_lfcSE = res_strain_MAX$lfcSE,
  strain_effect_MAX_pval = res_strain_MAX$padj,
  strain_effect_MAX_pass = gene_id %in% genes_strain_MAX_pass,
  
  strain_effect_POST = res_strain_POST$log2FoldChange,
  strain_effect_POST_lfcSE = res_strain_POST$lfcSE,
  strain_effect_POST_pval = res_strain_POST$padj,
  strain_effect_POST_pass = gene_id %in% genes_strain_POST_pass,
  
  # Domain × strain interaction coefficients (full model)
  domain_int_MAX_lfc = res_domain_int_MAX$log2FoldChange,
  domain_int_MAX_lfcSE = res_domain_int_MAX$lfcSE,
  domain_int_MAX_pass = gene_id %in% genes_domain_int_MAX_pass,
  
  domain_int_POST_lfc = res_domain_int_POST$log2FoldChange,
  domain_int_POST_lfcSE = res_domain_int_POST$lfcSE,
  domain_int_POST_pass = gene_id %in% genes_domain_int_POST_pass,
  
  # Time × strain interaction (full model)
  time_int_slope = res_time_int$log2FoldChange,
  time_int_slope_lfcSE = res_time_int$lfcSE,
  time_int_slope_pval = res_time_int$padj,
  time_int_pass = gene_id %in% genes_time_int_pass,
  
  # Three-way interaction coefficients (full model)
  threeway_MAX_slope = res_3way_MAX$log2FoldChange,
  threeway_MAX_slope_lfcSE = res_3way_MAX$lfcSE,
  threeway_MAX_pass = gene_id %in% genes_3way_MAX_pass,
  
  threeway_POST_slope = res_3way_POST$log2FoldChange,
  threeway_POST_slope_lfcSE = res_3way_POST$lfcSE,
  threeway_POST_pass = gene_id %in% genes_3way_POST_pass
) |>
  tidyr::replace_na(list(
    any_strain_sig = FALSE, domain_int_sig = FALSE,
    time_int_sig = FALSE, threeway_int_sig = FALSE,
    main_strain_pass = FALSE, has_any_interaction = FALSE,
    strain_effect_PM_pass = FALSE, strain_effect_MAX_pass = FALSE,
    strain_effect_POST_pass = FALSE,
    domain_int_MAX_pass = FALSE, domain_int_POST_pass = FALSE,
    time_int_pass = FALSE, threeway_MAX_pass = FALSE, threeway_POST_pass = FALSE
  )) |>
  dplyr::left_join(all_genes_palate |> dplyr::select(gene_id, symbol, gene_biotype), by = "gene_id") |>
  dplyr::mutate(
    # Overall filter: significant strain LRT, passes QC
    passes_strain_filter = any_strain_sig & main_strain_pass, #& !has_any_interaction,
    
    # Main strain pattern
    # Note: min_abs_lfc = 0.5 filter ensures no genes fall below 0.5
    main_strain_pattern = dplyr::case_when(
      !passes_strain_filter ~ "No effect / filtered",
      main_strain_lfc > 1 ~ "CAST >> B6 (strong)",
      main_strain_lfc > 0.5 ~ "CAST > B6 (moderate)",
      main_strain_lfc < -1 ~ "B6 >> CAST (strong)",
      main_strain_lfc < -0.5 ~ "B6 > CAST (moderate)"
    ),
    
    # Domain with largest strain effect (only for genes WITH domain interaction)
    max_domain_effect = dplyr::case_when(
      !domain_int_sig ~ NA_character_,
      is.na(strain_effect_PM) | is.na(strain_effect_MAX) | is.na(strain_effect_POST) ~ NA_character_,
      abs(strain_effect_PM) >= abs(strain_effect_MAX) & 
        abs(strain_effect_PM) >= abs(strain_effect_POST) ~ "PM",
      abs(strain_effect_MAX) >= abs(strain_effect_POST) & 
        abs(strain_effect_MAX) >= abs(strain_effect_PM) ~ "MAX",
      abs(strain_effect_POST) >= abs(strain_effect_MAX) & 
        abs(strain_effect_POST) >= abs(strain_effect_PM) ~ "POST"
    ),
    
    max_domain_direction = dplyr::case_when(
      is.na(max_domain_effect) ~ NA_character_,
      max_domain_effect == "PM" & strain_effect_PM > 0 ~ "PM: CAST > B6",
      max_domain_effect == "PM" & strain_effect_PM < 0 ~ "PM: B6 > CAST",
      max_domain_effect == "MAX" & strain_effect_MAX > 0 ~ "MAX: CAST > B6",
      max_domain_effect == "MAX" & strain_effect_MAX < 0 ~ "MAX: B6 > CAST",
      max_domain_effect == "POST" & strain_effect_POST > 0 ~ "POST: CAST > B6",
      max_domain_effect == "POST" & strain_effect_POST < 0 ~ "POST: B6 > CAST"
    ),
    
    domain_consistency = dplyr::case_when(
      is.na(strain_effect_PM) | is.na(strain_effect_MAX) | is.na(strain_effect_POST) ~ NA_character_,
      sign(strain_effect_PM) == sign(strain_effect_MAX) & 
        sign(strain_effect_MAX) == sign(strain_effect_POST) ~ "Consistent",
      TRUE ~ "Variable"
    ),
    
    time_pattern = dplyr::case_when(
      !time_int_sig ~ "No time interaction",
      !time_int_pass ~ "Filtered",
      is.na(time_int_slope) ~ NA_character_,
      time_int_slope > 0.1 ~ "Diverging (CAST changing faster than B6)",
      time_int_slope < -0.1 ~ "Diverging (B6 changing faster than CAST)",
      TRUE ~ "Parallel - Both changing similarly over time"
    ),
    
    # Overall pattern type — now mutually exclusive by construction
    pattern_type = dplyr::case_when(
      !any_strain_sig ~ "No strain effect",
      main_strain_pass ~ "Main strain effect only",
      threeway_int_sig ~ "Three-way interaction",
      domain_int_sig & time_int_sig ~ "Domain + Time interactions",
      domain_int_sig ~ "Domain × Strain interaction",
      time_int_sig ~ "Time × Strain interaction",
      TRUE ~ "Filtered (QC)"
    )
  )

# Filtered dataset: genes with consistent strain effect (no interactions)
strain_effects_filtered <- strain_effects_classified |>
  filter(passes_strain_filter)


# # source the pattern identification script
# source("DEgene_pattern_identification_CAST_B6.R")
# ── Load pattern classification results ────────────────────────────────────
pattern_summary <- read_csv(here("strain_pattern_summary.csv"))
main_patterns <- read_csv(here("main_strain_effect_patterns.csv"))
domain_patterns <- read_csv(here("domain_strain_patterns.csv"))
time_patterns <- read_csv(here("time_strain_patterns.csv"))
threeway_patterns <- read_csv(here("threeway_patterns.csv"))
domain_specificity <- read_csv(here("domain_specificity_patterns.csv"))
gene_assignments <- read_csv(here("strain_gene_pattern_assignments.csv"))

# ── Extract gene lists ─────────────────────────────────────────────────────

# Main strain effects (from additive model - average across contexts)
genes_CAST_high_strong <- main_patterns |>
  filter(main_strain_pattern == "CAST >> B6 (strong)") |> pull(gene_id)
genes_CAST_high_mod <- main_patterns |>
  filter(main_strain_pattern == "CAST > B6 (moderate)") |> pull(gene_id)
genes_B6_high_strong <- main_patterns |>
  filter(main_strain_pattern == "B6 >> CAST (strong)") |> pull(gene_id)
genes_B6_high_mod <- main_patterns |>
  filter(main_strain_pattern == "B6 > CAST (moderate)") |> pull(gene_id)
genes_main_only <- main_patterns |>
  filter(!has_any_interaction) |> pull(gene_id)

# Domain × strain patterns (spatial differences)
genes_PM_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "PM") |> pull(gene_id)
genes_MAX_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "MAX") |> pull(gene_id)
genes_POST_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "POST") |> pull(gene_id)
genes_opposite_directions <- domain_patterns |>
  filter(domain_consistency == "Variable") |> pull(gene_id)

# Domain × strain: temporal stability
genes_domain_static <- domain_patterns |>
  filter(temporal_stability == "Static: Domain pattern stable over time") |> pull(gene_id)
genes_domain_time_affected <- domain_patterns |>
  filter(temporal_stability == "Dynamic: Main time effect present, domain pattern may shift") |> pull(gene_id)
genes_domain_time_varying <- domain_patterns |>
  filter(temporal_stability == "Dynamic: Domain pattern changes over time (3-way)") |> pull(gene_id)

# Time × strain patterns (temporal dynamics)
genes_diverging_CAST_faster <- time_patterns |>
  filter(base_slope > 0.1) |> pull(gene_id)
genes_diverging_B6_faster <- time_patterns |>
  filter(base_slope < -0.1) |> pull(gene_id)

# Time × strain: spatial uniformity
genes_time_uniform <- time_patterns |>
  filter(spatial_pattern == "Uniform: Same temporal dynamics across all domains") |> pull(gene_id)
genes_time_domain_affected <- time_patterns |>
  filter(spatial_pattern == "Domain-affected: Domains differ at t=0, parallel trajectories") |> pull(gene_id)
genes_time_domain_specific <- time_patterns |>
  filter(spatial_pattern == "Domain-specific: Different temporal dynamics across domains") |> pull(gene_id)

# Three-way patterns (domain-specific temporal dynamics)
genes_threeway_complex <- threeway_patterns |>
  filter(str_detect(pattern_type, "Complex")) |> pull(gene_id)
genes_threeway_initial <- threeway_patterns |>
  filter(str_detect(pattern_type, "Primarily initial")) |> pull(gene_id)
genes_threeway_dynamic <- threeway_patterns |>
  filter(str_detect(pattern_type, "Primarily dynamic")) |> pull(gene_id)

# Domain specificity (which domains show effects at t=0)
genes_strain_PM_only <- domain_specificity |>
  filter(specificity_pattern == "PM only") |> pull(gene_id)
genes_strain_MAX_only <- domain_specificity |>
  filter(specificity_pattern == "MAX only") |> pull(gene_id)
genes_strain_POST_only <- domain_specificity |>
  filter(specificity_pattern == "POST only") |> pull(gene_id)
genes_strain_all_domains <- domain_specificity |>
  filter(specificity_pattern == "All domains") |> pull(gene_id)

# Domain specificity: temporal stability
genes_strain_domain_specific_static <- domain_specificity |>
  filter(!has_time_int & specificity_pattern != "All domains" & 
         specificity_pattern != "None (filtered)") |> pull(gene_id)
genes_strain_domain_specific_dynamic <- domain_specificity |>
  filter(has_time_int & specificity_pattern != "All domains" & 
         specificity_pattern != "None (filtered)") |> pull(gene_id)

# ── Pre-compute VST for plotting ──────────────────────────────────────────
vsd_global <- vst(strain_dds_full, blind = FALSE)
```

This analysis investigates how gene expression differs between B6 and CAST mouse strains across three palate domains (PM, MAX, POST) and developmental time. We use a statistical model that decomposes strain effects into:
 
- **Main strain effect**: Average CAST vs B6 difference, consistent across domains and time
- **Domain × Strain interaction**: How strain effect differs between domains
- **Time × Strain interaction**: How strain effect changes per unit developmental time
- **Three-way interaction**: Domain-specific changes in temporal strain dynamics


**Sample preparation and model fitting**

I merge technical replicates arising from re-sequencing of several samples using the `collapseReplicates()` function from DESeq2. I then calculate the relative or centered limb bud (LB) stage (`LB_stage_from_start`) by subtracting the minimum observed stage so that 0 represents the earliest developmental timepoint in the dataset.

I fit two models. The **full interaction model** (`~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + strain + strain:AP_domain + strain:LB_stage_from_start + strain:AP_domain:LB_stage_from_start`) includes strain and all its interactions with domain and developmental time, and is used to extract interaction coefficients and domain-specific strain effects. A separate **additive model** (`~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + strain`) without strain interaction terms is used to estimate the marginal (average) strain effect, representing the overall CAST vs B6 log2 fold change averaged across all domains and developmental timepoints. In the full interaction model, the `strain_CAST_vs_B6` coefficient would instead represent the strain effect conditional on the reference domain (PM) at time = 0, which is not the quantity of interest for genes with consistent effects.

**Identifying significant effects with LRT**

I use likelihood ratio tests (LRT) to identify genes with significant strain-related effects. The overall strain LRT compares the full model to a reduced model lacking all strain terms (`~ sex + AP_domain + LB_stage_from_start + AP_domain:ind`). Separate LRT tests identify genes with significant domain × strain interactions (removing `strain:AP_domain` and `strain:AP_domain:LB_stage_from_start` terms), time × strain interactions (removing `strain:LB_stage_from_start` and `strain:AP_domain:LB_stage_from_start` terms), and three-way interactions (removing only `strain:AP_domain:LB_stage_from_start` terms).

**Extracting effect sizes**

The LRT tests identify **which** genes have significant effects but do not provide effect sizes. To get the magnitude and direction of effects, I extract coefficients using Wald tests:

- **Main strain effect**: From the additive model — represents the average CAST vs B6 log2 fold change.
- **Domain-specific strain effects**: From the full model using contrasts. The strain effect in PM is the `strain_CAST_vs_B6` coefficient (PM is the reference level at time = 0). The strain effect in MAX combines the main effect with the `strainCAST:AP_domainMAX` interaction term, and similarly for POST.
- **Domain interaction coefficients**: From the full model — `strainCAST:AP_domainMAX` and `strainCAST:AP_domainPOST` indicate how much the strain effect in MAX and POST differs from PM.
- **Time interaction coefficient**: From the full model — `strainCAST:LB_stage_from_start` represents the change in log2FC(CAST/B6) per unit of developmental time.
- **Three-way coefficients**: From the full model — indicate how the temporal slope of the strain effect differs by domain.

**Quality filtering**

I apply quality filters to each result object individually, removing genes with low expression (baseMean below the 10th percentile), high uncertainty (lfcSE ≥ 1.5 for effect estimates, ≥ 0.5 for slope coefficients), unrealistically large effects (|log2FC| ≥ 10), or very small effects (|log2FC| < 0.5 for main effects, < 0.3 for interaction coefficients, < 0.1 for slope coefficients). This per-result filtering ensures that each effect estimate meets quality standards appropriate for its scale.

**Results summary**

From this analysis, I have identified **`r sum(strain_effects_classified$any_strain_sig, na.rm = TRUE)`** genes with significant strain effects (padj < 0.05). Of these, **`r nrow(strain_effects_filtered)`** have a main strain effect — they pass quality filters on the additive model's strain coefficient and the strain effect LRT. These significant genes are further classified into interaction categories based on their LRT results.

**Pattern classification**

I use the `DEgene_pattern_identification_CAST_B6.R` script to classify differentially expressed genes into expression pattern categories. The classification is hierarchical: genes are first separated into those with consistent effects versus those with interactions, and interaction genes are further sub-classified by the nature of their interaction.

**Domain × Strain interaction patterns** are assigned to genes with a significant domain × strain LRT (padj < 0.05) and at least one domain interaction coefficient that passes quality filters. These genes are classified by which domain shows the largest strain effect magnitude (PM, MAX, or POST) and whether the strain effect direction is consistent or variable across domains. A subset of particular interest are genes with opposite directions — where the strain effect reverses sign between domains. Additionally, these patterns are characterized by their **temporal stability**: (1) Static patterns show stable domain differences throughout development (no time or three-way interactions); (2) Time-affected patterns may shift as both strains change over time (time interaction present, but no three-way interaction); (3) Time-varying patterns actively change their spatial structure during development (three-way interaction significant).

**Domain specificity patterns** further characterize domain × strain interaction genes by identifying which specific domains show significant and well-estimated strain effects at t=0: PM only, MAX only, POST only, pairwise combinations, or all three domains. These patterns are additionally flagged as **temporally static** (no time or three-way interactions — domain specificity remains constant throughout development) or **temporally dynamic** (time or three-way interactions present — domain specificity pattern evolves during development).

**Time × Strain interaction patterns** are assigned to genes with a significant time × strain LRT and a time interaction slope that passes quality filters. These are classified as diverging with CAST increasing faster (positive slope > 0.1), diverging with B6 increasing faster (negative slope < -0.1), or parallel/weak. Strong divergence (|slope| > 0.2) is distinguished from moderate divergence. Additionally, these patterns are characterized by their **spatial uniformity**: (1) Uniform patterns show the same temporal dynamics across all domains (no domain or three-way interactions); (2) Domain-affected patterns have domains that differ at t=0 but show parallel temporal trajectories (domain interaction present, no three-way interaction); (3) Domain-specific patterns show different temporal dynamics in different domains (three-way interaction present).

**Three-way interaction patterns** are assigned to genes with a significant three-way LRT and at least one three-way coefficient passing quality filters. The domain-specific temporal slopes are reconstructed by combining the base time × strain slope with the three-way adjustments. These genes are classified based on **whether domains differ primarily in their initial state (t=0), their temporal dynamics (slopes), or both**: (1) Complex patterns show substantial differences across both initial effects and temporal trajectories; (2) Primarily initial patterns have domains that differ at t=0 but follow similar temporal trajectories; (3) Primarily dynamic patterns start similar but diverge over time with domain-specific slopes. Within each pattern type, genes are further characterized by which domain is most distinct when considering both initial state and temporal dynamics combined.


```{r}
#| label: pattern-summary-table
#| message: false
#| warning: false

pattern_summary |> 
  create_dt()
```

# Main Strain Effects

These genes show consistent expression differences between CAST and B6. The log2FC values represent the CAST vs B6 effect estimated from the model. 

:::: {#main-strain-tabs .panel-tabset .nav-pills}

## CAST >> B6 (Strong)

Genes with strong upregulation in CAST (log2FC > 1). Here are some examples:

```{r}
#| label: main-CAST-strong-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6


count_data <- get_normalized_counts(strain_dds_full, genes_CAST_high_strong, vsd = vsd_global)
plot_main_strain_effect(genes_CAST_high_strong, "CAST >> B6 (strong)", count_data, n = 2)


```

Here is the full list of genes:
```{r}
#| label: main-CAST-strong-table
#| message: false
#| warning: false

main_patterns |> 
  filter(gene_id %in% genes_CAST_high_strong) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```

Here is the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-CAST-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_strong_ora <- run_enrichment_hybrid(genes_CAST_high_strong, 
                                                    description = "CAST >> B6 (strong)")

```

```{r}
#| label: main-CAST-strong-ora-table
#| message: false
#| warning: false

if (!is.null(genes_CAST_high_strong_ora) && !is.null(genes_CAST_high_strong_ora$result)) {
  genes_CAST_high_strong_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}

```


## CAST > B6 (Moderate)

Genes with moderate upregulation in CAST (0.5 < log2FC ≤ 1).

```{r}
#| label: main-CAST-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_normalized_counts(strain_dds_full, 
                                    genes_CAST_high_mod, vsd = vsd_global)

plot_main_strain_effect(genes_CAST_high_mod, "CAST > B6 (moderate)", 
                        count_data, n = 2)

```


```{r}
#| label: main-CAST-mod-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_CAST_high_mod) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-CAST-mod-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_mod_ora <- run_enrichment_hybrid(genes_CAST_high_mod,
                                                 description = "CAST > B6 (moderate)")

```

```{r}
#| label: main-CAST-mod-ora-table
#| message: false
#| warning: false

if (!is.null(genes_CAST_high_mod_ora) && !is.null(genes_CAST_high_mod_ora$result)) {
  genes_CAST_high_mod_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}

```

## B6 >> CAST (Strong)

Genes with strong upregulation in B6 (log2FC < -1).

```{r}
#| label: main-B6-strong-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_normalized_counts(strain_dds_full, 
                                    genes_B6_high_strong, 
                                    vsd = vsd_global)
plot_main_strain_effect(genes_B6_high_strong, "B6 >> CAST (strong)", count_data, n = 2)

```

```{r}
#| label: main-B6-strong-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_B6_high_strong) |>
  select(gene_id,symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-B6-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true


genes_B6_high_strong_ora <- run_enrichment_hybrid(genes_B6_high_strong,
                                                  description = "B6 >> CAST (strong)")

```

```{r}
#| label: main-B6-strong-ora-table
#| message: false
#| warning: false


if (!is.null(genes_B6_high_strong_ora) && !is.null(genes_B6_high_strong_ora$result)) {
  genes_B6_high_strong_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}

```


## B6 > CAST (Moderate)

Genes with moderate upregulation in B6 (-1 ≤ log2FC < -0.5).

```{r}
#| label: main-B6-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_B6_high_mod, vsd = vsd_global)
plot_main_strain_effect(genes_B6_high_mod, "B6 > CAST (moderate)", count_data, n = 2)

```

```{r}
#| label: main-B6-mod-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_B6_high_mod) |>
  select(gene_id,symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-B6-mod-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_B6_high_mod_ora <- run_enrichment_hybrid(genes_B6_high_mod, 
                                               description = "B6 > CAST (moderate)")
  
```

```{r}
#| label: main-B6-mod-ora-table
#| message: false
#| warning: false

if (!is.null(genes_B6_high_mod_ora) && !is.null(genes_B6_high_mod_ora$result)) {
  genes_B6_high_mod_ora$result |> 
    filter(p_value < 0.01) |> 
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
  
```

::::

# Domain × Strain Interactions

These genes show strain differences that vary across palate domains. The model provides strain effects within each domain and interaction coefficients showing how MAX/POST differ from PM.

:::: {#domain-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-strain-overview
#| message: false
#| warning: false

domain_patterns |> 
  count(detailed_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |> 
  create_dt()

```

## PM: Biggest Effect

Genes where the strain effect is largest in the PM domain. The bar plots show the percent difference between CAST and B6 mean expression in each domain, highlighting that PM has the dominant strain effect for these genes.

```{r}
#| label: domain-PM-biggest-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_PM_biggest_effect, vsd = vsd_global)
plot_domain_strain_interaction(genes_PM_biggest_effect, "Biggest strain effect in PM", count_data, n = 2)

```

```{r}
#| label: domain-PM-biggest-table
#| message: false
#| warning: false

domain_patterns |> 
  filter(gene_id %in% genes_PM_biggest_effect) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_int_MAX_lfc, domain_int_POST_lfc, max_domain_direction, temporal_stability) |> 
  create_dt()

```

```{r}
#| label: domain-PM-biggest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in PM
genes_PM_biggest_up <- domain_patterns |> 
  filter(gene_id %in% genes_PM_biggest_effect, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_PM_biggest_down <- domain_patterns |> 
  filter(gene_id %in% genes_PM_biggest_effect, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_PM_biggest_effect_ora_up <- run_enrichment_hybrid(
  genes_PM_biggest_up, 
  description = "PM biggest effect - Up in CAST"
)

genes_PM_biggest_effect_ora_down <- run_enrichment_hybrid(
  genes_PM_biggest_down, 
  description = "PM biggest effect - Down in CAST"
)
  
```

```{r}
#| label: domain-PM-biggest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_PM_biggest_effect_ora_up$result)) genes_PM_biggest_effect_ora_up$result,
  if (!is.null(genes_PM_biggest_effect_ora_down$result)) genes_PM_biggest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}
  
```

## MAX: Biggest Effect

Genes where the strain effect is largest in the MAX domain. These genes show the most pronounced CAST vs B6 differences in the middle portion of the palate.

```{r}
#| label: domain-MAX-biggest-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_MAX_biggest_effect, vsd = vsd_global)
plot_domain_strain_interaction(genes_MAX_biggest_effect, "Biggest strain effect in MAX", count_data, n = 2)

```

```{r}
#| label: domain-MAX-biggest-table
#| message: false
#| warning: false

domain_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_effect) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_int_MAX_lfc, domain_int_POST_lfc, max_domain_direction, temporal_stability) |> 
  create_dt()

```

```{r}
#| label: domain-MAX-biggest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in MAX
genes_MAX_biggest_up <- domain_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_effect, strain_effect_MAX > 0) |> 
  pull(gene_id)

genes_MAX_biggest_down <- domain_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_effect, strain_effect_MAX < 0) |> 
  pull(gene_id)

genes_MAX_biggest_effect_ora_up <- run_enrichment_hybrid(
  genes_MAX_biggest_up, 
  description = "MAX biggest effect - Up in CAST"
)

genes_MAX_biggest_effect_ora_down <- run_enrichment_hybrid(
  genes_MAX_biggest_down, 
  description = "MAX biggest effect - Down in CAST"
)
  
```

```{r}
#| label: domain-MAX-biggest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_MAX_biggest_effect_ora_up$result)) genes_MAX_biggest_effect_ora_up$result,
  if (!is.null(genes_MAX_biggest_effect_ora_down$result)) genes_MAX_biggest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}
  
```

## POST: Biggest Effect

Genes where the strain effect is largest in the POST domain. These genes show the most pronounced CAST vs B6 differences in the posterior palate.

```{r}
#| label: domain-POST-biggest-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_POST_biggest_effect, vsd = vsd_global)
plot_domain_strain_interaction(genes_POST_biggest_effect, "Biggest strain effect in POST", count_data, n = 2)

```

```{r}
#| label: domain-POST-biggest-table
#| message: false
#| warning: false

domain_patterns |> 
  filter(gene_id %in% genes_POST_biggest_effect) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_int_MAX_lfc, domain_int_POST_lfc, max_domain_direction, temporal_stability) |> 
  create_dt()

```

```{r}
#| label: domain-POST-biggest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in POST
genes_POST_biggest_up <- domain_patterns |> 
  filter(gene_id %in% genes_POST_biggest_effect, strain_effect_POST > 0) |> 
  pull(gene_id)

genes_POST_biggest_down <- domain_patterns |> 
  filter(gene_id %in% genes_POST_biggest_effect, strain_effect_POST < 0) |> 
  pull(gene_id)

genes_POST_biggest_effect_ora_up <- run_enrichment_hybrid(
  genes_POST_biggest_up, 
  description = "POST biggest effect - Up in CAST"
)

genes_POST_biggest_effect_ora_down <- run_enrichment_hybrid(
  genes_POST_biggest_down, 
  description = "POST biggest effect - Down in CAST"
)

```

```{r}
#| label: domain-POST-biggest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_POST_biggest_effect_ora_up$result)) genes_POST_biggest_effect_ora_up$result,
  if (!is.null(genes_POST_biggest_effect_ora_down$result)) genes_POST_biggest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Opposite Directions

Genes where the strain effect reverses direction across domains — for example, higher in CAST in one domain but higher in B6 in another. These are particularly interesting as they suggest domain-specific regulatory divergence between strains.

```{r}
#| label: domain-opposite-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_opposite_directions, vsd = vsd_global)
plot_domain_strain_interaction(genes_opposite_directions, "Opposite directions across domains", count_data, n = 2)

```

```{r}
#| label: domain-opposite-table
#| message: false
#| warning: false

domain_patterns |> 
  filter(gene_id %in% genes_opposite_directions) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         domain_consistency, temporal_stability) |> 
  create_dt()

```

```{r}
#| label: domain-opposite-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction using PM as reference
genes_opposite_up_in_PM <- domain_patterns |> 
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_opposite_down_in_PM <- domain_patterns |> 
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_opposite_directions_ora_up <- run_enrichment_hybrid(
  genes_opposite_up_in_PM, 
  description = "Opposite directions - Up in CAST in PM"
)

genes_opposite_directions_ora_down <- run_enrichment_hybrid(
  genes_opposite_down_in_PM, 
  description = "Opposite directions - Down in CAST in PM"
)
 
```

```{r}
#| label: domain-opposite-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_opposite_directions_ora_up$result)) genes_opposite_directions_ora_up$result,
  if (!is.null(genes_opposite_directions_ora_down$result)) genes_opposite_directions_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Static Pattern

Domain differences that remain stable over developmental time (no time or three-way interaction). These represent fixed spatial differences in strain effects.

```{r}
#| label: domain-static-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_domain_static, vsd = vsd_global)
plot_domain_strain_interaction(genes_domain_static, "Static domain pattern", count_data, n = 2)

```

```{r}
#| label: domain-static-table
#| message: false
#| warning: false

domain_patterns |> 
  filter(gene_id %in% genes_domain_static) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         max_domain_effect, domain_consistency, temporal_stability) |> 
  create_dt()

```

## Time-Affected Pattern

Domain differences where main time effects are present, so the spatial pattern may shift over development, but without domain-specific temporal dynamics (no three-way interaction).

```{r}
#| label: domain-time-affected-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_domain_time_affected, vsd = vsd_global)
plot_domain_strain_interaction(genes_domain_time_affected, "Time-affected domain pattern", count_data, n = 2)

```

```{r}
#| label: domain-time-affected-table
#| message: false
#| warning: false

domain_patterns |> 
  filter(gene_id %in% genes_domain_time_affected) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         max_domain_effect, domain_consistency, temporal_stability) |> 
  create_dt()

```

## Time-Varying Pattern

Domain differences that actively change over developmental time (three-way interaction significant). The spatial pattern of strain effects evolves during development.

```{r}
#| label: domain-time-varying-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_domain_time_varying, vsd = vsd_global)
plot_domain_strain_interaction(genes_domain_time_varying, "Time-varying domain pattern", count_data, n = 2)

```

```{r}
#| label: domain-time-varying-table
#| message: false
#| warning: false

domain_patterns |> 
  filter(gene_id %in% genes_domain_time_varying) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         max_domain_effect, domain_consistency, temporal_stability) |> 
  create_dt()

```

::::

# Time × Strain Interactions

These genes show strain differences that change over developmental time. The temporal slope coefficient represents the change in log2FC(CAST/B6) per unit of developmental time.

:::: {#time-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: time-strain-overview
#| message: false
#| warning: false

time_patterns |> 
  count(detailed_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |> 
  create_dt()

```

## Diverging: CAST Faster

Genes where strains diverge over developmental time, with CAST expression increasing (or decreasing less) relative to B6 as development progresses. The positive time interaction slope indicates that the log2FC(CAST/B6) becomes more positive over time.

```{r}
#| label: time-CAST-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_diverging_CAST_faster, vsd = vsd_global)
plot_time_strain_interaction(genes_diverging_CAST_faster, "Diverging: CAST changing faster", count_data, n = 2)

```

```{r}
#| label: time-CAST-faster-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster) |>
  select(gene_id, symbol, base_slope, time_int_pval, main_strain_lfc, 
         trajectory_pattern, spatial_pattern) |>
  create_dt()

```

```{r}
#| label: time-CAST-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_CAST_faster_up <- time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc > 0) |> 
  pull(gene_id)

genes_CAST_faster_down <- time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc < 0) |> 
  pull(gene_id)

genes_diverging_CAST_faster_ora_up <- run_enrichment_hybrid(
  genes_CAST_faster_up, 
  description = "CAST faster - Up in CAST at baseline"
)

genes_diverging_CAST_faster_ora_down <- run_enrichment_hybrid(
  genes_CAST_faster_down, 
  description = "CAST faster - Down in CAST at baseline"
)

```

```{r}
#| label: time-CAST-faster-ora-table
#| message: false
#| warning: false

combined_diverging_cast_ora_results <- bind_rows(
  genes_diverging_CAST_faster_ora_up$result,
  genes_diverging_CAST_faster_ora_down$result
) 

if (!is.null(combined_diverging_cast_ora_results) && nrow(combined_diverging_cast_ora_results) > 0) {
  combined_diverging_cast_ora_results |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Diverging: B6 Faster

Genes where strains diverge over developmental time, with B6 expression increasing (or decreasing less) relative to CAST as development progresses. The negative time interaction slope indicates that the log2FC(CAST/B6) becomes more negative over time.

```{r}
#| label: time-B6-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_diverging_B6_faster, vsd = vsd_global)
plot_time_strain_interaction(genes_diverging_B6_faster, "Diverging: B6 changing faster", count_data, n = 4)

```

```{r}
#| label: time-B6-faster-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster) |>
  select(gene_id, symbol, base_slope, time_int_pval, main_strain_lfc, 
         trajectory_pattern, spatial_pattern) |> 
  create_dt()

```

```{r}
#| label: time-B6-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_B6_faster_up <- time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc > 0) |> 
  pull(gene_id)

genes_B6_faster_down <- time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc < 0) |> 
  pull(gene_id)

genes_diverging_B6_faster_ora_up <- run_enrichment_hybrid(
  genes_B6_faster_up, 
  description = "B6 faster - Up in CAST at baseline"
)

genes_diverging_B6_faster_ora_down <- run_enrichment_hybrid(
  genes_B6_faster_down, 
  description = "B6 faster - Down in CAST at baseline"
)

```

```{r}
#| label: time-B6-faster-ora-table
#| message: false
#| warning: false

combined_diverging_b6_ora_results <- bind_rows(
  genes_diverging_B6_faster_ora_up$result,
  genes_diverging_B6_faster_ora_down$result
) 

if (!is.null(combined_diverging_b6_ora_results) && nrow(combined_diverging_b6_ora_results) > 0) {
  combined_diverging_b6_ora_results |> 
    filter(p_value < 0.01) |> 
    create_dt()
}

```

## Uniform Across Domains

Temporal strain dynamics that are consistent across all palate domains (no domain or three-way interaction). These represent spatially uniform temporal changes.

```{r}
#| label: time-uniform-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_time_uniform, vsd = vsd_global)
plot_time_strain_interaction(genes_time_uniform, "Uniform temporal dynamics", count_data, n = 2)

```

```{r}
#| label: time-uniform-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_time_uniform) |>
  select(gene_id, symbol, base_slope, main_strain_lfc, trajectory_pattern, spatial_pattern) |> 
  create_dt()

```

## Domain-Affected

Temporal dynamics where domains differ at t=0 but show parallel trajectories over time (domain interaction present, but no three-way interaction).

```{r}
#| label: time-domain-affected-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_time_domain_affected, vsd = vsd_global)
plot_time_strain_interaction(genes_time_domain_affected, "Domain-affected temporal dynamics", count_data, n = 2)

```

```{r}
#| label: time-domain-affected-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_time_domain_affected) |>
  select(gene_id, symbol, base_slope, main_strain_lfc, trajectory_pattern, spatial_pattern) |> 
  create_dt()

```

## Domain-Specific Dynamics

Temporal dynamics that differ across domains (three-way interaction present). Different domains show different temporal trajectories of strain effects.

```{r}
#| label: time-domain-specific-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_time_domain_specific, vsd = vsd_global)
plot_time_strain_interaction(genes_time_domain_specific, "Domain-specific temporal dynamics", count_data, n = 2)

```

```{r}
#| label: time-domain-specific-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_time_domain_specific) |>
  select(gene_id, symbol, base_slope, slope_range, main_strain_lfc, 
         trajectory_pattern, spatial_pattern) |> 
  create_dt()

```

::::

# Three-Way Interactions

These genes show domain-specific strain differences that also change over developmental time. We classify them based on whether domains primarily differ in their initial state (t=0), their temporal dynamics (slopes), or both.

:::: {#threeway-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: threeway-overview
#| message: false
#| warning: false

threeway_patterns |> 
  count(detailed_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |> 
  create_dt()

```

## Complex Pattern

Genes where domains differ in BOTH their initial state (t=0) AND their temporal dynamics. These show substantial differences across both dimensions.

```{r}
#| label: threeway-complex-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_threeway_complex, vsd = vsd_global)
plot_threeway_interaction(genes_threeway_complex, "Complex: Differ in initial state & dynamics", count_data, n = 2)

```

```{r}
#| label: threeway-complex-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex) |>
  select(gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST, 
         init_range, slope_range, most_distinct_domain) |> 
  create_dt()

```

```{r}
#| label: threeway-complex-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_complex_PM <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "PM") |> 
  pull(gene_id)

genes_complex_MAX <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "MAX") |> 
  pull(gene_id)

genes_complex_POST <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "POST") |> 
  pull(gene_id)

genes_threeway_complex_ora_PM <- run_enrichment_hybrid(
  genes_complex_PM, 
  description = "Complex pattern - PM most distinct"
)

genes_threeway_complex_ora_MAX <- run_enrichment_hybrid(
  genes_complex_MAX, 
  description = "Complex pattern - MAX most distinct"
)

genes_threeway_complex_ora_POST <- run_enrichment_hybrid(
  genes_complex_POST, 
  description = "Complex pattern - POST most distinct"
)

```

```{r}
#| label: threeway-complex-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_complex_ora_PM$result)) {
    genes_threeway_complex_ora_PM$result |> 
      mutate(distinct_domain = paste0("PM (n=", length(genes_complex_PM), ")"))
  },
  if (!is.null(genes_threeway_complex_ora_MAX$result)) {
    genes_threeway_complex_ora_MAX$result |> 
      mutate(distinct_domain = paste0("MAX (n=", length(genes_complex_MAX), ")"))
  },
  if (!is.null(genes_threeway_complex_ora_POST$result)) {
    genes_threeway_complex_ora_POST$result |> 
      mutate(distinct_domain = paste0("POST (n=", length(genes_complex_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(distinct_domain, everything()) |> 
    create_dt()
}

```

## Primarily Initial Differences

Genes where domains differ mainly at t=0, but show similar temporal trajectories. The spatial pattern is established early and maintained over time.

```{r}
#| label: threeway-initial-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_threeway_initial, vsd = vsd_global)
plot_threeway_interaction(genes_threeway_initial, "Primarily initial differences", count_data, n = 2)

```

```{r}
#| label: threeway-initial-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial) |>
  select(gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST, 
         init_range, slope_range, most_distinct_domain) |> 
  create_dt()

```

```{r}
#| label: threeway-initial-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_initial_PM <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "PM") |> 
  pull(gene_id)

genes_initial_MAX <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "MAX") |> 
  pull(gene_id)

genes_initial_POST <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "POST") |> 
  pull(gene_id)

genes_threeway_initial_ora_PM <- run_enrichment_hybrid(
  genes_initial_PM, 
  description = "Initial differences - PM most distinct"
)

genes_threeway_initial_ora_MAX <- run_enrichment_hybrid(
  genes_initial_MAX, 
  description = "Initial differences - MAX most distinct"
)

genes_threeway_initial_ora_POST <- run_enrichment_hybrid(
  genes_initial_POST, 
  description = "Initial differences - POST most distinct"
)

```

```{r}
#| label: threeway-initial-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_initial_ora_PM$result)) {
    genes_threeway_initial_ora_PM$result |> 
      mutate(distinct_domain = paste0("PM (n=", length(genes_initial_PM), ")"))
  },
  if (!is.null(genes_threeway_initial_ora_MAX$result)) {
    genes_threeway_initial_ora_MAX$result |> 
      mutate(distinct_domain = paste0("MAX (n=", length(genes_initial_MAX), ")"))
  },
  if (!is.null(genes_threeway_initial_ora_POST$result)) {
    genes_threeway_initial_ora_POST$result |> 
      mutate(distinct_domain = paste0("POST (n=", length(genes_initial_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(distinct_domain, everything()) |> 
    create_dt()
}

```

## Primarily Dynamic Differences

Genes where domains start similar at t=0 but diverge over time with different temporal trajectories. The spatial pattern emerges during development.

```{r}
#| label: threeway-dynamic-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_threeway_dynamic, vsd = vsd_global)
plot_threeway_interaction(genes_threeway_dynamic, "Primarily dynamic differences", count_data, n = 2)

```

```{r}
#| label: threeway-dynamic-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic) |>
  select(gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST, 
         init_range, slope_range, most_distinct_domain) |> 
  create_dt()

```

```{r}
#| label: threeway-dynamic-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_dynamic_PM <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "PM") |> 
  pull(gene_id)

genes_dynamic_MAX <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "MAX") |> 
  pull(gene_id)

genes_dynamic_POST <- threeway_patterns |> 
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "POST") |> 
  pull(gene_id)

genes_threeway_dynamic_ora_PM <- run_enrichment_hybrid(
  genes_dynamic_PM, 
  description = "Dynamic differences - PM most distinct"
)

genes_threeway_dynamic_ora_MAX <- run_enrichment_hybrid(
  genes_dynamic_MAX, 
  description = "Dynamic differences - MAX most distinct"
)

genes_threeway_dynamic_ora_POST <- run_enrichment_hybrid(
  genes_dynamic_POST, 
  description = "Dynamic differences - POST most distinct"
)

```

```{r}
#| label: threeway-dynamic-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_dynamic_ora_PM$result)) {
    genes_threeway_dynamic_ora_PM$result |> 
      mutate(distinct_domain = paste0("PM (n=", length(genes_dynamic_PM), ")"))
  },
  if (!is.null(genes_threeway_dynamic_ora_MAX$result)) {
    genes_threeway_dynamic_ora_MAX$result |> 
      mutate(distinct_domain = paste0("MAX (n=", length(genes_dynamic_MAX), ")"))
  },
  if (!is.null(genes_threeway_dynamic_ora_POST$result)) {
    genes_threeway_dynamic_ora_POST$result |> 
      mutate(distinct_domain = paste0("POST (n=", length(genes_dynamic_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(distinct_domain, everything()) |> 
    create_dt()
}

```

::::

# Domain-Specific Strain Effects

For genes with domain × strain interactions, we identify which specific domains show significant strain effects (at t=0) and whether these patterns are stable or change over time.

:::: {#domain-specific-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-specific-overview
#| message: false
#| warning: false

domain_specificity |> 
  count(specificity_with_time) |> 
  arrange(desc(n)) |>
  rename(`Specificity Pattern` = specificity_with_time, `Number of Genes` = n) |> 
  create_dt()

```

## PM Only

Genes where the strain difference passes significance and effect size thresholds only in the PM domain (at t=0). These represent PM-specific regulatory divergence between B6 and CAST.

```{r}
#| label: domain-specific-PM-only-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

count_data <- get_normalized_counts(strain_dds_full, genes_strain_PM_only, vsd = vsd_global)
plot_domain_specificity(genes_strain_PM_only, "PM only", count_data, n = 2)

```

```{r}
#| label: domain-specific-PM-only-table
#| message: false
#| warning: false

domain_specificity |> 
  filter(gene_id %in% genes_strain_PM_only) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         specificity_pattern, has_time_int) |> 
  create_dt()

```

```{r}
#| label: domain-specific-PM-only-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by strain effect direction in PM
genes_PM_only_up <- domain_specificity |> 
  filter(gene_id %in% genes_strain_PM_only, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_PM_only_down <- domain_specificity |> 
  filter(gene_id %in% genes_strain_PM_only, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_strain_PM_only_ora_up <- run_enrichment_hybrid(
  genes_PM_only_up, 
  description = "PM only - Up in CAST"
)

genes_strain_PM_only_ora_down <- run_enrichment_hybrid(
  genes_PM_only_down, 
  description = "PM only - Down in CAST"
)

```

```{r}
#| label: domain-specific-PM-only-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  genes_strain_PM_only_ora_up$result |> 
    mutate(direction = paste0("Up in CAST (n=", length(genes_PM_only_up), ")")),
  genes_strain_PM_only_ora_down$result |> 
    mutate(direction = paste0("Down in CAST (n=", length(genes_PM_only_down), ")"))
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(direction, everything()) |> 
    create_dt()
}

```

## MAX Only

Genes where the strain difference passes significance and effect size thresholds only in the MAX domain (at t=0).

```{r}
#| label: domain-specific-MAX-only-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

if (length(genes_strain_MAX_only) > 0) {
  count_data <- get_normalized_counts(strain_dds_full, genes_strain_MAX_only, vsd = vsd_global)
  plot_domain_specificity(genes_strain_MAX_only, "MAX only", count_data, n = 2)
}

```

```{r}
#| label: domain-specific-MAX-only-table
#| message: false
#| warning: false

if (length(genes_strain_MAX_only) > 0) {
  domain_specificity |> 
    filter(gene_id %in% genes_strain_MAX_only) |>
    select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
           specificity_pattern, has_time_int) |> 
    create_dt()
}

```

```{r}
#| label: domain-specific-MAX-only-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_strain_MAX_only_ora_up <- NULL
genes_strain_MAX_only_ora_down <- NULL

if (length(genes_strain_MAX_only) > 0) {
  genes_MAX_only_up <- domain_specificity |> 
    filter(gene_id %in% genes_strain_MAX_only, strain_effect_MAX > 0) |> 
    pull(gene_id)
  
  genes_MAX_only_down <- domain_specificity |> 
    filter(gene_id %in% genes_strain_MAX_only, strain_effect_MAX < 0) |> 
    pull(gene_id)
  
  if (length(genes_MAX_only_up) > 0) {
    genes_strain_MAX_only_ora_up <- run_enrichment_hybrid(
      genes_MAX_only_up, 
      description = "MAX only - Up in CAST"
    )
  }
  
  if (length(genes_MAX_only_down) > 0) {
    genes_strain_MAX_only_ora_down <- run_enrichment_hybrid(
      genes_MAX_only_down, 
      description = "MAX only - Down in CAST"
    )
  }
}

```

```{r}
#| label: domain-specific-MAX-only-ora-table
#| message: false
#| warning: false

if (!is.null(genes_strain_MAX_only_ora_up) || !is.null(genes_strain_MAX_only_ora_down)) {
  combined <- bind_rows(
    if (!is.null(genes_strain_MAX_only_ora_up)) {
      genes_strain_MAX_only_ora_up$result |> 
        mutate(direction = paste0("Up in CAST (n=", length(genes_MAX_only_up), ")"))
    },
    if (!is.null(genes_strain_MAX_only_ora_down)) {
      genes_strain_MAX_only_ora_down$result |> 
        mutate(direction = paste0("Down in CAST (n=", length(genes_MAX_only_down), ")"))
    }
  )
  
  if (!is.null(combined) && nrow(combined) > 0) {
    combined |> 
      filter(p_value < 0.01) |> 
      select(direction, everything()) |> 
      create_dt()
  }
}

```

## POST Only

Genes where the strain difference passes significance and effect size thresholds only in the POST domain (at t=0).

```{r}
#| label: domain-specific-POST-only-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

if (length(genes_strain_POST_only) > 0) {
  count_data <- get_normalized_counts(strain_dds_full, genes_strain_POST_only, vsd = vsd_global)
  plot_domain_specificity(genes_strain_POST_only, "POST only", count_data, n = 2)
}

```

```{r}
#| label: domain-specific-POST-only-table
#| message: false
#| warning: false

if (length(genes_strain_POST_only) > 0) {
  domain_specificity |> 
    filter(gene_id %in% genes_strain_POST_only) |>
    select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
           specificity_pattern, has_time_int) |> 
    create_dt()
}

```

```{r}
#| label: domain-specific-POST-only-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_strain_POST_only_ora_up <- NULL
genes_strain_POST_only_ora_down <- NULL

if (length(genes_strain_POST_only) > 0) {
  genes_POST_only_up <- domain_specificity |> 
    filter(gene_id %in% genes_strain_POST_only, strain_effect_POST > 0) |> 
    pull(gene_id)
  
  genes_POST_only_down <- domain_specificity |> 
    filter(gene_id %in% genes_strain_POST_only, strain_effect_POST < 0) |> 
    pull(gene_id)
  
  if (length(genes_POST_only_up) > 0) {
    genes_strain_POST_only_ora_up <- run_enrichment_hybrid(
      genes_POST_only_up, 
      description = "POST only - Up in CAST"
    )
  }
  
  if (length(genes_POST_only_down) > 0) {
    genes_strain_POST_only_ora_down <- run_enrichment_hybrid(
      genes_POST_only_down, 
      description = "POST only - Down in CAST"
    )
  }
}

```

```{r}
#| label: domain-specific-POST-only-ora-table
#| message: false
#| warning: false

if (!is.null(genes_strain_POST_only_ora_up) || !is.null(genes_strain_POST_only_ora_down)) {
  combined <- bind_rows(
    if (!is.null(genes_strain_POST_only_ora_up)) {
      genes_strain_POST_only_ora_up$result |> 
        mutate(direction = paste0("Up in CAST (n=", length(genes_POST_only_up), ")"))
    },
    if (!is.null(genes_strain_POST_only_ora_down)) {
      genes_strain_POST_only_ora_down$result |> 
        mutate(direction = paste0("Down in CAST (n=", length(genes_POST_only_down), ")"))
    }
  )
  
  if (!is.null(combined) && nrow(combined) > 0) {
    combined |> 
      filter(p_value < 0.01) |> 
      select(direction, everything()) |> 
      create_dt()
  }
}

```

## All Domains

Genes where the strain difference passes significance and effect size thresholds in all three domains (PM, MAX, POST) at t=0. These represent pervasive strain effects across the entire palate.

```{r}
#| label: domain-specific-all-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

count_data <- get_normalized_counts(strain_dds_full, genes_strain_all_domains, vsd = vsd_global)
plot_domain_specificity(genes_strain_all_domains, "All domains", count_data, n = 2)

```

```{r}
#| label: domain-specific-all-table
#| message: false
#| warning: false

domain_specificity |> 
  filter(gene_id %in% genes_strain_all_domains) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         specificity_pattern, has_time_int) |> 
  create_dt()

```

```{r}
#| label: domain-specific-all-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by strain effect direction in PM (reference domain)
genes_all_domains_up <- domain_specificity |> 
  filter(gene_id %in% genes_strain_all_domains, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_all_domains_down <- domain_specificity |> 
  filter(gene_id %in% genes_strain_all_domains, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_strain_all_domains_ora_up <- run_enrichment_hybrid(
  genes_all_domains_up, 
  description = "All domains - Up in CAST"
)

genes_strain_all_domains_ora_down <- run_enrichment_hybrid(
  genes_all_domains_down, 
  description = "All domains - Down in CAST"
)

```

```{r}
#| label: domain-specific-all-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  genes_strain_all_domains_ora_up$result |> 
    mutate(direction = paste0("Up in CAST (n=", length(genes_all_domains_up), ")")),
  genes_strain_all_domains_ora_down$result |> 
    mutate(direction = paste0("Down in CAST (n=", length(genes_all_domains_down), ")"))
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |> 
    filter(p_value < 0.01) |> 
    select(direction, everything()) |> 
    create_dt()
}

```

## Static Domain Specificity

Domain-specific strain effects that remain stable over time (no time or three-way interactions). These genes show consistent domain-restricted patterns throughout development.

```{r}
#| label: domain-specific-static-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

count_data <- get_normalized_counts(strain_dds_full, genes_strain_domain_specific_static, vsd = vsd_global)
plot_domain_specificity(genes_strain_domain_specific_static, "Static domain specificity", count_data, n = 2)

```

```{r}
#| label: domain-specific-static-table
#| message: false
#| warning: false

domain_specificity |> 
  filter(gene_id %in% genes_strain_domain_specific_static) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         specificity_pattern, has_time_int) |> 
  create_dt()

```

## Dynamic Domain Specificity

Domain-specific strain effects that change over developmental time (time or three-way interactions present). The domain specificity pattern evolves during development.

```{r}
#| label: domain-specific-dynamic-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

count_data <- get_normalized_counts(strain_dds_full, genes_strain_domain_specific_dynamic, vsd = vsd_global)
plot_domain_specificity(genes_strain_domain_specific_dynamic, "Dynamic domain specificity", count_data, n = 2)

```

```{r}
#| label: domain-specific-dynamic-table
#| message: false
#| warning: false

domain_specificity |> 
  filter(gene_id %in% genes_strain_domain_specific_dynamic) |>
  select(gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
         specificity_pattern, has_time_int) |> 
  create_dt()

```

::::
# Comprehensive Tables

This section provides downloadable tables summarizing all analysis results.

:::: {#comprehensive-tables-tabs .panel-tabset .nav-pills}

## Gene-Level Pattern Assignments

All genes with their assigned expression patterns across strain, domain, and time.

```{r}
#| label: comprehensive-gene-assignments
#| message: false
#| warning: false

gene_assignments |>
  left_join(all_genes_palate |> select(gene_id, symbol, gene_biotype, chromosome)) |>
  select(gene_id, symbol, gene_biotype, chromosome, everything()) |>
  create_dt()

```

## Full Strain Effects Classification

Complete results table with all coefficients and classifications.

```{r}
#| label: comprehensive-strain-effects
#| message: false
#| warning: false

strain_effects_classified |>
  filter(any_strain_sig) |>
  select(
    gene_id, symbol, gene_biotype, baseMean,
    # Pattern classifications
    pattern_type, main_strain_pattern, 
    # Main strain effect
    main_strain_lfc, main_strain_pval,
    # Domain-specific effects
    strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    max_domain_effect, domain_consistency,
    # Time effects
    time_int_slope, time_pattern,
    # Significance flags
    domain_int_sig, time_int_sig, threeway_int_sig
  ) |>
  create_dt()
```

## Combined ORA Results

All over-representation analysis results combined into a single table.

```{r}
#| label: comprehensive-ora-combine
#| message: false
#| warning: false

# Combine all ORA results into one table
all_ora_results <- bind_rows(
  # Main strain effects
  if (!is.null(genes_CAST_high_strong_ora$result)) 
    genes_CAST_high_strong_ora$result |> 
      mutate(category = "Main Effect", pattern = "CAST >> B6 (strong)", direction = "Up in CAST"),
  if (!is.null(genes_CAST_high_mod_ora$result)) 
    genes_CAST_high_mod_ora$result |> 
      mutate(category = "Main Effect", pattern = "CAST > B6 (moderate)", direction = "Up in CAST"),
  if (!is.null(genes_B6_high_strong_ora$result)) 
    genes_B6_high_strong_ora$result |> 
      mutate(category = "Main Effect", pattern = "B6 >> CAST (strong)", direction = "Down in CAST"),
  if (!is.null(genes_B6_high_mod_ora$result)) 
    genes_B6_high_mod_ora$result |> 
      mutate(category = "Main Effect", pattern = "B6 > CAST (moderate)", direction = "Down in CAST"),
  
  # Domain effects - PM biggest
  if (!is.null(genes_PM_biggest_effect_ora_up$result)) 
    genes_PM_biggest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "PM biggest effect", direction = "Up in CAST"),
  if (!is.null(genes_PM_biggest_effect_ora_down$result)) 
    genes_PM_biggest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "PM biggest effect", direction = "Down in CAST"),
  
  # Domain effects - MAX biggest
  if (!is.null(genes_MAX_biggest_effect_ora_up$result)) 
    genes_MAX_biggest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "MAX biggest effect", direction = "Up in CAST"),
  if (!is.null(genes_MAX_biggest_effect_ora_down$result)) 
    genes_MAX_biggest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "MAX biggest effect", direction = "Down in CAST"),
  
  # Domain effects - POST biggest
  if (!is.null(genes_POST_biggest_effect_ora_up$result)) 
    genes_POST_biggest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "POST biggest effect", direction = "Up in CAST"),
  if (!is.null(genes_POST_biggest_effect_ora_down$result)) 
    genes_POST_biggest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "POST biggest effect", direction = "Down in CAST"),
  
  # Domain effects - Opposite directions
  if (!is.null(genes_opposite_directions_ora_up$result)) 
    genes_opposite_directions_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "Opposite directions", direction = "Up in CAST (PM)"),
  if (!is.null(genes_opposite_directions_ora_down$result)) 
    genes_opposite_directions_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "Opposite directions", direction = "Down in CAST (PM)"),
  
  # Time effects - CAST faster
  if (!is.null(genes_diverging_CAST_faster_ora_up$result)) 
    genes_diverging_CAST_faster_ora_up$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: CAST faster", direction = "Up in CAST"),
  if (!is.null(genes_diverging_CAST_faster_ora_down$result)) 
    genes_diverging_CAST_faster_ora_down$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: CAST faster", direction = "Down in CAST"),
  
  # Time effects - B6 faster
  if (!is.null(genes_diverging_B6_faster_ora_up$result)) 
    genes_diverging_B6_faster_ora_up$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: B6 faster", direction = "Up in CAST"),
  if (!is.null(genes_diverging_B6_faster_ora_down$result)) 
    genes_diverging_B6_faster_ora_down$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: B6 faster", direction = "Down in CAST"),
  
  # Three-way effects - Complex pattern
  if (!is.null(genes_threeway_complex_ora_PM$result)) 
    genes_threeway_complex_ora_PM$result |> 
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "PM most distinct"),
  if (!is.null(genes_threeway_complex_ora_MAX$result)) 
    genes_threeway_complex_ora_MAX$result |> 
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "MAX most distinct"),
  if (!is.null(genes_threeway_complex_ora_POST$result)) 
    genes_threeway_complex_ora_POST$result |> 
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "POST most distinct"),
  
  # Three-way effects - Primarily initial
  if (!is.null(genes_threeway_initial_ora_PM$result)) 
    genes_threeway_initial_ora_PM$result |> 
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "PM most distinct"),
  if (!is.null(genes_threeway_initial_ora_MAX$result)) 
    genes_threeway_initial_ora_MAX$result |> 
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "MAX most distinct"),
  if (!is.null(genes_threeway_initial_ora_POST$result)) 
    genes_threeway_initial_ora_POST$result |> 
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "POST most distinct"),
  
  # Three-way effects - Primarily dynamic
  if (!is.null(genes_threeway_dynamic_ora_PM$result)) 
    genes_threeway_dynamic_ora_PM$result |> 
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "PM most distinct"),
  if (!is.null(genes_threeway_dynamic_ora_MAX$result)) 
    genes_threeway_dynamic_ora_MAX$result |> 
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "MAX most distinct"),
  if (!is.null(genes_threeway_dynamic_ora_POST$result)) 
    genes_threeway_dynamic_ora_POST$result |> 
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "POST most distinct"),
  
  # Domain specificity - PM only
  if (!is.null(genes_strain_PM_only_ora_up$result)) 
    genes_strain_PM_only_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "PM only", direction = "Up in CAST"),
  if (!is.null(genes_strain_PM_only_ora_down$result)) 
    genes_strain_PM_only_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "PM only", direction = "Down in CAST"),
  
  # Domain specificity - MAX only
  if (!is.null(genes_strain_MAX_only_ora_up$result)) 
    genes_strain_MAX_only_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "MAX only", direction = "Up in CAST"),
  if (!is.null(genes_strain_MAX_only_ora_down$result)) 
    genes_strain_MAX_only_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "MAX only", direction = "Down in CAST"),
  
  # Domain specificity - POST only
  if (!is.null(genes_strain_POST_only_ora_up$result)) 
    genes_strain_POST_only_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "POST only", direction = "Up in CAST"),
  if (!is.null(genes_strain_POST_only_ora_down$result)) 
    genes_strain_POST_only_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "POST only", direction = "Down in CAST"),
  
  # Domain specificity - All domains
  if (!is.null(genes_strain_all_domains_ora_up$result)) 
    genes_strain_all_domains_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "All domains", direction = "Up in CAST"),
  if (!is.null(genes_strain_all_domains_ora_down$result)) 
    genes_strain_all_domains_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "All domains", direction = "Down in CAST")
)

```

```{r}
#| label: comprehensive-ora-table
#| message: false
#| warning: false

all_ora_results |>
  filter(p_value < 0.01) |>
  select(category, pattern, direction, source, term_name, p_value, 
         intersection_size, term_size, intersection) |>
  arrange(category, pattern, direction, p_value) |>
  create_dt()

```



::::

# Session Info

```{r}
#| label: session-info
#| message: false
#| warning: false

sessionInfo()
```