---
title: "Investigating strain, domain and developmental stage effects"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
      embed-resources: true
      standalone: true
code-fold: true
toc: true
toc-depth: 5
toc-expand: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| message: false
#| warning: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(gprofiler2)
set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e108_eg55_p17/")
library(enrichR)
library(DT)
library(here)

set.seed(123)
select <- dplyr::select 
rename <- dplyr::rename 
filter <- dplyr::filter 
summarize <- dplyr::summarize 
count <- dplyr::count

domain_colors <- c(PM = rgb(252,206,93, maxColorValue = 255), 
                   MAX = rgb(163,82,157, maxColorValue = 255), 
                   POST = rgb(158,156,201, maxColorValue = 255))

strain_colors <- c(B6 = "#555555", CAST = "#009E73")
```

```{r}
#| label: helper-functions
#| message: false
#| warning: false

create_dt <- function(x){
  format_num <- function(val) {
    case_when(
      is.na(val) ~ "NA",
      abs(val) < 0.01 ~ formatC(val, digits = 2),
      TRUE ~ formatC(round(val, 2))
    )
  }
  x |> 
    mutate(across(where(is.numeric), format_num)) |>
    DT::datatable(
      extensions = 'Buttons',
      rownames = FALSE, 
      filter = "top",
      options = list(dom = 'Blfrtip',
                     buttons = c('copy', 'csv', 'excel'),
                     pageLength = 10, 
                     scrollX = TRUE))
}

# Helper function for consistent filtering of DESeq2 results
filter_deseq_results <- function(res, 
                                  min_basemean = NULL,
                                  min_basemean_quantile = 0.1,
                                  max_lfcse = 1.5, 
                                  max_abs_lfc = 10, 
                                  min_abs_lfc = 0.5, 
                                  max_padj = 0.05) {
  
  # Convert to data frame
  res_df <- as.data.frame(res) |>
    rownames_to_column("gene_id")
  
  # Set baseMean threshold - use quantile if not specified
 if(is.null(min_basemean)) {
    min_basemean <- quantile(res_df$baseMean, min_basemean_quantile, na.rm = TRUE)
  }
  
  # Apply filters
  res_filtered <- res_df |>
    filter(
      baseMean > min_basemean,
      lfcSE < max_lfcse,
      abs(log2FoldChange) < max_abs_lfc,
      abs(log2FoldChange) > min_abs_lfc,
      padj < max_padj
    )
  
  return(res_filtered)
}

# Helper to get passing gene IDs from a filtered result
get_passing_genes <- function(res, ...) {
  filter_deseq_results(res, ...) |> pull(gene_id)
}
get_normalized_counts <- function(dds, genes) {
  vsd <- vst(dds, blind = FALSE)
  assay(vsd)[genes, , drop = FALSE] |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    pivot_longer(-gene_id, names_to = "sample", values_to = "count") |>
    left_join(colData(dds) |> as.data.frame() |> rownames_to_column("sample"), by = "sample") |>
    left_join(all_genes_palate |> select(gene_id, symbol), by = "gene_id")
}

# Function to run enrichment with both gprofiler2 AND Enrichr
run_enrichment_hybrid <- function(gene_ids, 
                                 gene_reference = all_genes_palate,
                                 description = "Gene Set",
                                 max_term_size = 660,
                                 min_term_size = 4) {
  
  # Convert gene IDs to symbols
  query_symbols <- gene_reference |>
    filter(gene_id %in% unique(gene_ids)) |>
    pull(symbol) |>
    unique()
  
  # Get background symbols
  background_symbols <- gene_reference$symbol
  
  # Skip if too few genes
  if(length(query_symbols) < 5) {
    message(paste0("Skipping ", description, ": fewer than 5 genes"))
    return(NULL)
  }
  
  # Query 1: gprofiler2 for GO, KEGG, Reactome, WikiPathways
  result_gprofiler <- gost(
    query = query_symbols,
    organism = "mmusculus",
    domain_scope = "custom",
    custom_bg = background_symbols,
    #sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP"),
    evcodes = TRUE
  )
  
  # Query 2: Enrichr for Mammalian Phenotype (full database)
  result_enrichr <- enrichr(
    genes = query_symbols,
    databases = c("MGI_Mammalian_Phenotype_Level_4_2021"),
                  #"MGI_Mammalian_Phenotype_Level_4_2019"),
    background =  background_symbols,
    include_overlap = TRUE
  )
  
  # Format gprofiler results
  gprofiler_df <- NULL
  if(!is.null(result_gprofiler$result)) {
    gprofiler_df <- result_gprofiler$result |>
      filter(term_size < max_term_size) |>
      filter(term_size > min_term_size) |>
      mutate(gene_set = description) |>
      select(gene_set, source, term_id, term_name, p_value, 
             query_size, term_size, intersection_size, intersection)
  }
  
  # Format Enrichr results to match gprofiler format
  enrichr_df <- NULL
  if(!is.null(result_enrichr) && length(result_enrichr) > 0) {
    # Process each database separately
    enrichr_dfs <- lapply(names(result_enrichr), function(db_name) {
      db_result <- result_enrichr[[db_name]]
      
      # Check if result has data (not empty or all NA)
      if(is.null(db_result) || nrow(db_result) == 0 || all(is.na(db_result$Term))) {
        return(NULL)
      }
      
      db_result |>
        filter(Adjusted.P.value < 0.05) |>
        mutate(
          gene_set = description,
          source = "Mammalian Phenotype",
          term_id = Term,
          term_name = Term,
          p_value = Adjusted.P.value,
          query_size = length(query_symbols),
          intersection_size = as.numeric(gsub("/.*", "", Overlap)),  # numerator (X from "X/Y")
          term_size = as.numeric(gsub(".*/", "", Overlap)),          # denominator (Y from "X/Y")
          intersection = Genes
        ) |>
        filter(term_size < max_term_size) |>
        filter(term_size > min_term_size) |>
        select(gene_set, source, term_id, term_name, p_value, 
               query_size, term_size, intersection_size, intersection)
    })
    
    # Remove NULL entries and combine
    enrichr_dfs <- enrichr_dfs[!sapply(enrichr_dfs, is.null)]
    if(length(enrichr_dfs) > 0) {
      enrichr_df <- bind_rows(enrichr_dfs)
    }
  }
  
  # Combine results
  combined_result <- bind_rows(gprofiler_df, enrichr_df)
  
  if(nrow(combined_result) > 0) {
    return(list(result = combined_result))
  } else {
    return(NULL)
  }
}

format_enrichment_result <- function(gost_result, max_terms = 50) {
  if(is.null(gost_result$result)) return(NULL)
  gost_result$result |>
    select(Source = source, Term_ID = term_id, Term_Name = term_name,
           P_Value = p_value, Term_Size = term_size, Query_Size = query_size,
           Intersection_Size = intersection_size, Genes = intersection) |>
    arrange(P_Value) |> head(max_terms) |>
    mutate(P_Value = signif(P_Value, 3),
           Gene_Ratio = paste0(Intersection_Size, "/", Query_Size))
}

# plotting-functions
plot_main_strain_effect <- function(gene_ids, direction, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = direction, x = "LB stage (centered)", y = "Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = strain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) + geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) + scale_fill_manual(values = strain_colors) +
      labs(x = "", y = "") + theme_pubclean(base_size = 16) + theme(legend.position = "none")
    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

plot_domain_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    strain_diffs <- gene_data |> group_by(AP_domain, strain) |> 
      summarize(mean_count = mean(count, na.rm = TRUE), .groups = "drop") |> 
      pivot_wider(names_from = strain, values_from = mean_count) |> 
      mutate(pct_diff = (CAST - B6) / B6 * 100)
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
    p2 <- ggplot(strain_diffs, aes(x = AP_domain, y = pct_diff, fill = AP_domain)) +
      geom_col(alpha = 0.8) + geom_hline(yintercept = 0, linetype = "dashed") +
      scale_fill_manual(values = domain_colors) + labs(y = "% Diff\n(CAST-B6)", x = "") +
      theme_pubclean(base_size = 16) + theme(legend.position = "none")
    p1 + p2 + plot_layout(widths = c(2.5, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1)
}

plot_time_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(aes(shape = AP_domain), alpha = 0.5, size = 2.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1.5) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "right", plot.title = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) + geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "none", strip.text = element_text(face = "bold"))
    p1 + p2 + plot_layout(nrow = 1, widths = c(1, 2.5))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect")
}

plot_threeway_interaction <- function(gene_ids, pattern_desc, count_data, n = 4) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.5, size = 2) + geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1, guides = "collect") & theme( legend.position = "bottom")
}

plot_domain_specificity <- function(gene_ids, domains_affected, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if(nrow(gene_data) == 0) return(NULL)
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain), position = position_jitterdodge(jitter.width = 0.1), alpha = 0.4, size = 1.5) +
      scale_y_log10(labels = scales::comma) + scale_fill_manual(values = strain_colors) + scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = paste("Strain effect in", domains_affected), x = "Domain", y = "Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) + geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      facet_wrap(~ AP_domain, nrow = 1) +
      scale_y_log10(labels = scales::comma) + scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Counts") +
      theme_pubclean(base_size = 16) + theme(legend.position = "none", strip.text = element_text(face = "bold"))
    p1 / p2 + plot_layout(heights = c(1, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  wrap_plots(plots, ncol = 1,  guides = "collect") & theme( legend.position = "bottom")
}
```

```{r}
#| label: load-data
#| message: false
#| warning: false

sample_details <- read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 
load("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/palate_gbrs_emase_counts_v4.RData")

contam_samples <- c("B6_12.5_PM_1","B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat_nondo[, !colnames(expr_mat_nondo) %in% contam_samples]
rm(expr_mat_do, expr_mat_nondo)

all_genes_palate <- read_tsv(here("_data","ENSMUSGid_to_symbol_v105.txt"),
                              col_names = c("gene_id","symbol","gene_biotype","chromosome","start","end"), 
                              skip = 1) |>  
  mutate(midpoint = (start + end) / 2) |>
  filter(gene_id %in% rownames(raw_expr_mat))

sample_annot <- sample_details |>
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |>
  separate(Strain_stock, into = c("strain","tmp"), remove = FALSE, sep = "_") |>
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, 
         palate_closure, L_R_palate) |>
  mutate(strain = if_else(strain == "B6J", "B6", strain)) |>
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |>
  select(-temp) |>
  mutate(Palate_ori = if_else(Palate_ori == "L", "Left", "Right"))
```

# Differential gene expression analysis overview

```{r}
#| label: prep-run-deseq2
#| message: false
#| warning: false

b6_cast_samples <- sample_annot |>
  filter(strain %in% c("B6", "CAST")) |>
  separate(sample, sep = "_", into = c("s","e","d","ind","rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = if_else(is.na(rep), "1", rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

coldata <- b6_cast_samples |>
  select(sample, unique_sampleid, sex, ind, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(coldata), drop = FALSE])

strain_dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~1)
strain_dds_coll <- collapseReplicates(strain_dds, strain_dds$unique_sampleid)

min_LB_stage <- min(colData(strain_dds_coll)$LB_stage)
colData(strain_dds_coll)$LB_stage_from_start <- colData(strain_dds_coll)$LB_stage - min_LB_stage
colData(strain_dds_coll)$sex <- factor(colData(strain_dds_coll)$sex)
colData(strain_dds_coll)$strain <- factor(colData(strain_dds_coll)$strain, levels = c("B6", "CAST"))
colData(strain_dds_coll)$ind <- factor(colData(strain_dds_coll)$ind)
# fit-full-model
# Set full model design
design(strain_dds_coll) <- ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start

# Fit full model - this estimates all coefficients
strain_dds_full <- DESeq(strain_dds_coll, parallel = TRUE)

# Check available coefficient names
coef_names <- resultsNames(strain_dds_full)

# lrt-tests
# LRT: Any strain effect at all
strain_dds_lrt <- DESeq(strain_dds_coll, test = "LRT", 
                        reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind,
                        parallel = TRUE)
res_any_strain <- results(strain_dds_lrt)

# LRT: Domain × Strain interaction
dds_domain_lrt <- DESeq(strain_dds_coll, test = "LRT",
                        full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                               strain + strain:AP_domain + strain:LB_stage_from_start + 
                               strain:AP_domain:LB_stage_from_start,
                        reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                   strain + strain:LB_stage_from_start,
                        parallel = TRUE)
res_domain_lrt <- results(dds_domain_lrt)

# LRT: Time × Strain interaction
dds_time_lrt <- DESeq(strain_dds_coll, test = "LRT",
                      full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start,
                      reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                 strain + strain:AP_domain,
                      parallel = TRUE)
res_time_lrt <- results(dds_time_lrt)

# LRT: Three-way interaction
dds_3way_lrt <- DESeq(strain_dds_coll, test = "LRT",
                      full = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                             strain + strain:AP_domain + strain:LB_stage_from_start + 
                             strain:AP_domain:LB_stage_from_start,
                      reduced = ~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + 
                                 strain + strain:AP_domain + strain:LB_stage_from_start,
                      parallel = TRUE)
res_3way_lrt <- results(dds_3way_lrt)

# extract-effects
# Main strain effect (CAST vs B6 in reference domain PM at time=0)
res_strain_main <- results(strain_dds_full, name = "strain_CAST_vs_B6")

# Domain interaction coefficients (how CAST effect differs from PM)
res_domain_MAX <- results(strain_dds_full, name = "AP_domainMAX.strainCAST")
res_domain_POST <- results(strain_dds_full, name = "AP_domainPOST.strainCAST")

# Time interaction coefficient (slope of strain effect over time)
res_time_int <- results(strain_dds_full, name = "LB_stage_from_start.strainCAST")

# Three-way interaction coefficients (how time slope differs by domain)
res_3way_MAX <- results(strain_dds_full, name = "AP_domainMAX.LB_stage_from_start.strainCAST")
res_3way_POST <- results(strain_dds_full, name = "AP_domainPOST.LB_stage_from_start.strainCAST")

# Calculate strain effect within each domain
# PM: just the main effect
# MAX: main effect + domain interaction
# POST: main effect + domain interaction

# strain-effects-by-domain
# Strain effect in PM (reference) = main strain coefficient
res_strain_PM <- res_strain_main

# Strain effect in MAX = main + strainCAST:AP_domainMAX
res_strain_MAX <- results(strain_dds_full,
                          contrast = list(c("strain_CAST_vs_B6", "AP_domainMAX.strainCAST")))

# Strain effect in POST = main + strainCAST:AP_domainPOST  
res_strain_POST <- results(strain_dds_full,
                           contrast = list(c("strain_CAST_vs_B6", "AP_domainPOST.strainCAST")))

# Apply filtering to each result object individually
# Each filter is based on that specific contrast's quality metrics

# Main strain effect
res_strain_main_filtered <- filter_deseq_results(res_strain_main)
genes_main_strain_pass <- res_strain_main_filtered$gene_id

# Strain effect in each domain
res_strain_PM_filtered <- filter_deseq_results(res_strain_PM)
res_strain_MAX_filtered <- filter_deseq_results(res_strain_MAX)
res_strain_POST_filtered <- filter_deseq_results(res_strain_POST)

genes_strain_PM_pass <- res_strain_PM_filtered$gene_id
genes_strain_MAX_pass <- res_strain_MAX_filtered$gene_id
genes_strain_POST_pass <- res_strain_POST_filtered$gene_id

# Domain interaction coefficients (different thresholds - these are differences)
res_domain_MAX_filtered <- filter_deseq_results(res_domain_MAX, min_abs_lfc = 0.3)
res_domain_POST_filtered <- filter_deseq_results(res_domain_POST, min_abs_lfc = 0.3)

genes_domain_int_MAX_pass <- res_domain_MAX_filtered$gene_id
genes_domain_int_POST_pass <- res_domain_POST_filtered$gene_id

# Time interaction coefficient (slope - different scale)
res_time_int_filtered <- filter_deseq_results(res_time_int, min_abs_lfc = 0.1, max_lfcse = 0.5)
genes_time_int_pass <- res_time_int_filtered$gene_id

# Three-way coefficients
res_3way_MAX_filtered <- filter_deseq_results(res_3way_MAX, min_abs_lfc = 0.1, max_lfcse = 0.5)
res_3way_POST_filtered <- filter_deseq_results(res_3way_POST, min_abs_lfc = 0.1, max_lfcse = 0.5)

genes_3way_MAX_pass <- res_3way_MAX_filtered$gene_id
genes_3way_POST_pass <- res_3way_POST_filtered$gene_id

# Summary of filtering
filter_summary <- tibble(
  Result = c("Main strain effect", "Strain in PM", "Strain in MAX", "Strain in POST",
             "Domain int (MAX)", "Domain int (POST)", "Time int", 
             "3-way (MAX)", "3-way (POST)"),
  Total_tested = c(nrow(res_strain_main), nrow(res_strain_PM), nrow(res_strain_MAX),
                   nrow(res_strain_POST), nrow(res_domain_MAX), nrow(res_domain_POST),
                   nrow(res_time_int), nrow(res_3way_MAX), nrow(res_3way_POST)),
  Passing_filter = c(length(genes_main_strain_pass), length(genes_strain_PM_pass),
                     length(genes_strain_MAX_pass), length(genes_strain_POST_pass),
                     length(genes_domain_int_MAX_pass), length(genes_domain_int_POST_pass),
                     length(genes_time_int_pass), length(genes_3way_MAX_pass),
                     length(genes_3way_POST_pass))
)

# combine-results
strain_effects_classified <- tibble::tibble(
  gene_id = rownames(res_any_strain),
  
  # Base expression level
  baseMean = res_any_strain$baseMean,
  
  # Significance from LRT tests
  any_strain_sig = res_any_strain$padj < 0.05,
  domain_int_sig = res_domain_lrt$padj < 0.05,
  time_int_sig = res_time_lrt$padj < 0.05,
  threeway_int_sig = res_3way_lrt$padj < 0.05,
  
  # P-values from LRT tests
  any_strain_pval = res_any_strain$padj,
  domain_int_pval = res_domain_lrt$padj,
  time_int_pval = res_time_lrt$padj,
  threeway_int_pval = res_3way_lrt$padj,
  
  # Main strain effect (Wald)
  main_strain_lfc = res_strain_main$log2FoldChange,
  main_strain_lfcSE = res_strain_main$lfcSE,
  main_strain_pval = res_strain_main$padj,
  main_strain_pass = gene_id %in% genes_main_strain_pass,
  
  # Strain effect within each domain
  strain_effect_PM = res_strain_PM$log2FoldChange,
  strain_effect_PM_lfcSE = res_strain_PM$lfcSE,
  strain_effect_PM_pval = res_strain_PM$padj,
  strain_effect_PM_pass = gene_id %in% genes_strain_PM_pass,
  
  strain_effect_MAX = res_strain_MAX$log2FoldChange,
  strain_effect_MAX_lfcSE = res_strain_MAX$lfcSE,
  strain_effect_MAX_pval = res_strain_MAX$padj,
  strain_effect_MAX_pass = gene_id %in% genes_strain_MAX_pass,
  
  strain_effect_POST = res_strain_POST$log2FoldChange,
  strain_effect_POST_lfcSE = res_strain_POST$lfcSE,
  strain_effect_POST_pval = res_strain_POST$padj,
  strain_effect_POST_pass = gene_id %in% genes_strain_POST_pass,
  
  # Domain interaction coefficients
  domain_int_MAX_lfc = res_domain_MAX$log2FoldChange,
  domain_int_MAX_lfcSE = res_domain_MAX$lfcSE,
  domain_int_MAX_pass = gene_id %in% genes_domain_int_MAX_pass,
  
  domain_int_POST_lfc = res_domain_POST$log2FoldChange,
  domain_int_POST_lfcSE = res_domain_POST$lfcSE,
  domain_int_POST_pass = gene_id %in% genes_domain_int_POST_pass,
  
  # Time interaction coefficient
  time_int_slope = res_time_int$log2FoldChange,
  time_int_slope_lfcSE = res_time_int$lfcSE,
  time_int_slope_pval = res_time_int$padj,
  time_int_pass = gene_id %in% genes_time_int_pass,
  
  # Three-way coefficients
  threeway_MAX_slope = res_3way_MAX$log2FoldChange,
  threeway_MAX_slope_lfcSE = res_3way_MAX$lfcSE,
  threeway_MAX_pass = gene_id %in% genes_3way_MAX_pass,
  
  threeway_POST_slope = res_3way_POST$log2FoldChange,
  threeway_POST_slope_lfcSE = res_3way_POST$lfcSE,
  threeway_POST_pass = gene_id %in% genes_3way_POST_pass
) |>
  tidyr::replace_na(list(
    any_strain_sig = FALSE, domain_int_sig = FALSE,
    time_int_sig = FALSE, threeway_int_sig = FALSE,
    main_strain_pass = FALSE, strain_effect_PM_pass = FALSE,
    strain_effect_MAX_pass = FALSE, strain_effect_POST_pass = FALSE,
    domain_int_MAX_pass = FALSE, domain_int_POST_pass = FALSE,
    time_int_pass = FALSE, threeway_MAX_pass = FALSE, threeway_POST_pass = FALSE
  )) |>
  dplyr::left_join(all_genes_palate |> dplyr::select(gene_id, symbol, gene_biotype), by = "gene_id") |>
  dplyr::mutate(
    # Overall filter: passes main strain filter AND has significant LRT
    passes_strain_filter = any_strain_sig & main_strain_pass,
    
    # Main strain pattern classification (only for genes passing filter)
    main_strain_pattern = dplyr::case_when(
      !passes_strain_filter ~ "No effect / filtered",
      main_strain_lfc > 1 ~ "CAST >> B6 (strong)",
      main_strain_lfc > 0.5 ~ "CAST > B6 (moderate)",
      main_strain_lfc > 0 ~ "CAST > B6 (weak)",
      main_strain_lfc < -1 ~ "B6 >> CAST (strong)",
      main_strain_lfc < -0.5 ~ "B6 > CAST (moderate)",
      TRUE ~ "B6 > CAST (weak)"
    ),
    
    # Domain with largest strain effect (among passing filters)
    max_domain_effect = dplyr::case_when(
      !domain_int_sig ~ NA_character_,
      is.na(strain_effect_PM) | is.na(strain_effect_MAX) | is.na(strain_effect_POST) ~ NA_character_,
      abs(strain_effect_PM) >= abs(strain_effect_MAX) & 
        abs(strain_effect_PM) >= abs(strain_effect_POST) ~ "PM",
      abs(strain_effect_MAX) >= abs(strain_effect_POST) ~ "MAX",
      TRUE ~ "POST"
    ),
    
    # Direction in max domain
    max_domain_direction = dplyr::case_when(
      is.na(max_domain_effect) ~ NA_character_,
      max_domain_effect == "PM" & strain_effect_PM > 0 ~ "PM: CAST > B6",
      max_domain_effect == "PM" ~ "PM: B6 > CAST",
      max_domain_effect == "MAX" & strain_effect_MAX > 0 ~ "MAX: CAST > B6",
      max_domain_effect == "MAX" ~ "MAX: B6 > CAST",
      max_domain_effect == "POST" & strain_effect_POST > 0 ~ "POST: CAST > B6",
      TRUE ~ "POST: B6 > CAST"
    ),
    
    # Domain consistency
    domain_consistency = dplyr::case_when(
      is.na(strain_effect_PM) | is.na(strain_effect_MAX) | is.na(strain_effect_POST) ~ NA_character_,
      sign(strain_effect_PM) == sign(strain_effect_MAX) & 
        sign(strain_effect_MAX) == sign(strain_effect_POST) ~ "Consistent",
      TRUE ~ "Variable"
    ),
    
    # Time pattern (only if passes time filter)
    time_pattern = dplyr::case_when(
      !time_int_sig ~ "No time interaction",
      !time_int_pass ~ "Filtered",
      is.na(time_int_slope) ~ NA_character_,
      time_int_slope > 0.1 ~ "Diverging (CAST faster)",
      time_int_slope < -0.1 ~ "Diverging (B6 faster)",
      TRUE ~ "Parallel"
    ),
    
    # Overall pattern type
    pattern_type = dplyr::case_when(
      !any_strain_sig ~ "No strain effect",
      !main_strain_pass ~ "Filtered (QC)",
      threeway_int_sig ~ "Three-way interaction",
      domain_int_sig & time_int_sig ~ "Domain + Time interactions",
      domain_int_sig ~ "Domain × Strain interaction",
      time_int_sig ~ "Time × Strain interaction",
      main_strain_pval < 0.05 ~ "Main strain effect only",
      TRUE ~ "Other"
    )
  )


# Create the filtered dataset for downstream analysis
strain_effects_filtered <- strain_effects_classified |>
  filter(passes_strain_filter)

# n_total_sig <- sum(strain_effects_classified$any_strain_sig, na.rm = TRUE)
# n_passing <- nrow(strain_effects_filtered)

# run-pattern-classification
# Source the pattern classification script
# source(here("DEgene_pattern_identification_CAST_B6.R"))
# Load pattern tables
pattern_summary <- read_csv(here("strain_pattern_summary.csv"))
main_patterns <- read_csv(here("main_strain_effect_patterns.csv"))
domain_patterns <- read_csv(here("domain_strain_patterns.csv"))
time_patterns <- read_csv(here("time_strain_patterns.csv"))
threeway_patterns <- read_csv(here("threeway_patterns.csv"))
domain_specificity <- read_csv(here("domain_specificity_patterns.csv"))
gene_assignments <- read_csv(here("strain_gene_pattern_assignments.csv"))

# Extract gene lists from main_patterns
genes_CAST_high_strong <- main_patterns |>
  filter(main_strain_pattern == "CAST >> B6 (strong)") |> pull(gene_id)
genes_CAST_high_mod <- main_patterns |>
  filter(main_strain_pattern == "CAST > B6 (moderate)") |> pull(gene_id)
genes_B6_high_strong <- main_patterns |>
  filter(main_strain_pattern == "B6 >> CAST (strong)") |> pull(gene_id)
genes_B6_high_mod <- main_patterns |>
  filter(main_strain_pattern == "B6 > CAST (moderate)") |> pull(gene_id)

# Extract gene lists from domain_patterns
genes_PM_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "PM") |> pull(gene_id)
genes_MAX_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "MAX") |> pull(gene_id)
genes_POST_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "POST") |> pull(gene_id)
genes_opposite_directions <- domain_patterns |>
  filter(domain_consistency == "Variable") |> pull(gene_id)

# Extract gene lists from time_patterns (already filtered for time_int_pass)
genes_diverging_CAST_faster <- time_patterns |>
  filter(time_int_slope > 0.1) |> pull(gene_id)
genes_diverging_B6_faster <- time_patterns |>
  filter(time_int_slope < -0.1) |> pull(gene_id)

# Extract gene lists from threeway_patterns
genes_PM_biggest_slope_diff <- threeway_patterns |>
  filter(max_slope_diff_domain == "PM") |> pull(gene_id)
genes_MAX_biggest_slope_diff <- threeway_patterns |>
  filter(max_slope_diff_domain == "MAX") |> pull(gene_id)
genes_POST_biggest_slope_diff <- threeway_patterns |>
  filter(max_slope_diff_domain == "POST") |> pull(gene_id)
genes_consistent_slopes <- threeway_patterns |>
  filter(slope_consistency == "Consistent") |> pull(gene_id)

# Extract gene lists from domain_specificity (uses per-domain pass flags)
genes_strain_PM_only <- domain_specificity |>
  filter(specificity_pattern == "PM only") |> pull(gene_id)
genes_strain_MAX_only <- domain_specificity |>
  filter(specificity_pattern == "MAX only") |> pull(gene_id)
genes_strain_POST_only <- domain_specificity |>
  filter(specificity_pattern == "POST only") |> pull(gene_id)
genes_strain_all_domains <- domain_specificity |>
  filter(specificity_pattern == "All domains") |> pull(gene_id)

# Extract gene lists from main_patterns
genes_CAST_high_strong <- main_patterns |>
  filter(main_strain_pattern == "CAST >> B6 (strong)") |> pull(gene_id)
genes_CAST_high_mod <- main_patterns |>
  filter(main_strain_pattern == "CAST > B6 (moderate)") |> pull(gene_id)
genes_B6_high_strong <- main_patterns |>
  filter(main_strain_pattern == "B6 >> CAST (strong)") |> pull(gene_id)
genes_B6_high_mod <- main_patterns |>
  filter(main_strain_pattern == "B6 > CAST (moderate)") |> pull(gene_id)

# Extract gene lists from domain_patterns
genes_PM_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "PM") |> pull(gene_id)
genes_MAX_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "MAX") |> pull(gene_id)
genes_POST_biggest_effect <- domain_patterns |>
  filter(max_domain_effect == "POST") |> pull(gene_id)
genes_opposite_directions <- domain_patterns |>
  filter(domain_consistency == "Variable") |> pull(gene_id)

# Extract gene lists from time_patterns
genes_diverging_CAST_faster <- time_patterns |>
  filter(time_int_slope > 0.1) |> pull(gene_id)
genes_diverging_B6_faster <- time_patterns |>
  filter(time_int_slope < -0.1) |> pull(gene_id)

# Extract gene lists from threeway_patterns
genes_PM_biggest_slope_diff <- threeway_patterns |>
  filter(max_slope_diff_domain == "PM") |> pull(gene_id)
genes_MAX_biggest_slope_diff <- threeway_patterns |>
  filter(max_slope_diff_domain == "MAX") |> pull(gene_id)
genes_POST_biggest_slope_diff <- threeway_patterns |>
  filter(max_slope_diff_domain == "POST") |> pull(gene_id)
genes_consistent_slopes <- threeway_patterns |>
  filter(slope_consistency == "Consistent") |> pull(gene_id)

# Extract gene lists from domain_specificity
genes_strain_PM_only <- domain_specificity |>
  filter(specificity_pattern == "PM only") |> pull(gene_id)
genes_strain_MAX_only <- domain_specificity |>
  filter(specificity_pattern == "MAX only") |> pull(gene_id)
genes_strain_POST_only <- domain_specificity |>
  filter(specificity_pattern == "POST only") |> pull(gene_id)
genes_strain_all_domains <- domain_specificity |>
  filter(specificity_pattern == "All domains") |> pull(gene_id)

```

This analysis investigates how gene expression differs between B6 and CAST mouse strains across three palate domains (PM, MAX, POST) and developmental time. We use a statistical model that decomposes strain effects into:
 
- **Main strain effect**: Average CAST vs B6 difference 
- **Domain × Strain interaction**: How strain effect differs between domains
- **Time × Strain interaction**: How strain effect changes per unit developmental time
- **Three-way interaction**: Domain-specific changes in temporal strain dynamics

Briefly, I merge the technical replicates due to re-sequencing of several samples using `collapseReplicates()` function from DESeq2. Next, I calculate the relative or centered limb bud (LB) stage (LB_stage_from_start), where I subtract the minimum observed stage so that 0 represents the earliest developmental timepoint in the dataset. Then, I use likelihood ratio test (LRT) to compare the linear model that lacks strain effects (`~ sex + AP_domain + LB_stage_from_start + AP_domain:ind`) to the full model (`~ sex + AP_domain + LB_stage_from_start + AP_domain:ind + strain + strain:AP_domain + strain:LB_stage_from_start + strain:AP_domain:LB_stage_from_start`) that includes strain and strain interaction effects with domain and developmental time. I run separate LRT tests to identify genes with significant domain × strain interactions (removing `strain:AP_domain` terms from the full model) and time × strain interactions (removing `strain:LB_stage_from_start` terms) and three-way interactions (removing `strain:AP_domain:LB_stage_from_start` terms).

The LRT tests identify **which** genes have significant effects but do not provide the actual effect sizes. To get the magnitude and direction of effects, I extract coefficients from the fitted full model using Wald tests. The main strain effect (`strain_CAST_vs_B6`) represents the CAST vs B6 log2 fold change in the reference domain (PM) at developmental time = 0. The domain interaction coefficients (`strainCAST.AP_domainMAX` and `strainCAST.AP_domainPOST`) tell us how much the strain effect differs in MAX and POST compared to PM. To get the actual strain effect within each domain, I use contrasts that combine the main effect with the appropriate interaction term (e.g., for MAX: `strain_CAST_vs_B6 + strainCAST.AP_domainMAX`). Similarly, the time interaction coefficient (`strainCAST.LB_stage_from_start`) represents the change in log2FC(CAST/B6) per unit of developmental time, and the three-way coefficients indicate how this temporal slope differs by domain.

From this analysis, I have identified **`r sum(strain_effects_classified$any_strain_sig, na.rm = TRUE)`** genes with significant strain effects (padj < 0.05). I then apply quality filters to each result object individually, removing genes with low expression (baseMean below 10th percentile), high uncertainty (lfcSE ≥ 1.5 for effect estimates, ≥ 0.5 for slopes), unrealistically large effects (|log2FC| ≥ 10), or very small effects (thresholds vary by coefficient type). This per-result filtering ensures that each effect estimate meets quality standards appropriate for its scale. After filtering the main strain effect, **`r nrow(strain_effects_filtered)`** high-confidence genes remain. I use the `strain_pattern_classification.R` script to classify these differentially expressed genes into various expression patterns based on the model coefficients: main strain effects (CAST vs B6 direction and magnitude), domain-specific patterns (which domain shows the largest effect, whether effects are consistent across domains), temporal patterns (diverging vs parallel trajectories), and domain specificity (which specific domains show significant strain differences).

```{r}
#| label: overview-summary
#| message: false
#| warning: false

strain_effects_filtered |>
  filter(any_strain_sig) |>
  count(pattern_type) |>
  arrange(desc(n)) |>
  rename(`Pattern Type` = pattern_type, `Number of Genes` = n) |>
  create_dt()
```

```{r}
#| label: pattern-summary-table
#| message: false
#| warning: false

pattern_summary |> 
  create_dt()
```

# Main Strain Effects

These genes show consistent expression differences between CAST and B6. The log2FC values represent the CAST vs B6 effect estimated from the model.

:::: {#main-strain-tabs .panel-tabset .nav-pills}

## CAST >> B6 (Strong)

Genes with strong upregulation in CAST (log2FC > 1). Here are some examples:

```{r}
#| label: main-CAST-strong-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6


count_data <- get_normalized_counts(strain_dds_full, genes_CAST_high_strong)
plot_main_strain_effect(genes_CAST_high_strong, "CAST >> B6 (strong)", count_data, n = 2)


```

Here is the full list of genes:
```{r}
#| label: main-CAST-strong-table
#| message: false
#| warning: false

main_patterns |> 
  filter(gene_id %in% genes_CAST_high_strong) |>
  select(symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```

Here is the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-CAST-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_strong_ora <- run_enrichment_hybrid(genes_CAST_high_strong, 
                                                    description = "CAST >> B6 (strong)")

```

```{r}
#| label: main-CAST-strong-ora-table
#| message: false
#| warning: false

genes_CAST_high_strong_ora$result |> 
  filter( p_value < 0.01) |> 
  create_dt()

```


## CAST > B6 (Moderate)

Genes with moderate upregulation in CAST (0.5 < log2FC ≤ 1).

```{r}
#| label: main-CAST-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_normalized_counts(strain_dds_full, 
                                    genes_CAST_high_mod)

plot_main_strain_effect(genes_CAST_high_mod, "CAST > B6 (moderate)", 
                        count_data, n = 2)

```


```{r}
#| label: main-CAST-mod-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_CAST_high_mod) |>
  select(symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-CAST-mod-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_mod_ora <- run_enrichment_hybrid(genes_CAST_high_mod,
                                                 description = "CAST > B6 (moderate)")

```

```{r}
#| label: main-CAST-mod-ora-table
#| message: false
#| warning: false

genes_CAST_high_mod_ora$result |> 
  filter( p_value < 0.01) |> 
  create_dt()

```

## B6 >> CAST (Strong)

Genes with strong upregulation in B6 (log2FC < -1).

```{r}
#| label: main-B6-strong-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_normalized_counts(strain_dds_full, 
                                    genes_B6_high_strong)
plot_main_strain_effect(genes_B6_high_strong, "B6 >> CAST (strong)", count_data, n = 2)

```

```{r}
#| label: main-B6-strong-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_B6_high_strong) |>
  select(symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-B6-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true


genes_B6_high_strong_ora <- run_enrichment_hybrid(genes_B6_high_strong,
                                                  description = "B6 >> CAST (strong)")

```

```{r}
#| label: main-B6-strong-ora-table
#| message: false
#| warning: false


genes_B6_high_strong_ora$result |> 
  filter( p_value < 0.01) |> 
  create_dt()

```


## B6 > CAST (Moderate)

Genes with moderate upregulation in B6 (-1 ≤ log2FC < -0.5).

```{r}
#| label: main-B6-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_B6_high_mod)
plot_main_strain_effect(genes_B6_high_mod, "B6 > CAST (moderate)", count_data, n = 2)

```

```{r}
#| label: main-B6-mod-table
#| message: false
#| warning: false


main_patterns |> 
  filter(gene_id %in% genes_B6_high_mod) |>
  select(symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |> 
  create_dt()

```


```{r}
#| label: main-B6-mod-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_B6_high_mod_ora <- run_enrichment_hybrid(genes_B6_high_mod, 
                                               description = "B6 > CAST (moderate)")
  
```

```{r}
#| label: main-B6-mod-ora-table
#| message: false
#| warning: false

genes_B6_high_mod_ora$result |> 
  filter( p_value < 0.01) |> 
  create_dt()
  
```

::::

# Domain × Strain Interactions

These genes show strain differences that vary across palate domains. The model provides strain effects within each domain and interaction coefficients showing how MAX/POST differ from PM.

:::: {#domain-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-strain-overview
#| message: false
#| warning: false

domain_patterns |> 
  count(detailed_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |> 
  create_dt()

```

## PM: Biggest Effect

Genes where the largest strain effect magnitude occurs in PM.

```{r}
#| label: domain-PM-biggest-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_PM_biggest_effect)
plot_domain_strain_interaction(genes_PM_biggest_effect, "Biggest strain effect in PM", count_data, n = 2)

```


```{r}
#| label: domain-PM-biggest-table
#| message: false
#| warning: false


domain_patterns |> 
  filter(gene_id %in% genes_PM_biggest_effect) |>
  select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
           domain_int_MAX_lfc, domain_int_POST_lfc, max_domain_direction) |> 
  create_dt()

```


```{r}
#| label: domain-PM-biggest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in PM
genes_PM_biggest_up <- domain_patterns |> 
  filter(gene_id %in% genes_PM_biggest_effect, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_PM_biggest_down <- domain_patterns |> 
  filter(gene_id %in% genes_PM_biggest_effect, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_PM_biggest_effect_ora_up <- run_enrichment_hybrid(
  genes_PM_biggest_up, 
  description = "PM biggest effect - Up in CAST"
)

genes_PM_biggest_effect_ora_down <- run_enrichment_hybrid(
  genes_PM_biggest_down, 
  description = "PM biggest effect - Down in CAST"
)
  
```

```{r}
#| label: domain-PM-biggest-ora-table
#| message: false
#| warning: false


genes_PM_biggest_effect_ora_up$result |> 
  rbind(
    genes_PM_biggest_effect_ora_down$result
  ) |> 
  filter(p_value < 0.01) |> 
  create_dt()
  
```

## MAX: Biggest Effect

Genes where the largest strain effect magnitude occurs in MAX.

```{r}
#| label: domain-MAX-biggest-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_MAX_biggest_effect)
plot_domain_strain_interaction(genes_MAX_biggest_effect, "Biggest strain effect in MAX", count_data, n = 2)

```

```{r}
#| label: domain-MAX-biggest-table
#| message: false
#| warning: false


domain_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_effect) |>
    select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
           domain_int_MAX_lfc, domain_int_POST_lfc, max_domain_direction) |> 
  create_dt()

```


```{r}
#| label: domain-MAX-biggest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in MAX
genes_MAX_biggest_up <- domain_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_effect, strain_effect_MAX > 0) |> 
  pull(gene_id)

genes_MAX_biggest_down <- domain_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_effect, strain_effect_MAX < 0) |> 
  pull(gene_id)

genes_MAX_biggest_effect_ora_up <- run_enrichment_hybrid(
  genes_MAX_biggest_up, 
  description = "MAX biggest effect - Up in CAST"
)

genes_MAX_biggest_effect_ora_down <- run_enrichment_hybrid(
  genes_MAX_biggest_down, 
  description = "MAX biggest effect - Down in CAST"
)
  
```


```{r}
#| label: domain-MAX-biggest-ora-table
#| message: false
#| warning: false

genes_MAX_biggest_effect_ora_up$result |> 
  rbind( 
    genes_MAX_biggest_effect_ora_down$result 
    ) |> 
  filter(p_value < 0.01) |> 
  create_dt()
  
```


## POST: Biggest Effect

Genes where the largest strain effect magnitude occurs in POST.

```{r}
#| label: domain-POST-biggest-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_POST_biggest_effect)
plot_domain_strain_interaction(genes_POST_biggest_effect, "Biggest strain effect in POST", count_data, n = 2)

```


```{r}
#| label: domain-POST-biggest-table
#| message: false
#| warning: false


domain_patterns |> 
  filter(gene_id %in% genes_POST_biggest_effect) |>
    select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, 
           domain_int_MAX_lfc, domain_int_POST_lfc, max_domain_direction) |> 
  create_dt()

```

```{r}
#| label: domain-POST-biggest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in POST
genes_POST_biggest_up <- domain_patterns |> 
  filter(gene_id %in% genes_POST_biggest_effect, strain_effect_POST > 0) |> 
  pull(gene_id)

genes_POST_biggest_down <- domain_patterns |> 
  filter(gene_id %in% genes_POST_biggest_effect, strain_effect_POST < 0) |> 
  pull(gene_id)

genes_POST_biggest_effect_ora_up <- run_enrichment_hybrid(
  genes_POST_biggest_up, 
  description = "POST biggest effect - Up in CAST"
)

genes_POST_biggest_effect_ora_down <- run_enrichment_hybrid(
  genes_POST_biggest_down, 
  description = "POST biggest effect - Down in CAST"
)

```

```{r}
#| label: domain-POST-biggest-ora-table
#| message: false
#| warning: false

genes_POST_biggest_effect_ora_up$result |> 
  rbind(
    genes_POST_biggest_effect_ora_down$result
  ) |> 
  filter(p_value < 0.01) |> 
  create_dt()

```

## Opposite Directions

Genes where strain effects go in opposite directions across domains.

```{r}
#| label: domain-opposite-plot
#| message: false
#| warning: false
#| fig-height: 12
#| fig-width: 14

count_data <- get_normalized_counts(strain_dds_full, genes_opposite_directions)
plot_domain_strain_interaction(genes_opposite_directions, "Opposite directions across domains", count_data, n = 2)

```

```{r}
#| label: domain-opposite-table
#| message: false
#| warning: false


domain_patterns |> 
  filter(gene_id %in% genes_opposite_directions) |>
  select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, domain_consistency) |> 
  create_dt()

```

```{r}
#| label: domain-opposite-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction using PM as reference
genes_opposite_up_in_PM <- domain_patterns |> 
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_opposite_down_in_PM <- domain_patterns |> 
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_opposite_directions_ora_up <- run_enrichment_hybrid(
  genes_opposite_up_in_PM, 
  description = "Opposite directions - Up in CAST in PM"
)

genes_opposite_directions_ora_down <- run_enrichment_hybrid(
  genes_opposite_down_in_PM, 
  description = "Opposite directions - Down in CAST in PM"
)
 
```

```{r}
#| label: domain-opposite-ora-table
#| message: false
#| warning: false

genes_opposite_directions_ora_up$result |> 
  rbind(
    genes_opposite_directions_ora_down$result
  ) |> 
  filter(p_value < 0.01) |> 
  create_dt()
 
```

::::

# Time × Strain Interactions

These genes show strain differences that change over developmental time. The `time_int_slope` coefficient represents the change in log2FC(CAST/B6) per unit of developmental time.

:::: {#time-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: time-strain-overview
#| message: false
#| warning: false

time_patterns |> 
  count(trajectory_pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = trajectory_pattern, `N Genes` = n) |> 
  create_dt()

```

## Diverging: CAST Faster

Strains diverge over time with CAST increasing faster (positive slope).

```{r}
#| label: time-CAST-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_diverging_CAST_faster)
plot_time_strain_interaction(genes_diverging_CAST_faster, "Diverging: CAST faster", count_data, n = 2)

```

```{r}
#| label: time-CAST-faster-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster) |>
  select(symbol, time_int_slope, time_int_pval, main_strain_lfc, trajectory_pattern) |>
  create_dt()

```

```{r}
#| label: time-CAST-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_CAST_faster_up <- time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc > 0) |> 
  pull(gene_id)

genes_CAST_faster_down <- time_patterns |> 
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc < 0) |> 
  pull(gene_id)

genes_diverging_CAST_faster_ora_up <- run_enrichment_hybrid(
  genes_CAST_faster_up, 
  description = "CAST faster - Up in CAST at baseline"
)

genes_diverging_CAST_faster_ora_down <- run_enrichment_hybrid(
  genes_CAST_faster_down, 
  description = "CAST faster - Down in CAST at baseline"
)

```

```{r}
#| label: time-CAST-faster-ora-table
#| message: false
#| warning: false

bind_rows(
  genes_diverging_CAST_faster_ora_up$result |> 
    mutate(direction = paste0("Up in CAST (n=", length(genes_CAST_faster_up), ")")),
  genes_diverging_CAST_faster_ora_down$result |> 
    mutate(direction = paste0("Down in CAST (n=", length(genes_CAST_faster_down), ")"))
) |> 
  filter(p_value < 0.01) |> 
  select(direction, everything()) |> 
  create_dt()

```

## Diverging: B6 Faster

Strains diverge over time with B6 increasing faster (negative slope).

```{r}
#| label: time-B6-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_normalized_counts(strain_dds_full, genes_diverging_B6_faster)
plot_time_strain_interaction(genes_diverging_B6_faster, "Diverging: B6 faster", count_data, n = 4)

```

```{r}
#| label: time-B6-faster-table
#| message: false
#| warning: false

time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster) |>
  select(symbol, time_int_slope, time_int_pval, main_strain_lfc, trajectory_pattern) |> 
  create_dt()

```

```{r}
#| label: time-B6-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_B6_faster_up <- time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc > 0) |> 
  pull(gene_id)

genes_B6_faster_down <- time_patterns |> 
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc < 0) |> 
  pull(gene_id)

genes_diverging_B6_faster_ora_up <- run_enrichment_hybrid(
  genes_B6_faster_up, 
  description = "B6 faster - Up in CAST at baseline"
)

genes_diverging_B6_faster_ora_down <- run_enrichment_hybrid(
  genes_B6_faster_down, 
  description = "B6 faster - Down in CAST at baseline"
)

```

```{r}
#| label: time-B6-faster-ora-table
#| message: false
#| warning: false

bind_rows(
  genes_diverging_B6_faster_ora_up$result |> 
    mutate(direction = paste0("Up in CAST (n=", length(genes_B6_faster_up), ")")),
  genes_diverging_B6_faster_ora_down$result |> 
    mutate(direction = paste0("Down in CAST (n=", length(genes_B6_faster_down), ")"))
) |> 
  filter(p_value < 0.01) |> 
  select(direction, everything()) |> 
  create_dt()

```

::::

# Three-Way Interactions

These genes show domain-specific strain differences that also change over developmental time. The three-way coefficients indicate how the temporal slope of strain effects differs by domain.

:::: {#threeway-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: threeway-overview
#| message: false
#| warning: false

threeway_patterns |> 
  count(pattern) |> 
  arrange(desc(n)) |>
  rename(Pattern = pattern, `N Genes` = n) |> 
  create_dt()

```

## PM Diverges Most

Genes where PM shows the most distinct temporal pattern.

```{r}
#| label: threeway-PM-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 10

count_data <- get_normalized_counts(strain_dds_full, genes_PM_biggest_slope_diff)
plot_threeway_interaction(genes_PM_biggest_slope_diff, "PM diverges most", count_data, n = 2)

```

```{r}
#| label: threeway-PM-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_PM_biggest_slope_diff) |>
  select(symbol, slope_PM, slope_MAX, slope_POST, pattern) |> 
  create_dt()

```

```{r}
#| label: threeway-PM-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by slope direction in PM
genes_PM_slope_up <- threeway_patterns |> 
  filter(gene_id %in% genes_PM_biggest_slope_diff, slope_PM > 0) |> 
  pull(gene_id)

genes_PM_slope_down <- threeway_patterns |> 
  filter(gene_id %in% genes_PM_biggest_slope_diff, slope_PM < 0) |> 
  pull(gene_id)

genes_PM_biggest_slope_diff_ora_up <- run_enrichment_hybrid(
  genes_PM_slope_up, 
  description = "PM diverges most - Positive slope"
)

genes_PM_biggest_slope_diff_ora_down <- run_enrichment_hybrid(
  genes_PM_slope_down, 
  description = "PM diverges most - Negative slope"
)
  
```

```{r}
#| label: threeway-PM-ora-table
#| message: false
#| warning: false

bind_rows(
  genes_PM_biggest_slope_diff_ora_up$result |> 
    mutate(direction = paste0("Positive slope (n=", length(genes_PM_slope_up), ")")),
  genes_PM_biggest_slope_diff_ora_down$result |> 
    mutate(direction = paste0("Negative slope (n=", length(genes_PM_slope_down), ")"))
) |> 
  filter(p_value < 0.01) |> 
  select(direction, everything()) |> 
  create_dt()
  
```

## MAX Diverges Most

Genes where MAX shows the most distinct temporal pattern.

```{r}
#| label: threeway-MAX-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 10

count_data <- get_normalized_counts(strain_dds_full, genes_MAX_biggest_slope_diff)
plot_threeway_interaction(genes_MAX_biggest_slope_diff, "MAX diverges most", count_data, n = 2)

```

```{r}
#| label: threeway-MAX-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_slope_diff) |>
  select(symbol, slope_PM, slope_MAX, slope_POST, pattern) |> 
  create_dt()

```

```{r}
#| label: threeway-MAX-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by slope direction in MAX
genes_MAX_slope_up <- threeway_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_slope_diff, slope_MAX > 0) |> 
  pull(gene_id)

genes_MAX_slope_down <- threeway_patterns |> 
  filter(gene_id %in% genes_MAX_biggest_slope_diff, slope_MAX < 0) |> 
  pull(gene_id)

genes_MAX_biggest_slope_diff_ora_up <- run_enrichment_hybrid(
  genes_MAX_slope_up, 
  description = "MAX diverges most - Positive slope"
)

genes_MAX_biggest_slope_diff_ora_down <- run_enrichment_hybrid(
  genes_MAX_slope_down, 
  description = "MAX diverges most - Negative slope"
)

```

```{r}
#| label: threeway-MAX-ora-table
#| message: false
#| warning: false

bind_rows(
  genes_MAX_biggest_slope_diff_ora_up$result |> 
    mutate(direction = paste0("Positive slope (n=", length(genes_MAX_slope_up), ")")),
  genes_MAX_biggest_slope_diff_ora_down$result |> 
    mutate(direction = paste0("Negative slope (n=", length(genes_MAX_slope_down), ")"))
) |> 
  filter(p_value < 0.01) |> 
  select(direction, everything()) |> 
  create_dt()

```

## POST Diverges Most

Genes where POST shows the most distinct temporal pattern.

```{r}
#| label: threeway-POST-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 10

count_data <- get_normalized_counts(strain_dds_full, genes_POST_biggest_slope_diff)
plot_threeway_interaction(genes_POST_biggest_slope_diff, "POST diverges most", count_data, n = 2)

```

```{r}
#| label: threeway-POST-table
#| message: false
#| warning: false

threeway_patterns |> 
  filter(gene_id %in% genes_POST_biggest_slope_diff) |>
  select(symbol, slope_PM, slope_MAX, slope_POST, pattern) |> 
  create_dt()

```

```{r}
#| label: threeway-POST-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by slope direction in POST
genes_POST_slope_up <- threeway_patterns |> 
  filter(gene_id %in% genes_POST_biggest_slope_diff, slope_POST > 0) |> 
  pull(gene_id)

genes_POST_slope_down <- threeway_patterns |> 
  filter(gene_id %in% genes_POST_biggest_slope_diff, slope_POST < 0) |> 
  pull(gene_id)

genes_POST_biggest_slope_diff_ora_up <- run_enrichment_hybrid(
  genes_POST_slope_up, 
  description = "POST diverges most - Positive slope"
)

genes_POST_biggest_slope_diff_ora_down <- run_enrichment_hybrid(
  genes_POST_slope_down, 
  description = "POST diverges most - Negative slope"
)

```

```{r}
#| label: threeway-POST-ora-table
#| message: false
#| warning: false

bind_rows(
  genes_POST_biggest_slope_diff_ora_up$result |> 
    mutate(direction = paste0("Positive slope (n=", length(genes_POST_slope_up), ")")),
  genes_POST_biggest_slope_diff_ora_down$result |> 
    mutate(direction = paste0("Negative slope (n=", length(genes_POST_slope_down), ")"))
) |> 
  filter(p_value < 0.01) |> 
  select(direction, everything()) |> 
  create_dt()

```

::::

# Domain-Specific Strain Effects

For genes with domain × strain interactions, we identify which specific domains show significant strain effects (combining significance and effect size thresholds).

:::: {#domain-specific-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-specific-overview
#| message: false
#| warning: false

domain_specificity |> 
  count(specificity_pattern) |> 
  arrange(desc(n)) |>
  rename(`Specificity Pattern` = specificity_pattern, `Number of Genes` = n) |> 
  create_dt()

```

## PM Only

Strain differences significant only in PM.

```{r}
#| label: domain-specific-PM-only-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

count_data <- get_normalized_counts(strain_dds_full, genes_strain_PM_only)
plot_domain_specificity(genes_strain_PM_only, "PM only", count_data, n = 2)

```

```{r}
#| label: domain-specific-PM-only-table
#| message: false
#| warning: false

domain_specificity |> 
  filter(gene_id %in% genes_strain_PM_only) |>
  select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, specificity_pattern) |> 
  create_dt()

```

```{r}
#| label: domain-specific-PM-only-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by strain effect direction in PM
genes_PM_only_up <- domain_specificity |> 
  filter(gene_id %in% genes_strain_PM_only, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_PM_only_down <- domain_specificity |> 
  filter(gene_id %in% genes_strain_PM_only, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_strain_PM_only_ora_up <- run_enrichment_hybrid(
  genes_PM_only_up, 
  description = "PM only - Up in CAST"
)

genes_strain_PM_only_ora_down <- run_enrichment_hybrid(
  genes_PM_only_down, 
  description = "PM only - Down in CAST"
)

```

```{r}
#| label: domain-specific-PM-only-ora-table
#| message: false
#| warning: false

bind_rows(
  genes_strain_PM_only_ora_up$result |> 
    mutate(direction = paste0("Up in CAST (n=", length(genes_PM_only_up), ")")),
  genes_strain_PM_only_ora_down$result |> 
    mutate(direction = paste0("Down in CAST (n=", length(genes_PM_only_down), ")"))
) |> 
  filter(p_value < 0.01) |> 
  select(direction, everything()) |> 
  create_dt()

```

## MAX Only

Strain differences significant only in MAX.

```{r}
#| label: domain-specific-MAX-only-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

if (length(genes_strain_MAX_only) > 0) {
  count_data <- get_normalized_counts(strain_dds_full, genes_strain_MAX_only)
  plot_domain_specificity(genes_strain_MAX_only, "MAX only", count_data, n = 2)
}

```

```{r}
#| label: domain-specific-MAX-only-table
#| message: false
#| warning: false

if (length(genes_strain_MAX_only) > 0) {
  domain_specificity |> 
    filter(gene_id %in% genes_strain_MAX_only) |>
    select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, specificity_pattern) |> 
    create_dt()
}

```

```{r}
#| label: domain-specific-MAX-only-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_strain_MAX_only_ora_up <- NULL
genes_strain_MAX_only_ora_down <- NULL

if (length(genes_strain_MAX_only) > 0) {
  genes_MAX_only_up <- domain_specificity |> 
    filter(gene_id %in% genes_strain_MAX_only, strain_effect_MAX > 0) |> 
    pull(gene_id)
  
  genes_MAX_only_down <- domain_specificity |> 
    filter(gene_id %in% genes_strain_MAX_only, strain_effect_MAX < 0) |> 
    pull(gene_id)
  
  if (length(genes_MAX_only_up) > 0) {
    genes_strain_MAX_only_ora_up <- run_enrichment_hybrid(
      genes_MAX_only_up, 
      description = "MAX only - Up in CAST"
    )
  }
  
  if (length(genes_MAX_only_down) > 0) {
    genes_strain_MAX_only_ora_down <- run_enrichment_hybrid(
      genes_MAX_only_down, 
      description = "MAX only - Down in CAST"
    )
  }
}

```

```{r}
#| label: domain-specific-MAX-only-ora-table
#| message: false
#| warning: false

if (!is.null(genes_strain_MAX_only_ora_up) || !is.null(genes_strain_MAX_only_ora_down)) {
  bind_rows(
    if (!is.null(genes_strain_MAX_only_ora_up)) {
      genes_strain_MAX_only_ora_up$result |> 
        mutate(direction = paste0("Up in CAST (n=", length(genes_MAX_only_up), ")"))
    },
    if (!is.null(genes_strain_MAX_only_ora_down)) {
      genes_strain_MAX_only_ora_down$result |> 
        mutate(direction = paste0("Down in CAST (n=", length(genes_MAX_only_down), ")"))
    }
  ) |> 
    filter(p_value < 0.01) |> 
    select(direction, everything()) |> 
    create_dt()
}

```

## POST Only

Strain differences significant only in POST.

```{r}
#| label: domain-specific-POST-only-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

if (length(genes_strain_POST_only) > 0) {
  count_data <- get_normalized_counts(strain_dds_full, genes_strain_POST_only)
  plot_domain_specificity(genes_strain_POST_only, "POST only", count_data, n = 2)
}

```

```{r}
#| label: domain-specific-POST-only-table
#| message: false
#| warning: false

if (length(genes_strain_POST_only) > 0) {
  domain_specificity |> 
    filter(gene_id %in% genes_strain_POST_only) |>
    select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, specificity_pattern) |> 
    create_dt()
}

```

```{r}
#| label: domain-specific-POST-only-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_strain_POST_only_ora_up <- NULL
genes_strain_POST_only_ora_down <- NULL

if (length(genes_strain_POST_only) > 0) {
  genes_POST_only_up <- domain_specificity |> 
    filter(gene_id %in% genes_strain_POST_only, strain_effect_POST > 0) |> 
    pull(gene_id)
  
  genes_POST_only_down <- domain_specificity |> 
    filter(gene_id %in% genes_strain_POST_only, strain_effect_POST < 0) |> 
    pull(gene_id)
  
  if (length(genes_POST_only_up) > 0) {
    genes_strain_POST_only_ora_up <- run_enrichment_hybrid(
      genes_POST_only_up, 
      description = "POST only - Up in CAST"
    )
  }
  
  if (length(genes_POST_only_down) > 0) {
    genes_strain_POST_only_ora_down <- run_enrichment_hybrid(
      genes_POST_only_down, 
      description = "POST only - Down in CAST"
    )
  }
}

```

```{r}
#| label: domain-specific-POST-only-ora-table
#| message: false
#| warning: false

if (!is.null(genes_strain_POST_only_ora_up) || !is.null(genes_strain_POST_only_ora_down)) {
  bind_rows(
    if (!is.null(genes_strain_POST_only_ora_up)) {
      genes_strain_POST_only_ora_up$result |> 
        mutate(direction = paste0("Up in CAST (n=", length(genes_POST_only_up), ")"))
    },
    if (!is.null(genes_strain_POST_only_ora_down)) {
      genes_strain_POST_only_ora_down$result |> 
        mutate(direction = paste0("Down in CAST (n=", length(genes_POST_only_down), ")"))
    }
  ) |> 
    filter(p_value < 0.01) |> 
    select(direction, everything()) |> 
    create_dt()
}

```

## All Domains

Strain differences significant in all three domains.

```{r}
#| label: domain-specific-all-plot
#| message: false
#| warning: false
#| fig-height: 14
#| fig-width: 9

count_data <- get_normalized_counts(strain_dds_full, genes_strain_all_domains)
plot_domain_specificity(genes_strain_all_domains, "All domains", count_data, n = 2)

```

```{r}
#| label: domain-specific-all-table
#| message: false
#| warning: false

domain_specificity |> 
  filter(gene_id %in% genes_strain_all_domains) |>
  select(symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST, specificity_pattern) |> 
  create_dt()

```

```{r}
#| label: domain-specific-all-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by strain effect direction in PM (reference domain)
genes_all_domains_up <- domain_specificity |> 
  filter(gene_id %in% genes_strain_all_domains, strain_effect_PM > 0) |> 
  pull(gene_id)

genes_all_domains_down <- domain_specificity |> 
  filter(gene_id %in% genes_strain_all_domains, strain_effect_PM < 0) |> 
  pull(gene_id)

genes_strain_all_domains_ora_up <- run_enrichment_hybrid(
  genes_all_domains_up, 
  description = "All domains - Up in CAST"
)

genes_strain_all_domains_ora_down <- run_enrichment_hybrid(
  genes_all_domains_down, 
  description = "All domains - Down in CAST"
)

```

```{r}
#| label: domain-specific-all-ora-table
#| message: false
#| warning: false

bind_rows(
  genes_strain_all_domains_ora_up$result |> 
    mutate(direction = paste0("Up in CAST (n=", length(genes_all_domains_up), ")")),
  genes_strain_all_domains_ora_down$result |> 
    mutate(direction = paste0("Down in CAST (n=", length(genes_all_domains_down), ")"))
) |> 
  filter(p_value < 0.01) |> 
  select(direction, everything()) |> 
  create_dt()

```

::::

# Comprehensive Tables

This section provides downloadable tables summarizing all analysis results.
:::: {#comprehensive-tables-tabs .panel-tabset .nav-pills}

## Pattern Summary

```{r}
#| label: comprehensive-pattern-summary
#| message: false
#| warning: false

pattern_summary |>
  create_dt()
```

## Gene-Level Pattern Assignments

All genes with their assigned expression patterns across strain, domain, and time.

```{r}
#| label: comprehensive-gene-assignments
#| message: false
#| warning: false

gene_assignments |>
  left_join(all_genes_palate |> select(gene_id, symbol, gene_biotype, chromosome)) |>
  select(gene_id, symbol, gene_biotype, chromosome, everything()) |>
  create_dt()

```

## Full Strain Effects Classification

Complete results table with all coefficients and classifications.

```{r}
#| label: comprehensive-strain-effects
#| message: false
#| warning: false

strain_effects_classified |>
  filter(any_strain_sig) |>
  select(
    gene_id, symbol, gene_biotype, baseMean,
    # Pattern classifications
    pattern_type, main_strain_pattern, 
    # Main strain effect
    main_strain_lfc, main_strain_pval,
    # Domain-specific effects
    strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    max_domain_effect, domain_consistency,
    # Time effects
    time_int_slope, time_pattern,
    # Significance flags
    domain_int_sig, time_int_sig, threeway_int_sig
  ) |>
  create_dt()
```

## Combined ORA Results

All over-representation analysis results combined into a single table.

```{r}
#| label: comprehensive-ora-combine
#| message: false
#| warning: false

# Combine all ORA results into one table
all_ora_results <- bind_rows(
  # Main strain effects
  if (!is.null(genes_CAST_high_strong_ora$result)) 
    genes_CAST_high_strong_ora$result |> 
      mutate(category = "Main Effect", pattern = "CAST >> B6 (strong)", direction = "Up in CAST"),
  if (!is.null(genes_CAST_high_mod_ora$result)) 
    genes_CAST_high_mod_ora$result |> 
      mutate(category = "Main Effect", pattern = "CAST > B6 (moderate)", direction = "Up in CAST"),
  if (!is.null(genes_B6_high_strong_ora$result)) 
    genes_B6_high_strong_ora$result |> 
      mutate(category = "Main Effect", pattern = "B6 >> CAST (strong)", direction = "Down in CAST"),
  if (!is.null(genes_B6_high_mod_ora$result)) 
    genes_B6_high_mod_ora$result |> 
      mutate(category = "Main Effect", pattern = "B6 > CAST (moderate)", direction = "Down in CAST"),
  
 # Domain effects - PM biggest
  if (!is.null(genes_PM_biggest_effect_ora_up$result)) 
    genes_PM_biggest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "PM biggest effect", direction = "Up in CAST"),
  if (!is.null(genes_PM_biggest_effect_ora_down$result)) 
    genes_PM_biggest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "PM biggest effect", direction = "Down in CAST"),
  
  # Domain effects - MAX biggest
  if (!is.null(genes_MAX_biggest_effect_ora_up$result)) 
    genes_MAX_biggest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "MAX biggest effect", direction = "Up in CAST"),
  if (!is.null(genes_MAX_biggest_effect_ora_down$result)) 
    genes_MAX_biggest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "MAX biggest effect", direction = "Down in CAST"),
  
  # Domain effects - POST biggest
  if (!is.null(genes_POST_biggest_effect_ora_up$result)) 
    genes_POST_biggest_effect_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "POST biggest effect", direction = "Up in CAST"),
  if (!is.null(genes_POST_biggest_effect_ora_down$result)) 
    genes_POST_biggest_effect_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "POST biggest effect", direction = "Down in CAST"),
  
  # Domain effects - Opposite directions
  if (!is.null(genes_opposite_directions_ora_up$result)) 
    genes_opposite_directions_ora_up$result |> 
      mutate(category = "Domain Effect", pattern = "Opposite directions", direction = "Up in CAST (PM)"),
  if (!is.null(genes_opposite_directions_ora_down$result)) 
    genes_opposite_directions_ora_down$result |> 
      mutate(category = "Domain Effect", pattern = "Opposite directions", direction = "Down in CAST (PM)"),
  
  # Time effects - CAST faster
  if (!is.null(genes_diverging_CAST_faster_ora_up$result)) 
    genes_diverging_CAST_faster_ora_up$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: CAST faster", direction = "Up in CAST"),
  if (!is.null(genes_diverging_CAST_faster_ora_down$result)) 
    genes_diverging_CAST_faster_ora_down$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: CAST faster", direction = "Down in CAST"),
  
  # Time effects - B6 faster
  if (!is.null(genes_diverging_B6_faster_ora_up$result)) 
    genes_diverging_B6_faster_ora_up$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: B6 faster", direction = "Up in CAST"),
  if (!is.null(genes_diverging_B6_faster_ora_down$result)) 
    genes_diverging_B6_faster_ora_down$result |> 
      mutate(category = "Time Effect", pattern = "Diverging: B6 faster", direction = "Down in CAST"),
  
  # Three-way effects - PM diverges most
  if (!is.null(genes_PM_biggest_slope_diff_ora_up$result)) 
    genes_PM_biggest_slope_diff_ora_up$result |> 
      mutate(category = "Three-way", pattern = "PM diverges most", direction = "Positive slope"),
  if (!is.null(genes_PM_biggest_slope_diff_ora_down$result)) 
    genes_PM_biggest_slope_diff_ora_down$result |> 
      mutate(category = "Three-way", pattern = "PM diverges most", direction = "Negative slope"),
  
  # Three-way effects - MAX diverges most
  if (!is.null(genes_MAX_biggest_slope_diff_ora_up$result)) 
    genes_MAX_biggest_slope_diff_ora_up$result |> 
      mutate(category = "Three-way", pattern = "MAX diverges most", direction = "Positive slope"),
  if (!is.null(genes_MAX_biggest_slope_diff_ora_down$result)) 
    genes_MAX_biggest_slope_diff_ora_down$result |> 
      mutate(category = "Three-way", pattern = "MAX diverges most", direction = "Negative slope"),
  
  # Three-way effects - POST diverges most
  if (!is.null(genes_POST_biggest_slope_diff_ora_up$result)) 
    genes_POST_biggest_slope_diff_ora_up$result |> 
      mutate(category = "Three-way", pattern = "POST diverges most", direction = "Positive slope"),
  if (!is.null(genes_POST_biggest_slope_diff_ora_down$result)) 
    genes_POST_biggest_slope_diff_ora_down$result |> 
      mutate(category = "Three-way", pattern = "POST diverges most", direction = "Negative slope"),
  
  # Domain specificity - PM only
  if (!is.null(genes_strain_PM_only_ora_up$result)) 
    genes_strain_PM_only_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "PM only", direction = "Up in CAST"),
  if (!is.null(genes_strain_PM_only_ora_down$result)) 
    genes_strain_PM_only_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "PM only", direction = "Down in CAST"),
  
  # Domain specificity - MAX only
  if (!is.null(genes_strain_MAX_only_ora_up$result)) 
    genes_strain_MAX_only_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "MAX only", direction = "Up in CAST"),
  if (!is.null(genes_strain_MAX_only_ora_down$result)) 
    genes_strain_MAX_only_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "MAX only", direction = "Down in CAST"),
  
  # Domain specificity - POST only
  if (!is.null(genes_strain_POST_only_ora_up$result)) 
    genes_strain_POST_only_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "POST only", direction = "Up in CAST"),
  if (!is.null(genes_strain_POST_only_ora_down$result)) 
    genes_strain_POST_only_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "POST only", direction = "Down in CAST"),
  
  # Domain specificity - All domains
  if (!is.null(genes_strain_all_domains_ora_up$result)) 
    genes_strain_all_domains_ora_up$result |> 
      mutate(category = "Domain Specificity", pattern = "All domains", direction = "Up in CAST"),
  if (!is.null(genes_strain_all_domains_ora_down$result)) 
    genes_strain_all_domains_ora_down$result |> 
      mutate(category = "Domain Specificity", pattern = "All domains", direction = "Down in CAST")
)

```

```{r}
#| label: comprehensive-ora-table
#| message: false
#| warning: false

all_ora_results |>
  filter(p_value < 0.05) |>
  select(category, pattern, direction, source, term_name, p_value, 
         intersection_size, term_size, intersection) |>
  arrange(category, pattern, direction, p_value) |>
  create_dt()

```

## ORA Summary by Category

Summary of enriched terms per analysis category.

```{r}
#| label: comprehensive-ora-summary
#| message: false
#| warning: false

all_ora_results |>
  filter(p_value < 0.05) |>
  group_by(category, pattern, source) |>
  summarize(
    n_terms = n(),
    top_term = term_name[which.min(p_value)],
    min_pval = min(p_value),
    .groups = "drop"
  ) |>
  arrange(category, pattern, min_pval) |>
  create_dt()
```

::::

# Session Info

```{r}
#| label: session-info
#| message: false
#| warning: false

sessionInfo()
```