---
title: "Investigating strain, domain and developmental stage effects"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
      embed-resources: true
      standalone: true
code-fold: true
toc: true
toc-depth: 5
toc-expand: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| message: false
#| warning: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(enrichR)
library(DT)
library(here)

set.seed(123)
select <- dplyr::select
rename <- dplyr::rename
filter <- dplyr::filter
summarize <- dplyr::summarize
count <- dplyr::count

domain_colors <- c(
  PM = rgb(252, 206, 93, maxColorValue = 255),
  MAX = rgb(163, 82, 157, maxColorValue = 255),
  POST = rgb(158, 156, 201, maxColorValue = 255)
)

strain_colors <- c(B6 = "#555555", CAST = "#009E73")
```

```{r}
#| label: helper-functions
#| message: false
#| warning: false

create_dt <- function(x) {
  format_num <- function(val) {
    case_when(
      is.na(val) ~ "NA",
      abs(val) < 0.01 ~ formatC(val, digits = 2, format = "e"),
      TRUE ~ formatC(round(val, 2))
    )
  }
  x |>
    mutate(across(where(is.numeric), format_num)) |>
    DT::datatable(
      extensions = "Buttons",
      rownames = FALSE,
      filter = "top",
      options = list(
        dom = "Blfrtip",
        buttons = c("copy", "csv", "excel"),
        pageLength = 5,
        scrollX = TRUE
      )
    )
}

# Helper function for consistent filtering of DESeq2 results
filter_deseq_results <- function(res,
                                 min_basemean = NULL,
                                 min_basemean_quantile = 0.1,
                                 max_lfcse = 3,
                                 max_abs_lfc = 20,
                                 min_abs_lfc = 0.5,
                                 max_padj = 0.05) {
  # Convert to data frame
  res_df <- as.data.frame(res) |>
    rownames_to_column("gene_id")

  # Set baseMean threshold - use quantile if not specified
  if (is.null(min_basemean)) {
    min_basemean <- quantile(res_df$baseMean, min_basemean_quantile, na.rm = TRUE)
  }

  # Apply filters
  res_filtered <- res_df |>
    filter(
      baseMean > min_basemean,
      lfcSE < max_lfcse,
      abs(log2FoldChange) < max_abs_lfc,
      abs(log2FoldChange) > min_abs_lfc,
      padj < max_padj
    )

  return(res_filtered)
}

# Helper to get passing gene IDs from a filtered result
get_passing_genes <- function(res, ...) {
  filter_deseq_results(res, ...) |> pull(gene_id)
}

get_vst_counts <- function(dds, genes, vsd = NULL) { # renamed for clarity
  if (is.null(vsd)) vsd <- vst(dds, blind = FALSE)
  assay(vsd)[genes, , drop = FALSE] |>
    as.data.frame() |>
    rownames_to_column("gene_id") |>
    pivot_longer(-gene_id, names_to = "sample", values_to = "count") |>
    left_join(colData(dds) |> as.data.frame() |> rownames_to_column("sample"), by = "sample") |>
    left_join(all_genes_palate |> select(gene_id, symbol), by = "gene_id")
}


run_enrichment_hybrid_v2 <- function(gene_ids,
                                     gene_reference = all_genes_palate,
                                     description = "Gene Set",
                                     max_term_size = 1000,
                                     min_term_size = 4,
                                     retries = 3,
                                     retry_wait = 5) {
  databases <- c(
    "GO_Biological_Process_2025",
    "GO_Cellular_Component_2025",
    "GO_Molecular_Function_2025",
    "GWAS_Catalog_2025",
    "Human_Phenotype_Ontology",
    "JASPAR_PWM_Mouse_2025",
    "MGI_Mammalian_Phenotype_Level_4_2021",
    "CORUM",
    "KEGG_2019_Mouse",
    "Tabula_Muris",
    # "GTEx_Tissues_V8_2023", # results too broad, not useful
    # "PerturbAtlas_MouseGenePerturbationSigs", # too big, parser can't handle it!
    "Mouse_Gene_Atlas",
    # "DSigDB",
    "Jensen_DISEASES_Curated_2025",
    "KOMP2_Mouse_Phenotypes_2022",
    "WikiPathways_2024_Mouse"
  )

  # Convert gene IDs to symbols
  query_symbols <- gene_reference |>
    filter(gene_id %in% unique(gene_ids)) |>
    pull(symbol) |>
    unique()

  # Get background symbols
  background_symbols <- gene_reference$symbol

  # Skip if too few genes
  if (length(query_symbols) < 5) {
    message(paste0("Skipping ", description, ": fewer than 5 genes"))
    return(NULL)
  }

  # Query each database individually with tryCatch + retry
  result_enrichr <- list()

  for (db in databases) {
    res <- NULL
    for (attempt in seq_len(retries)) {
      res <- tryCatch(
        enrichr(
          genes = query_symbols,
          databases = db,
          background = background_symbols,
          include_overlap = TRUE
        ),
        error = function(e) {
          message(sprintf("  [%s] Attempt %d failed: %s", db, attempt, e$message))
          NULL
        }
      )
      if (!is.null(res)) break
      if (attempt < retries) {
        message(sprintf("  [%s] Retrying in %ds...", db, retry_wait))
        Sys.sleep(retry_wait)
      }
    }

    if (is.null(res)) {
      message(sprintf("  [%s] Skipped after %d failed attempts.", db, retries))
    } else {
      result_enrichr <- c(result_enrichr, res)
    }
  }

  # Format Enrichr results
  enrichr_df <- NULL
  if (length(result_enrichr) > 0) {
    enrichr_dfs <- lapply(names(result_enrichr), function(db_name) {
      db_result <- result_enrichr[[db_name]]

      if (is.null(db_result) || nrow(db_result) == 0 || all(is.na(db_result$Term))) {
        return(NULL)
      }

      db_result |>
        filter(Adjusted.P.value < 0.05) |>
        mutate(
          gene_set = description,
          source = db_name,
          term_name = Term,
          p_value = Adjusted.P.value,
          query_size = length(query_symbols),
          intersection_size = as.numeric(gsub("/.*", "", Overlap)),
          term_size = as.numeric(gsub(".*/", "", Overlap)),
          intersection = Genes
        ) |>
        filter(term_size < max_term_size, term_size > min_term_size) |>
        select(
          gene_set, source, term_name, p_value,
          query_size, term_size, intersection_size, intersection
        )
    })

    enrichr_dfs <- enrichr_dfs[!sapply(enrichr_dfs, is.null)]
    if (length(enrichr_dfs) > 0) {
      enrichr_df <- bind_rows(enrichr_dfs)
    }
  }

  combined_result <- enrichr_df

  if (!is.null(combined_result) && nrow(combined_result) > 0) {
    return(list(result = combined_result))
  } else {
    return(NULL)
  }
}

# plotting-functions
plot_main_strain_effect <- function(gene_ids, direction, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) {
      return(NULL)
    }
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~AP_domain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = direction, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none", plot.title = element_text(face = "bold"), strip.text = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = strain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_fill_manual(values = strain_colors) +
      labs(x = "", y = "") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")
    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1)
}

plot_main_domain_effect <- function(gene_ids, pattern, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) {
      return(NULL)
    }
    gene_name <- unique(gene_data$symbol)[1]

    # p1: expression by domain, faceted by strain
    # Both strains should show the same domain pattern for a main effect
    p1 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = AP_domain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      facet_wrap(~strain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_fill_manual(values = domain_colors) +
      labs(
        title = gene_name, subtitle = pattern,
        x = "AP Domain", y = "Normalized Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold")
      )

    # p2: domain means connected as a line to highlight the gradient
    domain_means <- gene_data |>
      group_by(AP_domain) |>
      summarise(mean_count = mean(count), .groups = "drop")

    p2 <- ggplot(domain_means, aes(x = AP_domain, y = mean_count, group = 1)) +
      geom_line(linewidth = 1.2, color = "grey40") +
      geom_point(aes(color = AP_domain), size = 4) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(x = "", y = "Mean counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1)
}


plot_main_time_effect <- function(gene_ids, pattern, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) {
      return(NULL)
    }
    gene_name <- unique(gene_data$symbol)[1]

    # p1: trajectory over time, faceted by domain, colored by strain
    # Both strains should track together for a main time effect
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      facet_wrap(~AP_domain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(
        title = gene_name, subtitle = pattern,
        x = "LB stage (centered)", y = "Normalized Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold")
      )

    # p2: early vs late boxplot to summarise the temporal shift
    median_stage <- median(gene_data$LB_stage_from_start)
    p2 <- gene_data |>
      mutate(time_group = factor(
        if_else(LB_stage_from_start <= median_stage, "Early", "Late"),
        levels = c("Early", "Late")
      )) |>
      ggplot(aes(x = time_group, y = count, fill = time_group)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_continuous(labels = scales::comma) +
      labs(x = "", y = "") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1)
}
# ── Helper: build strain effect bar data ──────────────────────────
get_strain_diffs <- function(eff_data) {
  tibble(
    AP_domain = factor(c("PM", "MAX", "POST"), levels = c("PM", "MAX", "POST")),
    effects = c(
      eff_data$strain_effect_PM,
      eff_data$strain_effect_MAX,
      eff_data$strain_effect_POST
    )
  )
}

plot_domain_time_interaction <- function(gene_ids, pattern, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) {
      return(NULL)
    }
    gene_name <- unique(gene_data$symbol)[1]

    # p1: trajectories over time, faceted by domain, both strains overlaid
    # Key story: slopes differ across domains, but strains track together within each domain
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(aes(shape = strain), alpha = 0.4, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 1) +
      facet_wrap(~AP_domain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name, subtitle = pattern,
        x = "LB stage (centered)", y = "Normalized Counts",
        shape = "Strain"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "right",
        plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold")
      )

    # p2: slope comparison across domains (one point per domain per strain,
    # averaged into a per-domain mean ± SE to summarise the slope story)
    # Uses a simple early/late contrast per domain to visually show slope differences
    median_stage <- median(gene_data$LB_stage_from_start)
    slope_summary <- gene_data |>
      mutate(time_group = factor(
        if_else(LB_stage_from_start <= median_stage, "Early", "Late"),
        levels = c("Early", "Late")
      )) |>
      group_by(AP_domain, time_group) |>
      summarise(mean_count = mean(count), .groups = "drop")

    p2 <- ggplot(slope_summary, aes(
      x = time_group, y = mean_count,
      color = AP_domain, group = AP_domain
    )) +
      geom_line(linewidth = 1.2) +
      geom_point(size = 3) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(x = "", y = "Mean counts", color = "Domain") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    p1 + p2 + plot_layout(widths = c(3, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1)
}
# ── FUNCTION 1: Base plot (boxplot + effect bars) ──────────────────
# Used by: PM/MAX/POST strongest, static, opposite directions
plot_domain_strain_base <- function(gene_ids, pattern_desc, count_data,
                                    n = 5, eff_data_full = domain_strain_patterns) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    eff_data <- eff_data_full |> filter(gene_id == gid)
    if (nrow(gene_data) == 0 || nrow(eff_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data$symbol)[1]
    strain_diffs <- get_strain_diffs(eff_data)

    p_box <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain),
        position = position_jitterdodge(jitter.width = 0.15),
        alpha = 0.4, size = 1.5
      ) +
      scale_y_continuous(labels = scales::comma) +
      scale_fill_manual(values = strain_colors) +
      scale_color_manual(values = strain_colors) +
      labs(
        title = gene_name, subtitle = pattern_desc,
        x = "Domain", y = "Normalized Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold")
      )

    p_bar <- ggplot(strain_diffs, aes(x = AP_domain, y = effects, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      scale_fill_manual(values = domain_colors) +
      labs(y = "Strain Effect\n(log2FC CAST vs B6)", x = "") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    p_box + p_bar + plot_layout(widths = c(2.5, 1))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}

# ── FUNCTION 2: Profile plot (mean±SE lines) + effect bars ────────
# Used by: high domain specificity, opposite directions
plot_domain_strain_profile <- function(gene_ids, pattern_desc, count_data,
                                       n = 5, eff_data_full = domain_strain_patterns) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    eff_data <- eff_data_full |> filter(gene_id == gid)
    if (nrow(gene_data) == 0 || nrow(eff_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data$symbol)[1]
    strain_diffs <- get_strain_diffs(eff_data)

    summ <- gene_data |>
      group_by(AP_domain, strain) |>
      summarize(
        mn = mean(count, na.rm = TRUE),
        se = sd(count, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
      )

    p_profile <- ggplot(summ, aes(
      x = AP_domain, y = mn,
      color = strain, group = strain
    )) +
      geom_line(linewidth = 1.2) +
      geom_pointrange(aes(ymin = mn - se, ymax = mn + se), size = 0.8) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(
        title = gene_name, subtitle = pattern_desc,
        x = "Domain", y = "Mean Normalized Counts (±SE)"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold")
      )

    p_bar <- ggplot(strain_diffs, aes(x = AP_domain, y = effects, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      scale_fill_manual(values = domain_colors) +
      labs(y = "Strain Effect\n(log2FC CAST vs B6)", x = "") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    p_profile + p_bar + plot_layout(widths = c(2.5, 1))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}


# ── FUNCTION 3: Boxplot + strain-faceted temporal plot (single row)
# Used by: time-affected pattern
# Left  : boxplot per domain colored by strain
# Right : loess over time, faceted by strain, colored by domain
plot_domain_strain_temporal <- function(gene_ids, pattern_desc, count_data,
                                        n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data$symbol)[1]

    p_box <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain),
        position = position_jitterdodge(jitter.width = 0.15),
        alpha = 0.4, size = 1.5
      ) +
      scale_y_continuous(labels = scales::comma) +
      scale_fill_manual(values = strain_colors) +
      scale_color_manual(values = strain_colors) +
      labs(
        title = gene_name, subtitle = pattern_desc,
        x = "Domain", y = "Normalized Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold")
      )

    p_loess <- ggplot(gene_data, aes(
      x = LB_stage_from_start, y = count,
      color = AP_domain
    )) +
      geom_point(alpha = 0.5, size = 1.5) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 1.2) +
      facet_wrap(~strain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts", color = "Domain") +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        strip.text = element_text(face = "bold")
      )

    p_box + p_loess + plot_layout(widths = c(1, 2.5))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}


# ── FUNCTION 4: Overview lm + strain-faceted domain-colored loess ─
# Used by: uniform, diverging CAST/B6 faster time patterns
# Left : all data together, color = strain, shape = domain, lm smooth
#         — shows the overall temporal divergence between strains
# Right: faceted by strain, color = domain, loess smooth
#         — shows whether all domains move in parallel within each strain
plot_time_strain_overview <- function(gene_ids, pattern_desc, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data$symbol)[1]

    # Left: all domains together, strain as color, domain as shape
    p_overview <- ggplot(
      gene_data,
      aes(x = LB_stage_from_start, y = count, color = strain)
    ) +
      geom_point(aes(shape = AP_domain), alpha = 0.5, size = 2.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1.5) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      scale_shape_manual(values = c(PM = 16, MAX = 17, POST = 15)) +
      labs(
        title = gene_name, subtitle = pattern_desc,
        x = "LB stage (centered)", y = "Normalized Counts",
        shape = "Domain"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold")
      )

    # Right: faceted by strain, domain as color, loess
    p_facet <- ggplot(
      gene_data,
      aes(x = LB_stage_from_start, y = count, color = AP_domain)
    ) +
      geom_point(alpha = 0.5, size = 1.5) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 1.2) +
      facet_wrap(~strain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts", color = "Domain") +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        strip.text = element_text(face = "bold")
      )

    p_overview + p_facet + plot_layout(widths = c(1, 2.5))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}


plot_time_strain_interaction <- function(gene_ids, pattern_desc, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data$symbol)[1]

    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(aes(shape = AP_domain), alpha = 0.5, size = 2.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1.5) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = pattern_desc, x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "right", plot.title = element_text(face = "bold"))

    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
      facet_wrap(~AP_domain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none", strip.text = element_text(face = "bold"))

    p1 + p2 + plot_layout(nrow = 1, widths = c(1, 2.5))
  })
  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1, guides = "collect")
}


plot_domain_specificity <- function(gene_ids, domains_affected, count_data, n = 5) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    if (nrow(gene_data) == 0) {
      return(NULL)
    }
    gene_name <- unique(gene_data$symbol)[1]
    p1 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = strain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_point(aes(color = strain), position = position_jitterdodge(jitter.width = 0.1), alpha = 0.4, size = 1.5) +
      scale_y_continuous(labels = scales::comma) +
      scale_fill_manual(values = strain_colors) +
      scale_color_manual(values = strain_colors) +
      labs(title = gene_name, subtitle = paste("Strain effect in", domains_affected), x = "Domain", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
    p2 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = strain)) +
      geom_point(alpha = 0.4, size = 1.5) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
      facet_wrap(~AP_domain, nrow = 1) +
      scale_y_continuous(labels = scales::comma) +
      scale_color_manual(values = strain_colors) +
      labs(x = "LB stage (centered)", y = "Normalized Counts") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none", strip.text = element_text(face = "bold"))
    p1 / p2 + plot_layout(heights = c(1, 1))
  })
  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  wrap_plots(plots, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
}
```

```{r}
#| label: load-data
#| message: false
#| warning: false

sample_details <- read_csv(here("../samplesheet.csv"))
load(here("../read_mapping/palate_gbrs_emase_counts_v4.RData"))

contam_samples <- c("B6_12.5_PM_1", "B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat_nondo[, !colnames(expr_mat_nondo) %in% contam_samples]
rm(expr_mat_do, expr_mat_nondo)

all_genes_palate <- read_tsv(here("_data", "ENSMUSGid_to_symbol_v105.txt"),
  col_names = c("gene_id", "symbol", "gene_biotype", "chromosome", "start", "end"),
  skip = 1
) |>
  mutate(midpoint = (start + end) / 2) |>
  filter(gene_id %in% rownames(raw_expr_mat))

sample_annot <- sample_details |>
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |>
  separate(Strain_stock, into = c("strain", "tmp"), remove = FALSE, sep = "_") |>
  select(
    sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage,
    palate_closure, L_R_palate
  ) |>
  mutate(strain = if_else(strain == "B6J", "B6", strain)) |>
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |>
  select(-temp) |>
  mutate(Palate_ori = if_else(Palate_ori == "L", "Left", "Right"))
```

# Differential gene expression analysis overview

```{r}
#| label: prep-run-deseq2
#| message: false
#| warning: false

# ── Prep data for DESeq2 ──────────────────────────────────────────────────────
b6_cast_samples <- sample_annot |>
  filter(strain %in% c("B6", "CAST")) |>
  separate(sample, sep = "_", into = c("s", "e", "d", "ind", "rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = if_else(is.na(rep), "1", rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

coldata <- b6_cast_samples |>
  select(sample, unique_sampleid, sex, ind, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(coldata), drop = FALSE])

strain_dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~1)
# Center LB stage
min_LB_stage <- min(colData(strain_dds)$LB_stage)
colData(strain_dds)$LB_stage_from_start <- colData(strain_dds)$LB_stage - min_LB_stage
# Update covariates to factors
colData(strain_dds)$sex <- factor(colData(strain_dds)$sex)
colData(strain_dds)$strain <- factor(colData(strain_dds)$strain, levels = c("B6", "CAST"))
colData(strain_dds)$ind <- factor(colData(strain_dds)$ind)

# Collapse replicates
strain_dds_coll <- collapseReplicates(strain_dds, strain_dds$unique_sampleid)


# ── Fit models ────────────────────────────────────────────────────────────────
# 1. Additive model — no interactions
strain_dds_additive <- strain_dds_coll
design(strain_dds_additive) <- ~ sex + AP_domain + LB_stage_from_start + strain
strain_dds_additive <- DESeq(strain_dds_additive, parallel = TRUE)

# 2. Two-way model — all pairwise interactions, no three-way
strain_dds_twoway <- strain_dds_coll
design(strain_dds_twoway) <- ~ sex + AP_domain + LB_stage_from_start + strain +
  strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start
strain_dds_twoway <- DESeq(strain_dds_twoway, parallel = TRUE)

# 3. Full model — including three-way interaction
strain_dds_full <- strain_dds_coll
design(strain_dds_full) <- ~ sex + AP_domain + LB_stage_from_start +
  strain + strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start +
  strain:AP_domain:LB_stage_from_start
strain_dds_full <- DESeq(strain_dds_full, parallel = TRUE)
coef_names <- resultsNames(strain_dds_full)


# ── LRT tests ─────────────────────────────────────────────────────────────────
# 1a. Any strain effect
strain_lrt <- DESeq(strain_dds_additive,
  test = "LRT",
  reduced = ~ sex + AP_domain + LB_stage_from_start,
  full = ~ sex + AP_domain + LB_stage_from_start + strain,
  parallel = TRUE
)
res_strain_lrt <- results(strain_lrt)
genes_w_strain_eff <- res_strain_lrt |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05)

# 1b. Any domain effect
domain_lrt <- DESeq(strain_dds_additive,
  test = "LRT",
  reduced = ~ sex + LB_stage_from_start + strain,
  full = ~ sex + AP_domain + LB_stage_from_start + strain,
  parallel = TRUE
)
res_domain_lrt <- results(domain_lrt)
genes_w_domain_eff <- res_domain_lrt |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05)

# 1c. Any time effect
time_lrt <- DESeq(strain_dds_additive,
  test = "LRT",
  reduced = ~ sex + AP_domain + strain,
  full = ~ sex + AP_domain + LB_stage_from_start + strain,
  parallel = TRUE
)
res_time_lrt <- results(time_lrt)
genes_w_time_eff <- res_time_lrt |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05)

# 2a. Domain × Strain interaction
strain_domain_int_lrt <- DESeq(strain_dds_twoway,
  test = "LRT",
  reduced = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start,
  full = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start,
  parallel = TRUE
)
res_strain_domain_int_lrt <- results(strain_domain_int_lrt)
genes_w_strain_by_domain_eff <- res_strain_domain_int_lrt |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05)

# 2b. Time × Strain interaction
strain_time_int_lrt <- DESeq(strain_dds_twoway,
  test = "LRT",
  reduced = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:AP_domain + AP_domain:LB_stage_from_start,
  full = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start,
  parallel = TRUE
)
res_strain_time_int_lrt <- results(strain_time_int_lrt)
genes_w_strain_by_time_eff <- res_strain_time_int_lrt |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05)

# 2c. Domain × Time interaction
domain_time_int_lrt <- DESeq(strain_dds_twoway,
  test = "LRT",
  reduced = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:AP_domain + strain:LB_stage_from_start,
  full = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start,
  parallel = TRUE
)
res_domain_time_int_lrt <- results(domain_time_int_lrt)
genes_w_domain_by_time_eff <- res_domain_time_int_lrt |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05)

# 3. Three-way interaction (domain × time × strain)
dds_3way_lrt <- DESeq(strain_dds_full,
  test = "LRT",
  reduced = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start,
  full = ~ sex + AP_domain + LB_stage_from_start +
    strain + strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start +
    strain:AP_domain:LB_stage_from_start,
  parallel = TRUE
)
res_3way_lrt <- results(dds_3way_lrt)
genes_w_3way_eff <- res_3way_lrt |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05)


# ── Gene sets: any interaction involving each variable ────────────────────────
genes_with_any_strain_interaction <- unique(c(
  genes_w_strain_by_time_eff$gene_id,
  genes_w_strain_by_domain_eff$gene_id,
  genes_w_3way_eff$gene_id
))

genes_with_any_domain_interaction <- unique(c(
  genes_w_domain_by_time_eff$gene_id,
  genes_w_strain_by_domain_eff$gene_id,
  genes_w_3way_eff$gene_id
))

genes_with_any_time_interaction <- unique(c(
  genes_w_strain_by_time_eff$gene_id,
  genes_w_domain_by_time_eff$gene_id,
  genes_w_3way_eff$gene_id
))


# ── Extract effect estimates ───────────────────────────────────────────────────
# 1a. Strain main effect (additive model)
res_strain_main <- results(strain_dds_additive, name = "strain_CAST_vs_B6")

# 1b. Domain main effect (additive model) — three pairwise contrasts
res_domain_main_MAX <- results(strain_dds_additive, name = "AP_domain_MAX_vs_PM")
res_domain_main_POST <- results(strain_dds_additive, name = "AP_domain_POST_vs_PM")
res_domain_main_MAX_vs_POST <- results(strain_dds_additive,
  contrast = list(c("AP_domain_MAX_vs_PM", "AP_domain_POST_vs_PM"))
)

# 1c. Main time slope (additive model)
res_time_main <- results(strain_dds_additive, name = "LB_stage_from_start")

# 2a. Domain × Strain — domain-specific strain effects (absolute) and interaction deltas
res_strain_PM <- results(strain_dds_twoway, name = "strain_CAST_vs_B6")
res_strain_MAX <- results(strain_dds_twoway,
  contrast = list(c("strain_CAST_vs_B6", "AP_domainMAX.strainCAST"))
)
res_strain_POST <- results(strain_dds_twoway,
  contrast = list(c("strain_CAST_vs_B6", "AP_domainPOST.strainCAST"))
)
# Interaction coefficients: how CAST strain effect in MAX/POST differs from PM
res_domain_int_MAX <- results(strain_dds_twoway, name = "AP_domainMAX.strainCAST")
res_domain_int_POST <- results(strain_dds_twoway, name = "AP_domainPOST.strainCAST")

# 2b. Time × Strain — average CAST-vs-B6 temporal slope difference (two-way model)
res_time_int_twoway <- results(strain_dds_twoway, name = "LB_stage_from_start.strainCAST")

# 2c. Domain × Time — domain-specific time slopes (absolute) and interaction deltas
# PM slope = LB_stage_from_start alone; MAX/POST = LB_stage_from_start + interaction term
res_time_slope_PM <- results(strain_dds_twoway, name = "LB_stage_from_start")
res_time_slope_MAX <- results(strain_dds_twoway,
  contrast = list(c("LB_stage_from_start", "AP_domainMAX.LB_stage_from_start"))
)
res_time_slope_POST <- results(strain_dds_twoway,
  contrast = list(c("LB_stage_from_start", "AP_domainPOST.LB_stage_from_start"))
)
# Interaction coefficients: how time slope in MAX/POST differs from PM
res_domain_time_int_MAX <- results(strain_dds_twoway, name = "AP_domainMAX.LB_stage_from_start")
res_domain_time_int_POST <- results(strain_dds_twoway, name = "AP_domainPOST.LB_stage_from_start")

# 3. Three-way — interaction deltas and domain-specific absolute CAST-vs-B6 time slopes
res_3way_MAX <- results(strain_dds_full, name = "AP_domainMAX.LB_stage_from_start.strainCAST")
res_3way_POST <- results(strain_dds_full, name = "AP_domainPOST.LB_stage_from_start.strainCAST")
# PM slope = LB_stage_from_start.strainCAST alone; MAX/POST = + three-way term
res_time_strain_slope_PM <- results(strain_dds_full, name = "LB_stage_from_start.strainCAST")
res_time_strain_slope_MAX <- results(strain_dds_full,
  contrast = list(c(
    "LB_stage_from_start.strainCAST",
    "AP_domainMAX.LB_stage_from_start.strainCAST"
  ))
)
res_time_strain_slope_POST <- results(strain_dds_full,
  contrast = list(c(
    "LB_stage_from_start.strainCAST",
    "AP_domainPOST.LB_stage_from_start.strainCAST"
  ))
)


# ── QC filters ────────────────────────────────────────────────────────────────
# Each result is filtered by effect size / uncertainty thresholds appropriate
# to its scale, then restricted to genes passing the relevant LRT.
# Main effects additionally exclude genes with any interaction involving that variable.

# 1a. Strain main effect
res_strain_main_filtered <- filter_deseq_results(res_strain_main) |>
  filter(gene_id %in% genes_w_strain_eff$gene_id)
genes_main_strain_pass <- res_strain_main_filtered$gene_id

# 1b. Domain main effect
res_domain_main_MAX_filtered <- filter_deseq_results(res_domain_main_MAX) |>
  filter(gene_id %in% genes_w_domain_eff$gene_id) |>
  filter(!gene_id %in% genes_with_any_domain_interaction)
res_domain_main_POST_filtered <- filter_deseq_results(res_domain_main_POST) |>
  filter(gene_id %in% genes_w_domain_eff$gene_id) |>
  filter(!gene_id %in% genes_with_any_domain_interaction)
res_domain_main_MAX_vs_POST_filtered <- filter_deseq_results(res_domain_main_MAX_vs_POST) |>
  filter(gene_id %in% genes_w_domain_eff$gene_id) |>
  filter(!gene_id %in% genes_with_any_domain_interaction)

genes_domain_main_MAX_pass <- res_domain_main_MAX_filtered$gene_id
genes_domain_main_POST_pass <- res_domain_main_POST_filtered$gene_id
genes_domain_main_MAX_vs_POST_pass <- res_domain_main_MAX_vs_POST_filtered$gene_id

# 1c. Time main effect (slope scale → tighter thresholds)
res_time_main_filtered <- filter_deseq_results(res_time_main, min_abs_lfc = 0.1, max_lfcse = 0.5) |>
  filter(gene_id %in% genes_w_time_eff$gene_id) |>
  filter(!gene_id %in% genes_with_any_time_interaction)
genes_time_main_pass <- res_time_main_filtered$gene_id

# 2a. Domain × Strain — domain-specific strain effects
res_strain_PM_filtered <- filter_deseq_results(res_strain_PM) |> filter(gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_strain_MAX_filtered <- filter_deseq_results(res_strain_MAX) |> filter(gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_strain_POST_filtered <- filter_deseq_results(res_strain_POST) |> filter(gene_id %in% genes_w_strain_by_domain_eff$gene_id)

genes_strain_PM_pass <- res_strain_PM_filtered$gene_id
genes_strain_MAX_pass <- res_strain_MAX_filtered$gene_id
genes_strain_POST_pass <- res_strain_POST_filtered$gene_id

# 2a. Domain × Strain — interaction deltas (difference-of-differences → lower min_abs_lfc)
res_domain_int_MAX_filtered <- filter_deseq_results(res_domain_int_MAX, min_abs_lfc = 0.3) |> filter(gene_id %in% genes_w_strain_by_domain_eff$gene_id)
res_domain_int_POST_filtered <- filter_deseq_results(res_domain_int_POST, min_abs_lfc = 0.3) |> filter(gene_id %in% genes_w_strain_by_domain_eff$gene_id)

genes_domain_int_MAX_pass <- res_domain_int_MAX_filtered$gene_id
genes_domain_int_POST_pass <- res_domain_int_POST_filtered$gene_id

# 2b. Time × Strain — average temporal slope difference
res_time_int_filtered <- filter_deseq_results(res_time_int_twoway, min_abs_lfc = 0.1, max_lfcse = 0.5) |>
  filter(gene_id %in% genes_w_strain_by_time_eff$gene_id)
genes_time_int_pass <- res_time_int_filtered$gene_id

# 2c. Domain × Time — interaction deltas and domain-specific slopes
res_domain_time_int_MAX_filtered <- filter_deseq_results(res_domain_time_int_MAX, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_domain_by_time_eff$gene_id)
res_domain_time_int_POST_filtered <- filter_deseq_results(res_domain_time_int_POST, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_domain_by_time_eff$gene_id)
res_time_slope_PM_filtered <- filter_deseq_results(res_time_slope_PM, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_domain_by_time_eff$gene_id)
res_time_slope_MAX_filtered <- filter_deseq_results(res_time_slope_MAX, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_domain_by_time_eff$gene_id)
res_time_slope_POST_filtered <- filter_deseq_results(res_time_slope_POST, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_domain_by_time_eff$gene_id)

genes_domain_time_int_MAX_pass <- res_domain_time_int_MAX_filtered$gene_id
genes_domain_time_int_POST_pass <- res_domain_time_int_POST_filtered$gene_id
genes_time_slope_PM_pass <- res_time_slope_PM_filtered$gene_id
genes_time_slope_MAX_pass <- res_time_slope_MAX_filtered$gene_id
genes_time_slope_POST_pass <- res_time_slope_POST_filtered$gene_id

# 3. Three-way — interaction deltas and domain-specific CAST-vs-B6 time slopes
res_3way_MAX_filtered <- filter_deseq_results(res_3way_MAX, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_3way_eff$gene_id)
res_3way_POST_filtered <- filter_deseq_results(res_3way_POST, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_3way_eff$gene_id)
res_time_strain_slope_PM_filtered <- filter_deseq_results(res_time_strain_slope_PM, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_3way_eff$gene_id)
res_time_strain_slope_MAX_filtered <- filter_deseq_results(res_time_strain_slope_MAX, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_3way_eff$gene_id)
res_time_strain_slope_POST_filtered <- filter_deseq_results(res_time_strain_slope_POST, min_abs_lfc = 0.1, max_lfcse = 0.5) |> filter(gene_id %in% genes_w_3way_eff$gene_id)

genes_3way_MAX_pass <- res_3way_MAX_filtered$gene_id
genes_3way_POST_pass <- res_3way_POST_filtered$gene_id
genes_time_strain_slope_PM_pass <- res_time_strain_slope_PM_filtered$gene_id
genes_time_strain_slope_MAX_pass <- res_time_strain_slope_MAX_filtered$gene_id
genes_time_strain_slope_POST_pass <- res_time_strain_slope_POST_filtered$gene_id


# ── Helper ────────────────────────────────────────────────────────────────────
deseq_to_tibble <- function(res, lfc_col, lfcse_col, pval_col, keep_basemean = FALSE) {
  df <- res |>
    as_tibble(rownames = "gene_id") |>
    select(
      gene_id,
      {{ lfc_col }} := log2FoldChange,
      {{ lfcse_col }} := lfcSE,
      {{ pval_col }} := padj
    )
  if (keep_basemean) {
    df <- res |>
      as_tibble(rownames = "gene_id") |>
      select(
        gene_id, baseMean,
        {{ lfc_col }} := log2FoldChange,
        {{ lfcse_col }} := lfcSE,
        {{ pval_col }} := padj
      )
  }
  df
}


# ── Assemble classified tibble ────────────────────────────────────────────────
strain_effects_classified <-
  # Base: gene universe + baseMean from strain LRT
  as_tibble(res_strain_lrt, rownames = "gene_id") |>
  select(gene_id, baseMean, any_strain_pval = padj) |>
  # ── LRT p-values ────────────────────────────────────────────────────────────
  left_join(as_tibble(res_domain_lrt, rownames = "gene_id") |> select(gene_id, domain_pval = padj), by = "gene_id") |>
  left_join(as_tibble(res_time_lrt, rownames = "gene_id") |> select(gene_id, time_pval = padj), by = "gene_id") |>
  left_join(as_tibble(res_strain_domain_int_lrt, rownames = "gene_id") |> select(gene_id, strain_domain_int_pval = padj), by = "gene_id") |>
  left_join(as_tibble(res_strain_time_int_lrt, rownames = "gene_id") |> select(gene_id, strain_time_int_pval = padj), by = "gene_id") |>
  left_join(as_tibble(res_domain_time_int_lrt, rownames = "gene_id") |> select(gene_id, domain_time_int_pval = padj), by = "gene_id") |>
  left_join(as_tibble(res_3way_lrt, rownames = "gene_id") |> select(gene_id, threeway_int_pval = padj), by = "gene_id") |>
  # ── Main effects (additive model) ───────────────────────────────────────────
  left_join(deseq_to_tibble(
    res_strain_main,
    "main_strain_lfc", "main_strain_lfcSE", "main_strain_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_domain_main_MAX,
    "domain_main_MAX_vs_PM_lfc", "domain_main_MAX_vs_PM_lfcSE", "domain_main_MAX_vs_PM_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_domain_main_POST,
    "domain_main_POST_vs_PM_lfc", "domain_main_POST_vs_PM_lfcSE", "domain_main_POST_vs_PM_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_domain_main_MAX_vs_POST,
    "domain_main_MAX_vs_POST_lfc", "domain_main_MAX_vs_POST_lfcSE", "domain_main_MAX_vs_POST_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_time_main,
    "time_main_slope", "time_main_slope_lfcSE", "time_main_slope_pval"
  ), by = "gene_id") |>
  # ── Domain × Time interaction (two-way model) ────────────────────────────────
  # Interaction deltas (MAX/POST slope − PM slope)
  left_join(deseq_to_tibble(
    res_domain_time_int_MAX,
    "domain_time_int_MAX_slope", "domain_time_int_MAX_slope_lfcSE", "domain_time_int_MAX_slope_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_domain_time_int_POST,
    "domain_time_int_POST_slope", "domain_time_int_POST_slope_lfcSE", "domain_time_int_POST_slope_pval"
  ), by = "gene_id") |>
  # Domain-specific absolute time slopes
  left_join(deseq_to_tibble(
    res_time_slope_PM,
    "time_slope_PM", "time_slope_PM_lfcSE", "time_slope_PM_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_time_slope_MAX,
    "time_slope_MAX", "time_slope_MAX_lfcSE", "time_slope_MAX_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_time_slope_POST,
    "time_slope_POST", "time_slope_POST_lfcSE", "time_slope_POST_pval"
  ), by = "gene_id") |>
  # ── Domain × Strain interaction (two-way model) ──────────────────────────────
  # Domain-specific absolute strain effects
  left_join(deseq_to_tibble(
    res_strain_PM,
    "strain_effect_PM", "strain_effect_PM_lfcSE", "strain_effect_PM_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_strain_MAX,
    "strain_effect_MAX", "strain_effect_MAX_lfcSE", "strain_effect_MAX_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_strain_POST,
    "strain_effect_POST", "strain_effect_POST_lfcSE", "strain_effect_POST_pval"
  ), by = "gene_id") |>
  # Interaction deltas (CAST strain effect in MAX/POST − PM)
  left_join(deseq_to_tibble(
    res_domain_int_MAX,
    "domain_int_MAX_lfc", "domain_int_MAX_lfcSE", "domain_int_MAX_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_domain_int_POST,
    "domain_int_POST_lfc", "domain_int_POST_lfcSE", "domain_int_POST_pval"
  ), by = "gene_id") |>
  # ── Time × Strain interaction (two-way model) ────────────────────────────────
  # Average CAST-vs-B6 temporal slope difference across domains
  left_join(deseq_to_tibble(
    res_time_int_twoway,
    "time_int_slope_avg", "time_int_slope_lfcSE", "time_int_slope_pval"
  ), by = "gene_id") |>
  # ── Three-way interaction (full model) ───────────────────────────────────────
  # Interaction deltas (how CAST-vs-B6 time slope in MAX/POST differs from PM)
  left_join(deseq_to_tibble(
    res_3way_MAX,
    "threeway_MAX_slope", "threeway_MAX_slope_lfcSE", "threeway_MAX_slope_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_3way_POST,
    "threeway_POST_slope", "threeway_POST_slope_lfcSE", "threeway_POST_slope_pval"
  ), by = "gene_id") |>
  # Domain-specific absolute CAST-vs-B6 time slopes
  left_join(deseq_to_tibble(
    res_time_strain_slope_PM,
    "time_strain_slope_PM", "time_strain_slope_PM_lfcSE", "time_strain_slope_PM_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_time_strain_slope_MAX,
    "time_strain_slope_MAX", "time_strain_slope_MAX_lfcSE", "time_strain_slope_MAX_pval"
  ), by = "gene_id") |>
  left_join(deseq_to_tibble(
    res_time_strain_slope_POST,
    "time_strain_slope_POST", "time_strain_slope_POST_lfcSE", "time_strain_slope_POST_pval"
  ), by = "gene_id") |>
  # ── Significance flags, pass flags, pattern labels ───────────────────────────
  mutate(
    # LRT significance flags
    any_strain_sig = replace_na(any_strain_pval < 0.05, FALSE),
    domain_sig = replace_na(domain_pval < 0.05, FALSE),
    time_sig = replace_na(time_pval < 0.05, FALSE),
    strain_domain_int_sig = replace_na(strain_domain_int_pval < 0.05, FALSE),
    strain_time_int_sig = replace_na(strain_time_int_pval < 0.05, FALSE),
    domain_time_int_sig = replace_na(domain_time_int_pval < 0.05, FALSE),
    threeway_int_sig = replace_na(threeway_int_pval < 0.05, FALSE),

    # QC pass flags — main effects
    main_strain_pass = gene_id %in% genes_main_strain_pass,
    domain_main_MAX_pass = gene_id %in% genes_domain_main_MAX_pass,
    domain_main_POST_pass = gene_id %in% genes_domain_main_POST_pass,
    domain_main_MAX_vs_POST_pass = gene_id %in% genes_domain_main_MAX_vs_POST_pass,
    time_main_pass = gene_id %in% genes_time_main_pass,

    # QC pass flags — domain × strain interaction
    has_any_strain_interact = gene_id %in% genes_with_any_strain_interaction,
    strain_effect_PM_pass = gene_id %in% genes_strain_PM_pass,
    strain_effect_MAX_pass = gene_id %in% genes_strain_MAX_pass,
    strain_effect_POST_pass = gene_id %in% genes_strain_POST_pass,
    domain_int_MAX_pass = gene_id %in% genes_domain_int_MAX_pass,
    domain_int_POST_pass = gene_id %in% genes_domain_int_POST_pass,

    # QC pass flags — time × strain interaction
    time_int_pass = gene_id %in% genes_time_int_pass,

    # QC pass flags — domain × time interaction
    domain_time_int_MAX_pass = gene_id %in% genes_domain_time_int_MAX_pass,
    domain_time_int_POST_pass = gene_id %in% genes_domain_time_int_POST_pass,
    time_slope_PM_pass = gene_id %in% genes_time_slope_PM_pass,
    time_slope_MAX_pass = gene_id %in% genes_time_slope_MAX_pass,
    time_slope_POST_pass = gene_id %in% genes_time_slope_POST_pass,

    # QC pass flags — three-way interaction
    threeway_MAX_pass = gene_id %in% genes_3way_MAX_pass,
    threeway_POST_pass = gene_id %in% genes_3way_POST_pass,
    time_strain_slope_PM_pass = gene_id %in% genes_time_strain_slope_PM_pass,
    time_strain_slope_MAX_pass = gene_id %in% genes_time_strain_slope_MAX_pass,
    time_strain_slope_POST_pass = gene_id %in% genes_time_strain_slope_POST_pass,

    # ── Top-level filter gates ──────────────────────────────────────────────
    passes_strain_filter = any_strain_sig & main_strain_pass,
    passes_domain_filter = domain_sig &
      (domain_main_MAX_pass | domain_main_POST_pass | domain_main_MAX_vs_POST_pass) &
      !gene_id %in% genes_with_any_domain_interaction,
    passes_time_filter = time_sig &
      time_main_pass &
      !gene_id %in% genes_with_any_time_interaction,
    passes_domain_time_filter = domain_time_int_sig &
      (domain_time_int_MAX_pass | domain_time_int_POST_pass),

    # ── Pattern labels ───────────────────────────────────────────────────────
    main_strain_pattern = case_when(
      !passes_strain_filter ~ "No effect / filtered",
      main_strain_lfc > 1 ~ "CAST >> B6 (strong)",
      main_strain_lfc > 0.5 ~ "CAST > B6 (moderate)",
      main_strain_lfc < -1 ~ "B6 >> CAST (strong)",
      main_strain_lfc < -0.5 ~ "B6 > CAST (moderate)",
      TRUE ~ "Weak effect"
    ),
    domain_main_pattern = case_when(
      !passes_domain_filter ~ "No effect / filtered",
      domain_main_MAX_pass & domain_main_POST_pass ~ "MAX & POST differ from PM",
      domain_main_MAX_pass & domain_main_MAX_vs_POST_pass ~ "MAX differs from PM & POST",
      domain_main_POST_pass & domain_main_MAX_vs_POST_pass ~ "POST differs from PM & MAX",
      domain_main_MAX_pass ~ "MAX differs from PM",
      domain_main_POST_pass ~ "POST differs from PM",
      domain_main_MAX_vs_POST_pass ~ "MAX differs from POST",
      TRUE ~ "Filtered (QC)"
    ),
    time_main_pattern = case_when(
      !passes_time_filter ~ "No time effect / filtered",
      time_main_slope > 0.1 ~ "Expression increases over time",
      time_main_slope < -0.1 ~ "Expression decreases over time",
      TRUE ~ "Flat / near-zero slope"
    ),
    time_strain_pattern = case_when(
      !strain_time_int_sig ~ "No time × strain interaction",
      !time_int_pass ~ "Filtered",
      is.na(time_int_slope_avg) ~ NA_character_,
      time_int_slope_avg > 0.1 ~ "Diverging (CAST changing faster than B6)",
      time_int_slope_avg < -0.1 ~ "Diverging (B6 changing faster than CAST)",
      TRUE ~ "Parallel - both changing similarly"
    ),
    domain_time_pattern = case_when(
      !passes_domain_time_filter ~ "No domain×time interaction / filtered",
      domain_time_int_MAX_pass & domain_time_int_POST_pass ~ "MAX & POST slopes differ from PM",
      domain_time_int_MAX_pass ~ "MAX slope differs from PM",
      domain_time_int_POST_pass ~ "POST slope differs from PM",
      TRUE ~ "Filtered (QC)"
    ),

    # ── Summary pattern type (priority order) ───────────────────────────────
    pattern_type = case_when(
      threeway_int_sig ~ "Three-way interaction",
      strain_domain_int_sig & strain_time_int_sig ~ "Domain + Time × Strain interactions",
      strain_domain_int_sig ~ "Domain × Strain interaction",
      strain_time_int_sig ~ "Time × Strain interaction",
      passes_domain_time_filter ~ "Domain × Time interaction",
      passes_strain_filter ~ "Strain main effect",
      passes_domain_filter ~ "Domain main effect",
      passes_time_filter ~ "Time main effect",
      any_strain_sig | domain_sig | time_sig ~ "Filtered (QC)",
      TRUE ~ "No effect"
    )
  ) |>
  left_join(all_genes_palate |> select(gene_id, symbol, gene_biotype), by = "gene_id")


# ── Filtered dataset ──────────────────────────────────────────────────────────
strain_effects_filtered <- strain_effects_classified |>
  filter(passes_strain_filter)

# # source the pattern identification script
# source("DEgene_pattern_identification_CAST_B6.R")
# ── Load pattern classification results ───────────────────────────────────────
pattern_summary <- read_csv(here("strain_pattern_summary.csv"))
main_patterns <- read_csv(here("main_strain_effect_patterns.csv"))
domain_main_patterns <- read_csv(here("domain_main_effect_patterns.csv"))
time_main_patterns <- read_csv(here("time_main_effect_patterns.csv"))
domain_time_patterns <- read_csv(here("domain_time_interaction_patterns.csv"))
domain_strain_patterns <- read_csv(here("domain_strain_patterns.csv"))
time_strain_patterns <- read_csv(here("time_strain_patterns.csv"))
threeway_patterns <- read_csv(here("threeway_patterns.csv"))
gene_assignments <- read_csv(here("strain_gene_pattern_assignments.csv"))


# ── Strain main effects (additive model — average across contexts) ────────────
genes_CAST_high_strong <- main_patterns |>
  filter(main_strain_pattern == "CAST >> B6 (strong)") |>
  pull(gene_id)
genes_CAST_high_mod <- main_patterns |>
  filter(main_strain_pattern == "CAST > B6 (moderate)") |>
  pull(gene_id)
genes_B6_high_strong <- main_patterns |>
  filter(main_strain_pattern == "B6 >> CAST (strong)") |>
  pull(gene_id)
genes_B6_high_mod <- main_patterns |>
  filter(main_strain_pattern == "B6 > CAST (moderate)") |>
  pull(gene_id)
genes_main_only <- main_patterns |>
  filter(!has_any_strain_interact) |>
  pull(gene_id)


# ── Domain main effects (no domain interactions) ──────────────────────────────
genes_domain_increasing_posterior <- domain_main_patterns |>
  filter(anterior_posterior_gradient == "Increasing posterior (PM lowest)") |>
  pull(gene_id)
genes_domain_decreasing_posterior <- domain_main_patterns |>
  filter(anterior_posterior_gradient == "Decreasing posterior (PM highest)") |>
  pull(gene_id)
genes_domain_MAX_peak <- domain_main_patterns |>
  filter(anterior_posterior_gradient == "MAX peak (non-monotone)") |>
  pull(gene_id)
genes_domain_high_specificity <- domain_main_patterns |>
  filter(domain_specificity == "High") |>
  pull(gene_id)


# ── Time main effects (no time interactions) ──────────────────────────────────
genes_time_increasing <- time_main_patterns |>
  filter(time_main_pattern == "Expression increases over time") |>
  pull(gene_id)
genes_time_decreasing <- time_main_patterns |>
  filter(time_main_pattern == "Expression decreases over time") |>
  pull(gene_id)


# ── Domain × Time interaction patterns ───────────────────────────────────────
genes_domain_time_MAX_diverges <- domain_time_patterns |>
  filter(domain_time_int_MAX_pass) |>
  pull(gene_id)
genes_domain_time_POST_diverges <- domain_time_patterns |>
  filter(domain_time_int_POST_pass) |>
  pull(gene_id)
genes_domain_time_both_diverge <- domain_time_patterns |>
  filter(domain_time_int_MAX_pass & domain_time_int_POST_pass) |>
  pull(gene_id)


# ── Domain × Strain interaction patterns ──────────────────────────────────────
# Which domain shows the strongest strain response
genes_PM_strongest_strain <- domain_strain_patterns |>
  filter(strongest_domain == "PM") |>
  pull(gene_id)
genes_MAX_strongest_strain <- domain_strain_patterns |>
  filter(strongest_domain == "MAX") |>
  pull(gene_id)
genes_POST_strongest_strain <- domain_strain_patterns |>
  filter(strongest_domain == "POST") |>
  pull(gene_id)
genes_opposite_directions <- domain_strain_patterns |>
  filter(direction_pattern == "Opposite directions") |>
  pull(gene_id)
genes_high_domain_specificity <- domain_strain_patterns |>
  filter(specificity_strength == "High") |>
  pull(gene_id)

# Temporal stability of domain × strain patterns
genes_domain_strain_static <- domain_strain_patterns |>
  filter(time_status == "Static") |>
  pull(gene_id)
genes_domain_strain_time_affected <- domain_strain_patterns |>
  filter(time_status == "Time main effect present") |>
  pull(gene_id)


# ── Time × Strain interaction patterns ───────────────────────────────────────
# Trajectory direction
genes_diverging_CAST_faster <- time_strain_patterns |>
  filter(base_slope > 0.1) |>
  pull(gene_id)
genes_diverging_B6_faster <- time_strain_patterns |>
  filter(base_slope < -0.1) |>
  pull(gene_id)

# Spatial uniformity of temporal dynamics
genes_time_uniform <- time_strain_patterns |>
  filter(!strain_domain_int_sig & !threeway_int_sig) |>
  pull(gene_id)
genes_time_domain_affected <- time_strain_patterns |>
  filter(strain_domain_int_sig & !threeway_int_sig) |>
  pull(gene_id)
# Note: domain-specific temporal dynamics are handled in the three-way section below


# ── Three-way interaction patterns (domain-specific temporal strain dynamics) ─
genes_threeway_complex <- threeway_patterns |>
  filter(str_detect(pattern_type, "Complex")) |>
  pull(gene_id)
genes_threeway_initial <- threeway_patterns |>
  filter(str_detect(pattern_type, "Primarily initial")) |>
  pull(gene_id)
genes_threeway_dynamic <- threeway_patterns |>
  filter(str_detect(pattern_type, "Primarily dynamic")) |>
  pull(gene_id)

# ── Pre-compute VST for plotting ──────────────────────────────────────────
vsd_global <- vst(strain_dds_full, blind = FALSE)
```


```{r}
#| label: test-ind-cont
#| eval: false

library(variancePartition)

# 1. Transform counts properly
vsd <- vst(strain_dds_coll, blind = TRUE)

# 2. Run variance partition
form <- ~ (1 | ind) + (1 | AP_domain) + (1 | strain) + LB_stage_from_start

# Use all genes (or sample if too slow)
varPart <- fitExtractVarPartModel(
  exprObj = assay(vsd),
  formula = form,
  data = as.data.frame(colData(strain_dds_coll))
)

# 3. Visualize
plotVarPart(varPart)

# 4. Get specific numbers
vp_summary <- as.data.frame(varPart) %>%
  summarize(
    ind_median = median(ind),
    ind_mean = mean(ind),
    strain_median = median(strain),
    strain_mean = mean(strain),
    domain_median = median(AP_domain),
    domain_mean = mean(AP_domain),
    stage_median = median(LB_stage_from_start),
    stage_mean = mean(LB_stage_from_start)
  )

print(vp_summary)

# 5. Look at genes where individual matters most
high_ind_variance <- as.data.frame(varPart) %>%
  rownames_to_column("gene_id") %>%
  arrange(desc(ind)) %>%
  head(20) %>%
  left_join(all_genes_palate)

cat("\nGenes with highest individual variance:\n")
print(high_ind_variance %>% select(symbol, ind, AP_domain, LB_stage_from_start))

saveRDS(varPart, here("_data/B6_CAST_variancePartition.Rds"))
```


This analysis investigates how gene expression differs between B6 and CAST mouse strains across three palate domains (PM, MAX, POST) and developmental time. I use a statistical model that decomposes expression variation into:

- **Strain main effect**: Average CAST vs B6 difference across all domains and timepoints
- **Domain main effect**: Average expression differences across AP domains, consistent across strains and time
- **Time main effect**: Average change in expression over developmental time, consistent across strains and domains
- **Domain × Time interaction**: Domain-specific temporal trajectories, consistent across strains
- **Domain × Strain interaction**: How the strain effect differs between domains
- **Time × Strain interaction**: How the strain effect changes per unit developmental time
- **Three-way interaction**: Domain-specific changes in temporal strain dynamics

**Sample preparation and model fitting**

I merge technical replicates arising from re-sequencing of several samples using the `collapseReplicates()` function from DESeq2, then calculate the relative or centered limb bud (LB) stage (`LB_stage_from_start`) by subtracting the minimum observed stage so that 0 represents the earliest developmental timepoint in the dataset. The domains are collected from the same individual resulting in a repeated measure setup; however, variance partition analysis showed that individual-level variation accounted for a negligible proportion of transcriptional variance (median < 0.01%, mean 1.5%), therefore I omitted the nesting as an effect in the DESeq2 model (`ind:AP_domain`).

I use three statistical models to extract effects. The **additive model** (`~ sex + AP_domain + LB_stage_from_start + strain`) contains no interaction terms and is used to estimate marginal effects: the average CAST vs B6 log2 fold change, the average expression differences across AP domains, and the average temporal slope — all estimated while holding the other variables constant. The **two-way interaction model** (`~ sex + AP_domain + LB_stage_from_start + strain + strain:AP_domain + strain:LB_stage_from_start + AP_domain:LB_stage_from_start`) includes all pairwise interactions and is used to extract domain-specific strain effects averaged over developmental time, the overall temporal strain slope averaged across domains, domain-specific absolute time slopes, and domain×time interaction coefficients. The **full interaction model** additionally includes the three-way term (`strain:AP_domain:LB_stage_from_start`) and is used to extract three-way interaction coefficients and domain-specific CAST-vs-B6 temporal slopes.

**Identifying significant effects with LRT**

I use likelihood ratio tests (LRT) to identify genes with significant effects. The **overall strain LRT** compares the additive model to a reduced model lacking the strain term, identifying any genes with strain-related variation. The **domain LRT** and **time LRT** similarly use the additive model, removing `AP_domain` or `LB_stage_from_start` respectively to identify genes with significant domain or temporal expression differences. The **domain × strain LRT**, **time × strain LRT**, and **domain × time LRT** each use the two-way model as the full model and remove a single interaction term, testing for each pairwise interaction independently of the others. The **three-way LRT** compares the full model to the two-way model, testing whether temporal strain dynamics differ across domains. Using the two-way model as the basis for all pairwise interaction LRTs ensures the tests are cleanly separated — a gene with only a three-way interaction will not appear in the two-way interaction gene lists.

**Extracting effect sizes**

The LRT tests identify **which** genes have significant effects but do not provide effect sizes. To get the magnitude and direction of effects, I extract coefficients using Wald tests:

- **Strain main effect**: From the additive model — average CAST vs B6 log2 fold change across all domains and timepoints.
- **Domain main effect**: From the additive model — `AP_domain_MAX_vs_PM` and `AP_domain_POST_vs_PM` coefficients give the average expression difference between domains across strains and time. The MAX vs POST contrast is obtained by combining these two coefficients.
- **Main time slope**: From the additive model — `LB_stage_from_start` represents the average change in log2 expression per unit of developmental time, shared across strains and domains.
- **Domain-specific strain effects**: From the two-way model using contrasts. The strain effect in PM is the `strain_CAST_vs_B6` coefficient; MAX and POST combine the strain main effect with their respective `strainCAST:AP_domain` interaction terms. Using the two-way model ensures these estimates represent the strain effect in each domain averaged over developmental time.
- **Domain × strain interaction coefficients**: From the two-way model — `strainCAST:AP_domainMAX` and `strainCAST:AP_domainPOST` indicate how much the strain effect in MAX and POST differs from PM, averaged over time.
- **Overall time × strain slope**: From the two-way model — `strainCAST:LB_stage_from_start` represents the average change in log2FC(CAST/B6) per unit of developmental time, averaged across all domains.
- **Domain × time interaction coefficients**: From the two-way model — `AP_domainMAX:LB_stage_from_start` and `AP_domainPOST:LB_stage_from_start` indicate how the time slope in MAX and POST differs from PM. Domain-specific absolute time slopes are reconstructed by combining the PM slope (`LB_stage_from_start`) with these interaction terms.
- **Three-way coefficients**: From the full model — `strainCAST:AP_domainMAX:LB_stage_from_start` and `strainCAST:AP_domainPOST:LB_stage_from_start` indicate how the temporal strain slope in MAX and POST deviates from the PM slope. Domain-specific absolute CAST-vs-B6 temporal slopes are reconstructed by combining the PM-specific slope (`strainCAST:LB_stage_from_start` from the full model) with these coefficients.

**Quality filtering**

I apply quality filters to each result object individually, removing genes with low expression (baseMean below the 10th percentile), high uncertainty (lfcSE ≥ 1.5 for effect estimates, ≥ 0.5 for slope coefficients), unrealistically large effects (|log2FC| ≥ 10), or very small effects (|log2FC| < 0.5 for main effects, < 0.3 for interaction coefficients, < 0.1 for slope coefficients). Genes passing quality filters for main domain or time effects are additionally required to have no significant interactions involving that variable — a gene with a domain × strain interaction, for example, is excluded from the domain main effect category because the additive model coefficient does not adequately summarise its domain pattern. This per-result filtering ensures that each effect estimate meets quality standards appropriate for its scale and biological interpretation.

**Results summary**

From this analysis, I have identified **`r sum(strain_effects_classified$any_strain_sig, na.rm = TRUE)`** genes with significant strain effects (padj < 0.05), **`r sum(strain_effects_classified$domain_sig, na.rm = TRUE)`** with significant domain effects, and **`r sum(strain_effects_classified$time_sig, na.rm = TRUE)`** with significant temporal effects. Of the strain-affected genes, **`r nrow(strain_effects_filtered)`** pass quality filters on the additive model strain coefficient. These are further classified into interaction categories based on their LRT results.

**Pattern classification**

I use the `DEgene_pattern_identification_CAST_B6.R` script to classify genes into expression pattern categories. The classification is hierarchical: genes are first separated by which effects are significant and pass quality filters, and interaction genes are further sub-classified by the nature of their interaction.

**Domain main effect patterns** are assigned to genes with a significant domain LRT, at least one domain contrast passing quality filters, and no significant domain interactions (domain × strain or domain × time). These genes are classified by the direction of expression differences across the AP axis: monotonically increasing posterior (PM lowest), monotonically decreasing posterior (PM highest), or non-monotone (MAX peak or PM/POST peak). Domain specificity is further quantified by the maximum pairwise log2FC across domain contrasts.

**Time main effect patterns** are assigned to genes with a significant time LRT, a time slope passing quality filters, and no significant time interactions (time × strain or domain × time). These genes are classified by the direction (increasing or decreasing over development) and magnitude (strong, moderate, or weak) of the shared temporal slope.

**Domain × Time interaction patterns** are assigned to genes with a significant domain × time LRT and at least one domain × time interaction coefficient passing quality filters. These genes have temporal trajectories that differ between AP domains, independent of strain. They are classified by which domains show slopes diverging from PM (MAX, POST, or both), the direction of divergence (faster or slower increase than PM), and the overall slope range across domains.

**Domain × Strain interaction patterns** are assigned to genes with a significant domain × strain LRT and at least one domain interaction coefficient passing quality filters. These genes are classified by which domain shows the largest strain effect magnitude (PM, MAX, or POST) and whether the strain effect direction is consistent or variable across domains. Genes with opposite directions — where the strain effect reverses sign between domains — are flagged as a subset of particular interest. These patterns are also characterized by their **temporal stability**: (1) Static patterns show stable domain differences throughout development (no time or three-way interactions); (2) Time-affected patterns may shift as both strains change over time (time × strain interaction present, no three-way interaction); (3) Time-varying patterns actively change their spatial structure during development (three-way interaction significant).

**Time × Strain interaction patterns** are assigned to genes with a significant time × strain LRT and an overall time interaction slope passing quality filters. These are classified as diverging with CAST increasing faster (positive slope > 0.1), diverging with B6 increasing faster (negative slope < -0.1), or parallel/weak, with strong (|slope| > 0.2) and moderate divergence distinguished. These patterns are also characterized by their **spatial uniformity**: (1) Uniform patterns show the same temporal dynamics across all domains (no domain or three-way interactions); (2) Domain-affected patterns have domains that differ at t=0 but show parallel temporal trajectories (domain × strain interaction present, no three-way interaction); (3) Domain-specific patterns show different temporal dynamics in different domains (three-way interaction present). For genes with domain-specific dynamics, per-domain slopes are reconstructed from the full model.

**Three-way interaction patterns** are assigned to genes with a significant three-way LRT and at least one three-way coefficient passing quality filters. Per-domain temporal slopes are reconstructed by combining the PM-specific slope from the full model with the three-way adjustment coefficients for MAX and POST. These genes are classified based on **whether domains differ primarily in their initial state (t=0), their temporal dynamics (slopes), or both**: (1) Complex patterns show substantial differences across both initial effects and temporal trajectories; (2) Primarily initial patterns have domains that differ at t=0 but follow similar temporal trajectories; (3) Primarily dynamic patterns start similarly but diverge over time with domain-specific slopes. Within each pattern type, genes are further characterized by which domain is most distinct when combining initial state and temporal dynamics.


```{r}
#| label: pattern-summary-table
#| message: false
#| warning: false

pattern_summary |>
  create_dt()
```

# Strain Main Effects

These genes show expression differences between CAST and B6. The log2FC values represent the CAST vs B6 effect estimated from the model. 

:::: {#main-strain-tabs .panel-tabset .nav-pills}

## CAST >> B6 (Strong)

Genes with strong upregulation in CAST (log2FC > 1). Here are some examples:

```{r}
#| label: main-CAST-strong-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6


count_data <- get_vst_counts(strain_dds_full, genes_CAST_high_strong, vsd = vsd_global)
plot_main_strain_effect(genes_CAST_high_strong, "CAST >> B6 (strong)", count_data, n = 2)
```

Here is the full list of genes:
```{r}
#| label: main-CAST-strong-table
#| message: false
#| warning: false

main_patterns |>
  filter(gene_id %in% genes_CAST_high_strong) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |>
  create_dt()
```

Here is the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-CAST-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_strong_ora <- run_enrichment_hybrid_v2(genes_CAST_high_strong,
  description = "CAST >> B6 (strong)"
)
```

```{r}
#| label: main-CAST-strong-ora-table
#| message: false
#| warning: false

if (!is.null(genes_CAST_high_strong_ora) && !is.null(genes_CAST_high_strong_ora$result)) {
  genes_CAST_high_strong_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```


## CAST > B6 (Moderate)

Genes with moderate upregulation in CAST (0.5 < log2FC ≤ 1).

```{r}
#| label: main-CAST-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_vst_counts(strain_dds_full,
  genes_CAST_high_mod,
  vsd = vsd_global
)

plot_main_strain_effect(genes_CAST_high_mod, "CAST > B6 (moderate)",
  count_data,
  n = 2
)
```


```{r}
#| label: main-CAST-mod-table
#| message: false
#| warning: false


main_patterns |>
  filter(gene_id %in% genes_CAST_high_mod) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |>
  create_dt()
```


```{r}
#| label: main-CAST-mod-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_CAST_high_mod_ora <- run_enrichment_hybrid_v2(genes_CAST_high_mod,
  description = "CAST > B6 (moderate)"
)
```

```{r}
#| label: main-CAST-mod-ora-table
#| message: false
#| warning: false

if (!is.null(genes_CAST_high_mod_ora) && !is.null(genes_CAST_high_mod_ora$result)) {
  genes_CAST_high_mod_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## B6 >> CAST (Strong)

Genes with strong upregulation in B6 (log2FC < -1).

```{r}
#| label: main-B6-strong-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14


count_data <- get_vst_counts(strain_dds_full,
  genes_B6_high_strong,
  vsd = vsd_global
)
plot_main_strain_effect(genes_B6_high_strong, "B6 >> CAST (strong)", count_data, n = 2)
```

```{r}
#| label: main-B6-strong-table
#| message: false
#| warning: false


main_patterns |>
  filter(gene_id %in% genes_B6_high_strong) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |>
  create_dt()
```


```{r}
#| label: main-B6-strong-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true


genes_B6_high_strong_ora <- run_enrichment_hybrid_v2(genes_B6_high_strong,
  description = "B6 >> CAST (strong)"
)
```

```{r}
#| label: main-B6-strong-ora-table
#| message: false
#| warning: false


if (!is.null(genes_B6_high_strong_ora) && !is.null(genes_B6_high_strong_ora$result)) {
  genes_B6_high_strong_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```


## B6 > CAST (Moderate)

Genes with moderate upregulation in B6 (-1 ≤ log2FC < -0.5).

```{r}
#| label: main-B6-mod-plot
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 14

count_data <- get_vst_counts(strain_dds_full, genes_B6_high_mod, vsd = vsd_global)
plot_main_strain_effect(genes_B6_high_mod, "B6 > CAST (moderate)", count_data, n = 2)
```

```{r}
#| label: main-B6-mod-table
#| message: false
#| warning: false


main_patterns |>
  filter(gene_id %in% genes_B6_high_mod) |>
  select(gene_id, symbol, main_strain_lfc, main_strain_pval, main_strain_pattern) |>
  create_dt()
```


```{r}
#| label: main-B6-mod-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_B6_high_mod_ora <- run_enrichment_hybrid_v2(genes_B6_high_mod,
  description = "B6 > CAST (moderate)"
)
```

```{r}
#| label: main-B6-mod-ora-table
#| message: false
#| warning: false

if (!is.null(genes_B6_high_mod_ora) && !is.null(genes_B6_high_mod_ora$result)) {
  genes_B6_high_mod_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

::::

# Domain Main Effects

These genes show consistent expression differences across the anterior-posterior palate domains (PM, MAX, POST), independent of strain and developmental stage. The log2FC values represent domain contrasts estimated from the additive model.

:::: {#main-domain-tabs .panel-tabset .nav-pills}

## Increasing Posterior (PM lowest)

Genes with expression that increases along the AP axis — lowest in PM and progressively higher in MAX and/or POST.

```{r}
#| label: main-domain-increasing-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6

count_data <- get_vst_counts(strain_dds_full, genes_domain_increasing_posterior, vsd = vsd_global)
plot_main_domain_effect(genes_domain_increasing_posterior, "Increasing posterior (PM lowest)", count_data, n = 2)
```

Here is the full list of genes:
```{r}
#| label: main-domain-increasing-table
#| message: false
#| warning: false

domain_main_patterns |>
  filter(gene_id %in% genes_domain_increasing_posterior) |>
  select(
    gene_id, symbol, domain_main_MAX_vs_PM_lfc, domain_main_POST_vs_PM_lfc,
    domain_main_MAX_vs_POST_lfc, domain_main_pattern
  ) |>
  create_dt()
```

Here are the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-domain-increasing-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_domain_increasing_posterior_ora <- run_enrichment_hybrid_v2(
  genes_domain_increasing_posterior,
  description = "Increasing posterior (PM lowest)"
)
```

```{r}
#| label: main-domain-increasing-ora-table
#| message: false
#| warning: false

if (!is.null(genes_domain_increasing_posterior_ora) && !is.null(genes_domain_increasing_posterior_ora$result)) {
  genes_domain_increasing_posterior_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## Decreasing Posterior (PM highest)

Genes with expression that decreases along the AP axis — highest in PM and progressively lower in MAX and/or POST.

```{r}
#| label: main-domain-decreasing-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6

count_data <- get_vst_counts(strain_dds_full, genes_domain_decreasing_posterior, vsd = vsd_global)
plot_main_domain_effect(genes_domain_decreasing_posterior, "Decreasing posterior (PM highest)", count_data, n = 2)
```

Here is the full list of genes:
```{r}
#| label: main-domain-decreasing-table
#| message: false
#| warning: false

domain_main_patterns |>
  filter(gene_id %in% genes_domain_decreasing_posterior) |>
  select(
    gene_id, symbol, domain_main_MAX_vs_PM_lfc, domain_main_POST_vs_PM_lfc,
    domain_main_MAX_vs_POST_lfc, domain_main_pattern
  ) |>
  create_dt()
```

Here are the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-domain-decreasing-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_domain_decreasing_posterior_ora <- run_enrichment_hybrid_v2(
  genes_domain_decreasing_posterior,
  description = "Decreasing posterior (PM highest)"
)
```

```{r}
#| label: main-domain-decreasing-ora-table
#| message: false
#| warning: false

if (!is.null(genes_domain_decreasing_posterior_ora) && !is.null(genes_domain_decreasing_posterior_ora$result)) {
  genes_domain_decreasing_posterior_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## MAX Peak (Non-monotone)

Genes with highest expression in the MAX domain — expression is lower in both PM and POST, suggesting a mid-axis peak.

```{r}
#| label: main-domain-MAX-peak-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6

count_data <- get_vst_counts(strain_dds_full, genes_domain_MAX_peak, vsd = vsd_global)
plot_main_domain_effect(genes_domain_MAX_peak, "MAX peak (non-monotone)", count_data, n = 2)
```

Here is the full list of genes:
```{r}
#| label: main-domain-MAX-peak-table
#| message: false
#| warning: false

domain_main_patterns |>
  filter(gene_id %in% genes_domain_MAX_peak) |>
  select(
    gene_id, symbol, domain_main_MAX_vs_PM_lfc, domain_main_POST_vs_PM_lfc,
    domain_main_MAX_vs_POST_lfc, domain_main_pattern
  ) |>
  create_dt()
```

Here are the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-domain-MAX-peak-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_domain_MAX_peak_ora <- run_enrichment_hybrid_v2(
  genes_domain_MAX_peak,
  description = "MAX peak (non-monotone)"
)
```

```{r}
#| label: main-domain-MAX-peak-ora-table
#| message: false
#| warning: false

if (!is.null(genes_domain_MAX_peak_ora) && !is.null(genes_domain_MAX_peak_ora$result)) {
  genes_domain_MAX_peak_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## High Domain Specificity

Genes with large expression differences across domains (max pairwise log2FC > 1), regardless of gradient direction.

```{r}
#| label: main-domain-high-spec-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6

count_data <- get_vst_counts(strain_dds_full, genes_domain_high_specificity, vsd = vsd_global)
plot_main_domain_effect(genes_domain_high_specificity, "High domain specificity", count_data, n = 2)
```

Here is the full list of genes:
```{r}
#| label: main-domain-high-spec-table
#| message: false
#| warning: false

domain_main_patterns |>
  filter(gene_id %in% genes_domain_high_specificity) |>
  select(
    gene_id, symbol, domain_main_MAX_vs_PM_lfc, domain_main_POST_vs_PM_lfc,
    domain_main_MAX_vs_POST_lfc, domain_lfc_range, domain_specificity, domain_main_pattern
  ) |>
  create_dt()
```

Here are the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-domain-high-spec-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_domain_high_specificity_ora <- run_enrichment_hybrid_v2(
  genes_domain_high_specificity,
  description = "High domain specificity"
)
```

```{r}
#| label: main-domain-high-spec-ora-table
#| message: false
#| warning: false

if (!is.null(genes_domain_high_specificity_ora) && !is.null(genes_domain_high_specificity_ora$result)) {
  genes_domain_high_specificity_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

::::

# Stage Main Effects

These genes show consistent changes in expression over developmental time (LB stage), shared across both strains and all AP domains. The slope values represent the change in log2 expression per unit of LB stage, estimated from the additive model.

:::: {#main-time-tabs .panel-tabset .nav-pills}

## Increasing Over Time

Genes whose expression increases over development (slope > 0.1).

```{r}
#| label: main-time-increasing-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6

count_data <- get_vst_counts(strain_dds_full, genes_time_increasing, vsd = vsd_global)
plot_main_time_effect(genes_time_increasing, "Expression increases over time", count_data, n = 2)
```

Here is the full list of genes:
```{r}
#| label: main-time-increasing-table
#| message: false
#| warning: false

time_main_patterns |>
  filter(gene_id %in% genes_time_increasing) |>
  select(
    gene_id, symbol, time_main_slope, time_main_slope_lfcSE,
    time_main_slope_pval, slope_magnitude, time_main_pattern
  ) |>
  create_dt()
```

Here are the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-time-increasing-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_time_increasing_ora <- run_enrichment_hybrid_v2(
  genes_time_increasing,
  description = "Expression increases over time"
)
```

```{r}
#| label: main-time-increasing-ora-table
#| message: false
#| warning: false

if (!is.null(genes_time_increasing_ora) && !is.null(genes_time_increasing_ora$result)) {
  genes_time_increasing_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## Decreasing Over Time

Genes whose expression decreases over development (slope < -0.1).

```{r}
#| label: main-time-decreasing-plot
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6

count_data <- get_vst_counts(strain_dds_full, genes_time_decreasing, vsd = vsd_global)
plot_main_time_effect(genes_time_decreasing, "Expression decreases over time", count_data, n = 2)
```

Here is the full list of genes:
```{r}
#| label: main-time-decreasing-table
#| message: false
#| warning: false

time_main_patterns |>
  filter(gene_id %in% genes_time_decreasing) |>
  select(
    gene_id, symbol, time_main_slope, time_main_slope_lfcSE,
    time_main_slope_pval, slope_magnitude, time_main_pattern
  ) |>
  create_dt()
```

Here are the over-represented biological processes, pathways and phenotypes:
```{r}
#| label: main-time-decreasing-ora
#| message: false
#| warning: false
#| results: hide
#| cache: true

genes_time_decreasing_ora <- run_enrichment_hybrid_v2(
  genes_time_decreasing,
  description = "Expression decreases over time"
)
```

```{r}
#| label: main-time-decreasing-ora-table
#| message: false
#| warning: false

if (!is.null(genes_time_decreasing_ora) && !is.null(genes_time_decreasing_ora$result)) {
  genes_time_decreasing_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

::::


# Domain × Strain Interactions

These genes show strain differences that vary across palate domains. The analysis focuses on which domain shows the strongest strain response and whether strain effects are consistent in direction across domains.

:::: {#domain-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-strain-overview
#| message: false
#| warning: false

domain_strain_patterns |>
  count(domain_pattern) |>
  arrange(desc(n)) |>
  rename(Pattern = domain_pattern, `N Genes` = n) |>
  create_dt()
```

## PM: Strongest Effect

Genes where the strain effect is strongest in the PM domain. 

```{r}
#| label: domain-PM-strongest-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_PM_strongest_strain, vsd = vsd_global)
plot_domain_strain_base(genes_PM_strongest_strain, "Strongest strain effect in PM", count_data, n = 2)
```

```{r}
#| label: domain-PM-strongest-table
#| message: false
#| warning: false

domain_strain_patterns |>
  filter(gene_id %in% genes_PM_strongest_strain) |>
  select(
    gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    domain_strain_range, specificity_strength, direction_pattern, time_status
  ) |>
  create_dt()
```

```{r}
#| label: domain-PM-strongest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in PM
genes_PM_strongest_up <- domain_strain_patterns |>
  filter(gene_id %in% genes_PM_strongest_strain, strain_effect_PM > 0) |>
  pull(gene_id)

genes_PM_strongest_down <- domain_strain_patterns |>
  filter(gene_id %in% genes_PM_strongest_strain, strain_effect_PM < 0) |>
  pull(gene_id)

genes_PM_strongest_effect_ora_up <- run_enrichment_hybrid_v2(
  genes_PM_strongest_up,
  description = "PM strongest effect - Up in CAST"
)

genes_PM_strongest_effect_ora_down <- run_enrichment_hybrid_v2(
  genes_PM_strongest_down,
  description = "PM strongest effect - Down in CAST"
)
```

```{r}
#| label: domain-PM-strongest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_PM_strongest_effect_ora_up$result)) genes_PM_strongest_effect_ora_up$result,
  if (!is.null(genes_PM_strongest_effect_ora_down$result)) genes_PM_strongest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

## MAX: Strongest Effect

Genes where the strain effect is strongest in the MAX domain. 

```{r}
#| label: domain-MAX-strongest-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_MAX_strongest_strain, vsd = vsd_global)
plot_domain_strain_base(genes_MAX_strongest_strain, "Strongest strain effect in MAX", count_data, n = 2)
```

```{r}
#| label: domain-MAX-strongest-table
#| message: false
#| warning: false

domain_strain_patterns |>
  filter(gene_id %in% genes_MAX_strongest_strain) |>
  select(
    gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    domain_strain_range, specificity_strength, direction_pattern, time_status
  ) |>
  create_dt()
```

```{r}
#| label: domain-MAX-strongest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in MAX
genes_MAX_strongest_up <- domain_strain_patterns |>
  filter(gene_id %in% genes_MAX_strongest_strain, strain_effect_MAX > 0) |>
  pull(gene_id)

genes_MAX_strongest_down <- domain_strain_patterns |>
  filter(gene_id %in% genes_MAX_strongest_strain, strain_effect_MAX < 0) |>
  pull(gene_id)

genes_MAX_strongest_effect_ora_up <- run_enrichment_hybrid_v2(
  genes_MAX_strongest_up,
  description = "MAX strongest effect - Up in CAST"
)

genes_MAX_strongest_effect_ora_down <- run_enrichment_hybrid_v2(
  genes_MAX_strongest_down,
  description = "MAX strongest effect - Down in CAST"
)
```

```{r}
#| label: domain-MAX-strongest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_MAX_strongest_effect_ora_up$result)) genes_MAX_strongest_effect_ora_up$result,
  if (!is.null(genes_MAX_strongest_effect_ora_down$result)) genes_MAX_strongest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

## POST: Strongest Effect

Genes where the strain effect is strongest in the POST domain. 

```{r}
#| label: domain-POST-strongest-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_POST_strongest_strain, vsd = vsd_global)
plot_domain_strain_base(genes_POST_strongest_strain, "Strongest strain effect in POST", count_data, n = 2)
```

```{r}
#| label: domain-POST-strongest-table
#| message: false
#| warning: false

domain_strain_patterns |>
  filter(gene_id %in% genes_POST_strongest_strain) |>
  select(
    gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    domain_strain_range, specificity_strength, direction_pattern, time_status
  ) |>
  create_dt()
```

```{r}
#| label: domain-POST-strongest-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction based on strain effect in POST
genes_POST_strongest_up <- domain_strain_patterns |>
  filter(gene_id %in% genes_POST_strongest_strain, strain_effect_POST > 0) |>
  pull(gene_id)

genes_POST_strongest_down <- domain_strain_patterns |>
  filter(gene_id %in% genes_POST_strongest_strain, strain_effect_POST < 0) |>
  pull(gene_id)

genes_POST_strongest_effect_ora_up <- run_enrichment_hybrid_v2(
  genes_POST_strongest_up,
  description = "POST strongest effect - Up in CAST"
)

genes_POST_strongest_effect_ora_down <- run_enrichment_hybrid_v2(
  genes_POST_strongest_down,
  description = "POST strongest effect - Down in CAST"
)
```

```{r}
#| label: domain-POST-strongest-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_POST_strongest_effect_ora_up$result)) genes_POST_strongest_effect_ora_up$result,
  if (!is.null(genes_POST_strongest_effect_ora_down$result)) genes_POST_strongest_effect_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

## High Domain Specificity

Genes with particularly strong domain-specific strain effects (range > 1.0 log2FC). These show dramatic differences in how strongly strain effects manifest across domains.

```{r}
#| label: domain-high-specificity-plot
#| message: false
#| warning: false
#| fig-height: 15
#| fig-width: 14

count_data <- get_vst_counts(strain_dds_full, genes_high_domain_specificity, vsd = vsd_global)
plot_domain_specificity(genes_high_domain_specificity, "High domain specificity", count_data, n = 2)
```

```{r}
#| label: domain-high-specificity-table
#| message: false
#| warning: false

domain_strain_patterns |>
  filter(gene_id %in% genes_high_domain_specificity) |>
  select(
    gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    domain_strain_range, strongest_domain, weakest_domain, direction_pattern, time_status
  ) |>
  arrange(desc(domain_strain_range)) |>
  create_dt()
```

```{r}
#| label: domain-high-specificity-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_high_domain_specificity_ora <- run_enrichment_hybrid_v2(
  genes_high_domain_specificity,
  description = "High domain specificity"
)
```

```{r}
#| label: domain-high-specificity-ora-table
#| message: false
#| warning: false

if (!is.null(genes_high_domain_specificity_ora$result)) {
  genes_high_domain_specificity_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

## Opposite Directions

Genes where the strain effect reverses direction across domains — for example, higher in CAST in one domain but higher in B6 in another. These are particularly interesting as they suggest domain-specific regulatory divergence between strains.

```{r}
#| label: domain-opposite-plot
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 10

count_data <- get_vst_counts(strain_dds_full, genes_opposite_directions, vsd = vsd_global)
plot_domain_strain_profile(genes_opposite_directions, "Opposite directions across domains", count_data, n = 2)
```

```{r}
#| label: domain-opposite-table
#| message: false
#| warning: false

domain_strain_patterns |>
  filter(gene_id %in% genes_opposite_directions) |>
  select(
    gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    domain_strain_range, strongest_domain, direction_pattern, time_status
  ) |>
  create_dt()
```

```{r}
#| label: domain-opposite-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction using PM as reference
genes_opposite_up_in_PM <- domain_strain_patterns |>
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM > 0) |>
  pull(gene_id)

genes_opposite_down_in_PM <- domain_strain_patterns |>
  filter(gene_id %in% genes_opposite_directions, strain_effect_PM < 0) |>
  pull(gene_id)

genes_opposite_directions_ora_up <- run_enrichment_hybrid_v2(
  genes_opposite_up_in_PM,
  description = "Opposite directions - Up in CAST in PM"
)

genes_opposite_directions_ora_down <- run_enrichment_hybrid_v2(
  genes_opposite_down_in_PM,
  description = "Opposite directions - Down in CAST in PM"
)
```

```{r}
#| label: domain-opposite-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_opposite_directions_ora_up$result)) genes_opposite_directions_ora_up$result,
  if (!is.null(genes_opposite_directions_ora_down$result)) genes_opposite_directions_ora_down$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

## Static Pattern

Domain differences that remain stable over developmental time (no time or three-way interaction). These represent fixed spatial differences in strain effects.

```{r}
#| label: domain-static-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_domain_strain_static, vsd = vsd_global)
plot_domain_strain_temporal(genes_domain_strain_static, "Static domain pattern", count_data, n = 2)
```

```{r}
#| label: domain-static-table
#| message: false
#| warning: false

domain_strain_patterns |>
  filter(gene_id %in% genes_domain_strain_static) |>
  select(
    gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    strongest_domain, direction_pattern, specificity_strength, time_status
  ) |>
  create_dt()
```

## Time-Affected Pattern

Domain differences where time main effects are present, so the spatial pattern may shift over development, but without domain-specific temporal dynamics (no three-way interaction).

```{r}
#| label: domain-time-affected-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_domain_strain_time_affected, vsd = vsd_global)
plot_domain_strain_temporal(genes_domain_strain_time_affected, "Time-affected domain pattern", count_data, n = 2)
```

```{r}
#| label: domain-time-affected-table
#| message: false
#| warning: false

domain_strain_patterns |>
  filter(gene_id %in% genes_domain_strain_time_affected) |>
  select(
    gene_id, symbol, strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    strongest_domain, direction_pattern, specificity_strength, time_status
  ) |>
  create_dt()
```

::::

# Time × Strain Interactions

These genes show strain differences that change over developmental time. The temporal slope coefficient represents the change in log2FC(CAST/B6) per unit of developmental time.

:::: {#time-strain-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: time-strain-overview
#| message: false
#| warning: false

time_strain_patterns |>
  count(detailed_pattern) |>
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |>
  create_dt()
```

## Diverging: CAST Faster

Genes where strains diverge over developmental time, with CAST expression increasing (or decreasing less) relative to B6 as development progresses. The positive time interaction slope indicates that the log2FC(CAST/B6) becomes more positive over time.

```{r}
#| label: time-CAST-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_diverging_CAST_faster, vsd = vsd_global)
plot_time_strain_interaction(genes_diverging_CAST_faster, "Diverging: CAST changing faster", count_data, n = 2)
```

```{r}
#| label: time-CAST-faster-table
#| message: false
#| warning: false

time_strain_patterns |>
  filter(gene_id %in% genes_diverging_CAST_faster) |>
  select(
    gene_id, symbol, base_slope, strain_time_int_pval, main_strain_lfc,
    trajectory_pattern, spatial_pattern, threeway_int_sig
  ) |>
  create_dt()
```

```{r}
#| label: time-CAST-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_CAST_faster_up <- time_strain_patterns |>
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc > 0) |>
  pull(gene_id)

genes_CAST_faster_down <- time_strain_patterns |>
  filter(gene_id %in% genes_diverging_CAST_faster, main_strain_lfc < 0) |>
  pull(gene_id)

genes_diverging_CAST_faster_ora_up <- run_enrichment_hybrid_v2(
  genes_CAST_faster_up,
  description = "CAST faster - Up in CAST at baseline"
)

genes_diverging_CAST_faster_ora_down <- run_enrichment_hybrid_v2(
  genes_CAST_faster_down,
  description = "CAST faster - Down in CAST at baseline"
)
```

```{r}
#| label: time-CAST-faster-ora-table
#| message: false
#| warning: false

combined_diverging_cast_ora_results <- bind_rows(
  genes_diverging_CAST_faster_ora_up$result,
  genes_diverging_CAST_faster_ora_down$result
)

if (!is.null(combined_diverging_cast_ora_results) && nrow(combined_diverging_cast_ora_results) > 0) {
  combined_diverging_cast_ora_results |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

## Diverging: B6 Faster

Genes where strains diverge over developmental time, with B6 expression increasing (or decreasing less) relative to CAST as development progresses. The negative time interaction slope indicates that the log2FC(CAST/B6) becomes more negative over time.

```{r}
#| label: time-B6-faster-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_diverging_B6_faster, vsd = vsd_global)
plot_time_strain_interaction(genes_diverging_B6_faster, "Diverging: B6 changing faster", count_data, n = 2)
```

```{r}
#| label: time-B6-faster-table
#| message: false
#| warning: false

time_strain_patterns |>
  filter(gene_id %in% genes_diverging_B6_faster) |>
  select(
    gene_id, symbol, base_slope, strain_time_int_pval, main_strain_lfc,
    trajectory_pattern, spatial_pattern, threeway_int_sig
  ) |>
  create_dt()
```

```{r}
#| label: time-B6-faster-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by baseline strain effect direction
genes_B6_faster_up <- time_strain_patterns |>
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc > 0) |>
  pull(gene_id)

genes_B6_faster_down <- time_strain_patterns |>
  filter(gene_id %in% genes_diverging_B6_faster, main_strain_lfc < 0) |>
  pull(gene_id)

genes_diverging_B6_faster_ora_up <- run_enrichment_hybrid_v2(
  genes_B6_faster_up,
  description = "B6 faster - Up in CAST at baseline"
)

genes_diverging_B6_faster_ora_down <- run_enrichment_hybrid_v2(
  genes_B6_faster_down,
  description = "B6 faster - Down in CAST at baseline"
)
```

```{r}
#| label: time-B6-faster-ora-table
#| message: false
#| warning: false

combined_diverging_b6_ora_results <- bind_rows(
  genes_diverging_B6_faster_ora_up$result,
  genes_diverging_B6_faster_ora_down$result
)

if (!is.null(combined_diverging_b6_ora_results) && nrow(combined_diverging_b6_ora_results) > 0) {
  combined_diverging_b6_ora_results |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

## Uniform Across Domains

Temporal strain dynamics that are consistent across all palate domains (no domain or three-way interaction). These represent spatially uniform temporal changes.

```{r}
#| label: time-uniform-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_time_uniform, vsd = vsd_global)
plot_time_strain_interaction(genes_time_uniform, "Uniform temporal dynamics", count_data, n = 2)
```

```{r}
#| label: time-uniform-table
#| message: false
#| warning: false

time_strain_patterns |>
  filter(gene_id %in% genes_time_uniform) |>
  select(gene_id, symbol, base_slope, main_strain_lfc, trajectory_pattern, spatial_pattern, threeway_int_sig) |>
  create_dt()
```

## Domain-Affected

Temporal dynamics where domains differ at t=0 but show parallel trajectories over time (domain interaction present, but no three-way interaction).

```{r}
#| label: time-domain-affected-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_time_domain_affected, vsd = vsd_global)
plot_domain_strain_temporal(genes_time_domain_affected, "Domain-affected temporal dynamics", count_data, n = 2)
```

```{r}
#| label: time-domain-affected-table
#| message: false
#| warning: false

time_strain_patterns |>
  filter(gene_id %in% genes_time_domain_affected) |>
  select(gene_id, symbol, base_slope, main_strain_lfc, trajectory_pattern, spatial_pattern, threeway_int_sig) |>
  create_dt()
```

::::


# Domain × Time Interactions

These genes show temporal trajectories that differ across palate domains, independent of strain. The interaction slope coefficients represent the difference in the per-unit LB stage change between a given domain and the PM reference domain.

:::: {#domain-time-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-time-overview
#| message: false
#| warning: false

domain_time_patterns |>
  count(domain_time_pattern) |>
  arrange(desc(n)) |>
  rename(Pattern = domain_time_pattern, `N Genes` = n) |>
  create_dt()
```

## MAX Slope Differs from PM

Genes where the temporal trajectory in the MAX domain is significantly different from PM (interaction coefficient passes QC), regardless of what POST shows.

```{r}
#| label: domain-time-MAX-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_domain_time_MAX_diverges, vsd = vsd_global)
plot_domain_time_interaction(genes_domain_time_MAX_diverges, "MAX slope differs from PM", count_data, n = 2)
```

```{r}
#| label: domain-time-MAX-table
#| message: false
#| warning: false

domain_time_patterns |>
  filter(gene_id %in% genes_domain_time_MAX_diverges) |>
  select(
    gene_id, symbol,
    time_slope_PM, time_slope_MAX, time_slope_POST,
    domain_time_int_MAX_slope, domain_time_int_POST_slope,
    MAX_dynamics, POST_dynamics,
    slope_range, slope_range_magnitude
  ) |>
  arrange(desc(abs(domain_time_int_MAX_slope))) |>
  create_dt()
```

```{r}
#| label: domain-time-MAX-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by direction: is MAX increasing faster or slower than PM?
genes_MAX_faster_than_PM <- domain_time_patterns |>
  filter(
    gene_id %in% genes_domain_time_MAX_diverges,
    domain_time_int_MAX_slope > 0
  ) |>
  pull(gene_id)

genes_MAX_slower_than_PM <- domain_time_patterns |>
  filter(
    gene_id %in% genes_domain_time_MAX_diverges,
    domain_time_int_MAX_slope < 0
  ) |>
  pull(gene_id)

genes_domain_time_MAX_ora_faster <- run_enrichment_hybrid_v2(
  genes_MAX_faster_than_PM,
  description = "MAX slope differs from PM - MAX faster"
)

genes_domain_time_MAX_ora_slower <- run_enrichment_hybrid_v2(
  genes_MAX_slower_than_PM,
  description = "MAX slope differs from PM - MAX slower"
)
```

```{r}
#| label: domain-time-MAX-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_domain_time_MAX_ora_faster$result)) genes_domain_time_MAX_ora_faster$result,
  if (!is.null(genes_domain_time_MAX_ora_slower$result)) genes_domain_time_MAX_ora_slower$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## POST Slope Differs from PM

Genes where the temporal trajectory in the POST domain is significantly different from PM, regardless of what MAX shows.

```{r}
#| label: domain-time-POST-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_domain_time_POST_diverges, vsd = vsd_global)
plot_domain_time_interaction(genes_domain_time_POST_diverges, "POST slope differs from PM", count_data, n = 2)
```

```{r}
#| label: domain-time-POST-table
#| message: false
#| warning: false

domain_time_patterns |>
  filter(gene_id %in% genes_domain_time_POST_diverges) |>
  select(
    gene_id, symbol,
    time_slope_PM, time_slope_MAX, time_slope_POST,
    domain_time_int_MAX_slope, domain_time_int_POST_slope,
    MAX_dynamics, POST_dynamics,
    slope_range, slope_range_magnitude
  ) |>
  arrange(desc(abs(domain_time_int_POST_slope))) |>
  create_dt()
```

```{r}
#| label: domain-time-POST-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_POST_faster_than_PM <- domain_time_patterns |>
  filter(
    gene_id %in% genes_domain_time_POST_diverges,
    domain_time_int_POST_slope > 0
  ) |>
  pull(gene_id)

genes_POST_slower_than_PM <- domain_time_patterns |>
  filter(
    gene_id %in% genes_domain_time_POST_diverges,
    domain_time_int_POST_slope < 0
  ) |>
  pull(gene_id)

genes_domain_time_POST_ora_faster <- run_enrichment_hybrid_v2(
  genes_POST_faster_than_PM,
  description = "POST slope differs from PM - POST faster"
)

genes_domain_time_POST_ora_slower <- run_enrichment_hybrid_v2(
  genes_POST_slower_than_PM,
  description = "POST slope differs from PM - POST slower"
)
```

```{r}
#| label: domain-time-POST-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_domain_time_POST_ora_faster$result)) genes_domain_time_POST_ora_faster$result,
  if (!is.null(genes_domain_time_POST_ora_slower$result)) genes_domain_time_POST_ora_slower$result
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## Both MAX & POST Differ from PM

Genes where both MAX and POST show temporal trajectories that significantly diverge from PM. These represent the most spatially structured temporal dynamics — the entire AP axis shows domain-specific developmental progression.

```{r}
#| label: domain-time-both-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_domain_time_both_diverge, vsd = vsd_global)
plot_domain_time_interaction(genes_domain_time_both_diverge, "MAX & POST slopes differ from PM", count_data, n = 2)
```

```{r}
#| label: domain-time-both-table
#| message: false
#| warning: false

domain_time_patterns |>
  filter(gene_id %in% genes_domain_time_both_diverge) |>
  select(
    gene_id, symbol,
    time_slope_PM, time_slope_MAX, time_slope_POST,
    domain_time_int_MAX_slope, domain_time_int_POST_slope,
    MAX_dynamics, POST_dynamics,
    slope_range, slope_range_magnitude
  ) |>
  arrange(desc(slope_range)) |>
  create_dt()
```

```{r}
#| label: domain-time-both-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_domain_time_both_diverge_ora <- run_enrichment_hybrid_v2(
  genes_domain_time_both_diverge,
  description = "MAX & POST slopes both differ from PM"
)
```

```{r}
#| label: domain-time-both-ora-table
#| message: false
#| warning: false

if (!is.null(genes_domain_time_both_diverge_ora) && !is.null(genes_domain_time_both_diverge_ora$result)) {
  genes_domain_time_both_diverge_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

## High Slope Range

Genes with the largest differences in temporal slope across the AP axis (slope range > 0.5), regardless of which specific domains diverge. These show the most dramatic domain-specific developmental dynamics.

```{r}
#| label: domain-time-high-range-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

genes_domain_time_high_range <- domain_time_patterns |>
  filter(slope_range_magnitude == "High") |>
  pull(gene_id)

count_data <- get_vst_counts(strain_dds_full, genes_domain_time_high_range, vsd = vsd_global)
plot_domain_time_interaction(genes_domain_time_high_range, "High slope range across domains", count_data, n = 2)
```

```{r}
#| label: domain-time-high-range-table
#| message: false
#| warning: false

domain_time_patterns |>
  filter(slope_range_magnitude == "High") |>
  select(
    gene_id, symbol,
    time_slope_PM, time_slope_MAX, time_slope_POST,
    domain_time_int_MAX_slope, domain_time_int_POST_slope,
    diverging_domains, slope_range
  ) |>
  arrange(desc(slope_range)) |>
  create_dt()
```

```{r}
#| label: domain-time-high-range-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

genes_domain_time_high_range_ora <- run_enrichment_hybrid_v2(
  genes_domain_time_high_range,
  description = "High slope range across domains"
)
```

```{r}
#| label: domain-time-high-range-ora-table
#| message: false
#| warning: false

if (!is.null(genes_domain_time_high_range_ora) && !is.null(genes_domain_time_high_range_ora$result)) {
  genes_domain_time_high_range_ora$result |>
    filter(p_value < 0.01) |>
    create_dt()
} else {
  cat("No significant enrichment results found.")
}
```

::::
# Three-Way Interactions

These genes show domain-specific strain differences that also change over developmental time. We classify them based on whether domains primarily differ in their initial state (t=0), their temporal dynamics (slopes), or both.

:::: {#threeway-tabs .panel-tabset .nav-pills}

## Overview

```{r}
#| label: threeway-overview
#| message: false
#| warning: false

threeway_patterns |>
  count(detailed_pattern) |>
  arrange(desc(n)) |>
  rename(Pattern = detailed_pattern, `N Genes` = n) |>
  create_dt()
```

## Complex Pattern

Genes where domains differ in BOTH their initial state (t=0) AND their temporal dynamics. These show substantial differences across both dimensions.

```{r}
#| label: threeway-complex-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_threeway_complex, vsd = vsd_global)
plot_time_strain_overview(genes_threeway_complex, "Complex: Differ in initial state & dynamics", count_data, n = 2)
```

```{r}
#| label: threeway-complex-table
#| message: false
#| warning: false

threeway_patterns |>
  filter(gene_id %in% genes_threeway_complex) |>
  select(
    gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST,
    init_range, slope_range, most_distinct_domain
  ) |>
  create_dt()
```

```{r}
#| label: threeway-complex-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_complex_PM <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "PM") |>
  pull(gene_id)

genes_complex_MAX <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "MAX") |>
  pull(gene_id)

genes_complex_POST <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_complex, most_distinct_domain == "POST") |>
  pull(gene_id)

genes_threeway_complex_ora_PM <- run_enrichment_hybrid_v2(
  genes_complex_PM,
  description = "Complex pattern - PM most distinct"
)

genes_threeway_complex_ora_MAX <- run_enrichment_hybrid_v2(
  genes_complex_MAX,
  description = "Complex pattern - MAX most distinct"
)

genes_threeway_complex_ora_POST <- run_enrichment_hybrid_v2(
  genes_complex_POST,
  description = "Complex pattern - POST most distinct"
)
```

```{r}
#| label: threeway-complex-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_complex_ora_PM$result)) {
    genes_threeway_complex_ora_PM$result |>
      mutate(distinct_domain = paste0("PM (n=", length(genes_complex_PM), ")"))
  },
  if (!is.null(genes_threeway_complex_ora_MAX$result)) {
    genes_threeway_complex_ora_MAX$result |>
      mutate(distinct_domain = paste0("MAX (n=", length(genes_complex_MAX), ")"))
  },
  if (!is.null(genes_threeway_complex_ora_POST$result)) {
    genes_threeway_complex_ora_POST$result |>
      mutate(distinct_domain = paste0("POST (n=", length(genes_complex_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    select(distinct_domain, everything()) |>
    create_dt()
}
```

## Primarily Initial Differences

Genes where domains differ mainly at t=0, but show similar temporal trajectories. The spatial pattern is established early and maintained over time.

```{r}
#| label: threeway-initial-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_threeway_initial, vsd = vsd_global)
plot_time_strain_overview(genes_threeway_initial, "Primarily initial differences", count_data, n = 2)
```

```{r}
#| label: threeway-initial-table
#| message: false
#| warning: false

threeway_patterns |>
  filter(gene_id %in% genes_threeway_initial) |>
  select(
    gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST,
    init_range, slope_range, most_distinct_domain
  ) |>
  create_dt()
```

```{r}
#| label: threeway-initial-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_initial_PM <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "PM") |>
  pull(gene_id)

genes_initial_MAX <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "MAX") |>
  pull(gene_id)

genes_initial_POST <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_initial, most_distinct_domain == "POST") |>
  pull(gene_id)

genes_threeway_initial_ora_PM <- run_enrichment_hybrid_v2(
  genes_initial_PM,
  description = "Initial differences - PM most distinct"
)

genes_threeway_initial_ora_MAX <- run_enrichment_hybrid_v2(
  genes_initial_MAX,
  description = "Initial differences - MAX most distinct"
)

genes_threeway_initial_ora_POST <- run_enrichment_hybrid_v2(
  genes_initial_POST,
  description = "Initial differences - POST most distinct"
)
```

```{r}
#| label: threeway-initial-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_initial_ora_PM$result)) {
    genes_threeway_initial_ora_PM$result |>
      mutate(distinct_domain = paste0("PM (n=", length(genes_initial_PM), ")"))
  },
  if (!is.null(genes_threeway_initial_ora_MAX$result)) {
    genes_threeway_initial_ora_MAX$result |>
      mutate(distinct_domain = paste0("MAX (n=", length(genes_initial_MAX), ")"))
  },
  if (!is.null(genes_threeway_initial_ora_POST$result)) {
    genes_threeway_initial_ora_POST$result |>
      mutate(distinct_domain = paste0("POST (n=", length(genes_initial_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    select(distinct_domain, everything()) |>
    create_dt()
}
```

## Primarily Dynamic Differences

Genes where domains start similar at t=0 but diverge over time with different temporal trajectories. The spatial pattern emerges during development.

```{r}
#| label: threeway-dynamic-plot
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 16

count_data <- get_vst_counts(strain_dds_full, genes_threeway_dynamic, vsd = vsd_global)
plot_time_strain_overview(genes_threeway_dynamic, "Primarily dynamic differences", count_data, n = 2)
```

```{r}
#| label: threeway-dynamic-table
#| message: false
#| warning: false

threeway_patterns |>
  filter(gene_id %in% genes_threeway_dynamic) |>
  select(
    gene_id, symbol, init_PM, init_MAX, init_POST, slope_PM, slope_MAX, slope_POST,
    init_range, slope_range, most_distinct_domain
  ) |>
  create_dt()
```

```{r}
#| label: threeway-dynamic-ora
#| message: false
#| warning: false
#| cache: true
#| results: hide

# Split by the most distinct domain
genes_dynamic_PM <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "PM") |>
  pull(gene_id)

genes_dynamic_MAX <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "MAX") |>
  pull(gene_id)

genes_dynamic_POST <- threeway_patterns |>
  filter(gene_id %in% genes_threeway_dynamic, most_distinct_domain == "POST") |>
  pull(gene_id)

genes_threeway_dynamic_ora_PM <- run_enrichment_hybrid_v2(
  genes_dynamic_PM,
  description = "Dynamic differences - PM most distinct"
)

genes_threeway_dynamic_ora_MAX <- run_enrichment_hybrid_v2(
  genes_dynamic_MAX,
  description = "Dynamic differences - MAX most distinct"
)

genes_threeway_dynamic_ora_POST <- run_enrichment_hybrid_v2(
  genes_dynamic_POST,
  description = "Dynamic differences - POST most distinct"
)
```

```{r}
#| label: threeway-dynamic-ora-table
#| message: false
#| warning: false

combined <- bind_rows(
  if (!is.null(genes_threeway_dynamic_ora_PM$result)) {
    genes_threeway_dynamic_ora_PM$result |>
      mutate(distinct_domain = paste0("PM (n=", length(genes_dynamic_PM), ")"))
  },
  if (!is.null(genes_threeway_dynamic_ora_MAX$result)) {
    genes_threeway_dynamic_ora_MAX$result |>
      mutate(distinct_domain = paste0("MAX (n=", length(genes_dynamic_MAX), ")"))
  },
  if (!is.null(genes_threeway_dynamic_ora_POST$result)) {
    genes_threeway_dynamic_ora_POST$result |>
      mutate(distinct_domain = paste0("POST (n=", length(genes_dynamic_POST), ")"))
  }
)

if (!is.null(combined) && nrow(combined) > 0) {
  combined |>
    filter(p_value < 0.01) |>
    select(distinct_domain, everything()) |>
    create_dt()
}
```

::::

# Comprehensive Tables

This section provides downloadable tables summarizing all analysis results.

:::: {#comprehensive-tables-tabs .panel-tabset .nav-pills}

## Gene-Level Pattern Assignments

All genes with their assigned expression patterns across strain, domain, and time.

```{r}
#| label: comprehensive-gene-assignments
#| message: false
#| warning: false

gene_assignments |>
  left_join(all_genes_palate |> select(gene_id, symbol, gene_biotype, chromosome)) |>
  select(gene_id, symbol, gene_biotype, chromosome, everything()) |>
  create_dt()
```

## Full Effects Classification

Complete results table with all coefficients and classifications for all genes with any
significant effect (strain, domain, or time).

```{r}
#| label: comprehensive-strain-effects
#| message: false
#| warning: false

strain_effects_classified |>
  filter(any_strain_sig | domain_sig | time_sig) |>
  select(
    gene_id, symbol, gene_biotype, baseMean,
    # Top-level pattern
    pattern_type,
    # Main strain effect
    main_strain_pattern, main_strain_lfc, main_strain_pval,
    # Main domain effect
    domain_main_pattern,
    domain_main_MAX_vs_PM_lfc, domain_main_POST_vs_PM_lfc, domain_main_MAX_vs_POST_lfc,
    # Main time effect
    time_main_pattern, time_main_slope, time_main_slope_pval,
    # Domain × Time interaction
    domain_time_pattern,
    domain_time_int_MAX_slope, domain_time_int_POST_slope,
    time_slope_PM, time_slope_MAX, time_slope_POST,
    # Domain × Strain interaction
    strain_effect_PM, strain_effect_MAX, strain_effect_POST,
    strain_effect_PM_pass, strain_effect_MAX_pass, strain_effect_POST_pass,
    # Time × Strain interaction
    time_strain_pattern, time_int_slope_avg,
    # Three-way interaction
    threeway_MAX_slope, threeway_POST_slope,
    time_strain_slope_PM, time_strain_slope_MAX, time_strain_slope_POST,
    # Significance flags
    any_strain_sig, domain_sig, time_sig,
    strain_domain_int_sig, strain_time_int_sig, domain_time_int_sig, threeway_int_sig
  ) |>
  create_dt()
```

## Combined ORA Results

All over-representation analysis results combined into a single searchable table.

```{r}
#| label: comprehensive-ora-combine
#| message: false
#| warning: false

all_ora_results <- bind_rows(
  # ── Main strain effects ──────────────────────────────────────────────────
  if (!is.null(genes_CAST_high_strong_ora$result)) {
    genes_CAST_high_strong_ora$result |>
      mutate(category = "Main Strain", pattern = "CAST >> B6 (strong)", direction = "Up in CAST")
  },
  if (!is.null(genes_CAST_high_mod_ora$result)) {
    genes_CAST_high_mod_ora$result |>
      mutate(category = "Main Strain", pattern = "CAST > B6 (moderate)", direction = "Up in CAST")
  },
  if (!is.null(genes_B6_high_strong_ora$result)) {
    genes_B6_high_strong_ora$result |>
      mutate(category = "Main Strain", pattern = "B6 >> CAST (strong)", direction = "Down in CAST")
  },
  if (!is.null(genes_B6_high_mod_ora$result)) {
    genes_B6_high_mod_ora$result |>
      mutate(category = "Main Strain", pattern = "B6 > CAST (moderate)", direction = "Down in CAST")
  },

  # ── Main domain effects ──────────────────────────────────────────────────
  if (!is.null(genes_domain_increasing_posterior_ora$result)) {
    genes_domain_increasing_posterior_ora$result |>
      mutate(category = "Main Domain", pattern = "Increasing posterior (PM lowest)", direction = "All")
  },
  if (!is.null(genes_domain_decreasing_posterior_ora$result)) {
    genes_domain_decreasing_posterior_ora$result |>
      mutate(category = "Main Domain", pattern = "Decreasing posterior (PM highest)", direction = "All")
  },
  if (!is.null(genes_domain_MAX_peak_ora$result)) {
    genes_domain_MAX_peak_ora$result |>
      mutate(category = "Main Domain", pattern = "MAX peak (non-monotone)", direction = "All")
  },
  if (!is.null(genes_domain_high_specificity_ora$result)) {
    genes_domain_high_specificity_ora$result |>
      mutate(category = "Main Domain", pattern = "High domain specificity", direction = "All")
  },

  # ── Main time effects ────────────────────────────────────────────────────
  if (!is.null(genes_time_increasing_ora$result)) {
    genes_time_increasing_ora$result |>
      mutate(category = "Main Time", pattern = "Expression increases over time", direction = "Increasing")
  },
  if (!is.null(genes_time_decreasing_ora$result)) {
    genes_time_decreasing_ora$result |>
      mutate(category = "Main Time", pattern = "Expression decreases over time", direction = "Decreasing")
  },

  # ── Domain × Time interaction ────────────────────────────────────────────
  if (!is.null(genes_domain_time_MAX_ora_faster$result)) {
    genes_domain_time_MAX_ora_faster$result |>
      mutate(category = "Domain × Time", pattern = "MAX slope differs from PM", direction = "MAX faster")
  },
  if (!is.null(genes_domain_time_MAX_ora_slower$result)) {
    genes_domain_time_MAX_ora_slower$result |>
      mutate(category = "Domain × Time", pattern = "MAX slope differs from PM", direction = "MAX slower")
  },
  if (!is.null(genes_domain_time_POST_ora_faster$result)) {
    genes_domain_time_POST_ora_faster$result |>
      mutate(category = "Domain × Time", pattern = "POST slope differs from PM", direction = "POST faster")
  },
  if (!is.null(genes_domain_time_POST_ora_slower$result)) {
    genes_domain_time_POST_ora_slower$result |>
      mutate(category = "Domain × Time", pattern = "POST slope differs from PM", direction = "POST slower")
  },
  if (!is.null(genes_domain_time_both_diverge_ora$result)) {
    genes_domain_time_both_diverge_ora$result |>
      mutate(category = "Domain × Time", pattern = "MAX & POST slopes differ from PM", direction = "All")
  },
  if (!is.null(genes_domain_time_high_range_ora$result)) {
    genes_domain_time_high_range_ora$result |>
      mutate(category = "Domain × Time", pattern = "High slope range", direction = "All")
  },

  # ── Domain × Strain interaction ──────────────────────────────────────────
  if (!is.null(genes_PM_strongest_effect_ora_up$result)) {
    genes_PM_strongest_effect_ora_up$result |>
      mutate(category = "Domain × Strain", pattern = "PM strongest effect", direction = "Up in CAST")
  },
  if (!is.null(genes_PM_strongest_effect_ora_down$result)) {
    genes_PM_strongest_effect_ora_down$result |>
      mutate(category = "Domain × Strain", pattern = "PM strongest effect", direction = "Down in CAST")
  },
  if (!is.null(genes_MAX_strongest_effect_ora_up$result)) {
    genes_MAX_strongest_effect_ora_up$result |>
      mutate(category = "Domain × Strain", pattern = "MAX strongest effect", direction = "Up in CAST")
  },
  if (!is.null(genes_MAX_strongest_effect_ora_down$result)) {
    genes_MAX_strongest_effect_ora_down$result |>
      mutate(category = "Domain × Strain", pattern = "MAX strongest effect", direction = "Down in CAST")
  },
  if (!is.null(genes_POST_strongest_effect_ora_up$result)) {
    genes_POST_strongest_effect_ora_up$result |>
      mutate(category = "Domain × Strain", pattern = "POST strongest effect", direction = "Up in CAST")
  },
  if (!is.null(genes_POST_strongest_effect_ora_down$result)) {
    genes_POST_strongest_effect_ora_down$result |>
      mutate(category = "Domain × Strain", pattern = "POST strongest effect", direction = "Down in CAST")
  },
  if (!is.null(genes_high_domain_specificity_ora$result)) {
    genes_high_domain_specificity_ora$result |>
      mutate(category = "Domain × Strain", pattern = "High domain specificity", direction = "All")
  },
  if (!is.null(genes_opposite_directions_ora_up$result)) {
    genes_opposite_directions_ora_up$result |>
      mutate(category = "Domain × Strain", pattern = "Opposite directions", direction = "Up in CAST (PM)")
  },
  if (!is.null(genes_opposite_directions_ora_down$result)) {
    genes_opposite_directions_ora_down$result |>
      mutate(category = "Domain × Strain", pattern = "Opposite directions", direction = "Down in CAST (PM)")
  },

  # ── Time × Strain interaction ────────────────────────────────────────────
  if (!is.null(genes_diverging_CAST_faster_ora_up$result)) {
    genes_diverging_CAST_faster_ora_up$result |>
      mutate(category = "Time × Strain", pattern = "Diverging: CAST faster", direction = "Up in CAST at baseline")
  },
  if (!is.null(genes_diverging_CAST_faster_ora_down$result)) {
    genes_diverging_CAST_faster_ora_down$result |>
      mutate(category = "Time × Strain", pattern = "Diverging: CAST faster", direction = "Down in CAST at baseline")
  },
  if (!is.null(genes_diverging_B6_faster_ora_up$result)) {
    genes_diverging_B6_faster_ora_up$result |>
      mutate(category = "Time × Strain", pattern = "Diverging: B6 faster", direction = "Up in CAST at baseline")
  },
  if (!is.null(genes_diverging_B6_faster_ora_down$result)) {
    genes_diverging_B6_faster_ora_down$result |>
      mutate(category = "Time × Strain", pattern = "Diverging: B6 faster", direction = "Down in CAST at baseline")
  },

  # ── Three-way interaction ────────────────────────────────────────────────
  if (!is.null(genes_threeway_complex_ora_PM$result)) {
    genes_threeway_complex_ora_PM$result |>
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "PM most distinct")
  },
  if (!is.null(genes_threeway_complex_ora_MAX$result)) {
    genes_threeway_complex_ora_MAX$result |>
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "MAX most distinct")
  },
  if (!is.null(genes_threeway_complex_ora_POST$result)) {
    genes_threeway_complex_ora_POST$result |>
      mutate(category = "Three-way", pattern = "Complex pattern", direction = "POST most distinct")
  },
  if (!is.null(genes_threeway_initial_ora_PM$result)) {
    genes_threeway_initial_ora_PM$result |>
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "PM most distinct")
  },
  if (!is.null(genes_threeway_initial_ora_MAX$result)) {
    genes_threeway_initial_ora_MAX$result |>
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "MAX most distinct")
  },
  if (!is.null(genes_threeway_initial_ora_POST$result)) {
    genes_threeway_initial_ora_POST$result |>
      mutate(category = "Three-way", pattern = "Primarily initial differences", direction = "POST most distinct")
  },
  if (!is.null(genes_threeway_dynamic_ora_PM$result)) {
    genes_threeway_dynamic_ora_PM$result |>
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "PM most distinct")
  },
  if (!is.null(genes_threeway_dynamic_ora_MAX$result)) {
    genes_threeway_dynamic_ora_MAX$result |>
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "MAX most distinct")
  },
  if (!is.null(genes_threeway_dynamic_ora_POST$result)) {
    genes_threeway_dynamic_ora_POST$result |>
      mutate(category = "Three-way", pattern = "Primarily dynamic differences", direction = "POST most distinct")
  }
)
```

```{r}
#| label: comprehensive-ora-table
#| message: false
#| warning: false

all_ora_results |>
  filter(p_value < 0.01) |>
  select(
    category, pattern, direction, source, term_name, p_value,
    intersection_size, term_size, intersection
  ) |>
  arrange(category, pattern, direction, p_value) |>
  create_dt()
```

::::

# Session Info

```{r}
#| label: session-info
#| message: false
#| warning: false

sessionInfo()
```
