---
title: "Investigating A-P domain and developmental stage effects in B6"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    standalone: true
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| warning: false
#| message: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

# Load packages
suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(DT)
library(enrichR) # for mammalian phenotypes
library(here)

# Explicitly define common functions to avoid conflicts
select <- dplyr::select
rename <- dplyr::rename
filter <- dplyr::filter
summarize <- dplyr::summarize
count <- dplyr::count

set.seed(123) # for reproducibility

# Read sample details
sample_details <- read_csv(here("../samplesheet.csv"))

# Domain colors
domain_colors <- c(
  PM = rgb(252, 206, 93, maxColorValue = 255),
  MAX = rgb(163, 82, 157, maxColorValue = 255),
  POST = rgb(158, 156, 201, maxColorValue = 255)
)

# Strain colors
strain_colors <- c(
  B6 = "#555555",
  CAST = "#009E73",
  DO = rgb(163, 82, 157, maxColorValue = 255)
)

# helper functions
create_dt <- function(x) {
  format_num <- function(val) {
    case_when(
      is.na(val) ~ NA_character_,
      abs(val) < 0.01 ~ formatC(val, format = "e", digits = 2),
      TRUE ~ as.character(round(val, 2))
    )
  }

  x <- x |>
    mutate(across(where(is.numeric), format_num))

  DT::datatable(x,
    extensions = "Buttons",
    rownames = FALSE,
    filter = "top",
    options = list(
      dom = "Blfrtip",
      buttons = c("copy", "csv", "excel"),
      pageLength = 5,
      scrollX = TRUE
    )
  )
}

safe_result <- function(x) if (!is.null(x)) x$result else NULL

run_enrichment_hybrid_v2 <- function(gene_ids, 
                                     description = "Gene Set",
                                 gene_reference = all_genes_palate,
                                 max_term_size = 1000,
                                 min_term_size = 4,
                                 retries = 3,
                                 retry_wait = 5) {
  
  databases <- c(
     "GO_Biological_Process_2025",
      "GO_Cellular_Component_2025",
      "GO_Molecular_Function_2025",
      "GWAS_Catalog_2025", 
      "Human_Phenotype_Ontology",
      "JASPAR_PWM_Mouse_2025",
      "MGI_Mammalian_Phenotype_Level_4_2021",
      "CORUM",
      "KEGG_2019_Mouse",
      "Tabula_Muris",
      # "GTEx_Tissues_V8_2023", # results too broad, not useful
      # "PerturbAtlas_MouseGenePerturbationSigs", # too big, parser can't handle it!
      "Mouse_Gene_Atlas",
      # "DSigDB",
      "Jensen_DISEASES_Curated_2025",
      "KOMP2_Mouse_Phenotypes_2022",
      "WikiPathways_2024_Mouse"
  )
  
  # Convert gene IDs to symbols
  query_symbols <- gene_reference |>
    filter(gene_id %in% unique(gene_ids)) |>
    pull(symbol) |>
    unique()
  
  # Get background symbols
  background_symbols <- gene_reference$symbol
  
  # Skip if too few genes
  if(length(query_symbols) < 5) {
    message(paste0("Skipping ", description, ": fewer than 5 genes"))
    return(NULL)
  }
  
  # Query each database individually with tryCatch + retry
  result_enrichr <- list()
  
  for (db in databases) {
    res <- NULL
    for (attempt in seq_len(retries)) {
      res <- tryCatch(
        enrichr(
          genes      = query_symbols,
          databases  = db,
          background = background_symbols,
          include_overlap = TRUE
        ),
        error = function(e) {
          message(sprintf("  [%s] Attempt %d failed: %s", db, attempt, e$message))
          NULL
        }
      )
      if (!is.null(res)) break
      if (attempt < retries) {
        message(sprintf("  [%s] Retrying in %ds...", db, retry_wait))
        Sys.sleep(retry_wait)
      }
    }
    
    if (is.null(res)) {
      message(sprintf("  [%s] Skipped after %d failed attempts.", db, retries))
    } else {
      result_enrichr <- c(result_enrichr, res)
    }
  }
  
  # Format Enrichr results
  enrichr_df <- NULL
  if (length(result_enrichr) > 0) {
    enrichr_dfs <- lapply(names(result_enrichr), function(db_name) {
      db_result <- result_enrichr[[db_name]]
      
      if (is.null(db_result) || nrow(db_result) == 0 || all(is.na(db_result$Term))) {
        return(NULL)
      }
      
      db_result |>
        filter(Adjusted.P.value < 0.05) |>
        mutate(
          gene_set         = description,
          source           = db_name,
          term_name        = Term,
          p_value          = Adjusted.P.value,
          query_size       = length(query_symbols),
          intersection_size = as.numeric(gsub("/.*", "", Overlap)),
          term_size        = as.numeric(gsub(".*/", "", Overlap)),
          intersection     = Genes
        ) |>
        filter(term_size < max_term_size, term_size > min_term_size) |>
        select(gene_set, source, term_name, p_value,
               query_size, term_size, intersection_size, intersection)
    })
    
    enrichr_dfs <- enrichr_dfs[!sapply(enrichr_dfs, is.null)]
    if (length(enrichr_dfs) > 0) {
      enrichr_df <- bind_rows(enrichr_dfs)
    }
  }
  
  combined_result <- enrichr_df
  
  if (!is.null(combined_result) && nrow(combined_result) > 0) {
    return(list(result = combined_result))
  } else {
    return(NULL)
  }
}

# plotting-functions

# ===== FUNCTION 1: COMPARE UP VS DOWN IN SAME DOMAIN =====
plot_domain_direction_comparison <- function(genes_up, genes_down,
                                             domain_name, count_data,
                                             n_each = 2) {
  gene_col <- "symbol"

  # Get examples
  up_examples <- genes_up[1:min(n_each, length(genes_up))]
  down_examples <- genes_down[1:min(n_each, length(genes_down))]

  # Create plots for UP genes
  plots_up <- lapply(up_examples, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        subtitle = paste0("↑ in ", domain_name),
        x = "Development stage",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(color = "darkgreen")
      )
  })

  # Create plots for DOWN genes
  plots_down <- lapply(down_examples, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        subtitle = paste0("↓ in ", domain_name),
        x = "Development stage",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(color = "darkred")
      )
  })

  # Remove NULLs
  plots_up <- plots_up[!sapply(plots_up, is.null)]
  plots_down <- plots_down[!sapply(plots_down, is.null)]

  # Combine
  all_plots <- c(plots_up, plots_down)
  if (length(all_plots) == 0) {
    return(NULL)
  }

  if (length(all_plots) == 2) {
    combined <- all_plots[[1]] / all_plots[[2]] + plot_layout( guides = "collect") & theme(legend.position = "bottom")
  } else if (length(all_plots) == 4) {
    combined <- (all_plots[[1]] | all_plots[[2]]) / (all_plots[[3]] | all_plots[[4]]) + plot_layout( guides = "collect") & theme(legend.position = "bottom")
  } else {
    combined <- wrap_plots(all_plots, ncol = 2, guides = "collect") & theme(legend.position = "bottom")
  }

  combined <- combined +
    plot_annotation(
      title = paste0(domain_name, ": Higher (↑) vs Lower (↓)"),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    ) +
    plot_layout( guides = "collect") & theme(legend.position = "bottom")

  return(combined)
}

# ===== FUNCTION 2: PLOT DOMAIN HIERARCHY =====
plot_domain_hierarchy <- function(genes_highest, domain_name, count_data, n = 3) {
  gene_col <- "symbol"

  example_genes <- genes_highest[1:min(n, length(genes_highest))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    # Create both trajectory and boxplot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        subtitle = paste0("Highest in ", domain_name),
        x = "Development",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        plot.title = element_text(face = "bold")
      )

    p2 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = AP_domain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values = domain_colors) +
      labs(x = "", y = "") +
      theme_pubclean(base_size = 16) +
      theme(legend.position = "none")

    return(p1 + p2 + plot_layout(widths = c(2, 1)))
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }

  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Genes Highest in ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )

  return(combined)
}

# ===== FUNCTION 3: PLOT A-P GRADIENTS =====
plot_AP_gradient <- function(genes_decreasing_with_development, genes_increasing_with_development, count_data, n_each = 3) {
  gene_col <- "symbol"

  # Get examples
  dec_examples <- genes_decreasing_with_development[1:min(n_each, length(genes_decreasing_with_development))]
  inc_examples <- genes_increasing_with_development[1:min(n_each, length(genes_increasing_with_development))]

  # Function to create gradient plot
  make_gradient_plot <- function(gid, gradient_type) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    # Summary by domain
    domain_summary <- gene_data |>
      group_by(AP_domain) |>
      summarize(mean_count = mean(count, na.rm = TRUE), .groups = "drop")

    ggplot(domain_summary, aes(x = AP_domain, y = mean_count, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_point(
        data = gene_data, aes(y = count), alpha = 0.3, size = 1.5,
        position = position_jitter(width = 0.1)
      ) +
      scale_fill_manual(values = domain_colors) +
      scale_y_log10(labels = scales::comma) +
      labs(
        title = gene_name,
        subtitle = ifelse(gradient_type == "dec", "PM > MAX > POST", "POST > MAX > PM"),
        x = "",
        y = "Expression"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = ifelse(gradient_type == "dec",
          "darkred", "darkgreen"
        ))
      )
  }

  # Create plots
  plots_dec <- lapply(dec_examples, function(gid) make_gradient_plot(gid, "dec"))
  plots_inc <- lapply(inc_examples, function(gid) make_gradient_plot(gid, "inc"))

  # Remove NULLs
  plots_dec <- plots_dec[!sapply(plots_dec, is.null)]
  plots_inc <- plots_inc[!sapply(plots_inc, is.null)]

  # Combine
  all_plots <- c(plots_dec, plots_inc)
  if (length(all_plots) == 0) {
    return(NULL)
  }

  combined <- wrap_plots(all_plots, ncol = 3)
  combined <- combined +
    plot_annotation(
      title = "Anterior-Posterior Expression Gradients",
      subtitle = "Top: A→P decreasing | Bottom: A→P increasing",
      theme = theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12)
      )
    )

  return(combined)
}

# ===== FUNCTION 4: PLOT DOMAIN-SPECIFIC DYNAMICS =====
plot_domain_specific_dynamics <- function(gene_ids, domain_name, direction,
                                          count_data, slope_data, n = 3) {
  gene_col <- "symbol"

  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    # Get count data
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    # Get slope data
    gene_slopes <- slope_data |>
      filter(gene_id == gid) |>
      mutate(Domain = factor(Domain, levels = c("PM", "MAX", "POST")))

    # Trajectory plot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        subtitle = paste0(direction, " in ", domain_name, " only"),
        x = "Development",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold")
      )

    # Slope plot
    if (nrow(gene_slopes) > 0) {
      p2 <- ggplot(gene_slopes, aes(x = Domain, y = log2FoldChange, fill = Domain)) +
        geom_col(alpha = 0.8) +
        geom_errorbar(
          aes(
            ymin = log2FoldChange - lfcSE,
            ymax = log2FoldChange + lfcSE
          ),
          width = 0.3
        ) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_text(
          aes(label = ifelse(padj < 0.001, "***",
            ifelse(padj < 0.01, "**",
              ifelse(padj < 0.05, "*", "ns")
            )
          )),
          vjust = -0.5, size = 4
        ) +
        scale_fill_manual(values = domain_colors) +
        labs(y = "Slope", x = "") +
        theme_pubclean(base_size = 16) +
        theme(legend.position = "none")

      return(p1 + p2 + plot_layout(widths = c(2, 1)))
    } else {
      return(p1)
    }
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }

  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Domain-Specific Dynamics: ", domain_name, " ", direction),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )

  return(combined)
}

# ===== FUNCTION 5: PLOT DIVERGING PATTERNS =====
plot_diverging_detailed <- function(genes_faster, genes_opposite,
                                    domain_name, count_data, n_each = 2) {
  gene_col <- "symbol"

  faster_examples <- genes_faster[1:min(n_each, length(genes_faster))]
  opposite_examples <- genes_opposite[1:min(n_each, length(genes_opposite))]

  # Plot "faster" pattern
  plots_faster <- lapply(faster_examples, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        subtitle = paste0(domain_name, " increases faster"),
        x = "Development",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "darkgreen", size = 9)
      )
  })

  # Plot "opposite" pattern
  plots_opposite <- lapply(opposite_examples, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        subtitle = paste0(domain_name, " ↑, other ↓"),
        x = "Development",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "purple", size = 9)
      )
  })

  # Remove NULLs
  plots_faster <- plots_faster[!sapply(plots_faster, is.null)]
  plots_opposite <- plots_opposite[!sapply(plots_opposite, is.null)]

  all_plots <- c(plots_faster, plots_opposite)
  if (length(all_plots) == 0) {
    return(NULL)
  }

  if (length(all_plots) <= 2) {
    combined <- wrap_plots(all_plots, ncol = 1)
  } else {
    combined <- wrap_plots(all_plots, ncol = 2)
  }

  combined <- combined +
    plot_annotation(
      title = paste0("Diverging Patterns: ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )

  return(combined)
}

# ===== FUNCTION 6: PLOT SIMPLE TRAJECTORIES =====
plot_simple_trajectories <- function(gene_ids, count_data, title_text, n = 5) {
  gene_col <- "symbol"

  # Sample if needed
  if (length(gene_ids) > n) {
    set.seed(123)
    gene_ids <- sample(gene_ids, n)
  }

  plots <- lapply(gene_ids, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data[[gene_col]])[1]

    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        x = "Development",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        legend.text = element_text(size = 8),
        plot.title = element_text(face = "bold", size = 11)
      )
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  else if( length(plots) == 1){
      combined <- wrap_plots(plots, ncol = 1)
  }
  else if( length(plots) > 1){
      combined <- wrap_plots(plots, ncol = 3)
  }

  
  combined <- combined +
    plot_annotation(
      title = title_text,
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    ) +
    plot_layout( guides = "collect") & theme(legend.position = "bottom")

  return(combined)
}


# ===== FUNCTION: Plot trajectories with slope annotations =====
plot_trajectory_with_slopes <- function(gene_ids, count_data, slope_data,
                                        title_text, n = 4) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data$symbol)[1]

    # Get slopes for this gene
    gene_slopes <- slope_data |>
      filter(gene_id == gid) |>
      mutate(
        direction = case_when(
          log2FoldChange > 0.5 ~ "↑",
          log2FoldChange < -0.5 ~ "↓",
          TRUE ~ "→"
        ),
        slope_label = paste0(direction, round(log2FoldChange, 2))
      )

    # Create subtitle with slopes
    if (nrow(gene_slopes) > 0) {
      slope_text <- gene_slopes |>
        arrange(factor(Domain, levels = c("PM", "MAX", "POST"))) |>
        mutate(text = paste0(Domain, ": ", slope_label)) |>
        pull(text) |>
        paste(collapse = "  ")
    } else {
      slope_text <- ""
    }

    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(
        title = gene_name,
        subtitle = slope_text,
        x = "Development",
        y = "Counts"
      ) +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 11),
        plot.subtitle = element_text(size = 8, color = "gray30", family = "mono")
      )
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }
  else if( length(plots) == 1){
      combined <- wrap_plots(plots, ncol = 1)
  }
  else if( length(plots) > 1){
      combined <- wrap_plots(plots, ncol = 2)
  }
  
  combined <- combined +
    plot_annotation(
      title = title_text,
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )+
    plot_layout( guides = "collect") & theme(legend.position = "bottom")

  return(combined)
}

# ===== FUNCTION: Plot pairwise comparison with slope bars =====
plot_pairwise_trajectory <- function(gene_ids, count_data, slope_data,
                                     domain1, domain2, pattern_name, n = 4) {
  example_genes <- gene_ids[1:min(n, length(gene_ids))]

  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |>
      filter(gene_id == gid, AP_domain %in% c(domain1, domain2)) |>
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))

    if (nrow(gene_data) == 0) {
      return(NULL)
    }

    gene_name <- unique(gene_data$symbol)[1]

    # Get slopes for comparison
    gene_slopes <- slope_data |>
      filter(gene_id == gid, Domain %in% c(domain1, domain2))

    # Trajectory plot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name, x = "Development", y = "Counts") +
      theme_pubclean(base_size = 16) +
      theme(
        legend.position = "bottom",
        plot.title = element_text(face = "bold")
      )

    # Slope comparison bar plot
    if (nrow(gene_slopes) > 0) {
      p2 <- ggplot(gene_slopes, aes(x = Domain, y = log2FoldChange, fill = Domain)) +
        geom_col(alpha = 0.8, width = 0.6) +
        geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
        geom_text(aes(label = round(log2FoldChange, 2)),
          vjust = ifelse(gene_slopes$log2FoldChange > 0, -0.5, 1.5),
          size = 3
        ) +
        scale_fill_manual(values = domain_colors) +
        labs(y = "Slope (log2FC/stage)", x = "") +
        theme_pubclean(base_size = 16) +
        theme(legend.position = "none")

      return(p1 + p2 + plot_layout(widths = c(2, 1)))
    } else {
      return(p1)
    }
  })

  plots <- plots[!sapply(plots, is.null)]
  if (length(plots) == 0) {
    return(NULL)
  }

  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = pattern_name,
      subtitle = paste0("n = ", length(gene_ids), " genes"),
      theme = theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11)
      )
    )

  return(combined)
}

# ===== FUNCTION: Summary heatmap of trajectory patterns =====
plot_trajectory_heatmap <- function(trajectory_patterns, slope_data) {
  # Prepare data for heatmap
  heatmap_data <- trajectory_patterns |>
    select(gene_id, symbol, PM_log2FoldChange, MAX_log2FoldChange, POST_log2FoldChange) |>
    pivot_longer(
      cols = ends_with("_log2FoldChange"),
      names_to = "Domain",
      values_to = "Slope"
    ) |>
    mutate(Domain = gsub("_log2FoldChange", "", Domain)) |>
    mutate(Domain = factor(Domain, levels = c("PM", "MAX", "POST")))

  # Create slope matrix for clustering
  slope_matrix <- trajectory_patterns |>
    select(gene_id, PM_log2FoldChange, MAX_log2FoldChange, POST_log2FoldChange) |>
    column_to_rownames("gene_id") |>
    as.matrix()

  # Replace NAs with 0 for clustering
  slope_matrix[is.na(slope_matrix)] <- 0

  # Cluster genes
  if (nrow(slope_matrix) > 2) {
    gene_order <- hclust(dist(slope_matrix))$order
    gene_levels <- rownames(slope_matrix)[gene_order]
  } else {
    gene_levels <- rownames(slope_matrix)
  }

  heatmap_data <- heatmap_data |>
    mutate(gene_id = factor(gene_id, levels = gene_levels))

  ggplot(heatmap_data, aes(x = Domain, y = gene_id, fill = Slope)) +
    geom_tile(color = "white", linewidth = 0.1) +
    scale_fill_gradient2(
      low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0,
      limits = c(-3, 3),
      oob = scales::squish,
      name = "Slope\n(log2FC)"
    ) +
    labs(
      title = "Developmental Slopes by Domain",
      subtitle = paste0("n = ", n_distinct(heatmap_data$gene_id), " genes with interactions"),
      x = "",
      y = ""
    ) +
    theme_minimal() +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid = element_blank()
    )
}

# ===== FUNCTION: Trajectory pattern distribution plot =====
plot_pattern_distribution <- function(trajectory_patterns, pattern_col, title) {
  pattern_counts <- trajectory_patterns |>
    count(.data[[pattern_col]], name = "n") |>
    filter(!is.na(.data[[pattern_col]])) |>
    mutate(
      # Color by pattern type
      pattern_type = case_when(
        grepl("Opposite", .data[[pattern_col]]) ~ "Opposite",
        grepl("Co-increasing", .data[[pattern_col]]) ~ "Co-increasing",
        grepl("Co-decreasing", .data[[pattern_col]]) ~ "Co-decreasing",
        grepl("only", .data[[pattern_col]]) ~ "Single domain",
        TRUE ~ "Other"
      )
    ) |>
    mutate(!!pattern_col := fct_reorder(.data[[pattern_col]], n))

  ggplot(pattern_counts, aes(x = n, y = .data[[pattern_col]], fill = pattern_type)) +
    geom_col(alpha = 0.8) +
    geom_text(aes(label = n), hjust = -0.2, size = 3) +
    scale_fill_manual(
      values = c(
        "Opposite" = "#E41A1C",
        "Co-increasing" = "#4DAF4A",
        "Co-decreasing" = "#377EB8",
        "Single domain" = "#984EA3",
        "Other" = "#999999"
      ),
      name = "Pattern Type"
    ) +
    labs(
      title = title,
      x = "Number of genes",
      y = ""
    ) +
    theme_pubclean() +
    theme(legend.position = "right")
}


```

```{r}
#| label: load-data-emase
#| warning: false
#| message: false

# Loading the count data
load(here("../read_mapping/palate_gbrs_emase_counts_v4.RData") )

# Clean up expression matrix
expr_mat <- expr_mat_nondo
contam_samples <- c("B6_12.5_PM_1", "B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat[, !colnames(expr_mat) %in% contam_samples]

do_expr_mat <- expr_mat_do
remove_samples <- c(
  "DO_13.5_MAX_6", "DO_13.5_MAX_14", "DO_13.5_MAX_14_2",
  "DO_12.5_MAX_1", "DO_12.5_MAX_2", "DO_13.5_MAX_97", "DO_13.5_MAX_96"
)
do_expr_mat <- do_expr_mat[, !colnames(do_expr_mat) %in% remove_samples]

expr_mat_all <- cbind(
  expr_mat[intersect(rownames(expr_mat), rownames(do_expr_mat)), , drop = FALSE],
  do_expr_mat[intersect(rownames(expr_mat), rownames(do_expr_mat)), , drop = FALSE]
)
rm(expr_mat_do, expr_mat_nondo)

# Get v105 annotations from biomart
all.genes_v105 <- read_tsv(here("_data", "ENSMUSGid_to_symbol_v105.txt"),
  col_names = c("gene_id", "symbol", "gene_biotype", "chromosome", "start", "end"),
  skip = 1
) |>
  mutate(midpoint = (start + end) / 2)

all_genes_palate <- all.genes_v105 |>
  filter(gene_id %in% rownames(raw_expr_mat))

all_genes_palate_filtered <- all.genes_v105 |>
  filter(gene_id %in% rownames(expr_mat_all))

# Sample annotations
sample_annot <- sample_details |>
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |>
  separate(Strain_stock, into = c("strain", "tmp"), remove = FALSE, sep = "_") |>
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, palate_closure, L_R_palate) |>
  mutate(strain = case_when(
    strain == "B6J" ~ "B6",
    TRUE ~ strain
  )) |>
  mutate(LB_stage_binned = case_when(
    LB_stage < 12.25 ~ "exclude",
    between(LB_stage, 13.25, 13.5) ~ "exclude",
    between(LB_stage, 12.25, 12.75) ~ "1",
    between(LB_stage, 13.5, 14) ~ "2",
    between(LB_stage, 14.25, 14.75) ~ "3",
    between(LB_stage, 15.1, 15.6) ~ "4"
  )) |>
  mutate(LB_stage_binned = as_factor(LB_stage_binned)) |>
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |>
  select(-temp) |>
  mutate(Palate_ori = ifelse(Palate_ori == "L", "Left", "Right"))

sample_annot_all <- sample_details |>
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat_all)) |>
  separate(Strain_stock, into = c("strain", "tmp"), remove = FALSE, sep = "_") |>
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, palate_closure, L_R_palate) |>
  mutate(strain = case_when(
    strain == "B6J" ~ "B6",
    TRUE ~ strain
  )) |>
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |>
  select(-temp) |>
  mutate(Palate_ori = ifelse(Palate_ori == "L", "Left", "Right")) |>
  mutate(LB_stage_binned = case_when(
    LB_stage < 12.25 ~ "exclude",
    between(LB_stage, 13.25, 13.5) ~ "exclude",
    between(LB_stage, 12.25, 12.75) ~ "1",
    between(LB_stage, 13.5, 14) ~ "2",
    between(LB_stage, 14.25, 14.75) ~ "3",
    between(LB_stage, 15.1, 15.6) ~ "4"
  ))
```

# Differential gene expression analysis overview

```{r}
#| label: prep-run-b6-for-deseq2
#| warning: false
#| message: false

# Prep data for DESeq2
# Subsetting for B6 samples only and adding unique sample IDs
b6_comp_samples <- sample_annot |>
  filter(strain %in% c("B6")) |>
  separate(sample, sep = "_", into = c("s", "e", "d", "ind", "rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = ifelse(is.na(rep), 1, rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

# Making the coldata object to feed into DESeq2
b6_coldata <- b6_comp_samples |>
  select(sample, sex, ind, unique_sampleid, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate_if(is.character, as.factor) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

# Get the expression matrix to feed into DESeq2
b6_cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(b6_coldata), drop = FALSE])

# Make the dds object then collapse the technical replicates
b6_dds <- DESeqDataSetFromMatrix(
  countData = b6_cts,
  colData = b6_coldata,
  design = ~1
)
# Update LB-stage to have 0 as the starting point
min_LB_stage <- min(colData(b6_dds)$LB_stage)
colData(b6_dds)$LB_stage_from_start <- colData(b6_dds)$LB_stage - min_LB_stage

# Collapse the sequencing replicates
b6_dds_coll <- collapseReplicates(b6_dds, b6_dds$unique_sampleid)

# Define design matrices
b6_simple_model <- model.matrix(~ sex + LB_stage_from_start + AP_domain,
  data = colData(b6_dds_coll)
)
b6_comp_model <- model.matrix(~ sex + LB_stage_from_start * AP_domain,
  data = colData(b6_dds_coll)
)

# Run DESeq2 models
b6_dds_simple <- DESeq(b6_dds_coll, betaPrior = FALSE, full = b6_simple_model)
b6_dds_interaction <- DESeq(b6_dds_coll, betaPrior = FALSE, full = b6_comp_model)

# Likelihood ratio test to identify genes with significant interaction
b6_dds_lrt <- DESeq(b6_dds_coll,
  test = "LRT",
  full = b6_comp_model,
  reduced = b6_simple_model
)

# Identify genes with significant interaction
genes_with_interaction <- results(b6_dds_lrt) |>
  as_tibble(rownames = "gene_id") |>
  filter(padj < 0.05) |>
  pull(gene_id)


# Helper function for consistent filtering
filter_genes <- function(results_df, min_basemean = 50, max_lfcse = 1.5,
                         max_abs_lfc = 10, min_padj = 0.05, min_abs_lfc = 0.5) {
  results_df |>
    filter(
      baseMean > min_basemean,
      lfcSE < max_lfcse,
      abs(log2FoldChange) < max_abs_lfc,
      abs(log2FoldChange) > min_abs_lfc,
      padj < min_padj
    )
}

# Get genes with effects
# ===== MAIN EFFECTS (genes WITHOUT interaction) =====

# Domain effects from SIMPLE model
b6_domain_max_vs_pm <- results(b6_dds_simple, name = "AP_domainMAX") |>
  as_tibble(rownames = "gene_id") |>
  filter(!gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "MAX vs PM") |>
  left_join(all_genes_palate, by = "gene_id")

b6_domain_post_vs_pm <- results(b6_dds_simple, name = "AP_domainPOST") |>
  as_tibble(rownames = "gene_id") |>
  filter(!gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs PM") |>
  left_join(all_genes_palate, by = "gene_id")

b6_domain_post_vs_max <- results(b6_dds_simple,
  contrast = list("AP_domainPOST", "AP_domainMAX")
) |>
  as_tibble(rownames = "gene_id") |>
  filter(!gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs MAX") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine domain effects
b6_domain_main_effects <- bind_rows(
  b6_domain_max_vs_pm,
  b6_domain_post_vs_pm,
  b6_domain_post_vs_max
)

# Apply filtering
b6_domain_main_effects_clean <- filter_genes(
  b6_domain_main_effects,
  min_basemean = quantile(b6_domain_main_effects$baseMean, 0.1, na.rm = TRUE)
)

# Stage effects from SIMPLE model
b6_stage_main_effects <- results(b6_dds_simple, name = "LB_stage_from_start") |>
  as_tibble(rownames = "gene_id") |>
  filter(!gene_id %in% genes_with_interaction) |>
  mutate(Direction = ifelse(log2FoldChange > 0, "Increase", "Decrease")) |>
  left_join(all_genes_palate, by = "gene_id")

# Apply filtering
b6_stage_main_effects_clean <- filter_genes(
  b6_stage_main_effects,
  min_basemean = quantile(b6_stage_main_effects$baseMean, 0.1, na.rm = TRUE)
)

# Note that we need to be very careful when interpreting the main effects with a linear model that contains an interaction term, they do not represent the overall effect of the predictor (e.g. LB_stage) since the predictors involved in an interaction do not have a single overall effect anymore! For example the main effect for LB_stage will represent the effect when AP_domain is at the reference level which is the PM domain in our data set.
# ===== INTERACTION EFFECTS (genes WITH interaction) =====

# 1. INITIAL DOMAIN DIFFERENCES (at earliest stage)
initial_diff_MAX_vs_PM <- results(b6_dds_interaction, name = "AP_domainMAX") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "MAX vs PM at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

initial_diff_POST_vs_PM <- results(b6_dds_interaction, name = "AP_domainPOST") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs PM at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

initial_diff_POST_vs_MAX <- results(b6_dds_interaction,
  contrast = list(c("AP_domainPOST", "AP_domainMAX"))
) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs MAX at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine initial differences
b6_initial_differences <- bind_rows(
  initial_diff_MAX_vs_PM,
  initial_diff_POST_vs_PM,
  initial_diff_POST_vs_MAX
)

# 2. DEVELOPMENTAL SLOPES (rates of change in each domain)
slope_PM <- results(b6_dds_interaction, name = "LB_stage_from_start") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Domain = "PM") |>
  left_join(all_genes_palate, by = "gene_id")

slope_MAX <- results(b6_dds_interaction,
  contrast = list(c(
    "LB_stage_from_start",
    "LB_stage_from_start.AP_domainMAX" 
  ))
) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Domain = "MAX") |>
  left_join(all_genes_palate, by = "gene_id")

slope_POST <- results(b6_dds_interaction,
  contrast = list(c(
    "LB_stage_from_start",
    "LB_stage_from_start.AP_domainPOST"
  ))
) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Domain = "POST") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine all slopes
b6_domain_slopes <- bind_rows(slope_PM, slope_MAX, slope_POST)

# Apply filtering to slopes
b6_domain_slopes_clean <- filter_genes(
  b6_domain_slopes,
  min_basemean = quantile(b6_domain_slopes$baseMean, 0.1, na.rm = TRUE),
  max_lfcse = 2
)

# 3. SLOPE DIFFERENCES (interaction terms - how different are the slopes?)
slope_diff_MAX_vs_PM <- results(b6_dds_interaction,
  name = "LB_stage_from_start.AP_domainMAX"
) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "MAX vs PM slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

slope_diff_POST_vs_PM <- results(b6_dds_interaction,
  name = "LB_stage_from_start.AP_domainPOST"
) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs PM slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

slope_diff_POST_vs_MAX <- results(b6_dds_interaction,
  contrast = list(c(
    "LB_stage_from_start.AP_domainPOST",
    "LB_stage_from_start.AP_domainMAX"
  ))
) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs MAX slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine slope differences
b6_slope_differences <- bind_rows(
  slope_diff_MAX_vs_PM,
  slope_diff_POST_vs_PM,
  slope_diff_POST_vs_MAX
)

# Get normalized counts for plotting
b6_gene_counts <- counts(b6_dds_simple, normalized = TRUE) |>
  as_tibble(rownames = "gene_id") |>
  pivot_longer(-gene_id, names_to = "sample", values_to = "count") |>
  left_join(
    colData(b6_dds_simple) |>
      as_tibble(rownames = "sample"),
    by = "sample"
  ) |>
  left_join(all_genes_palate, by = "gene_id")


# Run pattern identification script that generates the pattern_summary, gene_assignments and trajectory_patterns files to read in.
# source(here("DEgene_pattern_identification_B6.R"))

# Load pattern assignments and summaries
pattern_summary <- read_csv(here("pattern_summary_detailed.csv"))
gene_assignments <- read_csv(here("gene_pattern_assignments_detailed.csv"))
trajectory_patterns <- read_csv(here("trajectory_patterns_detailed.csv"))


# Extract gene lists

# ===== EXTRACT GENE LISTS FROM TRAJECTORY PATTERNS =====

# --- MAX vs PM: Opposite trajectories ---
genes_MAX_up_PM_down <- trajectory_patterns |>
  filter(MAX_PM_pattern == "Opposite: MAX↑ PM↓") |>
  pull(gene_id)
genes_MAX_down_PM_up <- trajectory_patterns |>
  filter(MAX_PM_pattern == "Opposite: MAX↓ PM↑") |>
  pull(gene_id)

# --- MAX vs PM: Co-increasing ---
genes_coinc_MAX_faster_PM <- trajectory_patterns |>
  filter(MAX_PM_pattern == "Co-increasing: MAX faster") |>
  pull(gene_id)
genes_coinc_PM_faster_MAX <- trajectory_patterns |>
  filter(MAX_PM_pattern == "Co-increasing: PM faster") |>
  pull(gene_id)

# --- MAX vs PM: Co-decreasing ---
genes_codec_MAX_faster_PM <- trajectory_patterns |>
  filter(MAX_PM_pattern == "Co-decreasing: MAX faster") |>
  pull(gene_id)
genes_codec_PM_faster_MAX <- trajectory_patterns |>
  filter(MAX_PM_pattern == "Co-decreasing: PM faster") |>
  pull(gene_id)

# --- POST vs PM: Opposite trajectories ---
genes_POST_up_PM_down <- trajectory_patterns |>
  filter(POST_PM_pattern == "Opposite: POST↑ PM↓") |>
  pull(gene_id)
genes_POST_down_PM_up <- trajectory_patterns |>
  filter(POST_PM_pattern == "Opposite: POST↓ PM↑") |>
  pull(gene_id)

# --- POST vs PM: Co-increasing ---
genes_coinc_POST_faster_PM <- trajectory_patterns |>
  filter(POST_PM_pattern == "Co-increasing: POST faster") |>
  pull(gene_id)
genes_coinc_PM_faster_POST <- trajectory_patterns |>
  filter(POST_PM_pattern == "Co-increasing: PM faster") |>
  pull(gene_id)

# --- POST vs PM: Co-decreasing ---
genes_codec_POST_faster_PM <- trajectory_patterns |>
  filter(POST_PM_pattern == "Co-decreasing: POST faster") |>
  pull(gene_id)
genes_codec_PM_faster_POST <- trajectory_patterns |>
  filter(POST_PM_pattern == "Co-decreasing: PM faster") |>
  pull(gene_id)

# --- POST vs MAX: Opposite trajectories ---
genes_POST_up_MAX_down <- trajectory_patterns |>
  filter(POST_MAX_pattern == "Opposite: POST↑ MAX↓") |>
  pull(gene_id)
genes_POST_down_MAX_up <- trajectory_patterns |>
  filter(POST_MAX_pattern == "Opposite: POST↓ MAX↑") |>
  pull(gene_id)

# --- POST vs MAX: Co-increasing ---
genes_coinc_POST_faster_MAX <- trajectory_patterns |>
  filter(POST_MAX_pattern == "Co-increasing: POST faster") |>
  pull(gene_id)
genes_coinc_MAX_faster_POST <- trajectory_patterns |>
  filter(POST_MAX_pattern == "Co-increasing: MAX faster") |>
  pull(gene_id)

# --- POST vs MAX: Co-decreasing ---
genes_codec_POST_faster_MAX <- trajectory_patterns |>
  filter(POST_MAX_pattern == "Co-decreasing: POST faster") |>
  pull(gene_id)
genes_codec_MAX_faster_POST <- trajectory_patterns |>
  filter(POST_MAX_pattern == "Co-decreasing: MAX faster") |>
  pull(gene_id)

# --- Single domain dynamic (from overall_pattern) ---
genes_PM_only_up <- trajectory_patterns |>
  filter(overall_pattern == "PM↑ only") |>
  pull(gene_id)
genes_PM_only_down <- trajectory_patterns |>
  filter(overall_pattern == "PM↓ only") |>
  pull(gene_id)
genes_MAX_only_up <- trajectory_patterns |>
  filter(overall_pattern == "MAX↑ only") |>
  pull(gene_id)
genes_MAX_only_down <- trajectory_patterns |>
  filter(overall_pattern == "MAX↓ only") |>
  pull(gene_id)
genes_POST_only_up <- trajectory_patterns |>
  filter(overall_pattern == "POST↑ only") |>
  pull(gene_id)
genes_POST_only_down <- trajectory_patterns |>
  filter(overall_pattern == "POST↓ only") |>
  pull(gene_id)

# --- Two domains dynamic, same direction ---
genes_PM_MAX_up <- trajectory_patterns |>
  filter(overall_pattern == "PM+MAX↑, POST stable") |>
  pull(gene_id)
genes_PM_MAX_down <- trajectory_patterns |>
  filter(overall_pattern == "PM+MAX↓, POST stable") |>
  pull(gene_id)
genes_PM_POST_up <- trajectory_patterns |>
  filter(overall_pattern == "PM+POST↑, MAX stable") |>
  pull(gene_id)
genes_PM_POST_down <- trajectory_patterns |>
  filter(overall_pattern == "PM+POST↓, MAX stable") |>
  pull(gene_id)
genes_MAX_POST_up <- trajectory_patterns |>
  filter(overall_pattern == "MAX+POST↑, PM stable") |>
  pull(gene_id)
genes_MAX_POST_down <- trajectory_patterns |>
  filter(overall_pattern == "MAX+POST↓, PM stable") |>
  pull(gene_id)

# --- All three domains dynamic ---
genes_all_increasing <- trajectory_patterns |>
  filter(overall_pattern == "All↑") |>
  pull(gene_id)
genes_all_decreasing <- trajectory_patterns |>
  filter(overall_pattern == "All↓") |>
  pull(gene_id)

# --- Complex patterns ---
genes_opposite_slopes <- trajectory_patterns |>
  filter(grepl("Opposite", MAX_PM_pattern) |
    grepl("Opposite", POST_PM_pattern) |
    grepl("Opposite", POST_MAX_pattern)) |>
  pull(gene_id) |>
  unique()

genes_three_way_divergent <- trajectory_patterns |>
  filter(PM_sig & MAX_sig & POST_sig) |>
  filter(!(PM_up & MAX_up & POST_up) & !(PM_down & MAX_down & POST_down)) |>
  pull(gene_id)
```

Here, I analyze the B6 data on its own to identify genes differentially expressed across AP domains, over developmental time and change differently over time across domains.
The data from B6 animals includes males and females, and contain the three AP domains (PM, MAX and POST) obtained at various developmental stages (LB_stage).
I merge the technical replicates due to re-sequencing of several samples using `collapseReplicates()` function from DESeq2. Next, I calculate the relative or centered limb bud (LB) stage (LB_stage_from_start), where I subtract the minimum observed stage so that 0 represents the earliest developmental time point in the data set. 

In order to identify differentially expressed genes between domains and developmental time, I obtain the main effects for domain (`AP_domain`) and centered developmental time (`LB_stage_from_start`) while omiting a nesting for the domain since they are obtained from the same animal (`~ sex + AP_domain + LB_stage_from_start `).T he domains are collected from the same individual resulting in a repeated measure setup however variance partition analysis showed that individual-level variation accounted for a negligible proportion of transcriptional variance (median < 0.01%, mean 1.5%), therefore I omitted the nesting as an effect in the DeSeq2 model (`ind:AP_domain`). 

For identifying genes with significant interactions between domain and time, meaning they have different spatiotemporal gene expression patterns, I use the likelihood ratio test (LRT) to compare the initial simple model to the more complex linear model that includes the interaction term (`~ sex + AP_domain + LB_stage_from_start  + AP_domain:LB_stage_from_start`).
Once I have this differentially expressed gene list, I then use the complex model to estimate domain-specific developmental slopes.

From this analysis, I have identified **`r length(unique(b6_domain_main_effects_clean$gene_id))`** genes with domain main effects, **`r length(unique(b6_stage_main_effects_clean$gene_id))`** genes with stage main effects, and  **`r length(genes_with_interaction)`** genes with significant domain × stage interactions. I used `DEgene_pattern_identification_B6.R` script to classify these differentially expressed genes into sub-categories based on expression patterns.

- **Domain main effects**: Genes with different expression levels between domains but parallel developmental trajectories.
- **Stage main effects**: Genes with common developmental changes across all domains.
- **Trajectory patterns**: Genes with different developmental slopes between domains (interactions).

```{r}
#| label: summary-table
#| warning: false
#| message: false

pattern_summary |>
  group_by(Category) |>
  summarize(
    `Total Patterns` = n(),
    `Total Genes` = sum(N_Genes),
    .groups = "drop"
  ) |>
  create_dt()
```


# Domain Main Effects

Here I am highlighting genes that show differences in expression between A-P domains in B6.
In total there are **`r length(unique(b6_domain_main_effects_clean$gene_id))`** unique genes with a domain effect (i.e. differentially expressed between any two domains) that doesn't change over developmental time (i.e. maintain parallel developmental trajectories).
This means that while one domain might express a gene at higher or lower levels than another, all domains follow the same developmental pattern over time (all increasing, all decreasing, or all stable).

::: {#B6_DESeq2_domain_effects .panel-tabset .nav-pills}
## Overview

```{r}
#| label: domain-main-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Domain Main Effects") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

## MAX vs PM

```{r}
#| label: max-vs-pm-plot
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 10

genes_MAX_higher_than_PM <- gene_assignments |>
  filter(MAX_higher_PM) |>
  pull(gene_id)

genes_MAX_lower_than_PM <- gene_assignments |>
  filter(MAX_lower_PM) |>
  pull(gene_id)

if (length(genes_MAX_higher_than_PM) > 0 & length(genes_MAX_lower_than_PM) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_MAX_higher_than_PM,
    genes_down = genes_MAX_lower_than_PM,
    domain_name = "MAX vs PM",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**MAX higher than PM** contains `r length(genes_MAX_higher_than_PM)` genes and are over-representation analysis shows a mature neuronal phenotype with emphasis on synaptic connectivity, neurotransmission, and neural circuit function.


**MAX lower than PM** contains `r length(genes_MAX_lower_than_PM)` genes and over-representation analysis shows that it contains a developmental/embryonic phenotype dominated by morphogenetic processes, particularly skeletal patterning, organogenesis (kidney, heart, limb), and TGF-β superfamily signaling.

## POST vs PM

```{r}
#| label: post-vs-pm-plot
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 10

genes_POST_higher_than_PM <- gene_assignments |>
  filter(POST_higher_PM) |>
  pull(gene_id)

genes_POST_lower_than_PM <- gene_assignments |>
  filter(POST_lower_PM) |>
  pull(gene_id)

if (length(genes_POST_higher_than_PM) > 0 & length(genes_POST_lower_than_PM) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_POST_higher_than_PM,
    genes_down = genes_POST_lower_than_PM,
    domain_name = "POST vs PM",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**POST higher than PM** contains `r length(genes_POST_higher_than_PM)` genes and over-representation analysis shows that it exhibits a mature functional phenotype (synaptic transmission, ion channels, behavior) combined with a notable muscle/cardiac signature (sarcomere, myofibrils, cardiac conduction).


**POST lower than PM** contains `r length(genes_POST_lower_than_PM)` genes and over-representation analysis shows that they are enriched for early developmental processes including embryonic patterning (Wnt, Hedgehog/Gli signaling), organogenesis (tooth, kidney, skeleton), and epithelial differentiation (skin, desmosomes).

## POST vs MAX

```{r}
#| label: post-vs-max-plot
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 10

genes_POST_higher_than_MAX <- gene_assignments |>
  filter(POST_higher_MAX) |>
  pull(gene_id)

genes_POST_lower_than_MAX <- gene_assignments |>
  filter(POST_lower_MAX) |>
  pull(gene_id)

if (length(genes_POST_higher_than_MAX) > 0 & length(genes_POST_lower_than_MAX) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_POST_higher_than_MAX,
    genes_down = genes_POST_lower_than_MAX,
    domain_name = "POST vs MAX",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**POST higher than MAX** contain `r length(genes_POST_higher_than_MAX)` genes and over-representation analysis shows an enrichment for **muscle/contractile apparatus** genes—sarcomere components, myofibrils, and muscle contraction machinery.
The strong human phenotype associations with myopathies (Gowers sign, muscle weakness, abnormal muscle fibers) underscore this muscle identity.
There's also a cell adhesion signature involving protocadherins.


**POST lower than MAX** contain `r length(genes_POST_lower_than_MAX)` genes and over-representation analysis shows that it displays a mixed **neural + epithelial** signature. The neural component includes synapse organization and axon guidance, while the epithelial component features epidermis development, desmosomes, and cornified envelope.
Wnt signaling is prominent, along with developmental processes (cell fate commitment, kidney/nephron development).


## Highest in Each Domain

```{r}
#| label: combined-highest-genes
#| warning: false
#| message: false
#| results: hide

genes_PM_highest <- gene_assignments |>
  filter(Highest_PM) |>
  pull(gene_id)

genes_MAX_highest <- gene_assignments |>
  filter(Highest_MAX) |>
  pull(gene_id)

genes_POST_highest <- gene_assignments |>
  filter(Highest_POST) |>
  pull(gene_id)

if (length(genes_PM_highest) > 0) {
  n_show <- ifelse(length(genes_PM_highest) < 2, length(genes_PM_highest), 1)
  p1 <- plot_domain_hierarchy(
    genes_highest = genes_PM_highest,
    domain_name = "PM",
    count_data = b6_gene_counts,
    n = n_show
  )
}

if (length(genes_MAX_highest) > 0) {
  n_show <- ifelse(length(genes_MAX_highest) < 2, length(genes_MAX_highest), 1)
  p2 <- plot_domain_hierarchy(
    genes_highest = genes_MAX_highest,
    domain_name = "MAX",
    count_data = b6_gene_counts,
    n = n_show
  )
}

if (length(genes_POST_highest) > 0) {
  n_show <- ifelse(length(genes_POST_highest) < 2, length(genes_POST_highest), 1)
  p3 <- plot_domain_hierarchy(
    genes_highest = genes_POST_highest,
    domain_name = "POST",
    count_data = b6_gene_counts,
    n = n_show
  )
}
```

`r length(genes_PM_highest)` genes with highest expression at the PM compared to both MAX and POST domains.

`r length(genes_MAX_highest)` genes showing highest expression in MAX compared to both PM and POST domains.

`r length(genes_POST_highest)` genes with peak expression in POST in comparison to MAX and PM domains.

```{r}
#| label: combined-highest-plot
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 14

wrap_plots(p1, p2, p3, nrow = 3, ncol = 1, guides = "collect") & theme(legend.position = "bottom")
```


## A-P Gradients

```{r}
#| label: ap-gradients-plot
#| warning: false
#| message: false

genes_AP_decreasing <- gene_assignments |>
  filter(AP_decreasing) |>
  pull(gene_id)

genes_AP_increasing <- gene_assignments |>
  filter(AP_increasing) |>
  pull(gene_id)

if (length(genes_AP_decreasing) > 0 | length(genes_AP_increasing) > 0) {
  n_each <- 3
  if (length(genes_AP_decreasing) < 15) n_each <- min(n_each, length(genes_AP_decreasing))
  if (length(genes_AP_increasing) < 15) n_each <- min(n_each, length(genes_AP_increasing))

  plot_AP_gradient(
    genes_decreasing_with_development = genes_AP_decreasing,
    genes_increasing_with_development = genes_AP_increasing,
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

These genes show monotonic gradients along the A-P axis.

-   **A→P decreasing** (PM \> MAX \> POST): `r length(genes_AP_decreasing)` genes.
-   **A→P increasing** (POST \> MAX \> PM): `r length(genes_AP_increasing)` genes.

## Gene Tables


### MAX vs PM Genes
```{r}
#| label: max-vs-pm-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "MAX vs PM") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

### POST vs PM Genes

```{r}
#| label: post-vs-pm-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "POST vs PM") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

### POST vs MAX Genes

```{r}
#| label: post-vs-max-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "POST vs MAX") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```




### Highest in Each Domain Genes

```{r}
#| label: combined-highest-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% genes_PM_highest |
          gene_id %in% genes_MAX_highest |
           gene_id %in% genes_POST_highest ) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  distinct(gene_id, .keep_all = TRUE) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```



### A-P Gradients Genes
```{r}
#| label: ap-gradients-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% c(genes_AP_decreasing, genes_AP_increasing)) |>
  mutate(Gradient = ifelse(gene_id %in% genes_AP_decreasing, "A→P decreasing", "A→P increasing")) |>
  select(gene_id, symbol, chromosome, Gradient, Comparison, log2FoldChange, padj) |>
  arrange(Gradient, padj) |>
  create_dt()
```

## Over-representation Analysis Results

```{r}
#| label: domain-main-combined-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_MAX_higher_than_PM_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_MAX_higher_than_PM,
  description = "MAX higher than PM"
)

genes_MAX_lower_than_PM_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_MAX_lower_than_PM,
  description = "MAX lower than PM"
)

genes_POST_higher_than_PM_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_POST_higher_than_PM,
  description = "POST higher than PM"
)

genes_POST_lower_than_PM_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_POST_lower_than_PM,
  description = "POST lower than PM"
)

genes_POST_higher_than_MAX_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_POST_higher_than_MAX,
  description = "POST higher than MAX"
)

genes_POST_lower_than_MAX_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_POST_lower_than_MAX,
  description = "POST lower than MAX"
)

genes_PM_highest_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_PM_highest,
  description = "PM highest"
)

genes_MAX_highest_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_MAX_highest,
  description = "MAX highest"
)

genes_POST_highest_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_POST_highest,
  description = "POST highest"
)

genes_AP_decreasing_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_AP_decreasing,
  description = "A→P decreasing"
)
genes_AP_increasing_ora <- run_enrichment_hybrid_v2(
  gene_ids = genes_AP_increasing,
  description = "A→P increasing"
)
```


```{r}
#| label: domain-main-combined-ora-table
#| warning: false
#| message: false

genes_MAX_higher_than_PM_ora$result |>
  bind_rows(
    genes_MAX_lower_than_PM_ora$result
  ) |>
  bind_rows(
    genes_POST_higher_than_PM_ora$result
  ) |>
  bind_rows(
    genes_POST_lower_than_PM_ora$result
  ) |>
  bind_rows(
    genes_POST_higher_than_MAX_ora$result
  ) |>
  bind_rows(
    genes_POST_lower_than_MAX_ora$result
  ) |>
  bind_rows( 
    genes_PM_highest_ora$result 
         )|>
  bind_rows( 
    genes_MAX_highest_ora$result 
         )|>
  bind_rows( 
    genes_POST_highest_ora$result 
         )|>
  bind_rows( 
    genes_AP_decreasing_ora$result 
         )|>
  bind_rows(
    genes_AP_increasing_ora$result
  ) |>
  filter(p_value < 0.01) |>
  create_dt()
```
:::

# Stage Main Effects

These genes show developmental changes that are consistent across all three domains.
All domains increase or decrease together, representing shared developmental programs.

::: {#B6_DESeq2_stage_effects .panel-tabset .nav-pills}
## Overview

```{r}
#| label: stage-main-overview
#| warning: false
#| message: false

genes_increasing_with_development <- b6_stage_main_effects_clean |>
  filter(Direction == "Increase") |>
  pull(gene_id)

genes_decreasing_with_development <- b6_stage_main_effects_clean |>
  filter(Direction == "Decrease") |>
  pull(gene_id)

strong_increasers <- b6_stage_main_effects_clean |>
  filter(Direction == "Increase", abs(log2FoldChange) > 1) |>
  pull(gene_id)

strong_decreasers <- b6_stage_main_effects_clean |>
  filter(Direction == "Decrease", abs(log2FoldChange) > 1) |>
  pull(gene_id)

pattern_summary |>
  filter(Category == "Stage Main Effects") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

## Increasing with Development

```{r}
#| label: dev-increasing-plot
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 14

if (length(genes_increasing_with_development) > 0) {
  n_show <- ifelse(length(genes_increasing_with_development) < 15, length(genes_increasing_with_development), 6)
  plot_simple_trajectories(
    gene_ids = genes_increasing_with_development,
    count_data = b6_gene_counts,
    title_text = paste0("Increasing with Development (n=", length(genes_increasing_with_development), ")"),
    n = n_show
  )
}
```

**`r length(genes_increasing_with_development)`** genes increasing across all domains as development progresses showing an over-representation of genes involved in ciliary assembly, extracellular matrix, immune response, epithelial differentiation and ion transport.


## Decreasing with Development

```{r}
#| label: dev-decreasing-plot
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 14

if (length(genes_decreasing_with_development) > 0) {
  n_show <- ifelse(length(genes_decreasing_with_development) < 15, length(genes_decreasing_with_development), 6)
  plot_simple_trajectories(
    gene_ids = genes_decreasing_with_development,
    count_data = b6_gene_counts,
    title_text = paste0("Decreasing with Development (n=", length(genes_decreasing_with_development), ")"),
    n = n_show
  )
}
```

**`r length(genes_decreasing_with_development)`** genes decreasing across all domains that are over-represented for genes involved in erythropoiesis and early hematopoietic processes.

## Gene Lists

```{r}
#| label: dev-decreasing-table
#| warning: false
#| message: false

b6_stage_main_effects_clean |>
  filter(Direction == "Decrease" | Direction == "Increase") |>
  select(gene_id, symbol, chromosome, Direction, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()

```

## Over-representation Analysis Results

```{r}
#| label: dev-combined-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run ora on all sets
dev_decreasing_ora <- run_enrichment_hybrid_v2(
  gene_ids = b6_stage_main_effects_clean |>
    filter(Direction == "Decrease") |>
    pull(gene_id),
  description = "Decrease"
)

dev_increasing_ora <- run_enrichment_hybrid_v2(
  gene_ids = b6_stage_main_effects_clean |>
    filter(Direction == "Increase") |>
    pull(gene_id),
  description = "Increase"
)

```

```{r}
#| label: dev-combined-ora-table
#| warning: false
#| message: false

dev_increasing_ora$result |>
  bind_rows(
    dev_decreasing_ora$result
  ) |> 
  filter(p_value < 0.01) |>
  create_dt()

```


:::

# Developmental Trajectory Patterns Overview

These genes show significant **domain × stage interactions**, meaning their developmental trajectories differ between domains. I classify them by **what the slopes are actually doing** rather than just whether domains are converging or diverging.

**Pattern types:**

- **Opposite trajectories**: One domain increases while another decreases - the most biologically distinct patterns
- **Co-increasing/Co-decreasing**: Both domains change in the same direction but at different rates
- **Domain-restricted**: Only one or two domains show developmental changes


```{r}
#| label: trajectory-overview
#| warning: false
#| message: false

trajectory_patterns |>
  count(overall_pattern, name = "n_genes") |>
  arrange(desc(n_genes)) |>
  create_dt()
```

```{r}
#| label: trajectory-overview-plot
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6

trajectory_summary <- trajectory_patterns |>
  count(overall_pattern) |>
  filter(n > 0) |>
  mutate(overall_pattern = fct_reorder(overall_pattern, n))

ggplot(trajectory_summary, aes(x = n, y = overall_pattern, fill = overall_pattern)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "Distribution of Developmental Trajectory Patterns",
    subtitle = "Genes with significant domain × stage interactions",
    x = "Number of genes", y = ""
  ) +
  theme_pubclean() +
  theme(axis.text.y = element_text(size = 9)) +
  xlim(0, max(trajectory_summary$n) * 1.15)
```

# Pairwise Comparisons Between Domains

::: {#B6_trajectory_patterns .panel-tabset .nav-pills}

## MAX vs PM Trajectories

```{r}
#| label: max-pm-opposite-up
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_MAX_up_PM_down) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_MAX_up_PM_down, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Opposite: MAX↑ PM↓ (n=", length(genes_MAX_up_PM_down), ")"), n = 2
  )
}
```

```{r}
#| label: max-pm-opposite-down
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_MAX_down_PM_up) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_MAX_down_PM_up, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Opposite: MAX↓ PM↑ (n=", length(genes_MAX_down_PM_up), ")"), n = 2
  )
}
```

```{r}
#| label: max-pm-coinc-max-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_coinc_MAX_faster_PM) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_coinc_MAX_faster_PM, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-increasing: MAX faster (n=", length(genes_coinc_MAX_faster_PM), ")"), n = 2
  )
}
```

```{r}
#| label: max-pm-coinc-pm-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_coinc_PM_faster_MAX) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_coinc_PM_faster_MAX, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-increasing: PM faster (n=", length(genes_coinc_PM_faster_MAX), ")"), n = 2
  )
}
```

```{r}
#| label: max-pm-codec-max-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_codec_MAX_faster_PM) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_codec_MAX_faster_PM, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-decreasing: MAX faster (n=", length(genes_codec_MAX_faster_PM), ")"), n = 2
  )
}
```

```{r}
#| label: max-pm-codec-pm-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_codec_PM_faster_MAX) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_codec_PM_faster_MAX, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-decreasing: PM faster (n=", length(genes_codec_PM_faster_MAX), ")"), n = 2
  )
}
```

## POST vs PM Trajectories


```{r}
#| label: post-pm-opposite-up
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_POST_up_PM_down) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_POST_up_PM_down, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Opposite: POST↑ PM↓ (n=", length(genes_POST_up_PM_down), ")"), n = 2
  )
}
```

```{r}
#| label: post-pm-opposite-down
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_POST_down_PM_up) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_POST_down_PM_up, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Opposite: POST↓ PM↑ (n=", length(genes_POST_down_PM_up), ")"), n = 2
  )
}
```

```{r}
#| label: post-pm-coinc-post-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_coinc_POST_faster_PM) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_coinc_POST_faster_PM, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-increasing: POST faster (n=", length(genes_coinc_POST_faster_PM), ")"), n = 2
  )
}
```

```{r}
#| label: post-pm-coinc-pm-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_coinc_PM_faster_POST) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_coinc_PM_faster_POST, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-increasing: PM faster (n=", length(genes_coinc_PM_faster_POST), ")"), n = 2
  )
}
```

```{r}
#| label: post-pm-codec-post-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_codec_POST_faster_PM) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_codec_POST_faster_PM, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-decreasing: POST faster (n=", length(genes_codec_POST_faster_PM), ")"), n = 2
  )
}
```

```{r}
#| label: post-pm-codec-pm-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_codec_PM_faster_POST) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_codec_PM_faster_POST, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-decreasing: PM faster (n=", length(genes_codec_PM_faster_POST), ")"), n = 2
  )
}
```

## POST vs MAX Trajectories


```{r}
#| label: post-max-opposite-up
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_POST_up_MAX_down) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_POST_up_MAX_down, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Opposite: POST↑ MAX↓ (n=", length(genes_POST_up_MAX_down), ")"), n = 2
  )
}
```

```{r}
#| label: post-max-opposite-down
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_POST_down_MAX_up) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_POST_down_MAX_up, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Opposite: POST↓ MAX↑ (n=", length(genes_POST_down_MAX_up), ")"), n = 2
  )
}
```


```{r}
#| label: post-max-coinc-post-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_coinc_POST_faster_MAX) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_coinc_POST_faster_MAX, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-increasing: POST faster (n=", length(genes_coinc_POST_faster_MAX), ")"), n = 2
  )
}
```


```{r}
#| label: post-max-coinc-max-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_coinc_MAX_faster_POST) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_coinc_MAX_faster_POST, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-increasing: MAX faster (n=", length(genes_coinc_MAX_faster_POST), ")"), n = 2
  )
}
```

```{r}
#| label: post-max-codec-post-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_codec_POST_faster_MAX) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_codec_POST_faster_MAX, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-decreasing: POST faster (n=", length(genes_codec_POST_faster_MAX), ")"), n = 2
  )
}
```

```{r}
#| label: post-max-codec-max-faster
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 4

if (length(genes_codec_MAX_faster_POST) > 0) {
  plot_trajectory_with_slopes(
    gene_ids = genes_codec_MAX_faster_POST, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes, title_text = paste0("Co-decreasing: MAX faster (n=", length(genes_codec_MAX_faster_POST), ")"), n = 2
  )
}
```

## Gene Tables

### MAX vs PM Genes

```{r}
#| label: max-pm-table
#| warning: false
#| message: false

trajectory_patterns |>
  filter(!MAX_PM_pattern %in% c("Both stable", "Unclassified")) |>
  select(gene_id, symbol, MAX_PM_pattern,
    PM_slope = PM_log2FoldChange, MAX_slope = MAX_log2FoldChange,
    PM_padj, MAX_padj
  ) |>
  mutate(across(c(PM_slope, MAX_slope), ~ round(., 3))) |>
  arrange(MAX_PM_pattern, desc(abs(MAX_slope - PM_slope))) |>
  create_dt()
```

### POST vs PM Genes

```{r}
#| label: post-pm-table
#| warning: false
#| message: false

trajectory_patterns |>
  filter(!POST_PM_pattern %in% c("Both stable", "Unclassified")) |>
  select(gene_id, symbol, POST_PM_pattern,
    PM_slope = PM_log2FoldChange, POST_slope = POST_log2FoldChange,
    PM_padj, POST_padj
  ) |>
  mutate(across(c(PM_slope, POST_slope), ~ round(., 3))) |>
  arrange(POST_PM_pattern, desc(abs(POST_slope - PM_slope))) |>
  create_dt()
```

### POST vs MAX Genes

```{r}
#| label: post-max-table
#| warning: false
#| message: false

trajectory_patterns |>
  filter(!POST_MAX_pattern %in% c("Both stable", "Unclassified")) |>
  select(gene_id, symbol, POST_MAX_pattern,
    MAX_slope = MAX_log2FoldChange, POST_slope = POST_log2FoldChange,
    MAX_padj, POST_padj
  ) |>
  mutate(across(c(MAX_slope, POST_slope), ~ round(., 3))) |>
  arrange(POST_MAX_pattern, desc(abs(POST_slope - MAX_slope))) |>
  create_dt()
```

## Over-representation Analysis Results

```{r}
#| label: trajectory-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

ora_MAX_up_PM_down <- run_enrichment_hybrid_v2(gene_ids = genes_MAX_up_PM_down, description = "MAX↑ PM↓")
ora_MAX_down_PM_up <- run_enrichment_hybrid_v2(gene_ids = genes_MAX_down_PM_up, description = "MAX↓ PM↑")
ora_POST_up_PM_down <- run_enrichment_hybrid_v2(gene_ids = genes_POST_up_PM_down, description = "POST↑ PM↓")
ora_POST_down_PM_up <- run_enrichment_hybrid_v2(gene_ids = genes_POST_down_PM_up, description = "POST↓ PM↑")
ora_coinc_MAX_faster <- run_enrichment_hybrid_v2(gene_ids = genes_coinc_MAX_faster_PM, description = "Co-increasing: MAX faster")
ora_coinc_POST_faster <- run_enrichment_hybrid_v2(gene_ids = genes_coinc_POST_faster_PM, description = "Co-increasing: POST faster")
```

```{r}
#| label: trajectory-ora-table
#| warning: false
#| message: false

trajectory_ora_results <- bind_rows(
  ora_MAX_up_PM_down$result, ora_MAX_down_PM_up$result,
  ora_POST_up_PM_down$result, ora_POST_down_PM_up$result,
  ora_coinc_MAX_faster$result, ora_coinc_POST_faster$result
)

if (!is.null(trajectory_ora_results) && nrow(trajectory_ora_results) > 0) {
  trajectory_ora_results |>
    filter(p_value < 0.01) |>
    create_dt()
}
```

:::

# Domain-Specific Dynamics

Genes showing significant developmental changes in **only specific domains** while others remain stable.

::: {#B6_domain_dynamics .panel-tabset .nav-pills}

## Overview

```{r}
#| label: domain-dynamics-overview
#| warning: false
#| message: false

tibble(
  Pattern = c(
    "PM↑ only", "PM↓ only", "MAX↑ only", "MAX↓ only", "POST↑ only", "POST↓ only",
    "PM+MAX↑", "PM+MAX↓", "PM+POST↑", "PM+POST↓", "MAX+POST↑", "MAX+POST↓"
  ),
  N_Genes = c(
    length(genes_PM_only_up), length(genes_PM_only_down),
    length(genes_MAX_only_up), length(genes_MAX_only_down),
    length(genes_POST_only_up), length(genes_POST_only_down),
    length(genes_PM_MAX_up), length(genes_PM_MAX_down),
    length(genes_PM_POST_up), length(genes_PM_POST_down),
    length(genes_MAX_POST_up), length(genes_MAX_POST_down)
  )
) |>
  filter(N_Genes > 0) |>
  create_dt()
```

## PM Only Dynamic

`r length(genes_PM_only_up)` genes increase and `r length(genes_PM_only_down)` genes decrease specifically in PM while MAX and POST remain stable.

```{r}
#| label: pm-only-examples
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 10

plots <- list()

if (length(genes_PM_only_up) > 0) {
  plots[["PM↑"]] <- plot_domain_specific_dynamics(
    gene_ids = genes_PM_only_up, domain_name = "PM", direction = "↑",
    count_data = b6_gene_counts, slope_data = b6_domain_slopes, n = 1
  )
}

if (length(genes_PM_only_down) > 0) {
  plots[["PM↓"]] <- plot_domain_specific_dynamics(
    gene_ids = genes_PM_only_down, domain_name = "PM", direction = "↓",
    count_data = b6_gene_counts, slope_data = b6_domain_slopes, n = 1
  )
}

if (length(plots) > 0) wrap_plots(plots, nrow = 2)
```



## MAX Only Dynamic

`r length(genes_MAX_only_up)` genes increase and `r length(genes_MAX_only_down)` genes decrease specifically in MAX while PM and POST remain stable.

```{r}
#| label: max-only-examples
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 10

plots <- list()

if (length(genes_MAX_only_up) > 0) {
  plots[["MAX↑"]] <- plot_domain_specific_dynamics(
    gene_ids = genes_MAX_only_up, domain_name = "MAX", direction = "↑",
    count_data = b6_gene_counts, slope_data = b6_domain_slopes, n = 1
  )
}

if (length(genes_MAX_only_down) > 0) {
  plots[["MAX↓"]] <- plot_domain_specific_dynamics(
    gene_ids = genes_MAX_only_down, domain_name = "MAX", direction = "↓",
    count_data = b6_gene_counts, slope_data = b6_domain_slopes, n = 1
  )
}

if (length(plots) > 0) wrap_plots(plots, nrow = 2)
```



## POST Only Dynamic

`r length(genes_POST_only_up)` genes increase and `r length(genes_POST_only_down)` genes decrease specifically in POST while PM and MAX remain stable.

```{r}
#| label: post-only-examples
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 10

plots <- list()

if (length(genes_POST_only_up) > 0) {
  plots[["POST↑"]] <- plot_domain_specific_dynamics(
    gene_ids = genes_POST_only_up, domain_name = "POST", direction = "↑",
    count_data = b6_gene_counts, slope_data = b6_domain_slopes, n = 1
  )
}

if (length(genes_POST_only_down) > 0) {
  plots[["POST↓"]] <- plot_domain_specific_dynamics(
    gene_ids = genes_POST_only_down, domain_name = "POST", direction = "↓",
    count_data = b6_gene_counts, slope_data = b6_domain_slopes, n = 1
  )
}

if (length(plots) > 0) wrap_plots(plots, nrow = 2)
```


## Two Domains Dynamic

```{r}
#| label: two-domain-examples
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8

plots <- list()

if (length(genes_PM_MAX_up) > 0) {
  plots[["PM+MAX↑"]] <- plot_simple_trajectories(
    gene_ids = genes_PM_MAX_up, count_data = b6_gene_counts,
    title_text = paste0("PM + MAX↑, POST stable (n=", length(genes_PM_MAX_up), ")"), n = 1
  )
}

if (length(genes_PM_MAX_down) > 0) {
  plots[["PM+MAX↓"]] <- plot_simple_trajectories(
    gene_ids = genes_PM_MAX_down, count_data = b6_gene_counts,
    title_text = paste0("PM + MAX↓, POST stable (n=", length(genes_PM_MAX_down), ")"), n = 1
  )
}

if (length(genes_MAX_POST_up) > 0) {
  plots[["MAX+POST↑"]] <- plot_simple_trajectories(
    gene_ids = genes_MAX_POST_up, count_data = b6_gene_counts,
    title_text = paste0("MAX + POST↑, PM stable (n=", length(genes_MAX_POST_up), ")"), n = 1
  )
}

if (length(genes_MAX_POST_down) > 0) {
  plots[["MAX+POST↓"]] <- plot_simple_trajectories(
    gene_ids = genes_MAX_POST_down, count_data = b6_gene_counts,
    title_text = paste0("MAX + POST↓, PM stable (n=", length(genes_MAX_POST_down), ")"), n = 1
  )
}

if (length(genes_PM_POST_up) > 0) {
  plots[["PM+POST↑"]] <- plot_simple_trajectories(
    gene_ids = genes_PM_POST_up, count_data = b6_gene_counts,
    title_text = paste0("PM + POST↑, MAX stable (n=", length(genes_PM_POST_up), ")"), n = 1
  )
}

if (length(genes_PM_POST_down) > 0) {
  plots[["PM+POST↓"]] <- plot_simple_trajectories(
    gene_ids = genes_PM_POST_down, count_data = b6_gene_counts,
    title_text = paste0("PM + POST↓, MAX stable (n=", length(genes_PM_POST_down), ")"), n = 1
  )
}

if (length(plots) > 0) {
  wrap_plots(plots, ncol = 3, guides ="collect" ) & theme(legend.position = "bottom")
  }else NULL


```

## Gene Lists

```{r}
#| label: domain-dynamics-gene-lists
#| warning: false
#| message: false

gene_list_tbl <- bind_rows(
  if (length(genes_PM_only_up) > 0) tibble(Pattern = "PM↑ only", gene_id = genes_PM_only_up),
  if (length(genes_PM_only_down) > 0) tibble(Pattern = "PM↓ only", gene_id = genes_PM_only_down),
  if (length(genes_MAX_only_up) > 0) tibble(Pattern = "MAX↑ only", gene_id = genes_MAX_only_up),
  if (length(genes_MAX_only_down) > 0) tibble(Pattern = "MAX↓ only", gene_id = genes_MAX_only_down),
  if (length(genes_POST_only_up) > 0) tibble(Pattern = "POST↑ only", gene_id = genes_POST_only_up),
  if (length(genes_POST_only_down) > 0) tibble(Pattern = "POST↓ only", gene_id = genes_POST_only_down),
  if (length(genes_PM_MAX_up) > 0) tibble(Pattern = "PM+MAX↑", gene_id = genes_PM_MAX_up),
  if (length(genes_PM_MAX_down) > 0) tibble(Pattern = "PM+MAX↓", gene_id = genes_PM_MAX_down),
  if (length(genes_MAX_POST_up) > 0) tibble(Pattern = "MAX+POST↑", gene_id = genes_MAX_POST_up),
  if (length(genes_MAX_POST_down) > 0) tibble(Pattern = "MAX+POST↓", gene_id = genes_MAX_POST_down),
  if (length(genes_PM_POST_up) > 0) tibble(Pattern = "PM+POST↑", gene_id = genes_PM_POST_up),
  if (length(genes_PM_POST_down) > 0) tibble(Pattern = "PM+POST↓", gene_id = genes_PM_POST_down)
)

if (nrow(gene_list_tbl) > 0) {
  gene_list_tbl |>
    left_join(b6_domain_slopes_clean |> distinct(gene_id, symbol)) |>
    select(Pattern, gene_id, symbol) |>
    arrange(Pattern, symbol) |>
    create_dt()
}
```

## Over-representation Analysis Results

```{r}
#| label: combined-domain-specific-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

ora_PM_only_up <- run_enrichment_hybrid_v2(gene_ids = genes_PM_only_up, 
                                           description = "PM↑ only")
ora_PM_only_down <- run_enrichment_hybrid_v2(gene_ids = genes_PM_only_down, 
                                             description = "PM↓ only")
ora_MAX_only_up <- run_enrichment_hybrid_v2(gene_ids = genes_MAX_only_up, 
                                            description = "MAX↑ only")
ora_MAX_only_down <- run_enrichment_hybrid_v2(gene_ids = genes_MAX_only_down, 
                                              description = "MAX↓ only")
ora_POST_only_up <- run_enrichment_hybrid_v2(gene_ids = genes_POST_only_up, 
                                             description = "POST↑ only")
ora_POST_only_down <- run_enrichment_hybrid_v2(gene_ids = genes_POST_only_down, 
                                               description = "POST↓ only")
ora_PM_MAX_up <- run_enrichment_hybrid_v2(gene_ids = genes_PM_MAX_up, 
                                          description = "PM+MAX↑")
ora_MAX_POST_up <- run_enrichment_hybrid_v2(gene_ids = genes_MAX_POST_up, 
                                            description = "MAX+POST↑")
ora_PM_POST_up <- run_enrichment_hybrid_v2(gene_ids = genes_PM_POST_up, 
                                           description = "PM+POST↑")

```


```{r}
#| label: combined-domain-specific-ora-table
#| warning: false
#| message: false

domain_specific_ora_results <- bind_rows(
  safe_result(ora_PM_only_up), 
  safe_result(ora_PM_only_down),
  safe_result(ora_MAX_only_up), 
  safe_result(ora_MAX_only_down),
  safe_result(ora_POST_only_up), 
  safe_result(ora_POST_only_down),
  safe_result(ora_PM_MAX_up), 
  safe_result(ora_MAX_POST_up), 
  safe_result(ora_PM_POST_up)
)

if (!is.null(domain_specific_ora_results) && nrow(domain_specific_ora_results) > 0) {
  domain_specific_ora_results |>
    filter(p_value < 0.01) |>
    create_dt()
}
```



:::

# Complex Trajectory Patterns

Genes with particularly complex spatiotemporal dynamics including **all domains changing** (at different rates) or **opposite trajectories** across domain pairs.

::: {#B6_complex .panel-tabset .nav-pills}

## Overview

```{r}
#| label: complex-overview
#| warning: false
#| message: false

tibble(
  Pattern = c(
    "All domains↑ (different rates)", "All domains↓ (different rates)",
    "Any opposite slopes", "Three-way divergent"
  ),
  N_Genes = c(
    length(genes_all_increasing), length(genes_all_decreasing),
    length(genes_opposite_slopes), length(genes_three_way_divergent)
  )
) |> create_dt()
```

## Trajectory Examples

`r length(genes_all_increasing)` genes increase and `r length(genes_all_decreasing)` genes decrease in all three domains (at different rates). `r length(genes_opposite_slopes)` genes have at least one domain pair with opposite trajectories, and `r length(genes_three_way_divergent)` genes show significant but different slopes across all three domains.

```{r}
#| label: complex-examples
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 8

plots <- list()

if (length(genes_all_increasing) > 0) {
  plots[["All↑"]] <- plot_trajectory_with_slopes(
    gene_ids = genes_all_increasing, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes,
    title_text = paste0("All Domains↑ (n=", length(genes_all_increasing), ")"), n = 1
  )
}

if (length(genes_all_decreasing) > 0) {
  plots[["All↓"]] <- plot_trajectory_with_slopes(
    gene_ids = genes_all_decreasing, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes,
    title_text = paste0("All Domains↓ (n=", length(genes_all_decreasing), ")"), n = 1
  )
}

if (length(genes_opposite_slopes) > 0) {
  plots[["Opposite"]] <- plot_trajectory_with_slopes(
    gene_ids = genes_opposite_slopes, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes,
    title_text = paste0("Opposite Slopes (n=", length(genes_opposite_slopes), ")"), n = 1
  )
}

if (length(genes_three_way_divergent) > 0) {
  plots[["3-way"]] <- plot_trajectory_with_slopes(
    gene_ids = genes_three_way_divergent, count_data = b6_gene_counts,
    slope_data = b6_domain_slopes,
    title_text = paste0("Three-Way Divergent (n=", length(genes_three_way_divergent), ")"), n = 1
  )
}

if (length(plots) > 0) wrap_plots(plots, ncol = 2, guides = "collect") & theme(legend.position = "bottom")
```

## Gene Lists

```{r}
#| label: complex-gene-lists
#| warning: false
#| message: false

gene_list_tbl <- bind_rows(
  if (length(genes_all_increasing) > 0) tibble(Pattern = "All domains↑", gene_id = genes_all_increasing),
  if (length(genes_all_decreasing) > 0) tibble(Pattern = "All domains↓", gene_id = genes_all_decreasing),
  if (length(genes_opposite_slopes) > 0) tibble(Pattern = "Opposite slopes", gene_id = genes_opposite_slopes),
  if (length(genes_three_way_divergent) > 0) tibble(Pattern = "Three-way divergent", gene_id = genes_three_way_divergent)
)

if (nrow(gene_list_tbl) > 0) {
  gene_list_tbl |>
    left_join(
      trajectory_patterns |> distinct(gene_id, symbol, PM_log2FoldChange, MAX_log2FoldChange, POST_log2FoldChange),
      by = "gene_id"
    ) |>
    mutate(across(c(PM_log2FoldChange, MAX_log2FoldChange, POST_log2FoldChange), ~ round(., 2))) |>
    rename(PM = PM_log2FoldChange, MAX = MAX_log2FoldChange, POST = POST_log2FoldChange) |>
    select(Pattern, gene_id, symbol, PM, MAX, POST) |>
    arrange(Pattern, symbol) |>
    create_dt()
}
```

## Over-representation Analysis Results

```{r}
#| label: complex-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

ora_all_increasing <- run_enrichment_hybrid_v2(gene_ids = genes_all_increasing, 
                                               description = "All↑")
ora_all_decreasing <- run_enrichment_hybrid_v2(gene_ids = genes_all_decreasing, description = "All↓")
ora_opposite <- run_enrichment_hybrid_v2(gene_ids = genes_opposite_slopes, description = "Opposite slopes")
ora_three_way <- run_enrichment_hybrid_v2(gene_ids = genes_three_way_divergent, description = "Three-way divergent")
```

```{r}
#| label: complex-ora-table
#| warning: false
#| message: false

complex_ora <- bind_rows(
  safe_result(ora_all_increasing),
  safe_result(ora_all_decreasing),
  safe_result(ora_opposite),
  safe_result(ora_three_way)
)
if (!is.null(complex_ora) && nrow(complex_ora) > 0) {
  complex_ora |>
    filter(p_value < 0.01) |>
    create_dt()
}
```



:::

# Pattern Summary

::: {#B6_pattern_summary .panel-tabset .nav-pills}

## Overall Summary

```{r}
#| label: overall-summary
#| warning: false
#| message: false

# Create comprehensive summary
pattern_summary_all <- tibble(
  Category = c(
    rep("Domain Main Effects", 11),
    rep("Stage Main Effects", 4),
    rep("Trajectory: Opposite", 6),
    rep("Trajectory: Co-increasing", 6),
    rep("Trajectory: Co-decreasing", 6),
    rep("Domain-Specific: Single", 6),
    rep("Domain-Specific: Two", 6),
    rep("Complex", 4)
  ),
  Pattern = c(
    # Domain main effects
    "MAX higher than PM", "MAX lower than PM",
    "POST higher than PM", "POST lower than PM",
    "POST higher than MAX", "POST lower than MAX",
    "Highest in MAX", "Highest in POST", "Highest in PM",
    "A→P decreasing", "A→P increasing",
    # Stage main effects
    "All↑ (common)", "All↓ (common)", "Strong↑", "Strong↓",
    # Opposite trajectories
    "MAX↑ PM↓", "MAX↓ PM↑", "POST↑ PM↓", "POST↓ PM↑", "POST↑ MAX↓", "POST↓ MAX↑",
    # Co-increasing
    "MAX vs PM: MAX faster", "MAX vs PM: PM faster",
    "POST vs PM: POST faster", "POST vs PM: PM faster",
    "POST vs MAX: POST faster", "POST vs MAX: MAX faster",
    # Co-decreasing
    "MAX vs PM: MAX faster↓", "MAX vs PM: PM faster↓",
    "POST vs PM: POST faster↓", "POST vs PM: PM faster↓",
    "POST vs MAX: POST faster↓", "POST vs MAX: MAX faster↓",
    # Single domain
    "PM↑ only", "PM↓ only", "MAX↑ only", "MAX↓ only", "POST↑ only", "POST↓ only",
    # Two domain
    "PM+MAX↑", "PM+MAX↓", "PM+POST↑", "PM+POST↓", "MAX+POST↑", "MAX+POST↓",
    # Complex
    "All↑ (interaction)", "All↓ (interaction)", "Any opposite", "Three-way"
  ),
  N_Genes = c(
    # Domain main effects
    length(genes_MAX_higher_than_PM), length(genes_MAX_lower_than_PM),
    length(genes_POST_higher_than_PM), length(genes_POST_lower_than_PM),
    length(genes_POST_higher_than_MAX), length(genes_POST_lower_than_MAX),
    length(genes_MAX_highest), length(genes_POST_highest), length(genes_PM_highest),
    length(genes_AP_decreasing), length(genes_AP_increasing),
    # Stage main effects
    length(genes_increasing_with_development), length(genes_decreasing_with_development),
    length(strong_increasers), length(strong_decreasers),
    # Opposite
    length(genes_MAX_up_PM_down), length(genes_MAX_down_PM_up),
    length(genes_POST_up_PM_down), length(genes_POST_down_PM_up),
    length(genes_POST_up_MAX_down), length(genes_POST_down_MAX_up),
    # Co-increasing
    length(genes_coinc_MAX_faster_PM), length(genes_coinc_PM_faster_MAX),
    length(genes_coinc_POST_faster_PM), length(genes_coinc_PM_faster_POST),
    length(genes_coinc_POST_faster_MAX), length(genes_coinc_MAX_faster_POST),
    # Co-decreasing
    length(genes_codec_MAX_faster_PM), length(genes_codec_PM_faster_MAX),
    length(genes_codec_POST_faster_PM), length(genes_codec_PM_faster_POST),
    length(genes_codec_POST_faster_MAX),
    length(genes_codec_MAX_faster_POST),
    # Single domain
    length(genes_PM_only_up), length(genes_PM_only_down),
    length(genes_MAX_only_up), length(genes_MAX_only_down),
    length(genes_POST_only_up), length(genes_POST_only_down),
    # Two domain
    length(genes_PM_MAX_up), length(genes_PM_MAX_down),
    length(genes_PM_POST_up), length(genes_PM_POST_down),
    length(genes_MAX_POST_up), length(genes_MAX_POST_down),
    # Complex
    length(genes_all_increasing), length(genes_all_decreasing),
    length(genes_opposite_slopes), length(genes_three_way_divergent)
  )
) |>
  filter(N_Genes > 0)

pattern_summary_all |> create_dt()
```

```{r}
#| label: category-summary-plot
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6

# Summary by category
category_summary <- pattern_summary_all |>
  group_by(Category) |>
  summarize(Total_Genes = sum(N_Genes), N_Patterns = n(), .groups = "drop") |>
  mutate(Category = fct_reorder(Category, Total_Genes))

ggplot(category_summary, aes(x = Total_Genes, y = Category, fill = Category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(Total_Genes, " (", N_Patterns, " patterns)")), hjust = -0.05, size = 3.5) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Gene Counts by Pattern Category",
    subtitle = "Numbers show total genes (number of patterns)",
    x = "Number of genes", y = ""
  ) +
  theme_pubclean() +
  xlim(0, max(category_summary$Total_Genes) * 1.3)
```


## Trajectory Heatmap

```{r}
#| label: trajectory-heatmap
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 12

# Create slope matrix for heatmap
if (nrow(trajectory_patterns) > 10) {
  slope_matrix <- trajectory_patterns |>
    select(gene_id, symbol, PM_log2FoldChange, MAX_log2FoldChange, POST_log2FoldChange) |>
    column_to_rownames("gene_id") |>
    select(-symbol) |>
    as.matrix()

  colnames(slope_matrix) <- c("PM", "MAX", "POST")
  slope_matrix[is.na(slope_matrix)] <- 0

  # Limit to max 200 genes for visibility
  if (nrow(slope_matrix) > 200) {
    # Select genes with highest variance in slopes
    slope_var <- apply(slope_matrix, 1, var, na.rm = TRUE)
    top_genes <- names(sort(slope_var, decreasing = TRUE))[1:200]
    slope_matrix <- slope_matrix[top_genes, ]
  }

  # Create annotation for overall pattern
  row_annot <- trajectory_patterns |>
    filter(gene_id %in% rownames(slope_matrix)) |>
    select(gene_id, overall_pattern) |>
    column_to_rownames("gene_id")

  pheatmap(slope_matrix,
    color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
    breaks = seq(-3, 3, length.out = 101),
    cluster_cols = FALSE,
    show_rownames = FALSE,
    main = "Developmental Slopes by Domain\n(Genes with Interactions)",
    annotation_row = row_annot,
    fontsize = 10
  )
}
```

## Gene-Level Assignments

```{r}
#| label: gene-assignments-table
#| warning: false
#| message: false

# Create comprehensive gene-level table
all_genes_classified <- unique(c(
  b6_domain_main_effects_clean$gene_id,
  b6_stage_main_effects_clean$gene_id,
  trajectory_patterns$gene_id
))

gene_level_summary <- tibble(gene_id = all_genes_classified) |>
  left_join(all_genes_palate |> select(gene_id, symbol, chromosome), by = "gene_id") |>
  mutate(
    # Main effects
    Has_Domain_Effect = gene_id %in% b6_domain_main_effects_clean$gene_id,
    Has_Stage_Effect = gene_id %in% b6_stage_main_effects_clean$gene_id,
    Has_Interaction = gene_id %in% genes_with_interaction,

    # Domain patterns
    Domain_Pattern = case_when(
      gene_id %in% genes_PM_highest ~ "Highest in PM",
      gene_id %in% genes_MAX_highest ~ "Highest in MAX",
      gene_id %in% genes_POST_highest ~ "Highest in POST",
      gene_id %in% genes_AP_decreasing ~ "A→P decreasing",
      gene_id %in% genes_AP_increasing ~ "A→P increasing",
      TRUE ~ NA_character_
    ),

    # Stage direction
    Stage_Direction = case_when(
      gene_id %in% genes_increasing_with_development ~ "Increasing",
      gene_id %in% genes_decreasing_with_development ~ "Decreasing",
      TRUE ~ NA_character_
    )
  ) |>
  # Add trajectory pattern for interaction genes
  left_join(
    trajectory_patterns |> select(gene_id, overall_pattern, MAX_PM_pattern, POST_PM_pattern, POST_MAX_pattern),
    by = "gene_id"
  )

gene_level_summary |>
  select(
    gene_id, symbol, chromosome, Has_Domain_Effect, Has_Stage_Effect, Has_Interaction,
    Domain_Pattern, Stage_Direction, overall_pattern
  ) |>
  create_dt()
```

## Over-representation Results Combined

```{r}
#| label: all-ora-summary
#| warning: false
#| message: false

# Combine all ora results
all_ora_results <- bind_rows(
  # Domain effects
  genes_MAX_higher_than_PM_ora$result,
  genes_MAX_lower_than_PM_ora$result,
  genes_POST_higher_than_PM_ora$result,
  genes_POST_lower_than_PM_ora$result,
  genes_POST_higher_than_MAX_ora$result,
  genes_POST_lower_than_MAX_ora$result,
  genes_PM_highest_ora$result,
  genes_MAX_highest_ora$result,
  genes_POST_highest_ora$result,
  genes_AP_decreasing_ora$result,
  genes_AP_increasing_ora$result,
  # Stage effects
  dev_increasing_ora$result,
  dev_decreasing_ora$result,
  # Trajectory patterns
  ora_MAX_up_PM_down$result,
  ora_MAX_down_PM_up$result,
  ora_POST_up_PM_down$result,
  ora_POST_down_PM_up$result,
  ora_coinc_MAX_faster$result,
  ora_coinc_POST_faster$result,
  # Domain-specific
  ora_PM_only_up$result,
  ora_PM_only_down$result,
  ora_MAX_only_up$result,
  ora_MAX_only_down$result,
  ora_POST_only_up$result,
  ora_POST_only_down$result,
  ora_PM_MAX_up$result,
  ora_MAX_POST_up$result,
  ora_PM_POST_up$result,
  # Complex
  ora_all_increasing$result,
  ora_all_decreasing$result,
  ora_opposite$result,
  ora_three_way$result
)

if (!is.null(all_ora_results) && nrow(all_ora_results) > 0) {
  # Add category
  all_ora_results <- all_ora_results |>
    mutate(
      Category = case_when(
        gene_set %in% c(
          "MAX higher than PM", "MAX lower than PM",
          "POST higher than PM", "POST lower than PM",
          "POST higher than MAX", "POST lower than MAX",
          "PM highest", "MAX highest", "POST highest",
          "A→P decreasing", "A→P increasing"
        ) ~ "Domain Main Effects",
        gene_set %in% c("Increase", "Decrease") ~ "Stage Main Effects",
        gene_set %in% c("MAX↑ PM↓", "MAX↓ PM↑", "POST↑ PM↓", "POST↓ PM↑") ~ "Opposite Trajectories",
        gene_set %in% c("Co-increasing: MAX faster", "Co-increasing: POST faster") ~ "Co-increasing", 
        grepl("only", gene_set) ~ "Domain-Specific",
        grepl("\\+", gene_set) ~ "Two-Domain",
        gene_set %in% c("All↑", "All↓", "Opposite slopes", "Three-way divergent") ~ "Complex",
        TRUE ~ "Other"
      )
    ) |>
    select(Category, gene_set, source, term_name, p_value, intersection_size, term_size) |>
    filter(as.numeric(p_value) < 0.01) |>
    arrange(Category, gene_set, p_value)

  all_ora_results |> create_dt()
}
```

:::


# Over-representation Summary

::: {#B6_ora_summary .panel-tabset .nav-pills}
## Co-increasing

**MAX faster**
- ECM / collagen (dominant signal)
- Bone & skeletal: osteoblast morphology, bone mineral density, ossification, trabecular bone, skeletal physiology
- Skin & epidermis: hair follicle, keratinocyte, stratum granulosum, skin tensile strength
- Motile cilia / axoneme
- Wnt & Hippo signaling; ESC pluripotency
- Heart development, right ventricle hypertrophy, situs inversus / laterality defects
- Cancer pathways (breast, basal cell, gastric)
- Postnatal growth retardation

**POST faster**
- ECM (dominant signal)
- Nail dystrophy, blistering, keratoderma, skin barrier
- Dental abnormalities (hypoplasia, enamel)
- Alopecia
- Motile cilia; situs inversus
- Nociception (thermal, mechanical, chemical thresholds)
- Apical plasma membrane
- Hormone level regulation

---

## Complex

**All↑**
- ECM / collagen (dominant signal)
- Bone & skeletal: osteoclast/osteoblast, trabecular bone, ossification, bone mineral density (both increased and decreased), epiphyseal plate, osteopetrosis
- Muscle: contraction, myofibril, sarcomere, dystrophy, muscle fiber morphology
- Immune & inflammation: leukocyte/granulocyte chemotaxis & migration, cytokine activity, chemokine secretion, susceptibility to infection
- Skin: keratoderma, keratinocyte, cornified envelope, alopecia
- Calcium / ion homeostasis
- Joint contracture & stiffness (multiple joint sites)
- Cardiac: hypertrophy, fibrosis, dilated cardiomyopathy
- Dental abnormalities
- T cell / B cell / macrophage numbers
- Vascular: wound healing, angiogenesis, permeability

**Opposite slopes**
- Cervical vertebrae morphology (atlas, axis)
- Small olfactory bulb
- Neuroactive ligand-receptor interaction
- Organismal stress response

**Three-way divergent**
- NFATc4 transcription factor motif only

---

## Domain Main Effects

**A→P decreasing**
- Wnt signaling (canonical & non-canonical)
- Reproductive tract development (Wolffian duct, Müllerian duct, uterus)
- Hair / vibrissa morphology
- Nephric duct / mesonephric tubule morphogenesis

**MAX higher than PM**
- Synaptic: synaptic membrane, postsynaptic specialization, synapse organization, chemical synaptic transmission
- Neuronal: neuron projection development, growth cone, axon development, neuron differentiation
- Cognition, learning, memory
- Coordination, locomotion, gait
- Ion channels (cation, potassium)
- Seizures, tremors
- NRSF transcription factor motif

**MAX highest**
- Myelin sheath
- Muscle hypoplasia (intercostal, tongue)

**MAX lower than PM**
- Skeletal system development & morphogenesis
- Platelet function (aggregation, mean platelet volume)
- ECM / basement membrane
- Cartilage & connective tissue development
- Limb / appendage morphogenesis
- Heart morphogenesis; EMT
- TGF-beta signaling; Hippo signaling
- Kidney / renal development
- Elastic fibre formation

**PM highest**
- Xenobiotic response
- Postnatal lethality
- Hair shaft / skin barrier function
- Motor learning
- Cardiovascular physiology
- Platelet aggregation
- Apical plasma membrane

**POST higher than MAX**
- Cell-cell adhesion (plasma membrane)
- Muscle: contraction, myofibril, sarcomere, myopathy, diaphragm, Z line morphology
- Synapse organization
- Respiratory distress, cyanosis
- Bone mineral density
- ECM / collagen
- Wound healing

**POST higher than PM**
- Synaptic (dominant signal): postsynaptic density/membrane, synaptic membrane, chemical synaptic transmission
- Ion channels (broad: cation, gated)
- Muscle: contraction, myofibril, sarcomere
- ECM
- Neuronal behavior: anxiety, locomotion, learning, memory
- Membrane potential regulation
- Neuronal system (Reactome)

**POST highest**
- Respiratory distress, cyanosis, hypoxia
- Muscle: actin binding, contraction, sarcomere, Z line
- Diaphragm morphology
- Cell-cell adhesion

**POST lower than MAX**
- Synapse organization & assembly
- Epidermis / skin development; cornified envelope
- Basal cell carcinoma; Wnt signaling
- Neuron differentiation / cell fate commitment
- Kidney development (nephron, ureteric bud, mesonephric tubule)
- Axon guidance
- Desmosome

**POST lower than PM**
- Odontogenesis / dental development
- Neonatal lethality
- Embryonic morphogenesis & organ development
- ECM; cornified envelope; desmosome
- Skeletal development / ossification / osteoblast differentiation
- Kidney development (mesonephros, ureteric bud)
- Wnt signaling
- Cell junction assembly; transcription repressor activity
- Female fertility
- Skin: keratoderma, epidermal acanthosis

---

## Domain-Specific

**MAX↑ only**
- Lung morphology
- Neuronal system / neurotransmitter receptors
- Craniofacial: nasal capsule, basisphenoid, palate
- Hair follicle

**PM↑ only**
- Regulation of system process
- Response to cocaine; noise-induced hearing loss
- Blood vessel morphogenesis
- Synaptic transmission (positive regulation)
- ECM
- Glutamate:sodium symporter; integrin binding
- Intestine morphology; odontogenesis

**PM↓ only**
- Cation / metal ion transmembrane transport
- Symporter activity

**POST↑ only**
- Ion channels (dominant signal): presynaptic membrane, voltage-gated channels
- Synaptic transmission / modulation
- Membrane potential regulation
- Neurotransmitter transport & secretion; synaptic vesicle
- Gliogenesis / glial differentiation
- Heart contraction & circulation
- Coordination; exploration behavior

**POST↓ only**
- Tissue morphogenesis; embryonic morphogenesis
- Nervous system development
- Neural crest / pharyngeal arch derivatives: middle ear ossicles, craniofacial bones, Meckel's cartilage
- Vertebral / rib abnormalities; neural tube
- GnRH deficiency
- Skin pigmentation

---

## Opposite Trajectories

**MAX↑ PM↓**
- EEG multifocal slow activity

**MAX↓ PM↑**
- Nociception (inflammatory pain)
- Sympathetic nervous system physiology
- Sensory neuron innervation

---

## Stage Main Effects

**Decrease**
- Erythrocyte / hemoglobin (dominant signal): gas transport, oxygen carrier, hemoglobin complex, red cell osmotic fragility
- SRF transcription factor motifs
- Erythrocyte differentiation; myeloid cell differentiation
- Platelet (macrothrombocytopenia)

**Increase**
- Motile cilia / axoneme / flagellum (dominant signal): cilium movement, sperm motility, ciliary dyskinesia
- ECM / collagen (dominant signal)
- Immune & inflammation: leukocyte activation & migration, cytokine production, defense response, macrophage, innate & adaptive immunity
- Ion channels (broad: cation, gated, ligand-gated, voltage-gated)
- Skin: cornified envelope, keratoderma, epidermal thickening, acanthosis
- Bone: ossification, mineralization, trabecular bone, osteoblast
- GPCR signaling
- Respiratory: sinusitis, bronchiectasis, airway obstruction
- Situs / laterality defects
- Endopeptidase / serine protease activity
- Bacterial infection susceptibility
- MHC / antigen processing & presentation
- Collagen biosynthesis & degradation

---

## Two-Domain

**MAX+POST↑**
- Cilia / axoneme (dominant signal)
- Microtubule-based movement
- Sperm axoneme assembly
- Synaptic signaling (modulation)
- Seizure susceptibility

**PM+MAX↑**
- Skeletal muscle (dominant signal): sarcomere, myofibril, contraction, myopathy, dystrophy, muscle fiber types, atrophy
- Cardiac: cardiomyopathy, conduction defects, arrhythmia, heart failure, cardiac muscle development
- Ion channels: cation, potassium, calcium, ligand-gated
- Neuromuscular junction; acetylcholine receptor/signaling
- ECM / collagen
- Joint contractures
- Bone: remodeling, osteoblast, osteoclast
- Respiratory insufficiency (muscle weakness related)

**PM+POST↑**
- Neuroactive ligand-receptor interaction (GPCRs)
- Neurotransmitter receptors (serotonin, monoamine)
- Nociception (mechanical)
- Membrane potential regulation

---

## Cross-Category Patterns

- **ECM / collagen** appears across nearly every co-increasing and complex category, as well as stage-increasing and multiple domain main effect sets — the most pervasive signal overall.
- **Motile cilia / axoneme** is prominent in stage-increasing and the MAX+POST↑ two-domain set, and also recurs in co-increasing (both MAX and POST faster).
- **Skeletal muscle / sarcomere / myopathy** dominates the PM+MAX↑ two-domain set and appears in several POST-related domain main effect sets (POST higher than MAX, POST highest, POST higher than PM).
- **Synaptic / neuronal** terms concentrate in MAX-higher-than-PM and POST-higher-than-PM domain main effects, POST↑-only domain-specific, and the PM+POST↑ two-domain set.
- **Bone / skeletal development** phenotypes are widespread across co-increasing (MAX faster), complex (All↑), stage-increasing, and multiple domain main effect contrasts.
- **Skin / epidermis / keratinization** recurs across co-increasing, complex (All↑), stage-increasing, and POST-lower-than-PM domain main effects.
- **Immune / inflammation** (leukocyte migration, cytokine production, infection susceptibility) is concentrated in complex (All↑) and stage-increasing sets.
- **Ion channels** (cation, potassium, voltage-gated, ligand-gated) appear broadly in neuronal/synaptic sets (MAX higher, POST higher, POST↑ only) and in PM+MAX↑ muscle-related enrichments.
- **Erythrocyte / hemoglobin / gas transport** is specific to the stage-decreasing set — the only category with this signature.
- **Wnt signaling** concentrates in A→P decreasing domain main effects, POST-lower-than-MAX, and co-increasing MAX faster.
- **Cardiac** phenotypes (cardiomyopathy, conduction, arrhythmia, heart failure) are most prominent in PM+MAX↑ two-domain, with scattered appearances in complex (All↑) and co-increasing (MAX faster).
- **Laterality / situs** defects appear in co-increasing (both MAX and POST faster), complex (All↑), and stage-increasing.

:::

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
