---
title: "Investigating A-P domain and developmental stage effects in B6"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    standalone: true
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| warning: false
#| message: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

# Load packages
suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(DT)
library(here)

# Explicitly define common functions to avoid conflicts
select <- dplyr::select
rename <- dplyr::rename
filter <- dplyr::filter
summarize <- dplyr::summarize

set.seed(123) # for reproducibility

# DT table function
create_dt <- function(x){
  DT::datatable(x,
                extensions = 'Buttons',
                rownames = FALSE, 
                filter = "top",
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel'),
                               pageLength = 10, 
                               scrollX = TRUE
                ))
}

# Read sample details
sample_details <- read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 

# Domain colors
domain_colors <- c(PM = rgb(252, 206, 93, maxColorValue = 255), 
                   MAX = rgb(163, 82, 157, maxColorValue = 255), 
                   POST = rgb(158, 156, 201, maxColorValue = 255))

# Strain colors
strain_colors <- c(B6 = "#555555", 
                   CAST = "#009E73",
                   DO = rgb(163, 82, 157, maxColorValue = 255))
```

```{r}
#| label: plotting-functions
#| warning: false
#| message: false

# ===== FUNCTION 1: COMPARE UP VS DOWN IN SAME DOMAIN =====
plot_domain_direction_comparison <- function(genes_up, genes_down, 
                                             domain_name, count_data,
                                             n_each = 2) {
  
  gene_col <- "symbol"
  
  # Get examples
  up_examples <- genes_up[1:min(n_each, length(genes_up))]
  down_examples <- genes_down[1:min(n_each, length(genes_down))]
  
  # Create plots for UP genes
  plots_up <- lapply(up_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("↑ in ", domain_name),
           x = "Development stage",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold", size = 11),
            plot.subtitle = element_text(color = "darkgreen"))
  })
  
  # Create plots for DOWN genes
  plots_down <- lapply(down_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("↓ in ", domain_name),
           x = "Development stage",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold", size = 11),
            plot.subtitle = element_text(color = "darkred"))
  })
  
  # Remove NULLs
  plots_up <- plots_up[!sapply(plots_up, is.null)]
  plots_down <- plots_down[!sapply(plots_down, is.null)]
  
  # Combine
  all_plots <- c(plots_up, plots_down)
  if(length(all_plots) == 0) return(NULL)
  
  if(length(all_plots) == 2) {
    combined <- all_plots[[1]] / all_plots[[2]]
  } else if(length(all_plots) == 4) {
    combined <- (all_plots[[1]] | all_plots[[2]]) / (all_plots[[3]] | all_plots[[4]])
  } else {
    combined <- wrap_plots(all_plots, ncol = 2)
  }
  
  combined <- combined +
    plot_annotation(
      title = paste0(domain_name, ": Higher (↑) vs Lower (↓)"),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 2: PLOT DOMAIN HIERARCHY =====
plot_domain_hierarchy <- function(genes_highest, domain_name, count_data, n = 3) {
  
  gene_col <- "symbol"
  
  example_genes <- genes_highest[1:min(n, length(genes_highest))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Create both trajectory and boxplot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("Highest in ", domain_name),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8),
            plot.title = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = AP_domain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values = domain_colors) +
      labs(x = "", y = "") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none")
    
    return(p1 + p2 + plot_layout(widths = c(2, 1)))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Genes Highest in ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 3: PLOT A-P GRADIENTS =====
plot_AP_gradient <- function(genes_decreasing, genes_increasing, count_data, n_each = 3) {
  
  gene_col <- "symbol"
  
  # Get examples
  dec_examples <- genes_decreasing[1:min(n_each, length(genes_decreasing))]
  inc_examples <- genes_increasing[1:min(n_each, length(genes_increasing))]
  
  # Function to create gradient plot
  make_gradient_plot <- function(gid, gradient_type) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Summary by domain
    domain_summary <- gene_data |> 
      group_by(AP_domain) |> 
      summarize(mean_count = mean(count, na.rm = TRUE), .groups = "drop")
    
    ggplot(domain_summary, aes(x = AP_domain, y = mean_count, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_point(data = gene_data, aes(y = count), alpha = 0.3, size = 1.5,
                 position = position_jitter(width = 0.1)) +
      scale_fill_manual(values = domain_colors) +
      scale_y_log10(labels = scales::comma) +
      labs(title = gene_name,
           subtitle = ifelse(gradient_type == "dec", "PM > MAX > POST", "POST > MAX > PM"),
           x = "",
           y = "Expression") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = ifelse(gradient_type == "dec", 
                                                        "darkred", "darkgreen")))
  }
  
  # Create plots
  plots_dec <- lapply(dec_examples, function(gid) make_gradient_plot(gid, "dec"))
  plots_inc <- lapply(inc_examples, function(gid) make_gradient_plot(gid, "inc"))
  
  # Remove NULLs
  plots_dec <- plots_dec[!sapply(plots_dec, is.null)]
  plots_inc <- plots_inc[!sapply(plots_inc, is.null)]
  
  # Combine
  all_plots <- c(plots_dec, plots_inc)
  if(length(all_plots) == 0) return(NULL)
  
  combined <- wrap_plots(all_plots, ncol = 3)
  combined <- combined +
    plot_annotation(
      title = "Anterior-Posterior Expression Gradients",
      subtitle = "Top: A→P decreasing | Bottom: A→P increasing",
      theme = theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12)
      )
    )
  
  return(combined)
}

# ===== FUNCTION 4: PLOT DOMAIN-SPECIFIC DYNAMICS =====
plot_domain_specific_dynamics <- function(gene_ids, domain_name, direction,
                                          count_data, slope_data, n = 3) {
  
  gene_col <- "symbol"
  
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    # Get count data
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Get slope data
    gene_slopes <- slope_data |> 
      filter(gene_id == gid) |> 
      mutate(Domain = factor(Domain, levels = c("PM", "MAX", "POST")))
    
    # Trajectory plot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(direction, " in ", domain_name, " only"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"))
    
    # Slope plot
    if(nrow(gene_slopes) > 0) {
      p2 <- ggplot(gene_slopes, aes(x = Domain, y = log2FoldChange, fill = Domain)) +
        geom_col(alpha = 0.8) +
        geom_errorbar(aes(ymin = log2FoldChange - lfcSE,
                          ymax = log2FoldChange + lfcSE),
                      width = 0.3) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_text(aes(label = ifelse(padj < 0.001, "***",
                                     ifelse(padj < 0.01, "**",
                                            ifelse(padj < 0.05, "*", "ns")))),
                  vjust = -0.5, size = 4) +
        scale_fill_manual(values = domain_colors) +
        labs(y = "Slope", x = "") +
        theme_pubclean(base_size = 10) +
        theme(legend.position = "none")
      
      return(p1 + p2 + plot_layout(widths = c(2, 1)))
    } else {
      return(p1)
    }
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Domain-Specific Dynamics: ", domain_name, " ", direction),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 5: PLOT DIVERGING PATTERNS =====
plot_diverging_detailed <- function(genes_faster, genes_opposite, 
                                    domain_name, count_data, n_each = 2) {
  
  gene_col <- "symbol"
  
  faster_examples <- genes_faster[1:min(n_each, length(genes_faster))]
  opposite_examples <- genes_opposite[1:min(n_each, length(genes_opposite))]
  
  # Plot "faster" pattern
  plots_faster <- lapply(faster_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(domain_name, " increases faster"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = "darkgreen", size = 9))
  })
  
  # Plot "opposite" pattern
  plots_opposite <- lapply(opposite_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(domain_name, " ↑, other ↓"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = "purple", size = 9))
  })
  
  # Remove NULLs
  plots_faster <- plots_faster[!sapply(plots_faster, is.null)]
  plots_opposite <- plots_opposite[!sapply(plots_opposite, is.null)]
  
  all_plots <- c(plots_faster, plots_opposite)
  if(length(all_plots) == 0) return(NULL)
  
  if(length(all_plots) <= 2) {
    combined <- wrap_plots(all_plots, ncol = 1)
  } else {
    combined <- wrap_plots(all_plots, ncol = 2)
  }
  
  combined <- combined +
    plot_annotation(
      title = paste0("Diverging Patterns: ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 6: PLOT SIMPLE TRAJECTORIES =====
plot_simple_trajectories <- function(gene_ids, count_data, title_text, n = 5) {
  
  gene_col <- "symbol"
  
  # Sample if needed
  if(length(gene_ids) > n) {
    set.seed(123)
    gene_ids <- sample(gene_ids, n)
  }
  
  plots <- lapply(gene_ids, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8),
            plot.title = element_text(face = "bold", size = 11))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 3)
  combined <- combined +
    plot_annotation(
      title = title_text,
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}
```

```{r}
#| label: load-data-emase
#| warning: false
#| message: false

# Loading the count data
load("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/palate_gbrs_emase_counts_v4.RData")

# Clean up expression matrix
expr_mat <- expr_mat_nondo
contam_samples <- c("B6_12.5_PM_1", "B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat[, !colnames(expr_mat) %in% contam_samples]

do_expr_mat <- expr_mat_do
remove_samples <- c("DO_13.5_MAX_6", "DO_13.5_MAX_14", "DO_13.5_MAX_14_2",
                    "DO_12.5_MAX_1", "DO_12.5_MAX_2", "DO_13.5_MAX_97", "DO_13.5_MAX_96")
do_expr_mat <- do_expr_mat[, !colnames(do_expr_mat) %in% remove_samples]

expr_mat_all <- cbind(expr_mat[intersect(rownames(expr_mat), rownames(do_expr_mat)), , drop = FALSE],
                       do_expr_mat[intersect(rownames(expr_mat), rownames(do_expr_mat)), , drop = FALSE])
rm(expr_mat_do, expr_mat_nondo)

# Get v105 annotations from biomart
all.genes_v105 <- read_tsv(here("_data", "ENSMUSGid_to_symbol_v105.txt"),
                           col_names = c("gene_id", "symbol", "gene_biotype", "chromosome", "start", "end"), 
                           skip = 1) |>  
  mutate(midpoint = (start + end) / 2)

all_genes_palate <- all.genes_v105 |>
  filter(gene_id %in% rownames(raw_expr_mat))

all_genes_palate_filtered <- all.genes_v105 |>
  filter(gene_id %in% rownames(expr_mat_all))

# Sample annotations
sample_annot <- sample_details |> 
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |> 
  separate(Strain_stock, into = c("strain", "tmp"), remove = FALSE, sep = "_") |> 
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, palate_closure, L_R_palate) |> 
  mutate(strain = case_when(strain == "B6J" ~ "B6",
                            TRUE ~ strain)) |> 
  mutate(LB_stage_binned = case_when(
    LB_stage < 12.25 ~ "exclude",
    between(LB_stage, 13.25, 13.5) ~ "exclude",
    between(LB_stage, 12.25, 12.75) ~ "1",
    between(LB_stage, 13.5, 14) ~ "2",
    between(LB_stage, 14.25, 14.75) ~ "3",
    between(LB_stage, 15.1, 15.6) ~ "4"
  )) |> 
  mutate(LB_stage_binned = as_factor(LB_stage_binned)) |> 
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |> 
  select(-temp) |> 
  mutate(Palate_ori = ifelse(Palate_ori == "L", "Left", "Right")) 

sample_annot_all <- sample_details |> 
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat_all)) |> 
  separate(Strain_stock, into = c("strain", "tmp"), remove = FALSE, sep = "_") |> 
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, palate_closure, L_R_palate) |> 
  mutate(strain = case_when(strain == "B6J" ~ "B6",
                            TRUE ~ strain)) |> 
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |> 
  select(-temp) |> 
  mutate(Palate_ori = ifelse(Palate_ori == "L", "Left", "Right")) |> 
  mutate(LB_stage_binned = case_when(
    LB_stage < 12.25 ~ "exclude",
    between(LB_stage, 13.25, 13.5) ~ "exclude",
    between(LB_stage, 12.25, 12.75) ~ "1",
    between(LB_stage, 13.5, 14) ~ "2",
    between(LB_stage, 14.25, 14.75) ~ "3",
    between(LB_stage, 15.1, 15.6) ~ "4"
  ))
```

# Differential gene expression analysis overview

Briefly, I analyzed the B6 data on its own to identify genes differentially expressed across AP domains, over developmental time and change differently over time across domains.

For the individual strain analysis with B6:

I am focusing on the data from only the B6 animals that includes males and females, and contain the three AP domains (PM, MAX and POST) obtained at various developmental stages (LB_stage). I merge the technical replicates due to resequencing of several samples using `collapseReplicates()` function from DESeq2. We expect to have genes that show significant differences between AP domains, over developmental time and genes with different trends over time across the three tissues.

In order to identify differentially expressed genes between AP domains and over time I obtain the main effects for AP_domain and LB_stage while using a nesting for the AP domain since they are obtained from the same animal (~ AP_domain + LB_stage + AP_domain:ind).

For identifying genes with different trends over time across the three AP domains and genes differentially expressed between domains variably over time I use the likelihood ratio test (LRT) to compare the initial simple model to the one that includes an interaction term between AP domain and limb bud stage (~ AP_domain + LB_stage + AP_domain:ind + AP_domain:LB_stage). Once I have this DE gene list, I then use the full model to get the effects of developmental time in all three domains.

Note that we need to be very careful when interpreting the main effects with a linear model that contains an interaction term, they do not represent the overall effect of the predictor (e.g. LB_stage) since the predictors involved in an interaction do not have a single overall effect anymore! For example the main effect for LB_stage will represent the effect when AP_domain is at the reference level which is the PM domain in our data set.

```{r}
#| label: prep-b6-for-deseq2
#| warning: false
#| message: false

# Prep data for DESeq2
# Subsetting for B6 samples only and adding unique sample IDs
b6_comp_samples <- sample_annot |>
  filter(strain %in% c("B6")) |>
  separate(sample, sep = "_", into = c("s", "e", "d", "ind", "rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = ifelse(is.na(rep), 1, rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

# Making the coldata object to feed into DESeq2
b6_coldata <- b6_comp_samples |>
  select(sample, sex, ind, unique_sampleid, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate_if(is.character, as.factor) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

# Get the expression matrix to feed into DESeq2
b6_cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(b6_coldata), drop = FALSE])

# Make the dds object then collapse the technical replicates
b6_dds <- DESeqDataSetFromMatrix(countData = b6_cts,
                                  colData = b6_coldata,
                                  design = ~1)

# Collapse the sequencing replicates
b6_dds_coll <- collapseReplicates(b6_dds, b6_dds$unique_sampleid)

# Update LB-stage to have 0 as the starting point
min_LB_stage <- min(colData(b6_dds_coll)$LB_stage)
colData(b6_dds_coll)$LB_stage_from_start <- colData(b6_dds_coll)$LB_stage - min_LB_stage
```

```{r}
#| label: run-deseq2-b6
#| warning: false
#| message: false

# Define design matrices
b6_simple_model <- model.matrix(~ sex + ind:AP_domain + LB_stage_from_start + AP_domain, 
                                data = colData(b6_dds_coll))
b6_comp_model <- model.matrix(~ sex + ind:AP_domain + LB_stage_from_start * AP_domain, 
                               data = colData(b6_dds_coll))

# Run DESeq2 models
b6_dds_simple <- DESeq(b6_dds_coll, betaPrior = FALSE, full = b6_simple_model)
b6_dds_interaction <- DESeq(b6_dds_coll, betaPrior = FALSE, full = b6_comp_model)

# Likelihood ratio test to identify genes with significant interaction
b6_dds_lrt <- DESeq(b6_dds_coll, test = "LRT",
                    full = b6_comp_model,
                    reduced = b6_simple_model)

# Identify genes with significant interaction
genes_with_interaction <- results(b6_dds_lrt) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(padj < 0.05) |> 
  pull(gene_id)
```

```{r}
#| label: get-degenes-b6
#| warning: false
#| message: false

# Helper function for consistent filtering
filter_genes <- function(results_df, min_basemean = 50, max_lfcse = 1.5, 
                         max_abs_lfc = 10, min_padj = 0.05, min_abs_lfc = 0.5) {
  results_df |>
    filter(
      baseMean > min_basemean,
      lfcSE < max_lfcse,
      abs(log2FoldChange) < max_abs_lfc,
      abs(log2FoldChange) > min_abs_lfc,
      padj < min_padj
    )
}

# ===== MAIN EFFECTS (genes WITHOUT interaction) =====

# Domain effects from SIMPLE model
b6_domain_max_vs_pm <- results(b6_dds_simple, name = "AP_domainMAX") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Comparison = "MAX vs PM") |> 
  left_join(all_genes_palate, by = "gene_id")

b6_domain_post_vs_pm <- results(b6_dds_simple, name = "AP_domainPOST") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Comparison = "POST vs PM") |> 
  left_join(all_genes_palate, by = "gene_id")

b6_domain_post_vs_max <- results(b6_dds_simple, 
                                 contrast = list("AP_domainPOST", "AP_domainMAX")) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Comparison = "POST vs MAX") |> 
  left_join(all_genes_palate, by = "gene_id")

# Combine domain effects
b6_domain_main_effects <- bind_rows(
  b6_domain_max_vs_pm,
  b6_domain_post_vs_pm,
  b6_domain_post_vs_max
)

# Apply filtering
b6_domain_main_effects_clean <- filter_genes(
  b6_domain_main_effects,
  min_basemean = quantile(b6_domain_main_effects$baseMean, 0.1, na.rm = TRUE)
)

# Stage effects from SIMPLE model
b6_stage_main_effects <- results(b6_dds_simple, name = "LB_stage_from_start") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Direction = ifelse(log2FoldChange > 0, "Increase", "Decrease")) |> 
  left_join(all_genes_palate, by = "gene_id")

# Apply filtering
b6_stage_main_effects_clean <- filter_genes(
  b6_stage_main_effects,
  min_basemean = quantile(b6_stage_main_effects$baseMean, 0.1, na.rm = TRUE)
)

# ===== INTERACTION EFFECTS (genes WITH interaction) =====

# 1. INITIAL DOMAIN DIFFERENCES (at earliest stage)
initial_diff_MAX_vs_PM <- results(b6_dds_interaction, name = "AP_domainMAX") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "MAX vs PM at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

initial_diff_POST_vs_PM <- results(b6_dds_interaction, name = "AP_domainPOST") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs PM at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

initial_diff_POST_vs_MAX <- results(b6_dds_interaction, 
                                    contrast = list(c("AP_domainPOST", "AP_domainMAX"))) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs MAX at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine initial differences
b6_initial_differences <- bind_rows(
  initial_diff_MAX_vs_PM,
  initial_diff_POST_vs_PM,
  initial_diff_POST_vs_MAX
)

# 2. DEVELOPMENTAL SLOPES (rates of change in each domain)
slope_PM <- results(b6_dds_interaction, name = "LB_stage_from_start") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(gene_id %in% genes_with_interaction) |> 
  mutate(Domain = "PM") |> 
  left_join(all_genes_palate, by = "gene_id")

slope_MAX <- results(b6_dds_interaction, 
                     contrast = list(c("LB_stage_from_start", 
                                      "AP_domainMAX.LB_stage_from_start"))) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(gene_id %in% genes_with_interaction) |> 
  mutate(Domain = "MAX") |> 
  left_join(all_genes_palate, by = "gene_id")

slope_POST <- results(b6_dds_interaction, 
                      contrast = list(c("LB_stage_from_start", 
                                       "AP_domainPOST.LB_stage_from_start"))) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(gene_id %in% genes_with_interaction) |> 
  mutate(Domain = "POST") |> 
  left_join(all_genes_palate, by = "gene_id")

# Combine all slopes
b6_domain_slopes <- bind_rows(slope_PM, slope_MAX, slope_POST)

# Apply filtering to slopes
b6_domain_slopes_clean <- filter_genes(
  b6_domain_slopes,
  min_basemean = quantile(b6_domain_slopes$baseMean, 0.1, na.rm = TRUE),
  max_lfcse = 2
)

# 3. SLOPE DIFFERENCES (interaction terms - how different are the slopes?)
slope_diff_MAX_vs_PM <- results(b6_dds_interaction, 
                                name = "AP_domainMAX.LB_stage_from_start") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "MAX vs PM slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

slope_diff_POST_vs_PM <- results(b6_dds_interaction, 
                                 name = "AP_domainPOST.LB_stage_from_start") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs PM slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

slope_diff_POST_vs_MAX <- results(b6_dds_interaction, 
                                  contrast = list(c("AP_domainPOST.LB_stage_from_start",
                                                   "AP_domainMAX.LB_stage_from_start"))) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs MAX slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine slope differences
b6_slope_differences <- bind_rows(
  slope_diff_MAX_vs_PM,
  slope_diff_POST_vs_PM,
  slope_diff_POST_vs_MAX
)

# Get normalized counts for plotting
b6_gene_counts <- counts(b6_dds_simple, normalized = TRUE) |> 
  as_tibble(rownames = "gene_id") |>
  pivot_longer(-gene_id, names_to = "sample", values_to = "count") |> 
  left_join(
    colData(b6_dds_simple) |> 
      as_tibble(rownames = "sample"),
    by = "sample"
  ) |>
  left_join(all_genes_palate, by = "gene_id")
```

```{r}
#| label: load-pattern-results
#| warning: false
#| message: false

# Load pattern assignments and summaries
pattern_summary <- read_csv(here("pattern_summary_detailed.csv"))
gene_assignments <- read_csv(here("gene_pattern_assignments_detailed.csv"))
domain_dynamics <- read_csv(here("domain_specific_dynamics_detailed.csv"))
```

This report presents a comprehensive analysis of gene expression patterns during palate development across three anatomical domains (PM, MAX, POST) and developmental time. We identified **`r nrow(gene_assignments)` genes** with significant spatiotemporal expression patterns, classified into six major categories:

```{r}
#| label: summary-table
#| warning: false
#| message: false

pattern_summary |> 
  group_by(Category) |> 
  summarize(
    `Total Patterns` = n(),
    `Total Genes` = sum(N_Genes),
    .groups = "drop"
  ) |> 
  create_dt()
```

:::: {#B6_DESeq2_effects .panel-tabset .nav-pills}

## Domain Main Effects 

Here I am highlighting genes that show differences in expression between A-P domains in B6. In total there are `r length(unique(b6_domain_main_effects_clean$gene_id))` unique genes with a domain effect (i.e. differentially expressed between any two domains) that doesn't change over developmental time (i.e. maintain parallel developmental trajectories). This means that while one domain might express a gene at higher or lower levels than another, all domains follow the same developmental pattern over time (all increasing, all decreasing, or all stable).

::: {#B6_DESeq2_domain_effects .panel-tabset .nav-pills}

### Overview

```{r}
#| label: domain-main-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Domain Main Effects") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### MAX vs PM

```{r}
#| label: max-vs-pm-plot
#| warning: false
#| message: false

genes_MAX_higher_than_PM <- gene_assignments |>
  filter(MAX_higher_PM) |>
  pull(gene_id)

genes_MAX_lower_than_PM <- gene_assignments |>
  filter(MAX_lower_PM) |>
  pull(gene_id)

if(length(genes_MAX_higher_than_PM) > 0 & length(genes_MAX_lower_than_PM) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_MAX_higher_than_PM,
    genes_down = genes_MAX_lower_than_PM,
    domain_name = "MAX vs PM",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**Pattern:** Genes showing consistent differences between MAX and PM domains across development.

**Biological Interpretation:** These genes define constitutive differences between maxillary and palatine midline regions, potentially representing:

- **MAX higher than PM** (`r length(genes_MAX_higher_than_PM)` genes): MAX-specific identity or function
- **MAX lower than PM** (`r length(genes_MAX_lower_than_PM)` genes): PM-enriched programs

```{r}
#| label: max-vs-pm-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "MAX vs PM") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

### POST vs PM

```{r}
#| label: post-vs-pm-plot
#| warning: false
#| message: false

genes_POST_higher_than_PM <- gene_assignments |>
  filter(POST_higher_PM) |>
  pull(gene_id)

genes_POST_lower_than_PM <- gene_assignments |>
  filter(POST_lower_PM) |>
  pull(gene_id)

if(length(genes_POST_higher_than_PM) > 0 & length(genes_POST_lower_than_PM) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_POST_higher_than_PM,
    genes_down = genes_POST_lower_than_PM,
    domain_name = "POST vs PM",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**Pattern:** Consistent expression differences between posterior and palatine midline domains.

**Biological Interpretation:**

- **POST higher than PM** (`r length(genes_POST_higher_than_PM)` genes): Posterior-specific programs
- **POST lower than PM** (`r length(genes_POST_lower_than_PM)` genes): Anterior/midline enriched

```{r}
#| label: post-vs-pm-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "POST vs PM") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

### POST vs MAX

```{r}
#| label: post-vs-max-plot
#| warning: false
#| message: false

genes_POST_higher_than_MAX <- gene_assignments |>
  filter(POST_higher_MAX) |>
  pull(gene_id)

genes_POST_lower_than_MAX <- gene_assignments |>
  filter(POST_lower_MAX) |>
  pull(gene_id)

if(length(genes_POST_higher_than_MAX) > 0 & length(genes_POST_lower_than_MAX) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_POST_higher_than_MAX,
    genes_down = genes_POST_lower_than_MAX,
    domain_name = "POST vs MAX",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**Pattern:** Expression differences between posterior and maxillary domains.

```{r}
#| label: post-vs-max-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "POST vs MAX") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

### Highest in Each Domain

#### MAX Highest

```{r}
#| label: max-highest-plot
#| warning: false
#| message: false

genes_MAX_highest <- gene_assignments |>
  filter(Highest_MAX) |>
  pull(gene_id)

if(length(genes_MAX_highest) > 0) {
  n_show <- ifelse(length(genes_MAX_highest) < 15, length(genes_MAX_highest), 3)
  plot_domain_hierarchy(
    genes_highest = genes_MAX_highest,
    domain_name = "MAX",
    count_data = b6_gene_counts,
    n = n_show
  )
}
```

**Pattern:** `r length(genes_MAX_highest)` genes showing highest expression in MAX compared to both PM and POST.

**Biological Interpretation:** Core maxillary identity genes that distinguish lateral palate from both midline and posterior regions.

```{r}
#| label: max-highest-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% genes_MAX_highest) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  distinct(gene_id, .keep_all = TRUE) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

#### POST Highest

```{r}
#| label: post-highest-plot
#| warning: false
#| message: false

genes_POST_highest <- gene_assignments |>
  filter(Highest_POST) |>
  pull(gene_id)

if(length(genes_POST_highest) > 0) {
  n_show <- ifelse(length(genes_POST_highest) < 15, length(genes_POST_highest), 3)
  plot_domain_hierarchy(
    genes_highest = genes_POST_highest,
    domain_name = "POST",
    count_data = b6_gene_counts,
    n = n_show
  )
}
```

**Pattern:** `r length(genes_POST_highest)` genes with peak expression in posterior domain.

**Biological Interpretation:** Posterior identity genes, potentially involved in soft palate specification.

```{r}
#| label: post-highest-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% genes_POST_highest) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  distinct(gene_id, .keep_all = TRUE) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

#### PM Highest

```{r}
#| label: pm-highest-plot
#| warning: false
#| message: false

genes_PM_highest <- gene_assignments |>
  filter(Highest_PM) |>
  pull(gene_id)

if(length(genes_PM_highest) > 0) {
  n_show <- ifelse(length(genes_PM_highest) < 15, length(genes_PM_highest), 3)
  plot_domain_hierarchy(
    genes_highest = genes_PM_highest,
    domain_name = "PM",
    count_data = b6_gene_counts,
    n = n_show
  )
}
```

**Pattern:** `r length(genes_PM_highest)` genes with highest expression at the palatine midline.

**Biological Interpretation:** Midline-specific genes, potentially involved in fusion processes.

```{r}
#| label: pm-highest-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% genes_PM_highest) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  distinct(gene_id, .keep_all = TRUE) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

### A-P Gradients

```{r}
#| label: ap-gradients-plot
#| warning: false
#| message: false

genes_AP_decreasing <- gene_assignments |>
  filter(AP_decreasing) |>
  pull(gene_id)

genes_AP_increasing <- gene_assignments |>
  filter(AP_increasing) |>
  pull(gene_id)

if(length(genes_AP_decreasing) > 0 | length(genes_AP_increasing) > 0) {
  n_each <- 3
  if(length(genes_AP_decreasing) < 15) n_each <- min(n_each, length(genes_AP_decreasing))
  if(length(genes_AP_increasing) < 15) n_each <- min(n_each, length(genes_AP_increasing))
  
  plot_AP_gradient(
    genes_decreasing = genes_AP_decreasing,
    genes_increasing = genes_AP_increasing,
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** Monotonic gradients along the anterior-posterior axis.

**Biological Interpretation:**

- **A→P decreasing** (PM > MAX > POST): `r length(genes_AP_decreasing)` genes. Classic anterior specification or posterior repression.
- **A→P increasing** (POST > MAX > PM): `r length(genes_AP_increasing)` genes. Posterior specification or anterior repression.

```{r}
#| label: ap-gradients-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% c(genes_AP_decreasing, genes_AP_increasing)) |>
  mutate(Gradient = ifelse(gene_id %in% genes_AP_decreasing, "A→P decreasing", "A→P increasing")) |>
  select(gene_id, symbol, chromosome, Gradient, Comparison, log2FoldChange, padj) |>
  arrange(Gradient, padj) |>
  create_dt()
```

:::

## Stage Main Effects 

**These genes show developmental changes that are consistent across all three domains.** All domains increase or decrease together, representing shared developmental programs.

::: {#B6_DESeq2_stage_effects .panel-tabset .nav-pills}

### Overview

```{r}
#| label: stage-main-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Stage Main Effects") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### Increasing with Development

```{r}
#| label: dev-increasing-plot
#| warning: false
#| message: false

genes_increasing <- gene_assignments |>
  filter(Dev_increasing) |>
  pull(gene_id)

if(length(genes_increasing) > 0) {
  n_show <- ifelse(length(genes_increasing) < 15, length(genes_increasing), 5)
  plot_simple_trajectories(
    gene_ids = genes_increasing,
    count_data = b6_gene_counts,
    title_text = paste0("Increasing with Development (n=", length(genes_increasing), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_increasing)` genes increasing across all domains as development progresses.

**Biological Interpretation:** Core developmental programs intensifying during palatogenesis, such as cell proliferation, differentiation markers, or ECM production.

```{r}
#| label: dev-increasing-table
#| warning: false
#| message: false

b6_stage_main_effects_clean |>
  filter(Direction == "Increase") |>
  select(gene_id, symbol, chromosome, Direction, log2FoldChange, padj) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

### Decreasing with Development

```{r}
#| label: dev-decreasing-plot
#| warning: false
#| message: false

genes_decreasing <- gene_assignments |>
  filter(Dev_decreasing) |>
  pull(gene_id)

if(length(genes_decreasing) > 0) {
  n_show <- ifelse(length(genes_decreasing) < 15, length(genes_decreasing), 5)
  plot_simple_trajectories(
    gene_ids = genes_decreasing,
    count_data = b6_gene_counts,
    title_text = paste0("Decreasing with Development (n=", length(genes_decreasing), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_decreasing)` genes decreasing across all domains.

**Biological Interpretation:** Early patterning factors, progenitor markers, or genes involved in completed morphogenetic events.

```{r}
#| label: dev-decreasing-table
#| warning: false
#| message: false

b6_stage_main_effects_clean |>
  filter(Direction == "Decrease") |>
  select(gene_id, symbol, chromosome, Direction, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

:::

## Diverging Trajectories 

**These genes show different developmental trajectories between domains** - true domain × stage interactions where domains respond differently to developmental cues.

::: {#B6_DESeq2_diverging_traj .panel-tabset .nav-pills}

### Overview

```{r}
#| label: diverging-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Diverging") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### MAX Diverging from PM

```{r}
#| label: max-div-pm-plot
#| warning: false
#| message: false

genes_MAX_div <- gene_assignments |>
  filter(MAX_div_PM) |>
  pull(gene_id)

# Get faster vs opposite
genes_MAX_faster <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_MAX_div, Domain == "MAX", log2FoldChange > 0.5, padj < 0.05) |>
  pull(gene_id)

genes_MAX_opposite <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_MAX_div) |>
  select(gene_id, Domain, log2FoldChange) |>
  pivot_wider(names_from = Domain, values_from = log2FoldChange) |>
  filter(!is.na(MAX), !is.na(PM), MAX > 0, PM < 0) |>
  pull(gene_id)

if(length(genes_MAX_faster) > 0 | length(genes_MAX_opposite) > 0) {
  n_each <- 2
  if(length(genes_MAX_faster) < 15) n_each <- min(n_each, length(genes_MAX_faster))
  if(length(genes_MAX_opposite) < 15) n_each <- min(n_each, length(genes_MAX_opposite))
  
  plot_diverging_detailed(
    genes_faster = genes_MAX_faster,
    genes_opposite = genes_MAX_opposite,
    domain_name = "MAX vs PM",
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** MAX and PM follow different developmental trajectories.

**Biological Interpretation:**

- **MAX increases faster**: `r length(genes_MAX_faster)` genes becoming progressively enriched in MAX
- **MAX up, PM down**: `r length(genes_MAX_opposite)` genes with opposite responses

```{r}
#| label: max-div-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "MAX vs PM slope difference", padj < 0.05) |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

### POST Diverging from PM

```{r}
#| label: post-div-pm-plot
#| warning: false
#| message: false

genes_POST_div <- gene_assignments |>
  filter(POST_div_PM) |>
  pull(gene_id)

genes_POST_faster <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div, Domain == "POST", log2FoldChange > 0.5, padj < 0.05) |>
  pull(gene_id)

genes_POST_opposite <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div) |>
  select(gene_id, Domain, log2FoldChange) |>
  pivot_wider(names_from = Domain, values_from = log2FoldChange) |>
  filter(!is.na(POST), !is.na(PM), POST > 0, PM < 0) |>
  pull(gene_id)

if(length(genes_POST_faster) > 0 | length(genes_POST_opposite) > 0) {
  n_each <- 2
  if(length(genes_POST_faster) < 15) n_each <- min(n_each, length(genes_POST_faster))
  if(length(genes_POST_opposite) < 15) n_each <- min(n_each, length(genes_POST_opposite))
  
  plot_diverging_detailed(
    genes_faster = genes_POST_faster,
    genes_opposite = genes_POST_opposite,
    domain_name = "POST vs PM",
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** POST and PM show divergent trajectories.

```{r}
#| label: post-div-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs PM slope difference", padj < 0.05) |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

### POST Diverging from MAX

```{r}
#| label: post-div-max-plot
#| warning: false
#| message: false

genes_POST_div_MAX <- gene_assignments |>
  filter(POST_div_MAX) |>
  pull(gene_id)

genes_POST_faster_MAX <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div_MAX, Domain == "POST", log2FoldChange > 0.5, padj < 0.05) |>
  pull(gene_id)

genes_POST_opposite_MAX <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div_MAX) |>
  select(gene_id, Domain, log2FoldChange) |>
  pivot_wider(names_from = Domain, values_from = log2FoldChange) |>
  filter(!is.na(POST), !is.na(MAX), POST > 0, MAX < 0) |>
  pull(gene_id)

if(length(genes_POST_faster_MAX) > 0 | length(genes_POST_opposite_MAX) > 0) {
  n_each <- 2
  if(length(genes_POST_faster_MAX) < 15) n_each <- min(n_each, length(genes_POST_faster_MAX))
  if(length(genes_POST_opposite_MAX) < 15) n_each <- min(n_each, length(genes_POST_opposite_MAX))
  
  plot_diverging_detailed(
    genes_faster = genes_POST_faster_MAX,
    genes_opposite = genes_POST_opposite_MAX,
    domain_name = "POST vs MAX",
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** POST and MAX show divergent developmental responses.

```{r}
#| label: post-div-max-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs MAX slope difference", padj < 0.05) |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

:::

## Converging Trajectories 

**These genes show domains becoming more similar over time**, with initially different expression levels that converge as development progresses.

::: {#B6_DESeq2_conv_traj .panel-tabset .nav-pills}

### Overview

```{r}
#| label: converging-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Converging") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### MAX Converging to PM

```{r}
#| label: max-conv-pm-plot
#| warning: false
#| message: false

genes_MAX_conv <- gene_assignments |>
  filter(MAX_conv_PM) |>
  pull(gene_id)

if(length(genes_MAX_conv) > 0) {
  n_show <- ifelse(length(genes_MAX_conv) < 15, length(genes_MAX_conv), 5)
  plot_simple_trajectories(
    gene_ids = genes_MAX_conv,
    count_data = b6_gene_counts,
    title_text = paste0("MAX Converging to PM (n=", length(genes_MAX_conv), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_MAX_conv)` genes where MAX and PM become more similar over time.

**Biological Interpretation:** Equalization of expression suggesting tissue integration or coordinated maturation.

```{r}
#| label: max-conv-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "MAX vs PM slope difference", padj < 0.05, log2FoldChange < -0.5) |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

### POST Converging to PM

```{r}
#| label: post-conv-pm-plot
#| warning: false
#| message: false

genes_POST_conv <- gene_assignments |>
  filter(POST_conv_PM) |>
  pull(gene_id)

if(length(genes_POST_conv) > 0) {
  n_show <- ifelse(length(genes_POST_conv) < 15, length(genes_POST_conv), 5)
  plot_simple_trajectories(
    gene_ids = genes_POST_conv,
    count_data = b6_gene_counts,
    title_text = paste0("POST Converging to PM (n=", length(genes_POST_conv), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_POST_conv)` genes converging between POST and PM.

```{r}
#| label: post-conv-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs PM slope difference", padj < 0.05, log2FoldChange < -0.5) |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

### POST Converging to MAX

```{r}
#| label: post-conv-max-plot
#| warning: false
#| message: false

genes_POST_conv_MAX <- gene_assignments |>
  filter(POST_conv_MAX) |>
  pull(gene_id)

if(length(genes_POST_conv_MAX) > 0) {
  n_show <- ifelse(length(genes_POST_conv_MAX) < 15, length(genes_POST_conv_MAX), 5)
  plot_simple_trajectories(
    gene_ids = genes_POST_conv_MAX,
    count_data = b6_gene_counts,
    title_text = paste0("POST Converging to MAX (n=", length(genes_POST_conv_MAX), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_POST_conv_MAX)` genes converging between POST and MAX.

```{r}
#| label: post-conv-max-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs MAX slope difference", padj < 0.05, log2FoldChange < -0.5) |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

:::

## Domain-Specific Dynamics 

**These genes show significant developmental changes in only one or two domains** while remaining stable in others.

::: {#B6_DESeq2_domain_dyn .panel-tabset .nav-pills}

### Overview

```{r}
#| label: domain-specific-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Domain-Specific") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### PM Only

```{r}
#| label: pm-only-plot
#| warning: false
#| message: false

genes_PM_only_up <- gene_assignments |>
  filter(PM_only_up) |>
  pull(gene_id)

genes_PM_only_down <- domain_dynamics |>
  filter(PM_dynamic, !MAX_dynamic, !POST_dynamic, PM_slope < -0.5) |>
  pull(gene_id)

if(length(genes_PM_only_up) > 0) {
  n_show <- ifelse(length(genes_PM_only_up) < 15, length(genes_PM_only_up), 3)
  plot_domain_specific_dynamics(
    gene_ids = genes_PM_only_up,
    domain_name = "PM",
    direction = "Increasing",
    count_data = b6_gene_counts,
    slope_data = b6_domain_slopes_clean,
    n = n_show
  )
}
```

**Pattern:** Developmental changes only in PM domain.

**Biological Interpretation:** PM-specific processes like midline fusion. Found `r length(genes_PM_only_up)` increasing and `r length(genes_PM_only_down)` decreasing.

```{r}
#| label: pm-only-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% c(genes_PM_only_up, genes_PM_only_down), Domain == "PM") |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

### MAX Only

```{r}
#| label: max-only-plot
#| warning: false
#| message: false

genes_MAX_only_up <- gene_assignments |>
  filter(MAX_only_up) |>
  pull(gene_id)

genes_MAX_only_down <- domain_dynamics |>
  filter(!PM_dynamic, MAX_dynamic, !POST_dynamic, MAX_slope < -0.5) |>
  pull(gene_id)

if(length(genes_MAX_only_up) > 0) {
  n_show <- ifelse(length(genes_MAX_only_up) < 15, length(genes_MAX_only_up), 3)
  plot_domain_specific_dynamics(
    gene_ids = genes_MAX_only_up,
    domain_name = "MAX",
    direction = "Increasing",
    count_data = b6_gene_counts,
    slope_data = b6_domain_slopes_clean,
    n = n_show
  )
}
```

**Pattern:** Developmental changes only in MAX domain.

**Biological Interpretation:** MAX-specific processes like lateral shelf morphogenesis. Found `r length(genes_MAX_only_up)` increasing and `r length(genes_MAX_only_down)` decreasing.

```{r}
#| label: max-only-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% c(genes_MAX_only_up, genes_MAX_only_down), Domain == "MAX") |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

### POST Only

```{r}
#| label: post-only-plot
#| warning: false
#| message: false

genes_POST_only_up <- gene_assignments |>
  filter(POST_only_up) |>
  pull(gene_id)

genes_POST_only_down <- domain_dynamics |>
  filter(!PM_dynamic, !MAX_dynamic, POST_dynamic, POST_slope < -0.5) |>
  pull(gene_id)

if(length(genes_POST_only_up) > 0) {
  n_show <- ifelse(length(genes_POST_only_up) < 15, length(genes_POST_only_up), 3)
  plot_domain_specific_dynamics(
    gene_ids = genes_POST_only_up,
    domain_name = "POST",
    direction = "Increasing",
    count_data = b6_gene_counts,
    slope_data = b6_domain_slopes_clean,
    n = n_show
  )
}
```

**Pattern:** Developmental changes only in POST domain.

**Biological Interpretation:** Posterior-specific programs like soft palate differentiation. Found `r length(genes_POST_only_up)` increasing and `r length(genes_POST_only_down)` decreasing.

```{r}
#| label: post-only-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% c(genes_POST_only_up, genes_POST_only_down), Domain == "POST") |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

:::

## Complex Patterns 

**Genes showing particularly complex spatiotemporal dynamics** including opposite trajectories or three-way divergence.

::: {#B6_DESeq2_comp_pat .panel-tabset .nav-pills}

### Overview

```{r}
#| label: complex-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Complex") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### Opposite Slopes

```{r}
#| label: opposite-slopes-plot
#| warning: false
#| message: false

genes_opposite <- gene_assignments |>
  filter(Opposite_slopes) |>
  pull(gene_id)

if(length(genes_opposite) > 0) {
  n_show <- ifelse(length(genes_opposite) < 15, length(genes_opposite), 5)
  plot_simple_trajectories(
    gene_ids = genes_opposite,
    count_data = b6_gene_counts,
    title_text = paste0("Genes with Opposite Slopes (n=", length(genes_opposite), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_opposite)` genes where at least one domain increases while another decreases.

**Biological Interpretation:** Fundamentally different developmental programs between domains, potentially key regulators of domain identity.

```{r}
#| label: opposite-slopes-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% genes_opposite) |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(gene_id, Domain) |>
  create_dt()
```

### Three-Way Divergent

```{r}
#| label: three-way-plot
#| warning: false
#| message: false

genes_three_way <- gene_assignments |>
  filter(Three_way) |>
  pull(gene_id)

if(length(genes_three_way) > 0) {
  n_show <- ifelse(length(genes_three_way) < 15, length(genes_three_way), 5)
  plot_simple_trajectories(
    gene_ids = genes_three_way,
    count_data = b6_gene_counts,
    title_text = paste0("Three-Way Divergent Genes (n=", length(genes_three_way), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_three_way)` genes where all three domains show different developmental slopes.

**Biological Interpretation:** Maximum spatiotemporal complexity, likely central regulators of domain-specific identity and function.

```{r}
#| label: three-way-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% genes_three_way) |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(gene_id, Domain) |>
  create_dt()
```

:::

## Pattern Overlap Analysis

::: {#B6_DESeq2_pattern_overlap .panel-tabset .nav-pills}

### Multi-Pattern Genes

```{r}
#| label: pattern-overlap
#| warning: false
#| message: false

multi_pattern_genes <- gene_assignments |>
  filter(n_patterns > 1) |>
  arrange(desc(n_patterns))

multi_pattern_genes |>
  select(symbol, n_patterns) |>
  group_by(n_patterns) |>
  summarize(n_genes = n(), .groups = "drop") |>
  create_dt()
```

### Top Multi-Pattern Genes

```{r}
#| label: top-multi-pattern
#| warning: false
#| message: false

multi_pattern_genes |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, n_patterns) |>
  head(50) |>
  create_dt()
```

:::

## Comprehensive Tables

::: {#B6_DESeq2_tables .panel-tabset .nav-pills}

### All Patterns Summary

```{r}
#| label: all-patterns-table
#| warning: false
#| message: false

pattern_summary |>
  create_dt()
```

### Gene-Level Pattern Assignments

```{r}
#| label: gene-assignments-table
#| warning: false
#| message: false

gene_assignments |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, n_patterns, everything()) |>
  create_dt()
```

:::

## Conclusions

### Key Findings

1. **Domain Main Effects**: Identified spatial differences with parallel developmental trajectories across `r sum(pattern_summary$Category == "Domain Main Effects")` pattern types

2. **Developmental Dynamics**: Found `r sum(gene_assignments$Dev_increasing)` genes increasing and `r sum(gene_assignments$Dev_decreasing)` genes decreasing uniformly across all domains

3. **Diverging Trajectories**: Detected domain-specific developmental divergence in `r sum(pattern_summary$Category == "Diverging" & pattern_summary$N_Genes > 0)` pattern types

4. **Domain-Specific Programs**: 
   - PM only: `r sum(gene_assignments$PM_only)` genes
   - MAX only: `r sum(gene_assignments$MAX_only)` genes  
   - POST only: `r sum(gene_assignments$POST_only)` genes

5. **Complex Patterns**: Found `r sum(gene_assignments$Opposite_slopes)` genes with opposite slopes and `r sum(gene_assignments$Three_way)` with three-way divergent trajectories

### Biological Implications

- **Spatial Organization**: Strong A-P and M-L organization of gene expression programs
- **Temporal Dynamics**: Both shared and domain-specific developmental trajectories
- **Regulatory Complexity**: Extensive domain-specific and interactive regulation
- **Convergent Evolution**: Multiple genes showing convergent trajectories suggesting coordinated maturation

::::