---
title: "Investigating A-P domain and developmental stage effects in B6"
author: "Selcan Aydin"
date: "`r Sys.Date()`"
format: 
  html:
    embed-resources: true
    standalone: true
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: true
include-in-header:
  - text: |
      <style>
      .panel-tabset > .nav-tabs,
      .panel-tabset > .tab-content {
        border: none;
      }
      </style>
editor: 
  markdown: 
    wrap: sentence
comments:
  hypothesis: 
    theme: clean
---

```{r}
#| label: setup
#| warning: false
#| message: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)

# Load packages
suppressPackageStartupMessages(library(tidyverse))
library(pheatmap)
library(ggpubr)
library(patchwork)
library(DESeq2)
library(DT)
library(gprofiler2)
library(enrichR) # for mammalian phenotypes
library(here)

# Explicitly define common functions to avoid conflicts
select <- dplyr::select
rename <- dplyr::rename
filter <- dplyr::filter
summarize <- dplyr::summarize

set.seed(123) # for reproducibility

# Read sample details
sample_details <- read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 

# Domain colors
domain_colors <- c(PM = rgb(252, 206, 93, maxColorValue = 255), 
                   MAX = rgb(163, 82, 157, maxColorValue = 255), 
                   POST = rgb(158, 156, 201, maxColorValue = 255))

# Strain colors
strain_colors <- c(B6 = "#555555", 
                   CAST = "#009E73",
                   DO = rgb(163, 82, 157, maxColorValue = 255))

# helper functions
create_dt <- function(x){
  
  # Custom formatting function
  format_num <- function(val) {
    case_when(
      is.na(val) ~ NA_real_,
      abs(val) < 0.01 ~ signif(val, 2),
      TRUE ~ round(val, 2)
    )
  }
  
  x <- x |> 
    mutate(across(where(is.numeric), format_num))
  
  DT::datatable(x,
                extensions = 'Buttons',
                rownames = FALSE, 
                filter = "top",
                options = list(dom = 'Blfrtip',
                               buttons = c('copy', 'csv', 'excel'),
                               pageLength = 10, 
                               scrollX = TRUE
                ))
}

# Function to run enrichment with both gprofiler2 AND Enrichr
run_enrichment_hybrid <- function(gene_ids, 
                                 gene_reference = all_genes_palate,
                                 description = "Gene Set",
                                 max_term_size = 660,
                                 min_term_size = 4) {
  
  # Convert gene IDs to symbols
  query_symbols <- gene_reference |>
    filter(gene_id %in% unique(gene_ids)) |>
    pull(symbol) |>
    unique()
  
  # Get background symbols
  background_symbols <- gene_reference$symbol
  
  # Skip if too few genes
  if(length(query_symbols) < 5) {
    message(paste0("Skipping ", description, ": fewer than 5 genes"))
    return(NULL)
  }
  
  # Query 1: gprofiler2 for GO, KEGG, Reactome, WikiPathways
  result_gprofiler <- gost(
    query = query_symbols,
    organism = "mmusculus",
    domain_scope = "custom",
    custom_bg = background_symbols,
    #sources = c("GO:BP", "GO:MF", "GO:CC", "KEGG", "REAC", "WP"),
    evcodes = TRUE
  )
  
  # Query 2: Enrichr for Mammalian Phenotype (full database)
  result_enrichr <- enrichr(
    genes = query_symbols,
    databases = c("MGI_Mammalian_Phenotype_Level_4_2021"),
                  #"MGI_Mammalian_Phenotype_Level_4_2019"),
    background =  background_symbols,
    include_overlap = TRUE
  )
  
  # Format gprofiler results
  gprofiler_df <- NULL
  if(!is.null(result_gprofiler$result)) {
    gprofiler_df <- result_gprofiler$result |>
      filter(term_size < max_term_size) |>
      filter(term_size > min_term_size) |>
      mutate(gene_set = description) |>
      select(gene_set, source, term_id, term_name, p_value, 
             query_size, term_size, intersection_size, intersection)
  }
  
  # Format Enrichr results to match gprofiler format
  enrichr_df <- NULL
  if(!is.null(result_enrichr) && length(result_enrichr) > 0) {
    # Process each database separately
    enrichr_dfs <- lapply(names(result_enrichr), function(db_name) {
      db_result <- result_enrichr[[db_name]]
      
      # Check if result has data (not empty or all NA)
      if(is.null(db_result) || nrow(db_result) == 0 || all(is.na(db_result$Term))) {
        return(NULL)
      }
      
      db_result |>
        filter(Adjusted.P.value < 0.05) |>
        mutate(
          gene_set = description,
          source = "Mammalian Phenotype",
          term_id = Term,
          term_name = Term,
          p_value = Adjusted.P.value,
          query_size = length(query_symbols),
          term_size = as.numeric(gsub("/.*", "", Overlap)),  # Extract from "10/500" format
          intersection_size = as.numeric(gsub("/.*", "", Overlap)),
          intersection = Genes
        ) |>
        filter(term_size < max_term_size) |>
        filter(term_size > min_term_size) |>
        select(gene_set, source, term_id, term_name, p_value, 
               query_size, term_size, intersection_size, intersection)
    })
    
    # Remove NULL entries and combine
    enrichr_dfs <- enrichr_dfs[!sapply(enrichr_dfs, is.null)]
    if(length(enrichr_dfs) > 0) {
      enrichr_df <- bind_rows(enrichr_dfs)
    }
  }
  
  # Combine results
  combined_result <- bind_rows(gprofiler_df, enrichr_df)
  
  if(nrow(combined_result) > 0) {
    return(list(result = combined_result))
  } else {
    return(NULL)
  }
}

# Updated format function
format_enrichment_result <- function(enrichment_table) {
  
  all_results <- enrichment_table$result
  
  if(nrow(all_results) == 0) {
    message("No significant enrichment results found")
    return(NULL)
  }
  
  # Format for display
  all_results |>
    mutate(
      # Make source more readable
      source_type = case_when(
        source == "GO:BP" ~ "GO: Biological Process",
        source == "GO:MF" ~ "GO: Molecular Function",
        source == "GO:CC" ~ "GO: Cellular Component",
        source == "KEGG" ~ "KEGG Pathways",
        source == "REAC" ~ "Reactome Pathways",
        source == "WP" ~ "WikiPathways",
        source == "Mammalian Phenotype" ~ "Mammalian Phenotype (MGI)",
        TRUE ~ source
      )
    ) |>
    select(gene_set, source = source_type, term_id, term_name, p_value, 
           query_size, term_size, intersection_size, intersection) |>
    mutate(
      p_value_fmt = formatC(as.numeric(p_value), digits = 2, format = "e"),
      genes = sapply(strsplit(as.character(intersection), "[,;]"), function(x) {
        x <- trimws(x)
        if(length(x) > 5) {
          paste0(paste(x[1:5], collapse = ", "), ", ...")
        } else {
          paste(x, collapse = ", ")
        }
      })
    ) |>
    select(gene_set, source, term_name, p_value = p_value_fmt, 
           query_size, term_size, intersection_size, genes) |>
    arrange(gene_set, source, p_value)
}

# plotting-functions

# ===== FUNCTION 1: COMPARE UP VS DOWN IN SAME DOMAIN =====
plot_domain_direction_comparison <- function(genes_up, genes_down, 
                                             domain_name, count_data,
                                             n_each = 2) {
  
  gene_col <- "symbol"
  
  # Get examples
  up_examples <- genes_up[1:min(n_each, length(genes_up))]
  down_examples <- genes_down[1:min(n_each, length(genes_down))]
  
  # Create plots for UP genes
  plots_up <- lapply(up_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("↑ in ", domain_name),
           x = "Development stage",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold", size = 11),
            plot.subtitle = element_text(color = "darkgreen"))
  })
  
  # Create plots for DOWN genes
  plots_down <- lapply(down_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("↓ in ", domain_name),
           x = "Development stage",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold", size = 11),
            plot.subtitle = element_text(color = "darkred"))
  })
  
  # Remove NULLs
  plots_up <- plots_up[!sapply(plots_up, is.null)]
  plots_down <- plots_down[!sapply(plots_down, is.null)]
  
  # Combine
  all_plots <- c(plots_up, plots_down)
  if(length(all_plots) == 0) return(NULL)
  
  if(length(all_plots) == 2) {
    combined <- all_plots[[1]] / all_plots[[2]]
  } else if(length(all_plots) == 4) {
    combined <- (all_plots[[1]] | all_plots[[2]]) / (all_plots[[3]] | all_plots[[4]])
  } else {
    combined <- wrap_plots(all_plots, ncol = 2)
  }
  
  combined <- combined +
    plot_annotation(
      title = paste0(domain_name, ": Higher (↑) vs Lower (↓)"),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 2: PLOT DOMAIN HIERARCHY =====
plot_domain_hierarchy <- function(genes_highest, domain_name, count_data, n = 3) {
  
  gene_col <- "symbol"
  
  example_genes <- genes_highest[1:min(n, length(genes_highest))]
  
  plots <- lapply(example_genes, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Create both trajectory and boxplot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0("Highest in ", domain_name),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8),
            plot.title = element_text(face = "bold"))
    
    p2 <- ggplot(gene_data, aes(x = AP_domain, y = count, fill = AP_domain)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_fill_manual(values = domain_colors) +
      labs(x = "", y = "") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none")
    
    return(p1 + p2 + plot_layout(widths = c(2, 1)))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Genes Highest in ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 3: PLOT A-P GRADIENTS =====
plot_AP_gradient <- function(genes_decreasing, genes_increasing, count_data, n_each = 3) {
  
  gene_col <- "symbol"
  
  # Get examples
  dec_examples <- genes_decreasing[1:min(n_each, length(genes_decreasing))]
  inc_examples <- genes_increasing[1:min(n_each, length(genes_increasing))]
  
  # Function to create gradient plot
  make_gradient_plot <- function(gid, gradient_type) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Summary by domain
    domain_summary <- gene_data |> 
      group_by(AP_domain) |> 
      summarize(mean_count = mean(count, na.rm = TRUE), .groups = "drop")
    
    ggplot(domain_summary, aes(x = AP_domain, y = mean_count, fill = AP_domain)) +
      geom_col(alpha = 0.8) +
      geom_point(data = gene_data, aes(y = count), alpha = 0.3, size = 1.5,
                 position = position_jitter(width = 0.1)) +
      scale_fill_manual(values = domain_colors) +
      scale_y_log10(labels = scales::comma) +
      labs(title = gene_name,
           subtitle = ifelse(gradient_type == "dec", "PM > MAX > POST", "POST > MAX > PM"),
           x = "",
           y = "Expression") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = ifelse(gradient_type == "dec", 
                                                        "darkred", "darkgreen")))
  }
  
  # Create plots
  plots_dec <- lapply(dec_examples, function(gid) make_gradient_plot(gid, "dec"))
  plots_inc <- lapply(inc_examples, function(gid) make_gradient_plot(gid, "inc"))
  
  # Remove NULLs
  plots_dec <- plots_dec[!sapply(plots_dec, is.null)]
  plots_inc <- plots_inc[!sapply(plots_inc, is.null)]
  
  # Combine
  all_plots <- c(plots_dec, plots_inc)
  if(length(all_plots) == 0) return(NULL)
  
  combined <- wrap_plots(all_plots, ncol = 3)
  combined <- combined +
    plot_annotation(
      title = "Anterior-Posterior Expression Gradients",
      subtitle = "Top: A→P decreasing | Bottom: A→P increasing",
      theme = theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12)
      )
    )
  
  return(combined)
}

# ===== FUNCTION 4: PLOT DOMAIN-SPECIFIC DYNAMICS =====
plot_domain_specific_dynamics <- function(gene_ids, domain_name, direction,
                                          count_data, slope_data, n = 3) {
  
  gene_col <- "symbol"
  
  example_genes <- gene_ids[1:min(n, length(gene_ids))]
  
  plots <- lapply(example_genes, function(gid) {
    # Get count data
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    # Get slope data
    gene_slopes <- slope_data |> 
      filter(gene_id == gid) |> 
      mutate(Domain = factor(Domain, levels = c("PM", "MAX", "POST")))
    
    # Trajectory plot
    p1 <- ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(direction, " in ", domain_name, " only"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"))
    
    # Slope plot
    if(nrow(gene_slopes) > 0) {
      p2 <- ggplot(gene_slopes, aes(x = Domain, y = log2FoldChange, fill = Domain)) +
        geom_col(alpha = 0.8) +
        geom_errorbar(aes(ymin = log2FoldChange - lfcSE,
                          ymax = log2FoldChange + lfcSE),
                      width = 0.3) +
        geom_hline(yintercept = 0, linetype = "dashed") +
        geom_text(aes(label = ifelse(padj < 0.001, "***",
                                     ifelse(p_value < 0.01, "**",
                                            ifelse(padj < 0.05, "*", "ns")))),
                  vjust = -0.5, size = 4) +
        scale_fill_manual(values = domain_colors) +
        labs(y = "Slope", x = "") +
        theme_pubclean(base_size = 10) +
        theme(legend.position = "none")
      
      return(p1 + p2 + plot_layout(widths = c(2, 1)))
    } else {
      return(p1)
    }
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 1)
  combined <- combined +
    plot_annotation(
      title = paste0("Domain-Specific Dynamics: ", domain_name, " ", direction),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 5: PLOT DIVERGING PATTERNS =====
plot_diverging_detailed <- function(genes_faster, genes_opposite, 
                                    domain_name, count_data, n_each = 2) {
  
  gene_col <- "symbol"
  
  faster_examples <- genes_faster[1:min(n_each, length(genes_faster))]
  opposite_examples <- genes_opposite[1:min(n_each, length(genes_opposite))]
  
  # Plot "faster" pattern
  plots_faster <- lapply(faster_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(domain_name, " increases faster"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = "darkgreen", size = 9))
  })
  
  # Plot "opposite" pattern
  plots_opposite <- lapply(opposite_examples, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           subtitle = paste0(domain_name, " ↑, other ↓"),
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold"),
            plot.subtitle = element_text(color = "purple", size = 9))
  })
  
  # Remove NULLs
  plots_faster <- plots_faster[!sapply(plots_faster, is.null)]
  plots_opposite <- plots_opposite[!sapply(plots_opposite, is.null)]
  
  all_plots <- c(plots_faster, plots_opposite)
  if(length(all_plots) == 0) return(NULL)
  
  if(length(all_plots) <= 2) {
    combined <- wrap_plots(all_plots, ncol = 1)
  } else {
    combined <- wrap_plots(all_plots, ncol = 2)
  }
  
  combined <- combined +
    plot_annotation(
      title = paste0("Diverging Patterns: ", domain_name),
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}

# ===== FUNCTION 6: PLOT SIMPLE TRAJECTORIES =====
plot_simple_trajectories <- function(gene_ids, count_data, title_text, n = 5) {
  
  gene_col <- "symbol"
  
  # Sample if needed
  if(length(gene_ids) > n) {
    set.seed(123)
    gene_ids <- sample(gene_ids, n)
  }
  
  plots <- lapply(gene_ids, function(gid) {
    gene_data <- count_data |> 
      filter(gene_id == gid) |> 
      mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST")))
    
    if(nrow(gene_data) == 0) return(NULL)
    
    gene_name <- unique(gene_data[[gene_col]])[1]
    
    ggplot(gene_data, aes(x = LB_stage_from_start, y = count, color = AP_domain)) +
      geom_point(alpha = 0.5, size = 2) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2, linewidth = 1) +
      scale_y_log10(labels = scales::comma) +
      scale_color_manual(values = domain_colors) +
      labs(title = gene_name,
           x = "Development",
           y = "Counts") +
      theme_pubclean(base_size = 10) +
      theme(legend.position = "bottom",
            legend.text = element_text(size = 8),
            plot.title = element_text(face = "bold", size = 11))
  })
  
  plots <- plots[!sapply(plots, is.null)]
  if(length(plots) == 0) return(NULL)
  
  combined <- wrap_plots(plots, ncol = 3)
  combined <- combined +
    plot_annotation(
      title = title_text,
      theme = theme(plot.title = element_text(size = 14, face = "bold"))
    )
  
  return(combined)
}
```

```{r}
#| label: load-data-emase
#| warning: false
#| message: false

# Loading the count data
load("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/palate_gbrs_emase_counts_v4.RData")

# Clean up expression matrix
expr_mat <- expr_mat_nondo
contam_samples <- c("B6_12.5_PM_1", "B6_13.5_POST_3", "CAST_14.5_MAX_7", "B6_13.5_MAX_3")
expr_mat <- expr_mat[, !colnames(expr_mat) %in% contam_samples]

do_expr_mat <- expr_mat_do
remove_samples <- c("DO_13.5_MAX_6", "DO_13.5_MAX_14", "DO_13.5_MAX_14_2",
                    "DO_12.5_MAX_1", "DO_12.5_MAX_2", "DO_13.5_MAX_97", "DO_13.5_MAX_96")
do_expr_mat <- do_expr_mat[, !colnames(do_expr_mat) %in% remove_samples]

expr_mat_all <- cbind(expr_mat[intersect(rownames(expr_mat), rownames(do_expr_mat)), , drop = FALSE],
                       do_expr_mat[intersect(rownames(expr_mat), rownames(do_expr_mat)), , drop = FALSE])
rm(expr_mat_do, expr_mat_nondo)

# Get v105 annotations from biomart
all.genes_v105 <- read_tsv(here("_data", "ENSMUSGid_to_symbol_v105.txt"),
                           col_names = c("gene_id", "symbol", "gene_biotype", "chromosome", "start", "end"), 
                           skip = 1) |>  
  mutate(midpoint = (start + end) / 2)

all_genes_palate <- all.genes_v105 |>
  filter(gene_id %in% rownames(raw_expr_mat))

all_genes_palate_filtered <- all.genes_v105 |>
  filter(gene_id %in% rownames(expr_mat_all))

# Sample annotations
sample_annot <- sample_details |> 
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat)) |> 
  separate(Strain_stock, into = c("strain", "tmp"), remove = FALSE, sep = "_") |> 
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, palate_closure, L_R_palate) |> 
  mutate(strain = case_when(strain == "B6J" ~ "B6",
                            TRUE ~ strain)) |> 
  mutate(LB_stage_binned = case_when(
    LB_stage < 12.25 ~ "exclude",
    between(LB_stage, 13.25, 13.5) ~ "exclude",
    between(LB_stage, 12.25, 12.75) ~ "1",
    between(LB_stage, 13.5, 14) ~ "2",
    between(LB_stage, 14.25, 14.75) ~ "3",
    between(LB_stage, 15.1, 15.6) ~ "4"
  )) |> 
  mutate(LB_stage_binned = as_factor(LB_stage_binned)) |> 
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |> 
  select(-temp) |> 
  mutate(Palate_ori = ifelse(Palate_ori == "L", "Left", "Right")) 

sample_annot_all <- sample_details |> 
  filter(!sample %in% contam_samples & sample %in% colnames(expr_mat_all)) |> 
  separate(Strain_stock, into = c("strain", "tmp"), remove = FALSE, sep = "_") |> 
  select(sample, sex, AP_domain, Gest_stage, strain, batch, Harvest_date, LB_stage, palate_closure, L_R_palate) |> 
  mutate(strain = case_when(strain == "B6J" ~ "B6",
                            TRUE ~ strain)) |> 
  separate(L_R_palate, into = c("temp", "Palate_ori"), sep = "_") |> 
  select(-temp) |> 
  mutate(Palate_ori = ifelse(Palate_ori == "L", "Left", "Right")) |> 
  mutate(LB_stage_binned = case_when(
    LB_stage < 12.25 ~ "exclude",
    between(LB_stage, 13.25, 13.5) ~ "exclude",
    between(LB_stage, 12.25, 12.75) ~ "1",
    between(LB_stage, 13.5, 14) ~ "2",
    between(LB_stage, 14.25, 14.75) ~ "3",
    between(LB_stage, 15.1, 15.6) ~ "4"
  ))


```

# Differential gene expression analysis overview

```{r}
#| label: prep-run-b6-for-deseq2
#| warning: false
#| message: false

# Prep data for DESeq2
# Subsetting for B6 samples only and adding unique sample IDs
b6_comp_samples <- sample_annot |>
  filter(strain %in% c("B6")) |>
  separate(sample, sep = "_", into = c("s", "e", "d", "ind", "rep"), remove = FALSE) |>
  select(-e, -d) |>
  mutate(rep = ifelse(is.na(rep), 1, rep)) |>
  unite("unique_sampleid", c(s, ind, AP_domain, Gest_stage), sep = "_", remove = FALSE)

# Making the coldata object to feed into DESeq2
b6_coldata <- b6_comp_samples |>
  select(sample, sex, ind, unique_sampleid, rep, AP_domain, Gest_stage, strain, LB_stage) |>
  filter(!is.na(sex)) |>
  mutate_if(is.character, as.factor) |>
  mutate(AP_domain = factor(AP_domain, levels = c("PM", "MAX", "POST"))) |>
  column_to_rownames("sample")

# Get the expression matrix to feed into DESeq2
b6_cts <- round(raw_expr_mat[all_genes_palate$gene_id, rownames(b6_coldata), drop = FALSE])

# Make the dds object then collapse the technical replicates
b6_dds <- DESeqDataSetFromMatrix(countData = b6_cts,
                                  colData = b6_coldata,
                                  design = ~1)

# Collapse the sequencing replicates
b6_dds_coll <- collapseReplicates(b6_dds, b6_dds$unique_sampleid)

# Update LB-stage to have 0 as the starting point
min_LB_stage <- min(colData(b6_dds_coll)$LB_stage)
colData(b6_dds_coll)$LB_stage_from_start <- colData(b6_dds_coll)$LB_stage - min_LB_stage

# Define design matrices
b6_simple_model <- model.matrix(~ sex + ind:AP_domain + LB_stage_from_start + AP_domain, 
                                data = colData(b6_dds_coll))
b6_comp_model <- model.matrix(~ sex + ind:AP_domain + LB_stage_from_start * AP_domain, 
                               data = colData(b6_dds_coll))

# Run DESeq2 models
b6_dds_simple <- DESeq(b6_dds_coll, betaPrior = FALSE, full = b6_simple_model)
b6_dds_interaction <- DESeq(b6_dds_coll, betaPrior = FALSE, full = b6_comp_model)

# Likelihood ratio test to identify genes with significant interaction
b6_dds_lrt <- DESeq(b6_dds_coll, test = "LRT",
                    full = b6_comp_model,
                    reduced = b6_simple_model)

# Identify genes with significant interaction
genes_with_interaction <- results(b6_dds_lrt) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(padj < 0.05) |> 
  pull(gene_id)


# Helper function for consistent filtering
filter_genes <- function(results_df, min_basemean = 50, max_lfcse = 1.5, 
                         max_abs_lfc = 10, min_padj = 0.05, min_abs_lfc = 0.5) {
  results_df |>
    filter(
      baseMean > min_basemean,
      lfcSE < max_lfcse,
      abs(log2FoldChange) < max_abs_lfc,
      abs(log2FoldChange) > min_abs_lfc,
      padj < min_padj
    )
}

# Get genes with effects
# ===== MAIN EFFECTS (genes WITHOUT interaction) =====

# Domain effects from SIMPLE model
b6_domain_max_vs_pm <- results(b6_dds_simple, name = "AP_domainMAX") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Comparison = "MAX vs PM") |> 
  left_join(all_genes_palate, by = "gene_id")

b6_domain_post_vs_pm <- results(b6_dds_simple, name = "AP_domainPOST") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Comparison = "POST vs PM") |> 
  left_join(all_genes_palate, by = "gene_id")

b6_domain_post_vs_max <- results(b6_dds_simple, 
                                 contrast = list("AP_domainPOST", "AP_domainMAX")) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Comparison = "POST vs MAX") |> 
  left_join(all_genes_palate, by = "gene_id")

# Combine domain effects
b6_domain_main_effects <- bind_rows(
  b6_domain_max_vs_pm,
  b6_domain_post_vs_pm,
  b6_domain_post_vs_max
)

# Apply filtering
b6_domain_main_effects_clean <- filter_genes(
  b6_domain_main_effects,
  min_basemean = quantile(b6_domain_main_effects$baseMean, 0.1, na.rm = TRUE)
)

# Stage effects from SIMPLE model
b6_stage_main_effects <- results(b6_dds_simple, name = "LB_stage_from_start") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(!gene_id %in% genes_with_interaction) |> 
  mutate(Direction = ifelse(log2FoldChange > 0, "Increase", "Decrease")) |> 
  left_join(all_genes_palate, by = "gene_id")

# Apply filtering
b6_stage_main_effects_clean <- filter_genes(
  b6_stage_main_effects,
  min_basemean = quantile(b6_stage_main_effects$baseMean, 0.1, na.rm = TRUE)
)

# Note that we need to be very careful when interpreting the main effects with a linear model that contains an interaction term, they do not represent the overall effect of the predictor (e.g. LB_stage) since the predictors involved in an interaction do not have a single overall effect anymore! For example the main effect for LB_stage will represent the effect when AP_domain is at the reference level which is the PM domain in our data set.
# ===== INTERACTION EFFECTS (genes WITH interaction) =====

# 1. INITIAL DOMAIN DIFFERENCES (at earliest stage)
initial_diff_MAX_vs_PM <- results(b6_dds_interaction, name = "AP_domainMAX") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "MAX vs PM at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

initial_diff_POST_vs_PM <- results(b6_dds_interaction, name = "AP_domainPOST") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs PM at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

initial_diff_POST_vs_MAX <- results(b6_dds_interaction, 
                                    contrast = list(c("AP_domainPOST", "AP_domainMAX"))) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs MAX at earliest stage") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine initial differences
b6_initial_differences <- bind_rows(
  initial_diff_MAX_vs_PM,
  initial_diff_POST_vs_PM,
  initial_diff_POST_vs_MAX
)

# 2. DEVELOPMENTAL SLOPES (rates of change in each domain)
slope_PM <- results(b6_dds_interaction, name = "LB_stage_from_start") |> 
  as_tibble(rownames = "gene_id") |> 
  filter(gene_id %in% genes_with_interaction) |> 
  mutate(Domain = "PM") |> 
  left_join(all_genes_palate, by = "gene_id")

slope_MAX <- results(b6_dds_interaction, 
                     contrast = list(c("LB_stage_from_start", 
                                      "AP_domainMAX.LB_stage_from_start"))) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(gene_id %in% genes_with_interaction) |> 
  mutate(Domain = "MAX") |> 
  left_join(all_genes_palate, by = "gene_id")

slope_POST <- results(b6_dds_interaction, 
                      contrast = list(c("LB_stage_from_start", 
                                       "AP_domainPOST.LB_stage_from_start"))) |> 
  as_tibble(rownames = "gene_id") |> 
  filter(gene_id %in% genes_with_interaction) |> 
  mutate(Domain = "POST") |> 
  left_join(all_genes_palate, by = "gene_id")

# Combine all slopes
b6_domain_slopes <- bind_rows(slope_PM, slope_MAX, slope_POST)

# Apply filtering to slopes
b6_domain_slopes_clean <- filter_genes(
  b6_domain_slopes,
  min_basemean = quantile(b6_domain_slopes$baseMean, 0.1, na.rm = TRUE),
  max_lfcse = 2
)

# 3. SLOPE DIFFERENCES (interaction terms - how different are the slopes?)
slope_diff_MAX_vs_PM <- results(b6_dds_interaction, 
                                name = "AP_domainMAX.LB_stage_from_start") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "MAX vs PM slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

slope_diff_POST_vs_PM <- results(b6_dds_interaction, 
                                 name = "AP_domainPOST.LB_stage_from_start") |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs PM slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

slope_diff_POST_vs_MAX <- results(b6_dds_interaction, 
                                  contrast = list(c("AP_domainPOST.LB_stage_from_start",
                                                   "AP_domainMAX.LB_stage_from_start"))) |>
  as_tibble(rownames = "gene_id") |>
  filter(gene_id %in% genes_with_interaction) |>
  mutate(Comparison = "POST vs MAX slope difference") |>
  left_join(all_genes_palate, by = "gene_id")

# Combine slope differences
b6_slope_differences <- bind_rows(
  slope_diff_MAX_vs_PM,
  slope_diff_POST_vs_PM,
  slope_diff_POST_vs_MAX
)

# Get normalized counts for plotting
b6_gene_counts <- counts(b6_dds_simple, normalized = TRUE) |> 
  as_tibble(rownames = "gene_id") |>
  pivot_longer(-gene_id, names_to = "sample", values_to = "count") |> 
  left_join(
    colData(b6_dds_simple) |> 
      as_tibble(rownames = "sample"),
    by = "sample"
  ) |>
  left_join(all_genes_palate, by = "gene_id")


# Run pattern identification script that generates the pattern_summary, gene_assignments and domain_dynamics files to read in.
# source(here("DEgene_pattern_identification_B6.R"))

# Load pattern assignments and summaries
pattern_summary <- read_csv(here("pattern_summary_detailed.csv"))
gene_assignments <- read_csv(here("gene_pattern_assignments_detailed.csv"))
domain_dynamics <- read_csv(here("domain_specific_dynamics_detailed.csv"))
```

Here, I analyze the B6 data on its own to identify genes differentially expressed across AP domains, over developmental time and change differently over time across domains. The data from B6 animals includes males and females, and contain the three AP domains (PM, MAX and POST) obtained at various developmental stages (LB_stage).
I merge the technical replicates due to re-sequencing of several samples using `collapseReplicates()` function from DESeq2. 

In order to identify differentially expressed genes between domains and developmental time, I obtain the main effects for domain (AP_domain) and centered developmental time (LB_stage_from_the_start, minimum LB_stage value is subtracted to start the continous variable from 0) while using a nesting for the domain since they are obtained from the same animal (\~ sex + AP_domain + LB_stage + AP_domain:ind).


For identifying genes with significant interactions between domain and time meaning they have different spatiotemporal gene expression patterns I use the likelihood ratio test (LRT) to compare the initial simple model to the more complex linear model that includes the interaction term (\~ sex + AP_domain + LB_stage + AP_domain:ind + AP_domain:LB_stage).
Once I have this DE gene list, I then use the complex model to get the effects of developmental time in all three domains.

From this analysis, I have identified **`r nrow(gene_assignments)` genes** with significant spatiotemporal expression patterns across the three anatomical domains (PM, MAX, POST) and developmental time.
I used `DEgene_pattern_identification_B6.R` script to classify these differentially expressed genes into six major categories based on expression patterns:

```{r}
#| label: summary-table
#| warning: false
#| message: false

pattern_summary |> 
  group_by(Category) |> 
  summarize(
    `Total Patterns` = n(),
    `Total Genes` = sum(N_Genes),
    .groups = "drop"
  ) |> 
  create_dt()
```

Briefly, I explain how each gene list is generated below:

First, DESeq2 results were filtered to retain high-confidence calls: baseMean \> 10th percentile, log2FoldChange standard error \< 1.5 (or \< 2 for slope analyses), \|log2FoldChange\| between 0.5 and 10, and padj \< 0.05.

**1. Domain Main Effects (Parallel Trajectories)**\
For genes *without* significant interactions, domain differences were extracted from the simple model using coefficients for MAX vs PM (`AP_domainMAX`), POST vs PM (`AP_domainPOST`), and POST vs MAX (contrast: `list("AP_domainPOST", "AP_domainMAX")`).
Complex patterns (e.g., "highest in MAX") were identified by intersecting multiple pairwise comparisons.

**2. Stage Main Effects (Common Trajectories)**\
For genes *without* interactions, developmental changes common across domains were extracted from the simple model coefficient `LB_stage_from_start`.
Genes were classified as increasing or decreasing based on log2FoldChange sign, with strong changers defined as \|log2FC\| \> 1.

**3. Diverging Trajectories**\
For genes *with* significant interactions, domain-specific developmental slopes were extracted from the complex model: PM slope (`LB_stage_from_start`), MAX slope (contrast: `list(c("LB_stage_from_start", "AP_domainMAX.LB_stage_from_start"))`), and POST slope (contrast: `list(c("LB_stage_from_start", "AP_domainPOST.LB_stage_from_start"))`).
Slope differences (interaction terms) were extracted as: MAX vs PM (`AP_domainMAX.LB_stage_from_start`), POST vs PM (`AP_domainPOST.LB_stage_from_start`), and POST vs MAX (contrast: `list(c("AP_domainPOST.LB_stage_from_start", "AP_domainMAX.LB_stage_from_start"))`).
Diverging genes required padj \< 0.05 and log2FoldChange \> 0.5 for the slope difference.
Individual domain slopes (including non-significant ones) were examined to classify divergence patterns (e.g., "MAX increases faster" vs "MAX increases, PM decreases").

**4. Converging Trajectories**\
Converging genes were identified using the same slope difference tests as diverging genes but requiring negative slope differences (log2FoldChange \< -0.5).

**5. Domain-Specific Dynamics**\
Genes showing developmental changes in specific domains were identified from filtered domain slopes (padj \< 0.05, \|log2FC\| \> 0.5) and classified by the combination of dynamic domains.

**6. Complex Patterns**\
Genes with opposite slopes across domains or three-way divergent patterns were identified by analyzing the signs and significance of all three domain slopes simultaneously.

:::::::::: {#B6_DESeq2_effects .panel-tabset .nav-pills}
## Domain Main Effects

Here I am highlighting genes that show differences in expression between A-P domains in B6.
In total there are `r length(unique(b6_domain_main_effects_clean$gene_id))` unique genes with a domain effect (i.e. differentially expressed between any two domains) that doesn't change over developmental time (i.e. maintain parallel developmental trajectories).
This means that while one domain might express a gene at higher or lower levels than another, all domains follow the same developmental pattern over time (all increasing, all decreasing, or all stable).

Below is a summary of the over-representation results from the individual comparisons. Full ORA results are displayed in tables in each subsection tab.

| Category | PM (Anterior) | MAX (Middle) | POST (Posterior) | Interpretation |
|----------|---------------|--------------|------------------|----------------|
| **Wnt Signaling** | ↑↑↑ | ↑ | ↓↓ | A→P gradient, FNP identity |
| **TGF-β/BMP Signaling** | ↑↑↑ | ↓↓ | ↓ | PM bone formation |
| **Skeletal/Bone Development** | ↑↑↑ | ↓↓ | ↓ | Premaxillary ossification |
| **Embryonic/Developmental** | ↑↑↑ | ↓ | ↓↓ | PM progenitor state |
| **Synaptic Structure** | ↓↓↓ | ↑↑↑ | ↑↑↑ | Secondary palate neural differentiation |
| **Ion Channels** | ↓↓ | ↑↑↑ | ↑↑ | Neural function in MAX/POST |
| **Cognition/Behavior** | ↓↓ | ↑↑↑ | ↑↑ | Neural maturation markers |
| **Muscle/Sarcomere** | — | — | ↑↑↑ | Soft palate musculature |
| **Epithelial/Skin** | ↑↑ | ↑ | ↓↓ | Epithelial differentiation in PM/MAX |
| **EMT/Migration** | ↑↑↑ | ↓↓ | ↓ | Active in FNP-derived PM |

*Legend: ↑↑↑ strongly enriched, ↑↑ moderately enriched, ↑ slightly enriched, — neutral, ↓ slightly depleted, ↓↓ moderately depleted, ↓↓↓ strongly depleted*


::: {#B6_DESeq2_domain_effects .panel-tabset .nav-pills}

### Overview

```{r}
#| label: domain-main-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Domain Main Effects") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### MAX vs PM

```{r}
#| label: max-vs-pm-plot
#| warning: false
#| message: false

genes_MAX_higher_than_PM <- gene_assignments |>
  filter(MAX_higher_PM) |>
  pull(gene_id)

genes_MAX_lower_than_PM <- gene_assignments |>
  filter(MAX_lower_PM) |>
  pull(gene_id)

if(length(genes_MAX_higher_than_PM) > 0 & length(genes_MAX_lower_than_PM) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_MAX_higher_than_PM,
    genes_down = genes_MAX_lower_than_PM,
    domain_name = "MAX vs PM",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**Pattern:** Genes showing consistent differences between MAX and PM domains across development.

**MAX higher than PM** contains `r length(genes_MAX_higher_than_PM)` genes and are over-representation analysis shows that it exhibits a mature neuronal phenotype with emphasis on synaptic connectivity, neurotransmission, and neural circuit function.

| Category | Key Themes | Top Terms (by p-value) | Representative Genes |
|----------|-----------|----------------------|---------------------|
| **Synaptic Structure** | Postsynaptic specialization, synaptic membranes | Synaptic membrane (p=1.3e-9), Postsynaptic specialization (p=1.6e-6) | Snap25, Dlgap3, Bsn, Syn2, Nrxn2 |
| **Neuron Development** | Axonogenesis, growth cones, neurite morphology | Regulation of neuron projection development (p=3.7e-8), Growth cone (p=1.4e-6) | Map1b, Nefl, Kif21a, Plxna4 |
| **Ion Channels** | Potassium channels, cation transport | Neuronal System (p=1.1e-8), Potassium channels (p=0.0094) | Kcna4, Kcna5, Kcnd2, Kcnj12, Scn2a |
| **Cognition/Behavior** | Learning, memory, locomotion | Cognition (p=7.7e-6), Learning or memory (p=1.3e-5) | Chrnb2, Atp1a3, Plcb1, Gpr88 |
| **Mammalian Phenotypes** | Motor deficits, seizures, hyperactivity | Impaired coordination (p=3e-6), Ataxia (p=7.7e-6), Hyperactivity (p=3.7e-5) | Snap25, Atp1a3, Map1b, Plp1 |


**MAX lower than PM** contains `r length(genes_MAX_lower_than_PM)` genes and over-representation analysis shows that it contains a **developmental/embryonic phenotype** dominated by morphogenetic processes, particularly skeletal patterning, organogenesis (kidney, heart, limb), and TGF-β superfamily signaling.

| Category | Key Themes | Top Terms (by p-value) | Representative Genes |
|----------|-----------|----------------------|---------------------|
| **Skeletal Development** | Bone, cartilage, limb morphogenesis | Skeletal system development (p=2.4e-5), Cartilage development (p=7.9e-5) | Msx2, Tgfb1, Tgfb2, Bmp7, Runx3 |
| **Heart Development** | Cardiac morphogenesis, cardiovascular physiology | Heart morphogenesis (p=5.6e-4), Heart development (p=0.0097) | Gja1, Gata3, Smad6, Hopx, Rbm20 |
| **TGF-β Signaling** | BMP pathway, receptor serine/threonine kinase signaling | TGF-β receptor superfamily signaling (p=0.0039), TGF-β pathway (p=0.0092) | Tgfb1, Tgfb2, Smad6, Bmp7, Nbl1 |
| **EMT & Migration** | Epithelial-mesenchymal transition, endothelial migration | Positive regulation of EMT (p=0.0075), Endothelial cell migration (p=0.0015) | Loxl2, Emp2, Gata3 |
| **Extracellular Matrix** | Basement membrane, elastic fibers | Extracellular matrix (p=6.7e-5), Elastic fibre formation (p=1.3e-4) | Bgn, Loxl2, Serpinf1, Lamc3 |
| **Kidney Development** | Renal system morphogenesis | Renal system development (p=0.001) | Sall1, Wnt9b, Emx2 |

```{r}
#| label: max-vs-pm-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "MAX vs PM") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: mx-vs-pm-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on both sets
genes_MAX_higher_than_PM_ora <- run_enrichment_hybrid(gene_ids = genes_MAX_higher_than_PM,
                                     description = "MAX higher than PM"
                                     )

genes_MAX_lower_than_PM_ora <- run_enrichment_hybrid(gene_ids = genes_MAX_lower_than_PM,
                                     description = "MAX lower than PM"
                                     )
```

```{r}
#| label: mx-vs-pm-ora-table
#| warning: false
#| message: false

genes_MAX_higher_than_PM_ora$result |> 
  rbind(
    genes_MAX_lower_than_PM_ora$result
  ) |> 
  filter( p_value < 0.01) |> 
  create_dt()


```

### POST vs PM

```{r}
#| label: post-vs-pm-plot
#| warning: false
#| message: false

genes_POST_higher_than_PM <- gene_assignments |>
  filter(POST_higher_PM) |>
  pull(gene_id)

genes_POST_lower_than_PM <- gene_assignments |>
  filter(POST_lower_PM) |>
  pull(gene_id)

if(length(genes_POST_higher_than_PM) > 0 & length(genes_POST_lower_than_PM) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_POST_higher_than_PM,
    genes_down = genes_POST_lower_than_PM,
    domain_name = "POST vs PM",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**Pattern:** Consistent expression differences between posterior and palatine midline domains.

**POST higher than PM** contains `r length(genes_POST_higher_than_PM)` genes and over-representation analysis show that it exhibits a mature functional phenotype (synaptic transmission, ion channels, behavior) combined with a notable muscle/cardiac signature (sarcomere, myofibrils, cardiac conduction).

| Category | Key Themes | Top Terms (by p-value) | Representative Genes |
|----------|-----------|----------------------|---------------------|
| **Synaptic Structure** | Synaptic membranes, postsynaptic specialization, neuron-to-neuron synapses | Synaptic membrane (p=4.5e-13), Postsynaptic specialization (p=3.2e-9) | Nlgn1, Dlgap3, Shisa7, Nrxn2, Lrrtm1, Gria2 |
| **Synaptic Signaling** | Synapse organization, trans-synaptic signaling, chemical transmission | Synapse organization (p=1.8e-8), Modulation of chemical synaptic transmission (p=2.7e-7) | Camk2b, Dgkb, Nrn1, Apba1, Plcb1 |
| **Ion Channels** | Cation channels, membrane potential regulation, gated channels | Ion channel complex (p=2e-5), Cation channel activity (p=0.00099) | Kcnd2, Kcna4, Kcna5, Cacnb1, Trpc4, Trpc5 |
| **Muscle/Cardiac** | Muscle contraction, myofibril assembly, sarcomere, cardiac conduction | Muscle contraction (p=5.9e-6), Cardiac conduction (p=0.00031) | Ryr1, Tmod1, Lmod2, Myh7b, Tnnt3, Myoz1 |
| **Behavior/Cognition** | Locomotion, adult behavior, anxiety, learning | Locomotory behavior (p=0.00059), Decreased anxiety (p=0.00082) | Atp1a3, Gpr88, Ppp1r1b, Camk2b |
| **Mammalian Phenotypes** | Hyperactivity, seizures, impaired learning | Hyperactivity (p=3e-5), Seizures (p=0.0015) | Camk2b, Gria2, Kcnt1, Atp1a3, Cers1 |


**POST lower than PM** contains `r length(genes_POST_lower_than_PM)` genes and over-representation analysis shows that they are enriched for early developmental processes including embryonic patterning (Wnt, Hedgehog/Gli signaling), organogenesis (tooth, kidney, skeleton), and epithelial differentiation (skin, desmosomes). 

| Category | Key Themes | Top Terms (by p-value) | Representative Genes |
|----------|-----------|----------------------|---------------------|
| **Tooth Development** | Odontogenesis, dentin formation, enamel | Odontogenesis of dentin-containing tooth (p=1.4e-6) | Enam, Odam, Runx2, Msx1, Trp63 |
| **Embryonic Development** | Embryonic morphogenesis, organ development, pattern specification | Embryonic morphogenesis (p=4.1e-5), Embryonic organ development (p=4.6e-5) | Gli3, Gli1, Wnt3a, Wnt7a, Pax6, Gata3 |
| **Skeletal Development** | Bone, cartilage, ossification, osteoblast differentiation | Skeletal system development (p=0.00017), Osteoblast differentiation (p=0.00037) | Runx2, Satb2, Gja1, Tgfb1, Fam20c |
| **Wnt Signaling** | WNT ligand biogenesis, basal cell carcinoma pathway | WNT ligand biogenesis (p=0.0036), Basal cell carcinoma (p=0.0002) | Wnt3a, Wnt7a, Wnt9a, Wnt2b, Wnt9b |
| **Epithelial/Skin** | Epidermis development, cornified envelope, keratoderma | Cornified envelope formation (p=3.8e-5), Epidermis development (p=0.0012) | Trp63, Spink5, Ovol1, Krt79, Dsg1a, Dsc2 |
| **Kidney Development** | Ureteric bud, mesonephros development | Ureteric bud development (p=0.0029), Single kidney phenotype (p=0.00064) | Gdnf, Wnt9b, Gata3, Gli3, Npnt, Itga8 |
| **ECM/Cell Adhesion** | Desmosome, ECM-receptor interaction, integrins | Desmosome (p=0.00058), ECM-receptor interaction (p=3.3e-5) | Itgb3, Itgav, Sdc4, Cd44, Dsc2, Dsg2 |
| **Mammalian Phenotypes** | Neonatal lethality, skin barrier defects, fertility | Neonatal lethality (p=5.4e-6), Impaired skin barrier (p=0.0011) | Gli3, Msx1, Runx2, Spink5, Gja1 |



```{r}
#| label: post-vs-pm-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "POST vs PM") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-vs-pm-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on both sets
genes_POST_higher_than_PM_ora <- run_enrichment_hybrid(gene_ids = genes_POST_higher_than_PM,
                                     description = "POST higher than PM"
                                     )

genes_POST_lower_than_PM_ora <- run_enrichment_hybrid(gene_ids = genes_POST_lower_than_PM,
                                     description = "POST lower than PM"
                                     )
```

```{r}
#| label: post-vs-pm-ora-table
#| warning: false
#| message: false
genes_POST_higher_than_PM_ora$result |> 
  rbind(
    genes_POST_lower_than_PM_ora$result
  ) |> 
  filter( p_value < 0.01) |> 
  create_dt()
```

### POST vs MAX

```{r}
#| label: post-vs-max-plot
#| warning: false
#| message: false

genes_POST_higher_than_MAX <- gene_assignments |>
  filter(POST_higher_MAX) |>
  pull(gene_id)

genes_POST_lower_than_MAX <- gene_assignments |>
  filter(POST_lower_MAX) |>
  pull(gene_id)

if(length(genes_POST_higher_than_MAX) > 0 & length(genes_POST_lower_than_MAX) > 0) {
  plot_domain_direction_comparison(
    genes_up = genes_POST_higher_than_MAX,
    genes_down = genes_POST_lower_than_MAX,
    domain_name = "POST vs MAX",
    count_data = b6_gene_counts,
    n_each = 2
  )
}
```

**Pattern:** Expression differences between posterior and maxillary domains.

**POST higher than MAX** contain `r length(genes_POST_higher_than_MAX)` genes and over-representation analysis shows an enrichment for **muscle/contractile apparatus** genes—sarcomere components, myofibrils, and muscle contraction machinery. The strong human phenotype associations with myopathies (Gowers sign, muscle weakness, abnormal muscle fibers) underscore this muscle identity. There's also a cell adhesion signature involving protocadherins.

| Category | Key Themes | Top Terms (by p-value) | Representative Genes |
|----------|-----------|----------------------|---------------------|
| **Cell Adhesion** | Plasma membrane adhesion, protocadherins | Cell-cell adhesion via plasma-membrane adhesion molecules (p=4.8e-7) | Pcdh11x, Pcdh18, Pcdh15, Cdh18, Dchs2, Tenm4 |
| **Muscle Contraction** | Muscle system process, sarcomere function | Muscle contraction (p=2.2e-6), Muscle system process (p=0.0002) | Ryr1, Tmod1, Lmod2, Myh7b, Tnnt3, Tpm2 |
| **Sarcomere/Myofibril** | Thin filaments, contractile fibers | Myofibril (p=1.5e-5), Sarcomere (p=4.7e-5) | Tcap, Ldb3, Nexn, Ank1, Unc45b |
| **Synapse Organization** | Presynapse assembly, protein-protein interactions | Synapse organization (p=3.3e-5), Receptor-type tyrosine-protein phosphatases (p=2.3e-5) | Il1rap, Il1rapl1, Il1rapl2, Lrrc4b, Slitrk3 |
| **Extracellular Matrix** | Collagen, ECM components | Extracellular matrix (p=0.001), Collagen trimer (p=0.0049) | Col3a1, Col8a1, Col16a1, Lum, Thbs2 |
| **Human Phenotypes** | Myopathy, muscle weakness, respiratory distress | Gowers sign (p=5.2e-5), Proximal muscle weakness (p=0.00012), Myopathy (p=0.0002) | Sgcg, Ryr1, Scn4a, Dok7, Tpm2, Unc45b |
| **Mammalian Phenotypes** | Sarcomere defects, respiratory distress | Respiratory distress (p=0.00016), Abnormal sarcomere morphology (p=0.00023) | Ryr1, Lmod2, Stac3, Ldb3, Scn4a |


**POST lower than MAX** contain `r length(genes_POST_lower_than_MAX)` genes and over-representation analysis show that it displays a mixed **neural + epithelial** signature. The neural component includes synapse organization and axon guidance, while the epithelial component features epidermis development, desmosomes, and cornified envelope. Wnt signaling is prominent, along with developmental processes (cell fate commitment, kidney/nephron development).


| Category | Key Themes | Top Terms (by p-value) | Representative Genes |
|----------|-----------|----------------------|---------------------|
| **Synapse Organization** | Synapse assembly, synaptic regulation | Synapse organization (p=1.4e-5), Synapse assembly (p=0.00017) | Snap25, Snca, Cyfip2, Dock4, Celsr2, Oxtr |
| **Epidermis/Skin** | Epidermal differentiation, cornified envelope | Epidermis development (p=0.00011), Cornified envelope formation (p=0.00026) | Trp63, Spink5, Krt79, Gata6, Pax6, Tfap2c |
| **Wnt Signaling** | WNT ligand biogenesis, stem cell pluripotency | Basal cell carcinoma/Wnt (p=0.00018), WNT ligand biogenesis (p=0.00061) | Wnt3a, Wnt7a, Wnt9a, Wnt2b, Wnt9b |
| **Cell Adhesion** | Desmosomes, homophilic adhesion | Desmosome (p=0.0024), Homophilic cell adhesion (p=0.00088) | Dsc3, Dsg1a, Dsg1b, Dsg2, Pkp3, Nectin4 |
| **Cell Fate/Development** | Cell fate commitment, embryonic morphogenesis | Cell fate commitment (p=0.0013), Embryonic morphogenesis (p=0.0061) | Foxg1, Gata3, Gata6, Pax6, Gli1, Msx1 |
| **Kidney Development** | Nephron/tubule formation, ureteric bud | Mesonephric tubule formation (p=0.0027), Nephron development (p=0.0081) | Gdnf, Six1, Wnt9b, Gata3, Npnt |
| **Neuron Development** | Axon guidance, neuron projection | Cell morphogenesis in neuron differentiation (p=0.0011), Axon guidance (p=0.0039) | Cntn1, Sema4d, Cyfip2, Lama5, Pax6 |





```{r}
#| label: post-vs-max-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(Comparison == "POST vs MAX") |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(padj) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-vs-max-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on both sets
genes_POST_higher_than_MAX_ora <- run_enrichment_hybrid(gene_ids = genes_POST_higher_than_MAX,
                                     description = "POST higher than MAX"
                                     )

genes_POST_lower_than_MAX_ora <- run_enrichment_hybrid(gene_ids = genes_POST_lower_than_MAX,
                                     description = "POST lower than MAX"
                                     )
```

```{r}
#| label: post-vs-max-ora-table
#| warning: false
#| message: false

genes_POST_higher_than_MAX_ora$result |> 
  rbind(
    genes_POST_lower_than_MAX_ora$result
  ) |> 
  filter( p_value < 0.01) |> 
  create_dt()





```

### Highest in Each Domain

#### PM Highest

```{r}
#| label: pm-highest-plot
#| warning: false
#| message: false

genes_PM_highest <- gene_assignments |>
  filter(Highest_PM) |>
  pull(gene_id)

if(length(genes_PM_highest) > 0) {
  n_show <- ifelse(length(genes_PM_highest) < 15, length(genes_PM_highest), 3)
  plot_domain_hierarchy(
    genes_highest = genes_PM_highest,
    domain_name = "PM",
    count_data = b6_gene_counts,
    n = n_show
  )
}
```

`r length(genes_PM_highest)` genes with highest expression at the PM compared to both MAX and POST domains.

```{r}
#| label: pm-highest-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% genes_PM_highest) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  distinct(gene_id, .keep_all = TRUE) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: pm-highest-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_PM_highest_ora <- run_enrichment_hybrid(gene_ids = genes_PM_highest,
                                     description = "PM highest"
                                     )
```

```{r}
#| label: pm-highest-ora-table
#| warning: false
#| message: false

genes_PM_highest_ora$result |> 
  create_dt()
```

#### MAX Highest

```{r}
#| label: max-highest-plot
#| warning: false
#| message: false

genes_MAX_highest <- gene_assignments |>
  filter(Highest_MAX) |>
  pull(gene_id)

if(length(genes_MAX_highest) > 0) {
  n_show <- ifelse(length(genes_MAX_highest) < 15, length(genes_MAX_highest), 3)
  plot_domain_hierarchy(
    genes_highest = genes_MAX_highest,
    domain_name = "MAX",
    count_data = b6_gene_counts,
    n = n_show
  )
}
```

`r length(genes_MAX_highest)` genes showing highest expression in MAX compared to both PM and POST domains.


```{r}
#| label: max-highest-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% genes_MAX_highest) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  distinct(gene_id, .keep_all = TRUE) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: max-highest-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_MAX_highest_ora <- run_enrichment_hybrid(gene_ids = genes_MAX_highest,
                                     description = "MAX highest"
                                     )
```

```{r}
#| label: max-highest-ora-table
#| warning: false
#| message: false
genes_MAX_highest_ora$result |> 
  filter(p_value < 0.01) |> 
  create_dt()
```

#### POST Highest

```{r}
#| label: post-highest-plot
#| warning: false
#| message: false

genes_POST_highest <- gene_assignments |>
  filter(Highest_POST) |>
  pull(gene_id)

if(length(genes_POST_highest) > 0) {
  n_show <- ifelse(length(genes_POST_highest) < 15, length(genes_POST_highest), 3)
  plot_domain_hierarchy(
    genes_highest = genes_POST_highest,
    domain_name = "POST",
    count_data = b6_gene_counts,
    n = n_show
  )
}
```

`r length(genes_POST_highest)` genes with peak expression in POST in comparison to MAX and PM domains.


```{r}
#| label: post-highest-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% genes_POST_highest) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  distinct(gene_id, .keep_all = TRUE) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-highest-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_POST_highest_ora <- run_enrichment_hybrid(gene_ids = genes_POST_highest,
                                     description = "POST highest"
                                     )
```

```{r}
#| label: post-highest-ora-table
#| warning: false
#| message: false
genes_POST_highest_ora$result |> 
  filter( p_value < 0.01) |> 
  create_dt()
```



### A-P Gradients

```{r}
#| label: ap-gradients-plot
#| warning: false
#| message: false

genes_AP_decreasing <- gene_assignments |>
  filter(AP_decreasing) |>
  pull(gene_id)

genes_AP_increasing <- gene_assignments |>
  filter(AP_increasing) |>
  pull(gene_id)

if(length(genes_AP_decreasing) > 0 | length(genes_AP_increasing) > 0) {
  n_each <- 3
  if(length(genes_AP_decreasing) < 15) n_each <- min(n_each, length(genes_AP_decreasing))
  if(length(genes_AP_increasing) < 15) n_each <- min(n_each, length(genes_AP_increasing))
  
  plot_AP_gradient(
    genes_decreasing = genes_AP_decreasing,
    genes_increasing = genes_AP_increasing,
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** Monotonic gradients along the anterior-posterior axis.

-   **A→P decreasing** (PM \> MAX \> POST): `r length(genes_AP_decreasing)` genes. 
-   **A→P increasing** (POST \> MAX \> PM): `r length(genes_AP_increasing)` genes. 

```{r}
#| label: ap-gradients-table
#| warning: false
#| message: false

b6_domain_main_effects_clean |>
  filter(gene_id %in% c(genes_AP_decreasing, genes_AP_increasing)) |>
  mutate(Gradient = ifelse(gene_id %in% genes_AP_decreasing, "A→P decreasing", "A→P increasing")) |>
  select(gene_id, symbol, chromosome, Gradient, Comparison, log2FoldChange, padj) |>
  arrange(Gradient, padj) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: ap-gradients-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on both sets
genes_AP_decreasing_ora <- run_enrichment_hybrid(gene_ids = genes_AP_decreasing,
                                     description = "A→P decreasing"
                                     )

genes_AP_increasing_ora <- run_enrichment_hybrid(gene_ids = genes_AP_increasing,
                                     description = "A→P increasing"
                                     )

```

```{r}
#| label: ap-gradients-ora-table
#| warning: false
#| message: false
genes_AP_decreasing_ora$result |> 
  rbind(
    genes_AP_increasing_ora$result
  ) |> 
  create_dt()
```
:::

## Stage Main Effects

These genes show developmental changes that are consistent across all three domains. All domains increase or decrease together, representing shared developmental programs.

::: {#B6_DESeq2_stage_effects .panel-tabset .nav-pills}
### Overview

```{r}
#| label: stage-main-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Stage Main Effects") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### Increasing with Development

```{r}
#| label: dev-increasing-plot
#| warning: false
#| message: false

genes_increasing <- gene_assignments |>
  filter(Dev_increasing) |>
  pull(gene_id)

if(length(genes_increasing) > 0) {
  n_show <- ifelse(length(genes_increasing) < 15, length(genes_increasing), 5)
  plot_simple_trajectories(
    gene_ids = genes_increasing,
    count_data = b6_gene_counts,
    title_text = paste0("Increasing with Development (n=", length(genes_increasing), ")"),
    n = n_show
  )
}
```

`r length(genes_increasing)` genes increasing across all domains as development progresses showing an over-representation of genes involved in ciliary assembly, extracellular matrix, immune response, epithelial differentiation and ion transport. 

| Category | Key Terms | Top Genes | Best p-value |
|----------|-----------|-----------|--------------|
| Ciliary assembly/motility | Cilium movement, axoneme assembly, microtubule bundle formation | *Cfap91*, *Spag6l*, *Nme5*, *Spef2*, *Dnah11*, *Rsph4a* | 4.17×10⁻¹⁷ |
| Extracellular matrix | ECM organization, collagen formation, integrin interactions | *Col1a2*, *Col5a2*, *Col8a1*, *Bgn*, *Lum*, *Postn* | 4.36×10⁻¹¹ |
| Immune/inflammatory | Leukocyte migration, cytokine production, antigen presentation | *Cd74*, *Lgals3*, *Tlr9*, *Tgfb1*, *Il18*, *Tnf* | 1.29×10⁻⁹ |
| Epithelial differentiation | Cornified envelope, keratinization, intermediate filaments | *Krt6a*, *Krt16*, *Krt17*, *Dsc2*, *Dsg1a*, *Cdsn* | 1.23×10⁻¹¹ |
| Ion transport | Sodium/potassium channels, transmembrane transport | *Cftr*, *Slc6a19*, *Kcnj15*, *Scnn1b* | 1.62×10⁻⁴ |

```{r}
#| label: dev-increasing-table
#| warning: false
#| message: false

b6_stage_main_effects_clean |>
  filter(Direction == "Increase") |>
  select(gene_id, symbol, chromosome, Direction, log2FoldChange, padj) |>
  arrange(desc(log2FoldChange)) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: dev-increasing-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on all sets
dev_increasing_enrichment_results <- run_enrichment_hybrid(gene_ids = b6_stage_main_effects_clean |>
                                       filter(Direction == "Increase") |> 
                                         pull(gene_id),
                                     description = "Increase"
                                     )

# Format and combine results
dev_increasing_enrichment_table <- format_enrichment_result(dev_increasing_enrichment_results)
```

```{r}
#| label: dev-increasing-ora-table
#| warning: false
#| message: false

dev_increasing_enrichment_table |> 
  create_dt()

```

### Decreasing with Development

```{r}
#| label: dev-decreasing-plot
#| warning: false
#| message: false

genes_decreasing <- gene_assignments |>
  filter(Dev_decreasing) |>
  pull(gene_id)

if(length(genes_decreasing) > 0) {
  n_show <- ifelse(length(genes_decreasing) < 15, length(genes_decreasing), 5)
  plot_simple_trajectories(
    gene_ids = genes_decreasing,
    count_data = b6_gene_counts,
    title_text = paste0("Decreasing with Development (n=", length(genes_decreasing), ")"),
    n = n_show
  )
}
```

`r length(genes_decreasing)` genes decreasing across all domains that are over-represented for genes involved in  erythropoiesis and early hematopoietic processes.

| Category | Key Terms | Top Genes | Best p-value |
|----------|-----------|-----------|--------------|
| Erythropoiesis/gas transport | Hemoglobin complex, oxygen transport, erythrocyte development | *Hbb-bh1*, *Hbb-y*, *Hba-x*, *Hbq1b*, *Gata1*, *Rhag* | 5.17×10⁻⁵ |
| Platelet function | Platelet aggregation, megakaryocyte development | *Gp9*, *Gfi1b*, *Gp6*, *Tubb1*, *Mpl* | 1.73×10⁻³ |


```{r}
#| label: dev-decreasing-table
#| warning: false
#| message: false

b6_stage_main_effects_clean |>
  filter(Direction == "Decrease") |>
  select(gene_id, symbol, chromosome, Direction, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: dev-decreasing-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on all sets
dev_decreasing_enrichment_results <- run_enrichment_hybrid(gene_ids = b6_stage_main_effects_clean |>
                                       filter(Direction == "Decrease") |> 
                                         pull(gene_id),
                                     description = "Decrease"
                                     )

# Format and combine results
dev_decreasing_enrichment_table <- format_enrichment_result(dev_decreasing_enrichment_results)
```

```{r}
#| label: dev-decreasing-ora-table
#| warning: false
#| message: false
dev_decreasing_enrichment_table |> 
  create_dt()

```
:::

## Diverging Trajectories

**These genes show different developmental trajectories between domains** - true domain × stage interactions where domains respond differently to developmental cues.

::: {#B6_DESeq2_diverging_traj .panel-tabset .nav-pills}
### Overview

```{r}
#| label: diverging-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Diverging") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### MAX Diverging from PM

```{r}
#| label: max-div-pm-plot
#| warning: false
#| message: false

genes_MAX_div <- gene_assignments |>
  filter(MAX_div_PM) |>
  pull(gene_id)

# Get faster vs opposite
genes_MAX_faster <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_MAX_div, Domain == "MAX", log2FoldChange > 0.5, padj < 0.05) |>
  pull(gene_id)

genes_MAX_opposite <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_MAX_div) |>
  select(gene_id, Domain, log2FoldChange) |>
  pivot_wider(names_from = Domain, values_from = log2FoldChange) |>
  filter(!is.na(MAX), !is.na(PM), MAX > 0, PM < 0) |>
  pull(gene_id)

if(length(genes_MAX_faster) > 0 | length(genes_MAX_opposite) > 0) {
  n_each <- 2
  if(length(genes_MAX_faster) < 15) n_each <- min(n_each, length(genes_MAX_faster))
  if(length(genes_MAX_opposite) < 15) n_each <- min(n_each, length(genes_MAX_opposite))
  
  plot_diverging_detailed(
    genes_faster = genes_MAX_faster,
    genes_opposite = genes_MAX_opposite,
    domain_name = "MAX vs PM",
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** MAX and PM follow different developmental trajectories.

**Biological Interpretation:**

-   **MAX increases faster**: `r length(genes_MAX_faster)` genes becoming progressively enriched in MAX
-   **MAX up, PM down**: `r length(genes_MAX_opposite)` genes with opposite responses

```{r}
#| label: max-div-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "MAX vs PM slope difference", padj < 0.05) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: max-div-pm-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on all MAX diverging genes
genes_MAX_div_ora <- run_enrichment_hybrid(gene_ids = genes_MAX_div,
                                     description = "MAX diverging from PM"
                                     )
```

```{r}
#| label: max-div-pm-ora-table
#| warning: false
#| message: false
genes_MAX_div_ora$result |> 
  create_dt()
```

### POST Diverging from PM

```{r}
#| label: post-div-pm-plot
#| warning: false
#| message: false

genes_POST_div <- gene_assignments |>
  filter(POST_div_PM) |>
  pull(gene_id)

genes_POST_faster <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div, Domain == "POST", log2FoldChange > 0.5, padj < 0.05) |>
  pull(gene_id)

genes_POST_opposite <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div) |>
  select(gene_id, Domain, log2FoldChange) |>
  pivot_wider(names_from = Domain, values_from = log2FoldChange) |>
  filter(!is.na(POST), !is.na(PM), POST > 0, PM < 0) |>
  pull(gene_id)

if(length(genes_POST_faster) > 0 | length(genes_POST_opposite) > 0) {
  n_each <- 2
  if(length(genes_POST_faster) < 15) n_each <- min(n_each, length(genes_POST_faster))
  if(length(genes_POST_opposite) < 15) n_each <- min(n_each, length(genes_POST_opposite))
  
  plot_diverging_detailed(
    genes_faster = genes_POST_faster,
    genes_opposite = genes_POST_opposite,
    domain_name = "POST vs PM",
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** POST and PM show divergent trajectories.

```{r}
#| label: post-div-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs PM slope difference", padj < 0.05) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-div-pm-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on all POST diverging genes
genes_POST_div_ora <- run_enrichment_hybrid(gene_ids = genes_POST_div,
                                     description = "POST diverging from PM"
                                     )
```

```{r}
#| label: post-div-pm-ora-table
#| warning: false
#| message: false
genes_POST_div_ora$result |> 
  create_dt()
```

### POST Diverging from MAX

```{r}
#| label: post-div-max-plot
#| warning: false
#| message: false

genes_POST_div_MAX <- gene_assignments |>
  filter(POST_div_MAX) |>
  pull(gene_id)

genes_POST_faster_MAX <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div_MAX, Domain == "POST", log2FoldChange > 0.5, padj < 0.05) |>
  pull(gene_id)

genes_POST_opposite_MAX <- b6_domain_slopes_clean |>
  filter(gene_id %in% genes_POST_div_MAX) |>
  select(gene_id, Domain, log2FoldChange) |>
  pivot_wider(names_from = Domain, values_from = log2FoldChange) |>
  filter(!is.na(POST), !is.na(MAX), POST > 0, MAX < 0) |>
  pull(gene_id)

if(length(genes_POST_faster_MAX) > 0 | length(genes_POST_opposite_MAX) > 0) {
  n_each <- 2
  if(length(genes_POST_faster_MAX) < 15) n_each <- min(n_each, length(genes_POST_faster_MAX))
  if(length(genes_POST_opposite_MAX) < 15) n_each <- min(n_each, length(genes_POST_opposite_MAX))
  
  plot_diverging_detailed(
    genes_faster = genes_POST_faster_MAX,
    genes_opposite = genes_POST_opposite_MAX,
    domain_name = "POST vs MAX",
    count_data = b6_gene_counts,
    n_each = n_each
  )
}
```

**Pattern:** POST and MAX show divergent developmental responses.

```{r}
#| label: post-div-max-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs MAX slope difference", padj < 0.05) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-div-max-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on all POST diverging genes
genes_POST_div_MAX_ora <- run_enrichment_hybrid(gene_ids = genes_POST_div_MAX,
                                     description = "POST diverging from MAX"
                                     )
```

```{r}
#| label: post-div-max-ora-table
#| warning: false
#| message: false
genes_POST_div_MAX_ora$result |> 
  create_dt()
```
:::

## Converging Trajectories

**These genes show domains becoming more similar over time**, with initially different expression levels that converge as development progresses.

::: {#B6_DESeq2_conv_traj .panel-tabset .nav-pills}
### Overview

```{r}
#| label: converging-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Converging") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### MAX Converging to PM

```{r}
#| label: max-conv-pm-plot
#| warning: false
#| message: false

genes_MAX_conv <- gene_assignments |>
  filter(MAX_conv_PM) |>
  pull(gene_id)

if(length(genes_MAX_conv) > 0) {
  n_show <- ifelse(length(genes_MAX_conv) < 15, length(genes_MAX_conv), 5)
  plot_simple_trajectories(
    gene_ids = genes_MAX_conv,
    count_data = b6_gene_counts,
    title_text = paste0("MAX Converging to PM (n=", length(genes_MAX_conv), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_MAX_conv)` genes where MAX and PM become more similar over time.

**Biological Interpretation:** Equalization of expression suggesting tissue integration or coordinated maturation.

```{r}
#| label: max-conv-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "MAX vs PM slope difference", padj < 0.05, log2FoldChange < -0.5) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: max-conv-pm-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_MAX_conv_ora <- run_enrichment_hybrid(gene_ids = genes_MAX_conv,
                                     description = "MAX converging to PM"
                                     )
```

```{r}
#| label: max-conv-pm-ora-table
#| warning: false
#| message: false
genes_MAX_conv_ora$result |> 
  create_dt()
```

### POST Converging to PM

```{r}
#| label: post-conv-pm-plot
#| warning: false
#| message: false

genes_POST_conv <- gene_assignments |>
  filter(POST_conv_PM) |>
  pull(gene_id)

if(length(genes_POST_conv) > 0) {
  n_show <- ifelse(length(genes_POST_conv) < 15, length(genes_POST_conv), 5)
  plot_simple_trajectories(
    gene_ids = genes_POST_conv,
    count_data = b6_gene_counts,
    title_text = paste0("POST Converging to PM (n=", length(genes_POST_conv), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_POST_conv)` genes converging between POST and PM.

```{r}
#| label: post-conv-pm-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs PM slope difference", padj < 0.05, log2FoldChange < -0.5) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-conv-pm-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_POST_conv_ora <- run_enrichment_hybrid(gene_ids = genes_POST_conv,
                                     description = "POST converging to PM"
                                     )
```

```{r}
#| label: post-conv-pm-ora-table
#| warning: false
#| message: false
genes_POST_conv_ora$result |> 
  create_dt()
```

### POST Converging to MAX

```{r}
#| label: post-conv-max-plot
#| warning: false
#| message: false

genes_POST_conv_MAX <- gene_assignments |>
  filter(POST_conv_MAX) |>
  pull(gene_id)

if(length(genes_POST_conv_MAX) > 0) {
  n_show <- ifelse(length(genes_POST_conv_MAX) < 15, length(genes_POST_conv_MAX), 5)
  plot_simple_trajectories(
    gene_ids = genes_POST_conv_MAX,
    count_data = b6_gene_counts,
    title_text = paste0("POST Converging to MAX (n=", length(genes_POST_conv_MAX), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_POST_conv_MAX)` genes converging between POST and MAX.

```{r}
#| label: post-conv-max-table
#| warning: false
#| message: false

b6_slope_differences |>
  filter(Comparison == "POST vs MAX slope difference", padj < 0.05, log2FoldChange < -0.5) |>
  select(gene_id, symbol, chromosome, Comparison, log2FoldChange, padj) |>
  arrange(log2FoldChange) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-conv-max-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_POST_conv_MAX_ora <- run_enrichment_hybrid(gene_ids = genes_POST_conv_MAX,
                                     description = "POST converging to MAX"
                                     )
```

```{r}
#| label: post-conv-max-ora-table
#| warning: false
#| message: false
genes_POST_conv_MAX_ora$result |> 
  create_dt()
```
:::

## Domain-Specific Dynamics

**These genes show significant developmental changes in only one or two domains** while remaining stable in others.

::: {#B6_DESeq2_domain_dyn .panel-tabset .nav-pills}
### Overview

```{r}
#| label: domain-specific-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Domain-Specific") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### PM Only

```{r}
#| label: pm-only-plot
#| warning: false
#| message: false

genes_PM_only_up <- gene_assignments |>
  filter(PM_only_up) |>
  pull(gene_id)

genes_PM_only_down <- domain_dynamics |>
  filter(PM_dynamic, !MAX_dynamic, !POST_dynamic, PM_slope < -0.5) |>
  pull(gene_id)

if(length(genes_PM_only_up) > 0) {
  n_show <- ifelse(length(genes_PM_only_up) < 15, length(genes_PM_only_up), 3)
  plot_domain_specific_dynamics(
    gene_ids = genes_PM_only_up,
    domain_name = "PM",
    direction = "Increasing",
    count_data = b6_gene_counts,
    slope_data = b6_domain_slopes_clean,
    n = n_show
  )
}
```

**Pattern:** Developmental changes only in PM domain.

**Biological Interpretation:** PM-specific processes like midline fusion.
Found `r length(genes_PM_only_up)` increasing and `r length(genes_PM_only_down)` decreasing.

```{r}
#| label: pm-only-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% c(genes_PM_only_up, genes_PM_only_down), Domain == "PM") |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: pm-only-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on both sets
genes_PM_only_up_ora <- run_enrichment_hybrid(gene_ids = genes_PM_only_up,
                                     description = "PM only increasing"
                                     )

genes_PM_only_down_ora <- run_enrichment_hybrid(gene_ids = genes_PM_only_down,
                                     description = "PM only decreasing"
                                     )

```

```{r}
#| label: pm-only-ora-table
#| warning: false
#| message: false
genes_PM_only_up_ora$result |> 
  rbind(
    genes_PM_only_down_ora$result
  ) |> 
  create_dt()
```

### MAX Only

```{r}
#| label: max-only-plot
#| warning: false
#| message: false

genes_MAX_only_up <- gene_assignments |>
  filter(MAX_only_up) |>
  pull(gene_id)

genes_MAX_only_down <- domain_dynamics |>
  filter(!PM_dynamic, MAX_dynamic, !POST_dynamic, MAX_slope < -0.5) |>
  pull(gene_id)

if(length(genes_MAX_only_up) > 0) {
  n_show <- ifelse(length(genes_MAX_only_up) < 15, length(genes_MAX_only_up), 3)
  plot_domain_specific_dynamics(
    gene_ids = genes_MAX_only_up,
    domain_name = "MAX",
    direction = "Increasing",
    count_data = b6_gene_counts,
    slope_data = b6_domain_slopes_clean,
    n = n_show
  )
}
```

**Pattern:** Developmental changes only in MAX domain.

**Biological Interpretation:** MAX-specific processes like lateral shelf morphogenesis.
Found `r length(genes_MAX_only_up)` increasing and `r length(genes_MAX_only_down)` decreasing.

```{r}
#| label: max-only-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% c(genes_MAX_only_up, genes_MAX_only_down), Domain == "MAX") |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: max-only-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on both sets
genes_MAX_only_up_ora <- run_enrichment_hybrid(gene_ids = genes_MAX_only_up,
                                     description = "MAX only increasing"
                                     )

genes_MAX_only_down_ora <- run_enrichment_hybrid(gene_ids = genes_MAX_only_down,
                                     description = "MAX only decreasing"
                                     )
```


```{r}
#| label: max-only-ora-table
#| warning: false
#| message: false
genes_MAX_only_up_ora$result |> 
  rbind(
    genes_MAX_only_down_ora$result
  ) |> 
  create_dt()
```

### POST Only

```{r}
#| label: post-only-plot
#| warning: false
#| message: false

genes_POST_only_up <- gene_assignments |>
  filter(POST_only_up) |>
  pull(gene_id)

genes_POST_only_down <- domain_dynamics |>
  filter(!PM_dynamic, !MAX_dynamic, POST_dynamic, POST_slope < -0.5) |>
  pull(gene_id)

if(length(genes_POST_only_up) > 0) {
  n_show <- ifelse(length(genes_POST_only_up) < 15, length(genes_POST_only_up), 3)
  plot_domain_specific_dynamics(
    gene_ids = genes_POST_only_up,
    domain_name = "POST",
    direction = "Increasing",
    count_data = b6_gene_counts,
    slope_data = b6_domain_slopes_clean,
    n = n_show
  )
}
```

**Pattern:** Developmental changes only in POST domain.

**Biological Interpretation:** Posterior-specific programs like soft palate differentiation.
Found `r length(genes_POST_only_up)` increasing and `r length(genes_POST_only_down)` decreasing.

```{r}
#| label: post-only-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% c(genes_POST_only_up, genes_POST_only_down), Domain == "POST") |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(desc(abs(log2FoldChange))) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: post-only-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

# Run enrichment on both sets
genes_POST_only_up_ora <- run_enrichment_hybrid(gene_ids = genes_POST_only_up,
                                     description = "POST only increasing"
                                     )

genes_POST_only_down_ora <- run_enrichment_hybrid(gene_ids = genes_POST_only_down,
                                     description = "POST only decreasing"
                                     )

```

```{r}
#| label: post-only-ora-table
#| warning: false
#| message: false
genes_POST_only_up_ora$result |> 
  rbind(
    genes_POST_only_down_ora$result
  ) |> 
  create_dt()
```
:::

## Complex Patterns

**Genes showing particularly complex spatiotemporal dynamics** including opposite trajectories or three-way divergence.

::: {#B6_DESeq2_comp_pat .panel-tabset .nav-pills}
### Overview

```{r}
#| label: complex-overview
#| warning: false
#| message: false

pattern_summary |>
  filter(Category == "Complex") |>
  arrange(desc(N_Genes)) |>
  create_dt()
```

### Opposite Slopes

```{r}
#| label: opposite-slopes-plot
#| warning: false
#| message: false

genes_opposite <- gene_assignments |>
  filter(Opposite_slopes) |>
  pull(gene_id)

if(length(genes_opposite) > 0) {
  n_show <- ifelse(length(genes_opposite) < 15, length(genes_opposite), 5)
  plot_simple_trajectories(
    gene_ids = genes_opposite,
    count_data = b6_gene_counts,
    title_text = paste0("Genes with Opposite Slopes (n=", length(genes_opposite), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_opposite)` genes where at least one domain increases while another decreases.

**Biological Interpretation:** Fundamentally different developmental programs between domains, potentially key regulators of domain identity.

```{r}
#| label: opposite-slopes-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% genes_opposite) |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(gene_id, Domain) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: opposite-slopes-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_opposite_ora <- run_enrichment_hybrid(gene_ids = genes_opposite,
                                     description = "Opposite slopes"
                                     )
```

```{r}
#| label: opposite-slopes-ora-table
#| warning: false
#| message: false
genes_opposite_ora$result |> 
  create_dt()
```

### Three-Way Divergent

```{r}
#| label: three-way-plot
#| warning: false
#| message: false

genes_three_way <- gene_assignments |>
  filter(Three_way) |>
  pull(gene_id)

if(length(genes_three_way) > 0) {
  n_show <- ifelse(length(genes_three_way) < 15, length(genes_three_way), 5)
  plot_simple_trajectories(
    gene_ids = genes_three_way,
    count_data = b6_gene_counts,
    title_text = paste0("Three-Way Divergent Genes (n=", length(genes_three_way), ")"),
    n = n_show
  )
}
```

**Pattern:** `r length(genes_three_way)` genes where all three domains show different developmental slopes.

**Biological Interpretation:** Maximum spatiotemporal complexity, likely central regulators of domain-specific identity and function.

```{r}
#| label: three-way-table
#| warning: false
#| message: false

b6_domain_slopes_clean |>
  filter(gene_id %in% genes_three_way) |>
  select(gene_id, symbol, chromosome, Domain, log2FoldChange, padj) |>
  arrange(gene_id, Domain) |>
  create_dt()
```

Over-represented pathways and biological processes:

```{r}
#| label: three-way-ora
#| warning: false
#| message: false
#| results: hide
#| cache: true

genes_three_way_ora <- run_enrichment_hybrid(gene_ids = genes_three_way,
                                     description = "Three-way divergent"
                                     )
```

```{r}
#| label: three-way-ora-table
#| warning: false
#| message: false
genes_three_way_ora$result |> 
  create_dt()
```
:::

## Pattern Overlap Analysis

::: {#B6_DESeq2_pattern_overlap .panel-tabset .nav-pills}
### Multi-Pattern Genes

```{r}
#| label: pattern-overlap
#| warning: false
#| message: false

multi_pattern_genes <- gene_assignments |>
  filter(n_patterns > 1) |>
  arrange(desc(n_patterns))

multi_pattern_genes |>
  select(symbol, n_patterns) |>
  group_by(n_patterns) |>
  summarize(n_genes = n(), .groups = "drop") |>
  create_dt()
```

### Top Multi-Pattern Genes

```{r}
#| label: top-multi-pattern
#| warning: false
#| message: false

multi_pattern_genes |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, n_patterns) |>
  head(50) |>
  create_dt()
```

### Pattern Overlap UpSet Plot

```{r}
#| label: pattern-upset-plot
#| warning: false
#| message: false
#| fig.width: 14
#| fig.height: 10

# Prepare data for upset plot - select key pattern categories
upset_data <- gene_assignments |>
  select(
    gene_id,
    # Domain main effects
    MAX_higher_PM, MAX_lower_PM,
    POST_higher_PM, POST_lower_PM,
    POST_higher_MAX, POST_lower_MAX,
    Highest_MAX, Highest_POST, Highest_PM,
    AP_decreasing, AP_increasing,
    # Stage main effects
    Dev_increasing, Dev_decreasing,
    # Diverging
    MAX_div_PM, POST_div_PM, POST_div_MAX,
    # Converging
    MAX_conv_PM, POST_conv_PM, POST_conv_MAX,
    # Domain-specific
    PM_only_up, MAX_only_up, POST_only_up,
    # Complex
    Opposite_slopes, Three_way
  ) |>
  # Convert to data frame and set gene_id as rownames
  as.data.frame() |>
  column_to_rownames("gene_id") |>
  # Convert logical to numeric
  mutate(across(everything(), as.numeric))

# Create upset plot
UpSetR::upset(
  upset_data,
  sets = colnames(upset_data),
  order.by = "freq",
  decreasing = TRUE,
  mb.ratio = c(0.6, 0.4),
  number.angles = 0,
  point.size = 3,
  line.size = 1,
  mainbar.y.label = "Pattern Intersections",
  sets.x.label = "Genes per Pattern",
  text.scale = c(1.5, 1.2, 1.2, 1.2, 1.5, 1.2),
  set_size.show = TRUE,
  keep.order = TRUE,
  sets.bar.color = "#56B4E9"
)
```

This UpSet plot visualizes: - The size of each pattern category (left bar chart) - Which pattern combinations occur most frequently (top bar chart) - The degree of overlap between different patterns (connection lines) - Genes with multiple pattern assignments
:::
::::::::::

# Comprehensive Tables

::: {#B6_DESeq2_tables .panel-tabset .nav-pills}
## All Patterns Summary

```{r}
#| label: all-patterns-table
#| warning: false
#| message: false

pattern_summary |>
  create_dt()
```

## Gene-Level Pattern Assignments

```{r}
#| label: gene-assignments-table
#| warning: false
#| message: false

gene_assignments |>
  left_join(all_genes_palate |> select(gene_id, chromosome), by = "gene_id") |>
  select(gene_id, symbol, chromosome, n_patterns, everything()) |>
  create_dt()
```

## Over-representation analysis summary

```{r}
#| label: combined-ora-results
#| warning: false
#| message: false

# Collect all ORA results and combine them
all_ora_results <- bind_rows(
  # Domain Main Effects
  genes_MAX_higher_than_PM_ora$result,
  genes_MAX_lower_than_PM_ora$result,
  genes_POST_higher_than_PM_ora$result,
  genes_POST_lower_than_PM_ora$result,
  genes_POST_higher_than_MAX_ora$result,
  genes_POST_lower_than_MAX_ora$result,
  genes_MAX_highest_ora$result,
  genes_POST_highest_ora$result,
  genes_PM_highest_ora$result,
  genes_AP_decreasing_ora$result,
  genes_AP_increasing_ora$result,
  
  # Stage Main Effects
  dev_increasing_enrichment_results$result,
  dev_decreasing_enrichment_results$result,
  
  # Diverging Trajectories
  genes_MAX_div_ora$result,
  genes_POST_div_ora$result,
  genes_POST_div_MAX_ora$result,
  
  # Converging Trajectories
  genes_MAX_conv_ora$result,
  genes_POST_conv_ora$result,
  genes_POST_conv_MAX_ora$result,
  
  # Domain-Specific Dynamics
  genes_PM_only_up_ora$result,
  genes_PM_only_down_ora$result,
  genes_MAX_only_up_ora$result,
  genes_MAX_only_down_ora$result,
  genes_POST_only_up_ora$result,
  genes_POST_only_down_ora$result,
  
  # Complex Patterns
  genes_opposite_ora$result,
  genes_three_way_ora$result
) |>
  # Add category grouping for easier filtering
  mutate(
    Category = case_when(
      gene_set %in% c("MAX higher than PM", "MAX lower than PM", 
                      "POST higher than PM", "POST lower than PM",
                      "POST higher than MAX", "POST lower than MAX",
                      "MAX highest", "POST highest", "PM highest",
                      "A→P decreasing", "A→P increasing") ~ "Domain Main Effects",
      gene_set %in% c("Increase", "Decrease") ~ "Stage Main Effects",
      gene_set %in% c("MAX diverging from PM", "POST diverging from PM", 
                      "POST diverging from MAX") ~ "Diverging Trajectories",
      gene_set %in% c("MAX converging to PM", "POST converging to PM", 
                      "POST converging to MAX") ~ "Converging Trajectories",
      gene_set %in% c("PM only increasing", "PM only decreasing",
                      "MAX only increasing", "MAX only decreasing",
                      "POST only increasing", "POST only decreasing") ~ "Domain-Specific Dynamics",
      gene_set %in% c("Opposite slopes", "Three-way divergent") ~ "Complex Patterns",
      TRUE ~ "Other"
    )
  ) |>
  # Reorder columns
  select(Category, gene_set, source, term_id, term_name, p_value, 
         query_size, term_size, intersection_size, intersection) |>
  filter( p_value < 0.01) |> 
  arrange(Category, gene_set, p_value)
```

```{r}
#| label: combined-ora-results-table
#| warning: false
#| message: false
# Display the combined results
all_ora_results |>
  create_dt()

# # Summary by category
# all_ora_results |>
#   group_by(Category, gene_set) |>
#   summarize(
#     n_enriched_terms = n(),
#     .groups = "drop"
#   ) |>
#   arrange(Category, desc(n_enriched_terms)) |>
#   create_dt()

```

:::


# Session Info

```{r}
sessionInfo()
```
