---
title: "Data for Robyn"
output: html_document
author: Selcan Aydin
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(assertthat)
library(crunch)

```

This is what Robyn asked for:

> The datasets I need are
>
> 1.  meta data e.g., experimental_cleft_palate_DO-meta.csv
>
>     1.  columns are mouse_id that matches column names in expression and genoprobs file, sex (m, f),  generation, any other important covariates
>
> 2.  expression data (prefer TPM) \--\> experimental_cleft_palate_DO-data.csv.gz
>
>     1.  column 1: gene_id = ensemble gene_id followed by columns for mouse_ids
>
> 3.  genoprobs file, prefer RDS array or just a csv is fine \--\> experimental_cleft_palate_DO-genoprobs.RDS or .csv.gz
>
> 4.  eQTL results \--\> experimental_cleft_palate_DO-eQTL.csv.gz
>
>     1.  marker_id, marker_chr, gene_id, lod
>
>     2.  were corrections made for covariates?

Read in sample details and write the metadata to `/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-meta.csv` .

```{r write_metadata, warning=FALSE, message=FALSE}
# reading in the sample details + some phenotypes
sample_details <- readr::read_csv("/projects/munger-lab/projects/palate-project-murray-munger/samplesheet.csv") 
# remove contaminated samples then filter for only DO@13.5
contam_samples <- c("B6_12.5_PM_1","B6_13.5_POST_3", "CAST_14.5_MAX_7", "DO_13.5_MAX_6", "DO_13.5_MAX_14", "DO_13.5_MAX_14_2", "B6_13.5_MAX_3")

sample_annot <- sample_details |> 
  filter( !sample %in% contam_samples) |> 
  separate(Strain_stock, into = c("strain","tmp"), remove = F, sep = "_") |> 
  filter( strain =="DO") |> 
  filter(!( strain =="DO" & Gest_stage == "E12.5") ) |> 
  select( mouse_id = sample, 
          sex,
          generation = DO_gen,
          batch
          )

write_csv(sample_annot,
          file = "/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-meta.csv")


```

Read in tpm data and write to `/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-data.csv.gz`.

```{r write_tpm, warning=FALSE, message=FALSE}
# get the list of tpm files
tpm_files <- list.files("/projects/munger-lab/projects/palate-project-murray-munger/read_mapping/gbrs_emase", pattern = "DO_.*diploid.genes.tpm", full.names = T)
# read in and merge the tpm files
file1 <- read_tsv(tpm_files[1])
read_gbrs <- function(filename) {
  samplename <- paste(strsplit(basename(filename), "[.]")[[1]][1:2], collapse=".")
  dat <- read_tsv(filename) |> mutate(sample=samplename)
  dat <- unite(dat, id, sample, remove=FALSE)
  assert_that(are_equal(dat$locus, file1$locus))
  select(dat, locus, id, sample, total)
}
tpm <- bind_rows(lapply(tpm_files, read_gbrs)) |> 
  rename(tpm=total)

tpm |> 
  select(-sample) |>  
  pivot_wider( names_from = "id", values_from = "tpm") |> 
  rename("gene_id"="locus") |> 
  select(gene_id, sample_annot$mouse_id) |> 
  write.csv.gz(file = "/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-data.csv.gz")

```

Load in genoprobs and write to `/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-genoprobs.RDS`.

```{r write_genoprobs, message=FALSE, warning=FALSE}
load("/projects/munger-lab/projects/palate-project-murray-munger/qtl_mapping/DO_Palate_eQTL_forMapping.RData") # expr, exprZ, genoprobs ..etc.
write_rds(genoprobs, file = "/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-genoprobs.RDS")
rm(genoprobs, expr)
```


Load eQTL mapping results and write to `/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-eQTL.csv.gz`

```{r write_eqtl_results, message=FALSE, warning=FALSE}
load("/projects/munger-lab/projects/palate-project-murray-munger/qtl_mapping/DO_Palate_paired_eQTL_peaks.RData")

peaks |> 
  left_join(
    map_dat2 |> 
      mutate(pos = as.numeric(pos)),
    by=c("peak_chr"="chrom","peak_cM"="pos")
    ) |> 
  select(
    marker_id = marker,
    marker_chr = peak_chr,
    gene_id = phenotype,
    lod
  ) |> 
  write.csv.gz(file = "/projects/munger-lab/projects/palate-project-murray-munger/experimental_maxilla_DO-eQTL.csv.gz")



```

Details on data:

-   The data is from [maxilla](https://bioportal.bioontology.org/ontologies/UBERON?p=classes&conceptid=http%3A%2F%2Fpurl.obolibrary.org%2Fobo%2FUBERON_0002397) from E13.5 mouse embryos.

-   GBRS v0.1.6 is used (GRCm38).

-   Both the gene expression counts and genotypes are obtained from GBRS.

-   69k grid is used in mapping.

-   3 samples were removed due to contamination resulting in `r nrow(exprZ)` many samples.

-   For filtering I use median(tpm) \> 0.5 resulting in `r ncol(exprZ)` many genes.

-   The expected gene expression counts are upper quantile normalized then rankZ transformed before eQTL mapping.

-   Sex was used as a covariate in eQTL mapping.


